### 2020-06-18 covidWarehouse/manuscript/descriptives.Rmd
manuscript

---
title: "Incidence and period prevalence of confirmed COVID-19 cases in care home staff and residents: Cohort study of secondary use records in a large care home provider in England, Scotland and Northern Ireland (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
date: "16/06/2020"
bibliography: references.bib
csl: sage-vancouver.csl
output:
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
  pdf_document:
    citation_package: natbib
    df_print: kable
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
    reference_docx: template.docx
link-citations: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
options(OutDec = "·")
knit_output_pdf <- F
if(knit_output_pdf){
  options(knitr.table.format = "pdf")
} else {
  options(knitr.table.format = "pandoc")
}

source("../.Rprofile")
devtools::load_all("..")
load_data()
library(dplyr)
library(lubridate)
library(kableExtra)
library(captioner)
library(survival)
library(survminer)
library(ggplot2)
library(finalfit)

knit_output_pdf <- F
if(knit_output_pdf){
  options(knitr.table.format = "pdf")
} else {
  options(knitr.table.format = "pandoc")
}

date_min <- as.Date("2020-03-02")
date_max <- as.Date("2020-06-15")
date_max_tallies = as.Date("2020-06-07")

```

```{r descriptives}
homes_country <- select(reference_homes, home_code, postcode) %>% 
  left_join(transmute(reference_geography, postcode, 
                      country, region_england = region), by = "postcode") %>% 
  mutate(region_UK = if_else(country == "England", region_england, country)) %>% 
  select(-postcode)

incidents <- incidents %>% 
  dplyr::filter(dplyr::between(incident_date, date_min, date_max)) #%>% 
  # dplyr::filter(!resident_id %in% 
  #                 residents$resident_id[residents$funding_type %in% c("Young Chronic Sick/YPD", 
  #                                                                     "Younger Person")])

desc_before_exclusions <- list()
desc_before_exclusions$nrow_incidents <- nrow(incidents)
desc_before_exclusions$missing_id <- sum(grepl("^MISSING", incidents$resident_id))
desc_before_exclusions$missing_id_pct <- mean(grepl("^MISSING", incidents$resident_id))*100
desc_before_exclusions$link_fail <- sum(!incidents$resident_id[
  which(!grepl("^MISSING", incidents$resident_id))] %in% unique(residents$resident_id))
desc_before_exclusions$link_fail_pct <- tbrounding(
  sum(!incidents$resident_id[
    which(!grepl("^MISSING", incidents$resident_id))] %in% unique(residents$resident_id)) / 
    nrow(incidents)*100)

# removing erroneous incident reports (no identifier or not linking to residents file) 
incidents <- filter(incidents, !grepl("^MISSING", resident_id)) %>% 
  filter(resident_id %in% unique(residents$resident_id))

tline <- dplyr::bind_rows(timelines$timeline)
stopifnot(unique(tline$date_end)>=date_max)
tline <- filter(tline, between(date, date_min, date_max)) %>% 
  filter(!funding_type %in% c("Young Chronic Sick/YPD", "Younger Person"))

new_cases <- filter(new_cases, between(date, date_min, date_max_tallies))

approx_occupancy <- tline %>% 
  group_by(home_code, date) %>% 
  summarise(
    occupancy = sum(tidyr::replace_na(rday, 0)),
    susceptible_symptomatic = sum(susceptible_symptomatic, na.rm = T),
    susceptible_confirmed = sum(susceptible_confirmed, na.rm = T))

approx_occupancy <- merge(
  tibble(date = seq(date_min, date_max, 1)),
  tibble(home_code = unique(reference_homes$home_code)),
  all = TRUE) %>% 
  left_join(approx_occupancy)

desc_residents <- tline %>% 
  filter(rday == 1)  %>% 
  group_by(resident_id, home_code, dob, gender, 
           admission_type, home_type, user_type, status) %>% 
  summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 50, age.max = 100)) %>% 
  left_join(homes_country, by = "home_code")

res_counts <- desc_residents %>% 
  group_by(home_code) %>% 
  summarise(unique_residents = n_distinct(resident_id))

desc_homes <- reference_homes %>% 
  left_join(transmute(reference_geography, postcode, 
                      country, region_england = region), by = "postcode") %>% 
  left_join(filter(beds, year=="2020"), by = "home_code") %>% 
  left_join(res_counts, by = "home_code") %>%
  mutate(Scope = paste("FSHCG", country)) %>% 
  group_by(Scope) %>% 
  summarise(`Total homes` = n_distinct(home_code),
            `Total beds` = sum(beds),
            `Total contract beds` = sum(contract_beds, na.rm = T),
            `Total residents (excl. contract beds)` = sum(unique_residents, na.rm = T))

desc_national <- tibble(
  Scope = c("England", "Scotland (2017)", "Northern Ireland"),
  `Total homes` = c(9400, 1142, 483),
  `Total beds` = c(45000, 40926, 16095),
  `Total contract beds` = NA_integer_,
  `Total residents` = c(NA, 35989, NA)
)

desc_overall <- list()
desc_overall$total_residents <- dplyr::n_distinct(desc_residents$resident_id)
#COUNTS
desc_overall$CT_total_symptomatic <- sum(new_cases$new_symptomatic_home, na.rm=T) + 
  sum(new_cases$new_symptomatic_hospital, na.rm=T)
desc_overall$CT_p_symptomatic <- desc_overall$CT_total_symptomatic/desc_overall$total_residents
desc_overall$CT_p_symptomatic_ME <- 1.96 * sqrt(
  (desc_overall$CT_p_symptomatic*(1-desc_overall$CT_p_symptomatic)/desc_overall$total_residents)
)
#INCIDENT REPORTS
desc_overall$unique_residents_incidents <- dplyr::n_distinct(incidents$resident_id)

desc_overall$DAT_total_symptomatic <- n_distinct(filter(incidents, !is.na(covid_first_symptomatic))$resident_id)
desc_overall$DAT_p_symptomatic <- desc_overall$DAT_total_symptomatic/desc_overall$total_residents
desc_overall$DAT_p_symptomatic_ME <- 1.96 * sqrt(
  desc_overall$DAT_p_symptomatic * (1 - desc_overall$DAT_p_symptomatic) / 
    desc_overall$total_residents
)
desc_overall$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
desc_overall$unique_residents_tested <- dplyr::filter(incidents, covid_tested == 1) %>% 
  dplyr::n_distinct(.$resident_id)
desc_overall$unique_residents_test_result <- dplyr::filter(incidents, !is.na(covid_test_result)) %>% 
  dplyr::n_distinct(.$resident_id)
desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>% 
  dplyr::n_distinct(.$resident_id)

desc_homes <- bind_rows(desc_homes, desc_national)

## TABLES
table_nums <- captioner::captioner(prefix = "Table")
tab_national_descriptives <- table_nums(
  name = "tab_national_descriptives",
  caption = "Total homes, beds, and residents within FSHCG and nationwide")
tab_resident_descriptives <- table_nums(
  name = "tab_resident_descriptives",
  caption = paste0("Breakdown of FSHCG residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))

tab_CoxPH_confirmed <- table_nums(
  name = "tab_CoxPH_confirmed",
  caption =  paste0("Hazard ratios of becoming a confirmed case estimated from a Cox Proportional Hazards model (n=", 
  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
  ")"))

## FIGURES
fig_nums <- captioner(prefix = "Figure")

fig_KM_CT_residents_symptomatic <- fig_nums(
  name = "KM_CT_residents_symptomatic",
  caption = paste0("Kaplan-Meier estimator of the cumulative probability of residents becoming symptomatic based on care home manager daily counts")
)
fig_KM_confirmed_residents <- fig_nums(
  name = 'fig_KM_confirmed_residents',
  caption = paste0("Kaplan-Meier estimator of the cumulative probability of residents (n=", 
  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
  ") becoming confirmed cases based on Datix incidents report"))
fig_KM_CT_confirmed_residents <- fig_nums(
  name = 'fig_KM_CT_confirmed_residents',
  caption = "Kaplan-Meier estimator of the cumulative probability of residents becoming confirmed cases based on care home manager daily counts")

```

# Abstract

**Background**

**Methods:**

**Findings:** A total of `r formatbm(n_distinct(desc_residents$resident_id))` residents were examined across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)` of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).

**Interpretation**

**Funding:** This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].

# Background

COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes where one in seven individuals aged > 85 years live. However, there is no surveillance for infection in care homes, nor are there systems (or research studies) monitoring the impact of the pandemic on individuals or systems. Usual practices are disrupted during the pandemic, and care home staff are taking on new and unfamiliar roles, such as advanced care planning. Understanding the nature of these changes is critical to mitigate the impact of COVID-19 on residents, relatives and staff.

This study reports on data obtained from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care. In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`)). FSHCG residents accessed nose and throat swab antigen tests during hospitalisations in the first instance, but tests became available in care homes from XXXXX. From 11 May 2020, FSHCG home undertook testing every resident. ***(Haydn, more detail would be greatly appreciated)***

`r table_nums("tab_national_descriptives")` @cqcdata2020,@hscni2020,@isd2020

```{r table_national_descriptives, fig.cap = tab_national_descriptives, message=F}
mutate_all(desc_homes,  .funs = formatbm) %>% arrange(gsub("FSHC ", "", Scope)) %>% t %>% kable
```

# Methods

## Individual-level residents data

Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of:

- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly). This dataset does not include any records for occupants of local authority 'block contract' beds  ($n=$ `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))`).
- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff into FSHCG's Datix incidents management system. Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting. In addition, reports can include: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.

HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier. Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents dataset. Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) Datix reports did not contain enough information to generate an identifier and were excluded. Identifiers of a further `r desc_before_exclusions$link_fail` reports (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident. A proportion of these reports may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.

## Aggregate data

Daily counts reported by care home managers to FSHCG were obtained for every care home:

- new symptomatic cases (residents)
- new confirmed cases in home (residents)
- new confirmed cases in hospital (residents)
- deaths related to COVID-19 (Haydn please confirm what these correspond to: suspected/coroner report/tested positive before death?)
- new symptomatic cases (staff)
- new confirmed cases (staff)
- deaths related to COVID-19 (staff)

These counts include events relating to occupants of local authority contract beds.

## Analysis

We estimate the incidence of symptomatic and confirmed cases as well as deaths using the individual-level data. Doing so enables us to accurately account for the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died. In order to study the effect of potential risk factors, total cases and resident-days were aggregated by care home by care type and demographic characteristics and regressed in a Poisson hierarchical level with care home and time independent and identically distributed random effects. This provides measures of relative risks.

However, a Datix report may not have been filed for every event. We therefore also estimate the incidence and period prevalence of events using manager counts, using the total resident-days denominator derived from the individual-level data.

- figure for the cumulative number/proportion of staff and residents infected over time... this is another area of significant interest so we can get an estimate (with caveats) of the proportion of staff and residents who remain susceptible to infection. -> KM by age or care home type or nation/region
- how many care homes have not had any cases reported in staff or residents during the pandemic because these homes will presumably be at greatest risk in a second wave.
- individual and care home characteristics associated with high number of cases. Nationally the only 2 factors that have been proposed so far are care home size and provision of nursing care. -> **regression model on RR of infection** number of previous cases (lag effect) is important here. So the time effect

Data were analysed in `R`3·5·0 @RCoreTeam2018 and the `survival` package @therneau2015. All computer syntax is available on an online repository @Dutey2020.

# Results

## Descriptives

The study population consisted of `r formatbm(desc_overall$total_residents)` residents across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`). The vast majority (85·4%) were permanent residents. A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_overall$unique_residents_incidents)` unique residents.

Datix reports and manager counts respectively indicated that `r formatbm(desc_overall$DAT_total_symptomatic)` (`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% [95% confidence interval: `r tbrounding((desc_overall$DAT_p_symptomatic - desc_overall$DAT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$DAT_p_symptomatic + desc_overall$DAT_p_symptomatic_ME)*100)`]) and `r formatbm(desc_overall$CT_total_symptomatic)` (`r tbrounding(desc_overall$CT_p_symptomatic*100)`% [`r tbrounding((desc_overall$CT_p_symptomatic - desc_overall$CT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$CT_p_symptomatic + desc_overall$CT_p_symptomatic_ME)*100)`]) became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`. Temporal trends broken down by region (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`) indicate wide variation across the UK, with care homes in the South East and Northern Ireland reporting in excess of a third of residents with symptoms.

`r table_nums("tab_resident_descriptives")`

```{r table_resident_descriptives, message = FALSE, results="asis"}
arsenal::tableby(home_type  ~ gender + ageg + region_UK +
                                 user_type + admission_type + status, 
                               data = desc_residents) %>% summary()
```

`r fig_nums("fig_KM_CT_residents_symptomatic")`

```{r KM_CT_residents_symptomatic, fig.height = 5, fig.width = 10}

residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
  left_join(distinct(new_cases, date, home_code, 
                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
  filter(!is.na(home_code)) %>% 
  left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  left_join(approx_occupancy) %>% 
  arrange(date)

km_ct_residents_symp <- residents_ct_symp %>% 
  group_by(date, region_UK) %>% 
  summarise(
    symptomatic = sum(tidyr::replace_na(new_symptomatic_home, 0) + 
                      tidyr::replace_na(new_symptomatic_hospital, 0)),
    occupancy = sum(tidyr::replace_na(susceptible_symptomatic, 0))
  ) %>% 
  group_by(region_UK) %>% 
  km_ct_estimator(df = ., t = date, d = symptomatic, 
                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
  ungroup() %>% 
  mutate(
    St_max = S_t^exp(1.96*S_t_SElog),
    St_min = S_t^exp(-1.96*S_t_SElog)
  )

order_key <-km_ct_residents_symp %>% 
  group_by(region_UK) %>% 
  summarise(S_t = min(S_t)) %>% 
  arrange(S_t)
km_ct_residents_symp$region_UK <- factor(
  km_ct_residents_symp$region_UK,
  levels = order_key$region_UK
)
km_ct_residents_symp$region_group <- cut(as.integer(km_ct_residents_symp$region_UK), 2, labels = 1:2)

ggplot(aes(x = date, y = 1-S_t, 
           ymax = 1 - St_max,
           ymin = 1 - St_min,
           color = region_UK), data = km_ct_residents_symp) +
  # facet_grid(region_group ~ .) +
  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
  geom_path(lwd=1.2) +
  scale_y_continuous("Cum probability confirmed case",
                     breaks = seq(0,1,.05)) +
  scale_color_discrete("Region") +
  scale_x_date(name = "Date", 
               breaks = unique(get_monday_date(km_ct_residents_symp$date)),
               labels = format.Date(
                 unique(get_monday_date(km_ct_residents_symp$date)),
                 "%d/%m/%y"
               )) +
  theme_bw()

```

## Resident confirmed cases

The reports were returned by `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes  totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds. 
Datix incidents accounted for `r formatbm(desc_overall$unique_residents_tested)` residents tested at least once. Test results were known for `r formatbm(desc_overall$unique_residents_test_result)` residents: `r formatbm(desc_overall$unique_residents_test_result_positive)` residents had a record of a positive test result. This data cannot be treated as representative, as engagement with FSHCG confirmed that test coming back negative were not commonly recorded on Datix. Care home manager daily counts of new confirmed cases added up to `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` cases tested in homes and `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` tested in hospitals. The regional breakdown is presented in

The temporal trend of confirmed cases obtained from either Datix reports (`r fig_nums("fig_KM_confirmed_residents", display = "cite")`) or manager counts (`r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`)

`r fig_nums("fig_KM_confirmed_residents")`

```{r km_confirmed, fig.cap=fig_KM_confirmed_residents}
km_confirmed <- tline %>% 
  left_join(select(reference_homes, home_code, postcode)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  filter(rday == 1 & susceptible_confirmed == 1) %>% 
  mutate(home_type = factor(home_type, levels = c("Residential", "Nursing"))) %>% 
  group_by(resident_id, dob, gender, region_UK,
           home_type, user_type, funding_type, admission_type) %>% 
  summarise(
    start = as.numeric(min(date) - date_min),
    end = max(date)
  ) %>% 
  left_join(select(timelines, resident_id, covid_first_positive)) %>% 
  mutate(event = if_else(
    is.na(covid_first_positive),
    0L,
    as.integer(covid_first_positive == end)
  )) %>% 
  mutate(
    stop = as.numeric(end +1 - date_min),
    ageg = cut_age_denary(compute_age(dob, date_min), age.min = 50, age.max = 100)
  )

survfit(Surv(time = start, 
                   time2 = stop,
                   event = event) ~ region_UK, data = km_confirmed) %>% 
  ggsurvplot(., conf.int = TRUE, 
           fun = "event",
           legend = c(0.16, 0.6),
           ggtheme = theme_bw(),
           # TODO: double check labels at the end
           break.x.by	= 7,
           break.y.by = .02,
           ylim = c(0, .17),
           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
           ylab = "Cum probability of confirmed case",
           legend.labs = sort(unique(km_confirmed$region_UK)),
           conf.int.alpha = .1)

```

`r fig_nums("fig_KM_confirmed_residents")`

```{r KM_CT_residents_infection, fig.height = 5, fig.width = 10}

residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
  left_join(distinct(new_cases, date, home_code, 
                     new_confirmed_home, new_confirmed_hospital)) %>% 
  filter(!is.na(home_code)) %>% 
  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  left_join(approx_occupancy) %>% 
  arrange(date)

km_ct_residents_confirmed <- residents_ct %>% 
  group_by(date, region_UK) %>% 
  summarise(
    confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
                      tidyr::replace_na(new_confirmed_hospital, 0)),
    occupancy = sum(tidyr::replace_na(susceptible_confirmed, 0))
  ) %>% 
  group_by(region_UK) %>% 
  km_ct_estimator(df = ., t = date, d = confirmed, 
                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
  ungroup() %>% 
  mutate(
    St_max = S_t^exp(1.96*S_t_SElog),
    St_min = S_t^exp(-1.96*S_t_SElog)
  )

order_key <-km_ct_residents_confirmed %>% 
  group_by(region_UK) %>% 
  summarise(S_t = min(S_t)) %>% 
  arrange(S_t)
km_ct_residents_confirmed$region_UK <- factor(
  km_ct_residents_confirmed$region_UK,
  levels = order_key$region_UK
)
km_ct_residents_confirmed$region_group <- cut(as.integer(km_ct_residents_confirmed$region_UK), 2, labels = 1:2)

ggplot(aes(x = date, y = 1-S_t, 
           ymax = 1 - St_max,
           ymin = 1 - St_min,
           color = region_UK), data = km_ct_residents_confirmed) +
  # facet_grid(region_group ~ .) +
  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
  geom_path(lwd=1.2) +
  scale_y_continuous("Cum probability confirmed case",
                     breaks = seq(0,1,.02)) +
  scale_color_discrete("Region") +
  scale_x_date(name = "Date", 
               breaks = unique(get_monday_date(km_ct_residents_confirmed$date)),
               labels = format.Date(
                 unique(get_monday_date(km_ct_residents_confirmed$date)),
                 "%d/%m/%y"
               )) +
  theme_bw()

```

`r table_nums("tab_CoxPH_confirmed")`

```{r CPH_confirmed, results="asis"}
explanatory = c("gender", "ageg", "home_type", "region_UK")
dependent = "Surv(time = start, time2 = stop, event = event)"
km_confirmed %>%
  finalfit(dependent, explanatory) -> CPH_confirmed
knitr::kable(CPH_confirmed, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

## Staff symptomatic and confirmed cases

```{r KM_Staff_Symptomatic, fig.height = 5, fig.width = 10}

staff_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
  left_join(distinct(new_cases, date, home_code, 
                     new_colleague_confirmed, new_colleague_symptomatic)) %>% 
  filter(!is.na(home_code)) %>% 
  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
  left_join(select(reference_geography, postcode, region_UK)) %>% 
  arrange(date)

km_staff_symptomatic <- staff_ct %>% 
  group_by(date, region_UK) %>% 
  summarise(
    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
    staff_total = sum(staff_total)
  ) %>% 
  group_by(region_UK) %>% 
  km_ct_estimator(., t=date, d=symptomatic, pop=staff_total,subtract_cases = TRUE) %>% 
  ungroup() %>% 
  mutate(
    St_max = S_t^exp(1.96*S_t_SElog),
    St_min = S_t^exp(-1.96*S_t_SElog)
  )

order_key <-km_staff_symptomatic %>% 
  group_by(region_UK) %>% 
  summarise(S_t = min(S_t)) %>% 
  arrange(S_t)
km_staff_symptomatic$region_UK <- factor(
  km_staff_symptomatic$region_UK,
  levels = order_key$region_UK
)
km_staff_symptomatic$region_group <- cut(as.integer(km_staff_symptomatic$region_UK), 2, labels = 1:2)

ggplot(aes(x = date, y = 1-S_t, 
           ymax = 1 - St_max,
           ymin = 1 - St_min,
           color = region_UK), data = km_staff_symptomatic) +
  # facet_grid(region_group ~ .) +
  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
  geom_path(lwd=1.2) +
  scale_color_discrete("Region") +  
  scale_y_continuous("Cum probability symptomatic case",
                     breaks = seq(0,1,.02)) +
  scale_x_date(name = "Date", 
               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
               labels = format.Date(
                 unique(get_monday_date(km_staff_symptomatic$date)),
                 "%d/%m/%y"
               )) +
  theme_bw()

```

```{r KM_Staff_Confirmed, fig.height = 7, fig.width = 10}
km_staff_confirmed <- staff_ct %>% 
  group_by(date, region_UK) %>% 
  summarise(
    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
    staff_total = sum(staff_total)
  ) %>% 
  group_by(region_UK) %>% 
  km_ct_estimator(., date, confirmed, staff_total) %>% 
  ungroup() %>% 
  mutate(
    St_max = S_t^exp(1.96*S_t_SElog),
    St_min = S_t^exp(-1.96*S_t_SElog)
  )

order_key <- km_staff_confirmed %>% 
  group_by(region_UK) %>% 
  summarise(S_t = min(S_t)) %>% 
  arrange(S_t)
km_staff_confirmed$region_UK <- factor(
  km_staff_confirmed$region_UK,
  levels = order_key$region_UK
)
km_staff_confirmed$region_group <- cut(as.integer(km_staff_confirmed$region_UK), 2, labels = 1:2)

ggplot(aes(x = date, y = 1-S_t, 
           ymax = 1 - St_max,#(S_t + 1.96 * S_t_SE),
           ymin = 1 - St_min, #(S_t - 1.96 * S_t_SE),
           color = region_UK), data = km_staff_confirmed) +
  facet_grid(region_group ~ .) +
  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
  geom_path(lwd=1.2) +
  scale_color_discrete("Region") +
  ylab("Pr confirmed case") +
  scale_x_date(name = "Date", 
               breaks = unique(get_monday_date(km_staff_confirmed$date)),
               labels = format.Date(
                 unique(get_monday_date(km_staff_confirmed$date)),
                 "%d/%m/%y"
               )) +
  theme_bw()

```

# Declarations

## Funding

This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].

## Research ethics

This study was approved by the University College London Research Ethics Committee (project reference 13355/002).

# References

### 2020-06-18 covidWarehouse/R/compute.R

@@ -5,23 +5,24 @@ compute_age <- function(dob, date){
     , "year")))
 }
 
-cut_age_denary <- function(age) {
-  # as.character(
+cut_age_denary <- function(age, age.min = 40, age.max = 130) {
+
     cut(
       x = age,
-      breaks = c(seq(30, 130, 10), 150),
+      breaks = c(0, seq(age.min, age.max, 10), 150),
       include.lowest = FALSE,
       right = FALSE,
-      labels = c(paste0(
-        seq(30, 129, 10), "-",
-        seq(39, 129, 10), " years"),
-        "130+ years")
-    )#)
+      labels = c(
+        paste0("<", age.min, " years"),
+        paste0(
+          seq(age.min, age.max-1, 10), "-",
+          seq(age.min+9, age.max-1, 10), " years"),
+        paste0(age.max, "+ years"))
+    )
 }
 
 
 cut_age_quinary <- function(age) {
-  as.character(
     cut(
       x = age,
       breaks = c(seq(30, 135, 5), 150),
@@ -31,11 +32,11 @@ cut_age_quinary <- function(age) {
         seq(30, 130, 5), "-",
         seq(34, 134, 5), " years"),
         "135+ years")
-    ))
+    )
 }
 
 cut_age_pentadecimal <- function(age) {
-  as.character(
+
     cut(
       x = age,
       breaks = c(seq(30, 135, 15), 150),
@@ -45,8 +46,8 @@ cut_age_pentadecimal <- function(age) {
         seq(30, 129, 15), "-",
         seq(44, 135, 15), " years"),
         "135+ years")
-    ))
-  }
+    )
+}
 
 tbrounding <- function(x, places = 1){
   format(round(x, places), nsmall = places)
@@ -97,7 +98,7 @@ min_non_missing <- function(x) {
 #' @importFrom lubridate %m+% days
 get_time_span <- function(data = incidents) {
   dplyr::summarise(data,
-            start = get_monday_date(min(incident_date)),
+            start = as.Date("2020-03-02"),
             end = get_monday_date(max(incident_date)) %m+% days(6)) %>%
     as.list
 
@@ -146,6 +147,7 @@ resident_timelines <- function(residents_data = residents,
 
   incidents_data <- dplyr::group_by(incidents_data, resident_id) %>%
     dplyr::summarise(
+      covid_first_symptomatic = min_non_missing(covid_first_symptomatic),
[…]

### 2020-06-18 covidWarehouse/R/database.R

@@ -349,6 +349,9 @@ extract_residents <- function() {
       " residents records")
   }
 
+  residents <- dplyr::filter(residents,
+                             !funding_type %in% c("Young Chronic Sick/YPD", "Younger Person"))
+
   save(residents, file = file.path(getOption("fshc_files"), "residents.rda"))
   usethis::ui_done("residents.rda")
 }

### 2020-06-18 covidWarehouse/R/stats.R

@@ -117,16 +117,25 @@ plot_funnel <- function(data, y, denom, home_code, home_name) {
 #' @param pop unquoted name of the population (inclusive of those who previously
 #' experienced the event, but removing those that are censored, e.g. those who
 #' left the study without experiencing the event)
+#' @param subtract_cases a boolean indicated whether events count `d` should be
+#' substracted from pop as time passes in order to calculate the total exposed-to-risk.
+#' The default is `TRUE` for when `pop` number includes person-days for past cases.
 #' @param overwrite a boolean indicating whether the function can overwrite variables
 #' `n_t`, `S_t` and `S_t_SE` in `df` (default is TRUE)
 #'
 #' @return the input data frame `df` with additional variables `n_t` for the population
 #' at risk size, `S_t` for the survivor function point estimator, and `S_t_SE` for its
-#' standard error
-#' @references Greenwood M, Jr. \em{The Natural Duration of Cancer}. Reports of Public Health
-#' and Related Subjects Vol 33, HMSO, London; 1926.
+#' standard error. A further `S_t_SElog` corresponding to an asymptotic estimator of the
+#' standard deviation, which can be used to caculate confident interval bounds like so:
+#' \itemize{
+#'   \item \code{St_min = S_t^exp(-1.96*S_t_SElog)}
+#'   \item \code{St_max = S_t^exp(+1.96*S_t_SElog)}
+#' }
+#'
+#' @references section 1.1.4 of: Kalbfleisch JD, Prentice RL. Statistical Analysis of
+#' Failure Time Data. New York: Wiley 1980.
 #' @export
-km_ct_estimator <- function(df, t, d, pop, overwrite = TRUE) {
+km_ct_estimator <- function(df, t, d, pop, subtract_cases = TRUE, overwrite = TRUE) {
 
   overwriting <- c("n_t", "S_t", "S_t_SE")[which(
     sapply(c("n_t", "S_t", "S_t_SE"),
@@ -141,18 +150,34 @@ km_ct_estimator <- function(df, t, d, pop, overwrite = TRUE) {
 
   }
 
-  dplyr::arrange(df, {{t}}) %>%
-    dplyr::mutate(
+  df <- dplyr::arrange(df, {{t}})
[…]

### 2020-06-18 covidWarehouse/tests/testthat/test-data.R

@@ -9,19 +9,21 @@ test_that("min_non_missing", {
   expect_equal(min_non_missing(c(NA, NA)), NA)
   expect_equal(min_non_missing(c(lubridate::ymd("2020-05-08"), as.Date(NA))),
                lubridate::ymd("2020-05-08"))
-  expect_equal(min_non_missing(c(as.Date(NA), as.Date(NA))), NA)
+  expect_equal(min_non_missing(c(as.Date(NA), as.Date(NA))), as.Date(NA))
 })
 
 
 test_that("age_groups", {
-  expect_equal(cut_age_denary(102), "100-109 years")
-  expect_equal(cut_age_denary(150), NA_character_)
+  expect_equal(as.character(cut_age_denary(29)), "<40 years")
 
-  expect_equal(cut_age_quinary(102), "100-104 years")
-  expect_equal(cut_age_quinary(150), NA_character_)
+  expect_equal(as.character(cut_age_denary(102)), "100-109 years")
+  expect_equal(as.character(cut_age_denary(150)), NA_character_)
 
-  expect_equal(cut_age_pentadecimal(102), "90-104 years")
-  expect_equal(cut_age_pentadecimal(150), NA_character_)
+  expect_equal(as.character(cut_age_quinary(102)), "100-104 years")
+  expect_equal(as.character(cut_age_quinary(150)), NA_character_)
+
+  expect_equal(as.character(cut_age_pentadecimal(102)), "90-104 years")
+  expect_equal(as.character(cut_age_pentadecimal(150)), NA_character_)
 
 
 

### 2020-06-18 covidWarehouse/tests/testthat/test-stats.R

@@ -29,7 +29,7 @@ km_indiv <- data.frame(km_indiv[1:8])
 km_indiv$pop <- km_indiv$n - cumsum(dplyr::lag(km_indiv$n.censor, 1, default = 0))
 km_indiv$n_t <- km_indiv$n
 
-expect_error(km_ct_estimator(km_indiv, time, n.event, pop, overwrite = FALSE))
+expect_error(km_ct_estimator(km_indiv, time, n.event, pop, subtract_cases = TRUE, overwrite = FALSE))
 output <- expect_warning(km_ct_estimator(km_indiv, time, n.event, pop, overwrite = TRUE))
 expect_equal(output$surv, output$S_t)
 expect_equal(output$std.err, output$S_t_SE)

### 2020-06-19 covidWarehouse/manuscript/descriptives.Rmd
manuscript version 0.1

@@
 output:
-  html_document:
-    df_print: paged
+  word_document:
     toc: yes
-    code_folding: hide
+    reference_docx: template.docx
   pdf_document:
@@
     toc: yes
-  word_document:
+  html_document:
+    df_print: paged
     toc: yes
-    reference_docx: template.docx
+    code_folding: hide
 link-citations: yes
@@
 ```{r descriptives}
+reference_homes <- reference_homes %>% 
+  left_join(select(filter(beds, year=="2020"), -year)) %>% 
+  mutate(staff_to_bed_ratio = staff_total/beds,
+         carer_to_bed_ratio = (reference_homes$staff_care_assistants + reference_homes$staff_nurses)/beds,
+         bed_tot_category = cut(.$beds, 
+           breaks = seq(20,100,15), right = F, 
+           paste0(seq(20, 80, 15), "-", seq(34, 94, 15), " beds")))
+
 homes_country <- select(reference_homes, home_code, postcode) %>% 
@@
   dplyr::filter(dplyr::between(incident_date, date_min, date_max)) #%>%
-  # dplyr::filter(!resident_id %in% 
-  #                 residents$resident_id[residents$funding_type %in% c("Young Chronic Sick/YPD", 
-  #                                                                     "Younger Person")])
 
@@
 stopifnot(unique(tline$date_end)>=date_max)
-tline <- filter(tline, between(date, date_min, date_max)) %>% 
-  filter(!funding_type %in% c("Young Chronic Sick/YPD", "Younger Person"))
+tline <- filter(tline, between(date, date_min, date_max))
 
-new_cases <- filter(new_cases, between(date, date_min, date_max_tallies))
+new_cases <- filter(new_cases, between(date, date_min, date_max_tallies)) %>% 
+    filter(home_code %in% unique(residents$home_code))
 
@@
 desc_homes <- reference_homes %>% 
-  left_join(transmute(reference_geography, postcode, 
-                      country, region_england = region), by = "postcode") %>% 
-  left_join(filter(beds, year=="2020"), by = "home_code") %>% 
+  filter(home_code %in% unique(residents$home_code)) %>% 
+  left_join(transmute(reference_geography, country, postcode), by = "postcode") %>% 
   left_join(res_counts, by = "home_code") %>%
@@
   group_by(Scope) %>% 
-  summarise(`Total homes` = n_distinct(home_code),
+  summarise(., 
+            `Total homes` = n_distinct(home_code),
             `Total beds` = sum(beds),
@@
 desc_overall$total_residents <- dplyr::n_distinct(desc_residents$resident_id)
+desc_overall$total_staff <- reference_homes %>% filter(home_code %in% unique(residents$home_code)) %>% .$staff_total %>% sum()
+
 #COUNTS
@@
   (desc_overall$CT_p_symptomatic*(1-desc_overall$CT_p_symptomatic)/desc_overall$total_residents)
+)
+desc_overall$CT_total_confirmed <-  sum(new_cases$new_confirmed_home, na.rm = T) +
+  sum(new_cases$new_confirmed_hospital, na.rm = T)
+desc_overall$CT_p_confirmed <- desc_overall$CT_total_confirmed/desc_overall$total_residents
+desc_overall$CT_p_confirmed_ME <- 1.96 * sqrt(
+  desc_overall$CT_p_confirmed * (1-desc_overall$CT_p_confirmed) / 
+    desc_overall$total_residents
 )
@@
 )
+
+home_total_confirmed_DAT <- filter(incidents, !is.na(covid_first_symptomatic)) %>% 
+  group_by(home_code) %>% 
+  summarise(DAT_confirmed = tidyr::replace_na(n_distinct(resident_id), 0))
+home_total_confirmed_CT <- new_cases %>% group_by(home_code) %>% 
+  summarise(CT_confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
+                                 tidyr::replace_na(new_confirmed_hospital, 0)))
+
+home_total_confirmed <- distinct(reference_homes, home_code, home_name_clickview, postcode) %>% 
+  left_join(select(reference_geography, postcode, country, region_UK), by = "postcode") %>% 
+  left_join(res_counts, by = "home_code") %>% 
+  left_join(home_total_confirmed_CT) %>% 
+  left_join(home_total_confirmed_DAT) %>% 
+  mutate(CT_confirmed = tidyr::replace_na(CT_confirmed, 0),
+        DAT_confirmed = tidyr::replace_na(DAT_confirmed, 0))
+
+home_zero_confirmed <- home_total_confirmed %>% 
+  group_by(region_UK) %>% 
+  summarise(`Total homes` = n(),
+            zero_datix = sum(DAT_confirmed == 0),
+            pct_datix = sum(DAT_confirmed == 0)/n(),
+            zero_CT = sum(CT_confirmed == 0) ,
+            pct_CT =  sum(CT_confirmed == 0)/n()) %>% 
+  arrange(-pct_datix)
+
+desc_overall$total_homes <- n_distinct(home_total_confirmed$home_code)
+desc_overall$CT_zero_case <- sum(home_total_confirmed$CT_confirmed == 0)
+desc_overall$DAT_zero_case <- sum(home_total_confirmed$DAT_confirmed == 0)
+
 desc_overall$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
@@
   dplyr::n_distinct(.$resident_id)
-desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>% 
-  dplyr::n_distinct(.$resident_id)
+desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>%   dplyr::n_distinct(.$resident_id)
 
@@
   caption = paste0("Breakdown of FSHCG residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
-
+tab_zero_cases <- table_nums(
+  name = "tab_zero_cases",
+  caption = "Number and proportion of homes without confirmed cases by UK region"
+)
 tab_CoxPH_confirmed <- table_nums(
   name = "tab_CoxPH_confirmed",
-  caption =  paste0("Hazard ratios of becoming a confirmed case estimated from a Cox Proportional Hazards model (n=", 
+  caption =  paste0("Factors associated with the hazard of become a confirmed case: hazard ratios (HR) from a Cox proportional hazards model (n=", 
   formatbm(dplyr::n_distinct(desc_residents$resident_id)),
@@
 )
-fig_KM_confirmed_residents <- fig_nums(
-  name = 'fig_KM_confirmed_residents',
+fig_KM_residents_confirmed <- fig_nums(
+  name = 'fig_KM_residents_confirmed',
   caption = paste0("Kaplan-Meier estimator of the cumulative probability of residents (n=",
@@
 
+fig_KM_staff_symptomatic <- fig_nums(
+  name = 'fig_KM_staff_symptomatic',
+  caption = paste0("Kaplan-Meier estimator of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  ") becoming symptomatic cases based on manager-reported daily counts"))
+fig_KM_staff_confirmed <- fig_nums(
+  name = 'fig_KM_staff_confirmed',
+  caption = paste0("Kaplan-Meier estimator of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  ") becoming confirmed cases based on manager-reported daily counts"))
+
+```
+
+```{r period_prevalence}
+cis_data <- list(
+  n_positive = 66,
+  n_negative = 60618
+)
+
+england_results <- list()
+england_results$n_positive = sum(inner_join(new_cases, filter(homes_country, country == "England"))[c("new_confirmed_home", "new_confirmed_hospital")], na.rm = T)
+england_results$n_negative = n_distinct(inner_join(residents, filter(homes_country, country == "England"))$resident_id) - england_results$n_positive
+
+RR_england <- (
+  england_results$n_positive / (england_results$n_positive + england_results$n_negative)
+  ) / (
+  cis_data$n_positive / (cis_data$n_positive + cis_data$n_negative)
+)
+
+RR_SE <- sqrt(
+  1/england_results$n_positive  + 
+    1/cis_data$n_positive +
+    1/(england_results$n_positive + england_results$n_negative) +
+    1/(cis_data$n_positive + cis_data$n_negative)
+)
 ```
@@
 
-**Background**
+**Background** COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes.
+Scarce evidence is available on epidemiological trends in this setting.
+This study reports on data obtained from the Four Seasons Health Care Group (FSHCG), an independent provider of long-term and respite residential and nursing care representing 9% of all beds in England, Scotland and Northern Ireland.
 
-**Methods:**
+**Methods:** Secondary use datasets were obtained from FSHCG.
+Individual-level records on residents and COVID-19 related events are investigated to estimate the incidence and period prevalence of symptoms, confirmed case, and death.
+Findings are compared with aggregate counts reported daily by care home managers to FSHCG as well as with findings from a national community survey.
 
-**Findings:** A total of `r formatbm(n_distinct(desc_residents$resident_id))` residents were examined across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)` of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+**Findings:** Records of `r formatbm(n_distinct(desc_residents$resident_id))` residents present in FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")` were analysed.
+The period prevalence of COVID-19 symptoms is comprised between`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
+Across the period, managers registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`]) cases confirmed with a naropharyngeal RNA test.
+Datix incidents reported just `r formatbm(desc_overall$CT_total_confirmed)` confirmed cases.
+In England, `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
+Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
+The staff-to-bed ratio was the only factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
 
-**Interpretation**
+**Interpretation** Despite uncertain case ascertainment in care homes, internal surveillance suggests a greater transmission than in the community.
+Staffing may impact care homes' response to outbreaks and in turn influence rates of confirmed cases.
+The absence of high-quality case ascertainment hinders local infection prevention initiatives which could be better supported by laboratory data surveillance.
 
@@
 
-COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes where one in seven individuals aged > 85 years live.
+COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes where one in seven individuals aged 85 years and over live.
 However, there is no surveillance for infection in care homes, nor are there systems (or research studies) monitoring the impact of the pandemic on individuals or systems.
@@
 In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`)).
-FSHCG residents accessed nose and throat swab antigen tests during hospitalisations in the first instance, but tests became available in care homes from XXXXX.
+FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
 From 11 May 2020, FSHCG home undertook testing every resident.
@@
 
-These counts include events relating to occupants of local authority contract beds.
+These counts include events relating to occupants of local authority contract beds as well as occupants receiving a different category of support (eg.
+young people) who were excluded from the study population.
 
@@
 
-- figure for the cumulative number/proportion of staff and residents infected over time...
-this is another area of significant interest so we can get an estimate (with caveats) of the proportion of staff and residents who remain susceptible to infection.
--> KM by age or care home type or nation/region
-- how many care homes have not had any cases reported in staff or residents during the pandemic because these homes will presumably be at greatest risk in a second wave.
-- individual and care home characteristics associated with high number of cases.
-Nationally the only 2 factors that have been proposed so far are care home size and provision of nursing care.
--> **regression model on RR of infection** number of previous cases (lag effect) is important here.
-So the time effect
+The incidence of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020 [@cisbulletin2020; @cisdata2020].
 
@@
 The vast majority (85·4%) were permanent residents.
-A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_overall$unique_residents_incidents)` unique residents.
+A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported on Datix as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_overall$unique_residents_incidents)` unique residents.
+The reports were returned by `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes  totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
 
-Datix reports and manager counts respectively indicated that `r formatbm(desc_overall$DAT_total_symptomatic)` (`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% [95% confidence interval: `r tbrounding((desc_overall$DAT_p_symptomatic - desc_overall$DAT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$DAT_p_symptomatic + desc_overall$DAT_p_symptomatic_ME)*100)`]) and `r formatbm(desc_overall$CT_total_symptomatic)` (`r tbrounding(desc_overall$CT_p_symptomatic*100)`% [`r tbrounding((desc_overall$CT_p_symptomatic - desc_overall$CT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$CT_p_symptomatic + desc_overall$CT_p_symptomatic_ME)*100)`]) became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
+Datix reports and manager counts respectively indicated that `r formatbm(desc_overall$DAT_total_symptomatic)` (`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% [95% confidence interval: `r tbrounding((desc_overall$DAT_p_symptomatic - desc_overall$DAT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$DAT_p_symptomatic + desc_overall$DAT_p_symptomatic_ME)*100)`]) and `r formatbm(desc_overall$CT_total_symptomatic)` (`r tbrounding(desc_overall$CT_p_symptomatic*100)`% [`r tbrounding((desc_overall$CT_p_symptomatic - desc_overall$CT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$CT_p_symptomatic + desc_overall$CT_p_symptomatic_ME)*100)`]) residents became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
 Temporal trends broken down by region (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`) indicate wide variation across the UK, with care homes in the South East and Northern Ireland reporting in excess of a third of residents with symptoms.
@@
 
-`r fig_nums("fig_KM_CT_residents_symptomatic")`
-
-```{r KM_CT_residents_symptomatic, fig.height = 5, fig.width = 10}
+```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 7}
 
@@
   geom_path(lwd=1.2) +
-  scale_y_continuous("Cum probability confirmed case",
+  scale_y_continuous("Cum probability symptomatic case",
                      breaks = seq(0,1,.05)) +
@@
                )) +
-  theme_bw()
+  theme_bw()  +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
 
@@
 
+`r fig_nums("KM_CT_residents_symptomatic")`
+
 ## Resident confirmed cases
 
-The reports were returned by `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes  totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
-Datix incidents accounted for `r formatbm(desc_overall$unique_residents_tested)` residents tested at least once.
-Test results were known for `r formatbm(desc_overall$unique_residents_test_result)` residents: `r formatbm(desc_overall$unique_residents_test_result_positive)` residents had a record of a positive test result.
-This data cannot be treated as representative, as engagement with FSHCG confirmed that test coming back negative were not commonly recorded on Datix.
+Datix incidents reports accounted for `r formatbm(desc_overall$unique_residents_tested)` residents tested at least once.
+Test results were known for `r formatbm(desc_overall$unique_residents_test_result)` of them (`r tbrounding(desc_overall$unique_residents_test_result/desc_overall$unique_residents_tested*100)`%), including `r formatbm(desc_overall$unique_residents_test_result_positive)` residents with a positive test result.
+This data should be treated as representative, as tests coming back negative are not commonly recorded on Datix.
 Care home manager daily counts of new confirmed cases added up to `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` cases tested in homes and `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` tested in hospitals.
-The regional breakdown is presented in
+The temporal trend of confirmed cases broken down by region is presented using both Datix reports (`r fig_nums("fig_KM_residents_confirmed", display = "cite")`) and manager counts (`r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`).
 
-The temporal trend of confirmed cases obtained from either Datix reports (`r fig_nums("fig_KM_confirmed_residents", display = "cite")`) or manager counts (`r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`)
+For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
+In comparison, the survey of English community households @cisbulletin2020 found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
+This implies a relative risk of `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
 
-`r fig_nums("fig_KM_confirmed_residents")`
-
-```{r km_confirmed, fig.cap=fig_KM_confirmed_residents}
+```{r km_confirmed, fig.height=4,fig.width=8}
 km_confirmed <- tline %>% 
-  left_join(select(reference_homes, home_code, postcode)) %>% 
+  left_join(select(reference_homes, home_code, postcode, 
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
   left_join(select(reference_geography, postcode, region_UK)) %>% 
@@
   mutate(home_type = factor(home_type, levels = c("Residential", "Nursing"))) %>% 
-  group_by(resident_id, dob, gender, region_UK,
-           home_type, user_type, funding_type, admission_type) %>% 
+  group_by(resident_id, dob, gender, region_UK, home_code,
+           home_type, user_type, funding_type, admission_type, 
+           staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category) %>% 
   summarise(
@@
            fun = "event",
-           legend = c(0.16, 0.6),
+           legend = "right",
            ggtheme = theme_bw(),
@@
 
-`r fig_nums("fig_KM_confirmed_residents")`
+`r fig_nums("fig_KM_residents_confirmed")`
 
-```{r KM_CT_residents_infection, fig.height = 5, fig.width = 10}
+```{r KM_CT_residents_infection, fig.height = 4, fig.width = 7}
 
@@
                )) +
-  theme_bw()
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
 
 ```
+
+`r fig_nums("fig_KM_CT_confirmed_residents")`
+
+Between `r desc_overall$DAT_zero_case` (`r tbrounding( sum(home_zero_confirmed$zero_datix)/ sum(home_zero_confirmed[["Total homes"]])*100)`%) and  `r desc_overall$CT_zero_case` (`r tbrounding(sum(home_zero_confirmed$zero_CT)/sum(home_zero_confirmed[["Total homes"]])*100)`%) care homes had not accounted any confirmed cases whether in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
+Yet the significant conflict between the two sources of data makes it difficult to ascertain which care homes had not had any confirmed cases without querying the national laboratory information systems.
+
+`r table_nums("tab_zero_cases")`
+
+```{r tab_zero_cases, results="asis"}
+home_zero_confirmed %>% 
+  transmute(`UK region/nation` = region_UK,
+            `Total homes`,
+            `Homes with no Datix confirmed case` = paste0(zero_datix, " (", 
+                                          tbrounding( pct_datix*100), "%)"),
+            `Homes with no manager-reported case` = paste0(zero_CT, " (", 
+                                          tbrounding( pct_CT*100), "%)")) %>% 
+  bind_rows(
+    tibble(`UK region/nation` = "Total",
+           `Total homes` = sum(home_zero_confirmed$`Total homes`),
+           `Homes with no Datix confirmed case` = paste0(
+             sum(home_zero_confirmed$zero_datix), " (", 
+             tbrounding( sum(home_zero_confirmed$zero_datix) / 
+                           sum(home_zero_confirmed$`Total homes`)*100), "%)"),
+            `Homes with no manager-reported case` = paste0(
+             sum(home_zero_confirmed$zero_CT), " (", 
+             tbrounding( sum(home_zero_confirmed$zero_CT) / 
+                           sum(home_zero_confirmed$`Total homes`)*100), "%)"))
+  ) %>% kable(align = c("l", "r", "r", "r", "r"))
+
+```
+
+Using the individual-level data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+According to this, and other than geographical location, the staff-to-bed ratio was the only factor significantly associated with the rate of confirmed cases.
+An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 13% reduction [9%; 18%] in the rate of confirmed cases.
 
@@
 ```{r CPH_confirmed,   results="asis"}
-explanatory = c("gender", "ageg", "home_type", "region_UK")
+# coxph(Surv(time = start, time2 = stop, event = event) ~ gender + ageg+home_type + region_UK+carer_to_bed_ratio+bed_tot_category, data = km_confirmed) %>% summary()
+explanatory = c("gender", "ageg", "home_type", "region_UK" , 
+                "staff_to_bed_ratio", "bed_tot_category")
 dependent = "Surv(time = start, time2 = stop, event = event)"
-km_confirmed %>%
-  finalfit(dependent, explanatory) -> CPH_confirmed
+CPH_confirmed <-  finalfit(km_confirmed, dependent, 
+                           explanatory, column = T,
+                           add_dependent_label = F)  
 knitr::kable(CPH_confirmed, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
@@
 
-```{r KM_Staff_Symptomatic, fig.height = 5, fig.width = 10}
+```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 7}
 
@@
   scale_y_continuous("Cum probability symptomatic case",
-                     breaks = seq(0,1,.02)) +
+                     breaks = seq(0,1,.05)) +
   scale_x_date(name = "Date", 
@@
                )) +
-  theme_bw()
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
 
@@
 
-```{r KM_Staff_Confirmed, fig.height = 7, fig.width = 10}
+`r fig_nums("fig_KM_staff_symptomatic")`
+
+```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 7}
 km_staff_confirmed <- staff_ct %>% 
@@
            color = region_UK), data = km_staff_confirmed) +
-  facet_grid(region_group ~ .) +
+  # facet_grid(region_group ~ .) +
   geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
@@
   scale_color_discrete("Region") +
-  ylab("Pr confirmed case") +
+  scale_y_continuous("Cum probability confirmed case",
+                     breaks = seq(0,1,.01)) +
   scale_x_date(name = "Date", 
@@
                )) +
-  theme_bw()
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
 
 ```
+
+`r fig_nums("fig_KM_staff_confirmed")`
+
+# Interpretation
+
+Such data have limitations for care and prevention rather than epidemiological surveillance.
+Beyond uncertain on the exact number of cases, the viral antigen test campaign is a likely subject of underascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
+Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
+According to a meta-analysis of 7 studies @Kucirka2020, the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
+This factor appears to be one of the reasons why nursing homes have a higher rate of confirmed cases; a difference that is no longer significant once the staff staff_to_bed_ratio is present in the model.
+
+Consequently findings presented in the current study are likely to be underestimates, particularly estimates of the incidence staff_to_bed_ratio comparing care homes with the community in England.
 

### 2020-06-19 covidWarehouse/R/database.R

@@ -480,6 +480,11 @@ extract_tallies <- function() {
   if(!file.exists(file.path(getOption("fshc_files"), "reference_homes.rda"))) {
     extract_reference_homes()
   }
+  if(!file.exists(file.path(getOption("fshc_files"), "residents.rda"))) {
+    extract_residents()
+  }
+
+  load(file.path(getOption("fshc_files"), "residents.rda"))
 
   files <- list.files(file.path(getOption("FSHC_EXTRACTS_DIRECTORY"), "tallies"), full.names = T)
   new_filepath <- sort(grep("tally_new_cases.*csv$", files, value = T), decreasing = T)[1]
@@ -491,12 +496,12 @@ extract_tallies <- function() {
   names(new_cases) <- tolower(names(new_cases))
   new_cases$date <- lubridate::dmy(new_cases$date)
   new_cases$time <- lubridate::dmy_hm(new_cases$time)
-  new_cases <- dplyr::filter(new_cases, !home_code %in% c("LAH", "SIL"))
+  new_cases <- dplyr::filter(new_cases, home_code %in% residents$home_code)
 
   names(total_cases) <- tolower(names(total_cases))
   total_cases$date <- lubridate::dmy(total_cases$date)
   total_cases$time <- lubridate::dmy_hm(total_cases$time)
-  total_cases <- dplyr::filter(total_cases, !home_code %in% c("LAH", "SIL"))
+  total_cases <- dplyr::filter(total_cases, home_code %in% residents$home_code)
 
   # Merge Gilmerton NCC within 565	Gilmerton
   # Rid of Holybourne Day Centre (closed)
@@ -507,8 +512,8 @@ extract_tallies <- function() {
   new_cases$home_code[new_cases$home_code == "GIL"] <- "565"
   new_cases <- dplyr::filter(new_cases, home_code != "240")
 
-  check_homes(unique(total_cases$home_code))
-  check_homes(unique(new_cases$home_code))
+  check_homes(unique(total_cases$home_code), call = "`total_cases`")
+  check_homes(unique(new_cases$home_code), call = "`new_cases`")
 
   save(total_cases, file = file.path(getOption("fshc_files"), "total_cases.rda"))
   usethis::ui_done("total_cases.rda")

### 2020-06-24 covidWarehouse/manuscript/manuscript.Rmd
add mortality model

--- 2020-06-19 covidWarehouse/manuscript/descriptives.Rmd
@@
 ---
-title: "Incidence and period prevalence of confirmed COVID-19 cases in care home staff and residents: Cohort study of secondary use records in a large care home provider in England, Scotland and Northern Ireland (March-June 2020)"
+title: "Incidence and period prevalence of confirmed COVID-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
 # author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
@@
 bibliography: references.bib
-csl: sage-vancouver.csl
+csl: elsevier-vancouver.csl
 output:
-  word_document:
-    toc: yes
-    reference_docx: template.docx
-  pdf_document:
-    citation_package: natbib
-    df_print: kable
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
-    toc: yes
   html_document:
@@
     code_folding: hide
+  word_document:
+    toc: yes
+    reference_docx: template.docx
 link-citations: yes
@@
 ```{r descriptives}
+
+tline <- dplyr::bind_rows(timelines$timeline)
+stopifnot(unique(tline$date_end)>=date_max)
+tline <- filter(tline, between(date, date_min, date_max)) %>% 
+  mutate(user_type2 = case_when(
+    user_type %in% c("General", "Elderly") ~ "General/elderly",
+    user_type == "Dementia" ~ "Dementia",
+    TRUE ~ NA_character_ ))
+
+home_inclusion <- unique(tline$home_code)
+
 reference_homes <- reference_homes %>% 
+  filter(home_code %in% home_inclusion) %>% 
   left_join(select(filter(beds, year=="2020"), -year)) %>% 
   mutate(staff_to_bed_ratio = staff_total/beds,
-         carer_to_bed_ratio = (reference_homes$staff_care_assistants + reference_homes$staff_nurses)/beds,
+         carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
          bed_tot_category = cut(.$beds, 
-           breaks = seq(20,100,15), right = F, 
-           paste0(seq(20, 80, 15), "-", seq(34, 94, 15), " beds")))
+           breaks = seq(20,100,25), right = F, 
+           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds")))
 
@@
 
+reference_homes <- left_join(reference_homes, homes_country)
+
 incidents <- incidents %>% 
-  dplyr::filter(dplyr::between(incident_date, date_min, date_max)) #%>%
+  filter(dplyr::between(incident_date, date_min, date_max)) %>% 
+  filter(home_code %in% home_inclusion)
 
@@
 
-tline <- dplyr::bind_rows(timelines$timeline)
-stopifnot(unique(tline$date_end)>=date_max)
-tline <- filter(tline, between(date, date_min, date_max))
-
 new_cases <- filter(new_cases, between(date, date_min, date_max_tallies)) %>% 
-    filter(home_code %in% unique(residents$home_code))
+    filter(home_code %in% home_inclusion)
 
@@
   tibble(date = seq(date_min, date_max, 1)),
-  tibble(home_code = unique(reference_homes$home_code)),
+  tibble(home_code = home_inclusion),
   all = TRUE) %>% 
@@
   group_by(resident_id, home_code, dob, gender, 
-           admission_type, home_type, user_type, status) %>% 
+           admission_type, home_type, user_type2, status) %>% 
   summarise(resident_days = sum(rday),
@@
             susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
-  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 50, age.max = 100)) %>% 
+  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
   left_join(homes_country, by = "home_code")
@@
 desc_homes <- reference_homes %>% 
-  filter(home_code %in% unique(residents$home_code)) %>% 
-  left_join(transmute(reference_geography, country, postcode), by = "postcode") %>% 
   left_join(res_counts, by = "home_code") %>%
@@
   ")"))
-
+tab_CoxPH_death <- table_nums(
+  name = "tab_CoxPH_death",
+  caption =  paste0("Factors associated with all-cause mortality: hazard ratios (HR) from a Cox proportional hazards model (n=", 
+  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  ")"))
 ## FIGURES
@@
 fig_KM_CT_residents_symptomatic <- fig_nums(
-  name = "KM_CT_residents_symptomatic",
-  caption = paste0("Kaplan-Meier estimator of the cumulative probability of residents becoming symptomatic based on care home manager daily counts")
+  name = "fig_KM_CT_residents_symptomatic",
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of residents becoming symptomatic based on care home manager daily counts")
 )
@@
   name = 'fig_KM_residents_confirmed',
-  caption = paste0("Kaplan-Meier estimator of the cumulative probability of residents (n=",
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of residents (n=",
   formatbm(dplyr::n_distinct(desc_residents$resident_id)),
@@
   name = 'fig_KM_CT_confirmed_residents',
-  caption = "Kaplan-Meier estimator of the cumulative probability of residents becoming confirmed cases based on care home manager daily counts")
-
+  caption = "Kaplan-Meier estimates of the cumulative probability of residents becoming confirmed cases based on care home manager daily counts")
+fig_KM_mortality <- fig_nums(
+  "fig_KM_mortality",
+  caption = paste0("Kaplan-Meier estimates of resident (n=",
+  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  ") survival by SARS-COV-2 case type"))
 fig_KM_staff_symptomatic <- fig_nums(
   name = 'fig_KM_staff_symptomatic',
-  caption = paste0("Kaplan-Meier estimator of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
   ") becoming symptomatic cases based on manager-reported daily counts"))
@@
   name = 'fig_KM_staff_confirmed',
-  caption = paste0("Kaplan-Meier estimator of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
   ") becoming confirmed cases based on manager-reported daily counts"))
@@
 
-RR_england <- (
-  england_results$n_positive / (england_results$n_positive + england_results$n_negative)
-  ) / (
-  cis_data$n_positive / (cis_data$n_positive + cis_data$n_negative)
-)
+RR_england <- relative_risk(
+  a = england_results$n_positive,
+  b = england_results$n_negative,
+  c = cis_data$n_positive,
+  d = cis_data$n_negative)
 
-RR_SE <- sqrt(
-  1/england_results$n_positive  + 
-    1/cis_data$n_positive +
-    1/(england_results$n_positive + england_results$n_negative) +
-    1/(cis_data$n_positive + cis_data$n_negative)
-)
+RR_SE <- relative_risk_SE(
+  a = england_results$n_positive,
+  b = england_results$n_negative,
+  c = cis_data$n_positive,
+  d = cis_data$n_negative)
 ```
@@
 
-**Background** COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes.
-Scarce evidence is available on epidemiological trends in this setting.
-This study reports on data obtained from the Four Seasons Health Care Group (FSHCG), an independent provider of long-term and respite residential and nursing care representing 9% of all beds in England, Scotland and Northern Ireland.
+**Background** COVID-19 has caused substantial morbidity and mortality in care home residents and staff.
+National surveys suggest infection is declining in the general population in the UK, but outbreaks persist in care homes.
+Epidemiological data on disease trends in care home staff and residents are scarce.
+We used administrative data from the Four Seasons Health Care Group (FSHCG), a national care home chain representing 9% of all beds in England, Scotland and Northern Ireland.
 
-**Methods:** Secondary use datasets were obtained from FSHCG.
-Individual-level records on residents and COVID-19 related events are investigated to estimate the incidence and period prevalence of symptoms, confirmed case, and death.
-Findings are compared with aggregate counts reported daily by care home managers to FSHCG as well as with findings from a national community survey.
+**Methods:** Retrospective cohort study using individual-level electronic records from residents (incidents reports) and daily care home counts of infection for both residents and staff.
+We estimated the incidence and period prevalence of symptoms, confirmed infection, and death in residents and staff, and compare them with findings from a national community survey.
+We tested a range of risk factors for resident infections in multivariate regression models.
 
-**Findings:** Records of `r formatbm(n_distinct(desc_residents$resident_id))` residents present in FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")` were analysed.
-The period prevalence of COVID-19 symptoms is comprised between`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
-Across the period, managers registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`]) cases confirmed with a naropharyngeal RNA test.
-Datix incidents reported just `r formatbm(desc_overall$CT_total_confirmed)` confirmed cases.
-In England, `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
+**Findings:** Records were available for `r formatbm(n_distinct(desc_residents$resident_id))` residents present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
+The period prevalence of COVID-19 symptoms ranged between `r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
+Across the period, care homes registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]) infections confirmed with naropharyngeal RNA tests.
+Incident reports accounted just `r formatbm(desc_overall$unique_residents_test_result_positive)` confirmed infections.
+In England, `r england_results$n_positive` infections were confirmed of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
 Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
@@
 
-COVID-19 causes significant mortality in elderly and vulnerable people and spreads easily in care homes where one in seven individuals aged 85 years and over live.
-However, there is no surveillance for infection in care homes, nor are there systems (or research studies) monitoring the impact of the pandemic on individuals or systems.
-Usual practices are disrupted during the pandemic, and care home staff are taking on new and unfamiliar roles, such as advanced care planning.
-Understanding the nature of these changes is critical to mitigate the impact of COVID-19 on residents, relatives and staff.
+Care home residents are extremely vulnerable to infection due to their advanced age and high prevalence of comorbidity, and this is compounded by frequent exposure to infection, for example through contact with infected staff members.
+At the peak of the pandemic the number of deaths in UK care home residents was threefold higher than the equivalent period in 2019,[@onsDR2020] but estimates of infection incidence and prevalence over this time period are lacking due to delays in establishing universal testing for care home residents and staff.
 
-This study reports on data obtained from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care.
+For example, at the start of the pandemic, testing for SARS-CoV-2 infection testing using reverse transcriptase polymerase chain reaction (RT-PCR) was only routinely available for care home residents or staff who were admitted to hospital.
+Testing was also performed in care homes that notified an outbreak to Public Health England, but this was limited to a maximum of five test per care home.
+Widespread testing for residents first became available on 11 May 2020 through the whole care home testing programme.
+[@dhscTesting2020] Consequently national estimates of incidence and prevalence based on test results obtained prior to 11 May will substantially underestimate the burden of infection in care home residents.
+
+An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020] One such investigation estimated that 50% of residents and 20% of staff are infected during an outbreak in London, UK.[@Easter6] Estimates of this nature cannot be generalised to the unknown proportion of care homes that have not reported any outbreaks.
+
+Information on disease incidence and prevalence, and how this varies by care home is essential to inform the public health response to future waves of infection, and in particular decisions around the need for widespread universal testing in care home staff and residents.
+
+This study obtained data from `r formatbm(dplyr::n_distinct(residents$home_code))` care homes from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care.
+It estimated incidence and period prevalence of COVID-19 events.
+In addition, it tested and identified individual and organisational risk factors associated with the rate of confirmed infections.
+
+# Methods
+
 In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`)).
-FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
-From 11 May 2020, FSHCG home undertook testing every resident.
-***(Haydn, more detail would be greatly appreciated)***
+As of 22 June 2020, 90% of FSHCG care homes have participated in the whole care home testing programme.
 
-`r table_nums("tab_national_descriptives")` @cqcdata2020,@hscni2020,@isd2020
+FSHCG collects administrative data on their residents through a variety of electronic systems that are primarily used for billing purposes.
+This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as "Datix".
+
+`r table_nums("tab_national_descriptives")` [@cqcdata2020; @hscni2020; @isd2020]
 
@@
 
-# Methods
-
 ## Individual-level residents data
 
-Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of:
+Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of two datasets:
 
@@
 
-Daily counts reported by care home managers to FSHCG were obtained for every care home:
+We obtained care home organisational characteristics: staff headcounts, bed counts, ***latest Care Quality Commission rating***, and types of care available (nursing dementia; nursing general; residential dementia; residential general; younger person).
+
+On 24 March 2020, FSHCG introduced an additional system to monitor COVID-19 cases in each care home, requiring care home managers to report daily tallies of suspected and confirmed infection in staff and residents to their regional manager.
+Daily counts reported by care homes to FSHCG were obtained for every care home:
 
@@
 
-We estimate the incidence of symptomatic and confirmed cases as well as deaths using the individual-level data.
-Doing so enables us to accurately account for the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
-In order to study the effect of potential risk factors, total cases and resident-days were aggregated by care home by care type and demographic characteristics and regressed in a Poisson hierarchical level with care home and time independent and identically distributed random effects.
-This provides measures of relative risks.
+We used individual-level incident reports and daily tallies to generate estimates of disease incidence in residents and period prevalence in residents and staff.
+Death status as of `r format(date_max, "%e %B %Y")` was used to estimate the case-fatality rate and the relative risks of all-cause death in confirmed cases .
+The individual data provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
+A Cox proportional hazards model was used to test the effect of potential risk factors, whether individual- (sex, age, care type) or organisation-level (region, care home size, staff-to-bed ratio).
+
+We also investigated whether the number of confirmed infections among staff was associated with subsequent infections amongst residents, independently of time. For this, we regressed.
 
@@
 
-The incidence of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020 [@cisbulletin2020; @cisdata2020].
+The incidence of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
 
-Data were analysed in `R`3·5·0 @RCoreTeam2018 and the `survival` package @therneau2015. All computer syntax is available on an online repository @Dutey2020.
+Data were analysed in `R`3·5·0[@RCoreTeam2018] and the `survival` package.[@therneau2015] All computer syntax is available on an online repository.[@Dutey2020]
 
-# Results
+# Findings
 
@@
 arsenal::tableby(home_type  ~ gender + ageg + region_UK +
-                                 user_type + admission_type + status, 
+                                 user_type2 + admission_type + status, 
                                data = desc_residents) %>% summary()
@@
 
-`r fig_nums("KM_CT_residents_symptomatic")`
+`r fig_nums("fig_KM_CT_residents_symptomatic")`
 
 ## Resident confirmed cases
+
+FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
+From 11 May 2020, most FSHCG home undertook testing every resident as part of the whole home testing programme.[@dhscTesting2020]
 
@@
 For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
-In comparison, the survey of English community households @cisbulletin2020 found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
 This implies a relative risk of `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
@@
 ```{r km_confirmed, fig.height=4,fig.width=8}
+
 km_confirmed <- tline %>% 
@@
   group_by(resident_id, dob, gender, region_UK, home_code,
-           home_type, user_type, funding_type, admission_type, 
+           home_type, user_type2, funding_type, admission_type, 
            staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category) %>% 
@@
     stop = as.numeric(end +1 - date_min),
-    ageg = cut_age_denary(compute_age(dob, date_min), age.min = 50, age.max = 100)
+    ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)
   )
@@
 Between `r desc_overall$DAT_zero_case` (`r tbrounding( sum(home_zero_confirmed$zero_datix)/ sum(home_zero_confirmed[["Total homes"]])*100)`%) and  `r desc_overall$CT_zero_case` (`r tbrounding(sum(home_zero_confirmed$zero_CT)/sum(home_zero_confirmed[["Total homes"]])*100)`%) care homes had not accounted any confirmed cases whether in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
-Yet the significant conflict between the two sources of data makes it difficult to ascertain which care homes had not had any confirmed cases without querying the national laboratory information systems.
+The substantial lack of agreement between the two sources of data makes it difficult to ascertain which care homes had not had any confirmed cases.
 
@@
 ```{r CPH_confirmed,   results="asis"}
-# coxph(Surv(time = start, time2 = stop, event = event) ~ gender + ageg+home_type + region_UK+carer_to_bed_ratio+bed_tot_category, data = km_confirmed) %>% summary()
-explanatory = c("gender", "ageg", "home_type", "region_UK" , 
+# cph_conf <- coxph(Surv(time = start, time2 = stop, event = event) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category, data = km_confirmed)
+# summary(cph_conf)
+# cox.zph(cph_conf)
+# survfit(Surv(time = start, time2 = stop, event = event)~factor(home_type), data = km_confirmed) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
+explanatory = c("gender", "ageg", "home_type", "user_type2", "region_UK" ,  
                 "staff_to_bed_ratio", "bed_tot_category")
@@
                            add_dependent_label = F)  
-knitr::kable(CPH_confirmed, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
+knitr::kable(CPH_confirmed,
+             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
 ```
+
+## All-cause mortality
+
+```{r km_death}
+km_death <- tline %>% 
+  filter(rday==1) %>% 
+  mutate(home_type = factor(home_type, c("Residential", "Nursing"))) %>% 
+  group_by(resident_id, home_code, dob, gender, 
+           status, home_type, user_type2,
+           susceptible_symptomatic, 
+           susceptible_confirmed) %>% 
+  summarise(entry_date = min(date),
+            exit_date = max(date+1)) %>% 
+  mutate(date1 = as.numeric( entry_date - date_min),
+         date2 = as.numeric(exit_date - date_min),
+         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
+  mutate(symptomatic = -susceptible_symptomatic+1,
+         confirmed = -susceptible_confirmed+1,
+         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
+         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
+  mutate(covid_status = case_when(
+    symptomatic==0 & confirmed == 0 ~ "Negative",
+    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
+    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
+    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
+  )) %>% 
+  mutate(covid_status = factor(covid_status, 
+                               levels = c("Negative", "Symptomatic (not confirmed)",
+                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
+  left_join(select(reference_homes, home_code, region_UK,
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
+  group_by(resident_id) %>% 
+  mutate(death_status = if_else(
+    date1 == max(date1),
+    as.integer(status == "Deceased"),
+    0L
+  )) %>% 
+  ungroup()   
+# km_death <- bind_cols(km_death, data.frame(model.matrix(~-1+covid_status, data = km_death)))
+```
+
+The crude risk of all-cause death for the entire population of study was 19·4% (`r table_nums("tab_resident_descriptives", display = "cite")`).
+Time-varying covariate Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
+Mortality increased with age and men's rate of death was 47% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
+
+After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 3.6 [2·2; 5·9] and 8·3 [6·9; 10·0], confirming that symptomatic cases were more severe.
+As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 5·2 [4·3; 6·3].
+There was strong evidence that hazards were not proportional across these categories -- a Kaplan-Meier estimator of the unadjusted survival function illustrates this (`r fig_nums("fig_KM_mortality", display="cite")`).
+Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
+The 28% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
+This effect is likely attributable to a greater burden of comorbidities in nursing home residents.
+Beyond, there was no evidence of an association between care home size and all-cause mortality.
+After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·6; 1.0]).
+
+`r table_nums("tab_CoxPH_death")`
+
+```{r risk_death, results="asis"}
+# cph_death <- coxph(Surv(date1, date2, death_status) ~
+#                       covid_status + gender+ ageg+ home_type +
+#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+
+#                       cluster(resident_id), data = km_death)
+# summary(cph_death)
+# cox.zph(cph_death)
+# 
+explanatory = c("gender", "ageg", "home_type" , 
+                "staff_to_bed_ratio", "bed_tot_category", 
+                "covid_status", "region_UK", "cluster(resident_id)")
+dependent = "Surv(time = date1, time2 = date2, event = death_status)"
+CPH_death <-  finalfit(km_death, dependent, explanatory, 
+                       column = T, add_dependent_label = F)  
+knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
+
+```
+
+```{r fig_KM_mortality}
+survfit(Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + cluster(resident_id), data = km_death) %>% 
+  ggsurvplot(., conf.int = TRUE, 
+           legend = "right",
+           ggtheme = theme_bw(),
+           break.x.by	= 7,
+           break.y.by = .1,
+           legend.labs = levels(km_death$covid_status),
+           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
+           ylab = "Survival",
+           conf.int.alpha = .1) 
+```
+*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevatn Datix incident report.
+
+`r fig_nums("fig_KM_mortality")`
 
@@
 Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
-According to a meta-analysis of 7 studies @Kucirka2020, the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
-This factor appears to be one of the reasons why nursing homes have a higher rate of confirmed cases; a difference that is no longer significant once the staff staff_to_bed_ratio is present in the model.
+According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
+This factor appears to be one of the reasons why nursing homes have a higher rate of confirmed cases; a difference that is no longer significant once the staff-to-bed ratio is present in the model.
 
-Consequently findings presented in the current study are likely to be underestimates, particularly estimates of the incidence staff_to_bed_ratio comparing care homes with the community in England.
+Consequently findings presented in the current study are likely to be underestimates, particularly estimates of the incidence ratio comparing care homes with the community in England.
 

### 2020-06-24 covidWarehouse/R/compute.R

@@ -15,7 +15,7 @@ cut_age_denary <- function(age, age.min = 40, age.max = 130) {
       labels = c(
         paste0("<", age.min, " years"),
         paste0(
-          seq(age.min, age.max-1, 10), "-",
+          seq(age.min, age.max-1, 10), "–",
           seq(age.min+9, age.max-1, 10), " years"),
         paste0(age.max, "+ years"))
     )
@@ -29,24 +29,26 @@ cut_age_quinary <- function(age) {
       include.lowest = FALSE,
       right = FALSE,
       labels = c(paste0(
-        seq(30, 130, 5), "-",
+        seq(30, 130, 5), "–",
         seq(34, 134, 5), " years"),
         "135+ years")
     )
 }
 
-cut_age_pentadecimal <- function(age) {
+cut_age_pentadecimal <- function(age, age.min = 40, age.max = 130) {
 
-    cut(
-      x = age,
-      breaks = c(seq(30, 135, 15), 150),
-      include.lowest = FALSE,
-      right = FALSE,
-      labels = c(paste0(
-        seq(30, 129, 15), "-",
-        seq(44, 135, 15), " years"),
-        "135+ years")
-    )
+  cut(
+    x = age,
+    breaks = c(0, seq(age.min, age.max, 15), 150),
+    include.lowest = FALSE,
+    right = FALSE,
+    labels = c(
+      paste0("<", age.min, " years"),
+      paste0(
+        seq(age.min, age.max-1, 15), "–",
+        seq(age.min+14, age.max-1, 15), " years"),
+      paste0(age.max, "+ years"))
+  )
 }
 
 tbrounding <- function(x, places = 1){

### 2020-06-24 covidWarehouse/R/database.R

@@ -294,7 +294,7 @@ extract_residents <- function() {
   residents_src <- clean_variable_names(residents_src)
   # Removing Jersey
   residents_src <- dplyr::rename(residents_src, home_code = home_id)
-  residents_src <- dplyr::filter(residents_src, !home_code %in% c("LAH", "SIL"))
+  residents_src <- dplyr::filter(residents_src, !home_code %in% c("365", "LAH", "SIL"))
 
   residents <- residents_src
   for(i in grep("(_date)|(date_of)|(dob)", names(residents))){
@@ -336,6 +336,12 @@ extract_residents <- function() {
   residents$admission_type <- trimws(residents$admission_type)
   residents$funding_type <- trimws(residents$funding_type)
   residents <- merge(residents, home_types, all.x = TRUE)
+  residents <- dplyr::mutate(residents,
+                             gender = dplyr::case_when(
+                               trimws(gender) == "M" ~ "Male",
+                               trimws(gender) == "F" ~ "Female",
+                               TRUE ~ NA_character_
+                             ))
 
   residents <- dplyr::distinct(residents) %>%
     dplyr::rename(resident_id = encrypted_id)

### 2020-06-24 covidWarehouse/R/stats.R

@@ -182,3 +182,12 @@ km_ct_estimator <- function(df, t, d, pop, subtract_cases = TRUE, overwrite = TR
 
 }
 
+
+
+relative_risk <- function(a, b, c, d) {
+  (a/b)/(c/d)
+}
+
+relative_risk_SE <- function(a, b, c, d){
+  sqrt(1/a + 1/c + 1/(a + b) + 1/(c + d))
+}

### 2020-06-24 covidWarehouse/tests/testthat/test-data.R

@@ -16,13 +16,13 @@ test_that("min_non_missing", {
 test_that("age_groups", {
   expect_equal(as.character(cut_age_denary(29)), "<40 years")
 
-  expect_equal(as.character(cut_age_denary(102)), "100-109 years")
+  expect_equal(as.character(cut_age_denary(102)), "100–109 years")
   expect_equal(as.character(cut_age_denary(150)), NA_character_)
 
-  expect_equal(as.character(cut_age_quinary(102)), "100-104 years")
+  expect_equal(as.character(cut_age_quinary(102)), "100–104 years")
   expect_equal(as.character(cut_age_quinary(150)), NA_character_)
 
-  expect_equal(as.character(cut_age_pentadecimal(102)), "90-104 years")
+  expect_equal(as.character(cut_age_pentadecimal(102)), "100–114 years")
   expect_equal(as.character(cut_age_pentadecimal(150)), NA_character_)
 
 

### 2020-06-24 covidWarehouse/manuscript/manuscript.Rmd
setting max_date back to 2020-06-07

@@
 ---
-title: "Incidence and period prevalence of confirmed COVID-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
+title: "Incidence of confirmed COVID-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
 # author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
@@
 date_min <- as.Date("2020-03-02")
-date_max <- as.Date("2020-06-15")
+date_max <- as.Date("2020-06-07") #as.Date("2020-06-15")
 date_max_tallies = as.Date("2020-06-07")
@@
 tline <- filter(tline, between(date, date_min, date_max)) %>% 
-  mutate(user_type2 = case_when(
+  mutate(home_type = factor(home_type, c("Residential", "Nursing"))) %>% 
+  mutate(
+    user_type2 = case_when(
       user_type %in% c("General", "Elderly") ~ "General/elderly",
       user_type == "Dementia" ~ "Dementia",
-    TRUE ~ NA_character_ ))
+      TRUE ~ NA_character_ 
+      ),
+    admission_type2 = case_when(
+      admission_type %in% c("Continuing Care", "Independent Living") ~ "Continuing care/independent living", 
+      TRUE ~ admission_type
+    ))
 
@@
   select(-postcode)
-
 reference_homes <- left_join(reference_homes, homes_country)
+rm(homes_country)
 
@@
     susceptible_confirmed = sum(susceptible_confirmed, na.rm = T))
-
 approx_occupancy <- merge(
@@
 
-desc_residents <- tline %>% 
-  filter(rday == 1)  %>% 
+residents_episodes <- tline %>% 
+  filter(rday==1) %>% 
   group_by(resident_id, home_code, dob, gender, 
-           admission_type, home_type, user_type2, status) %>% 
+           status, home_type, user_type2,
+           susceptible_symptomatic, 
+           susceptible_confirmed) %>% 
+  summarise(entry_date = min(date),
+            exit_date = max(date+1),
+            resident_days = sum(rday),
+            susceptible_symptomatic_days = sum(susceptible_symptomatic),
+            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
+  mutate(date1 = as.numeric( entry_date - date_min),
+         date2 = as.numeric(exit_date - date_min),
+         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
+  mutate(symptomatic = -susceptible_symptomatic+1,
+         confirmed = -susceptible_confirmed+1,
+         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
+         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
+  mutate(covid_status = case_when(
+    symptomatic==0 & confirmed == 0 ~ "Negative",
+    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
+    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
+    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
+  )) %>% 
+  mutate(covid_status = factor(covid_status, 
+                               levels = c("Negative", "Symptomatic (not confirmed)",
+                                          "Asymptomatic confirmed", "Symptomatic confirmed")),
+         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
+  left_join(select(reference_homes, home_code, country, region_UK,
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
+  group_by(resident_id) %>% 
+  mutate(death_status = if_else(
+    date1 == max(date1),
+    as.integer(status == "Deceased"),
+    0L
+  )) %>% 
+  ungroup()
+
+residents_overview <- tline %>% 
+  filter(rday==1) %>%
+  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
+  group_by(resident_id, home_code, ageg, gender, 
+           admission_type2, home_type, user_type2, status) %>% 
   summarise(resident_days = sum(rday),
             susceptible_symptomatic_days = sum(susceptible_symptomatic),
-            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
-  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
-  left_join(homes_country, by = "home_code")
+            susceptible_confirmed_days = sum(susceptible_confirmed),
+            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
+            confirmed = max(-susceptible_confirmed+1, na.rm = T),
+            start_date = min(date, na.rm = T),
+            stop_date = max(date, na.rm = T),
+            start_int = as.numeric(min(date, na.rm = T) - date_min),
+            stop_int = as.numeric(max(date, na.rm = T)- date_min)
+    ) %>% 
+  mutate(covid_status = case_when(
+    symptomatic==0 & confirmed == 0 ~ "Negative",
+    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
+    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
+    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
+  )) %>% 
+  mutate(covid_status = factor(covid_status, 
+                               levels = c("Negative", "Symptomatic (not confirmed)",
+                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
+  left_join(select(reference_homes, home_code, country, region_UK,
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
+  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
 
-res_counts <- desc_residents %>% 
+res_counts <- residents_overview %>% 
   group_by(home_code) %>% 
@@
 desc_overall <- list()
-desc_overall$total_residents <- dplyr::n_distinct(desc_residents$resident_id)
-desc_overall$total_staff <- reference_homes %>% filter(home_code %in% unique(residents$home_code)) %>% .$staff_total %>% sum()
+desc_overall$total_residents <- dplyr::n_distinct(residents_overview$resident_id)
+desc_overall$total_staff <- reference_homes %>% filter(home_code %in% unique(residents_overview$home_code)) %>% .$staff_total %>% sum()
 
@@
     desc_overall$total_residents
+)
+
+desc_overall$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
+desc_overall$unique_residents_tested <- dplyr::filter(incidents, covid_tested == 1) %>% 
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+desc_overall$unique_residents_test_result <- dplyr::filter(incidents, !is.na(covid_test_result)) %>% 
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>%  
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+
+desc_overall$DAT_p_confirmed <- desc_overall$unique_residents_test_result_positive/desc_overall$total_residents
+desc_overall$DAT_p_confirmed_ME <- 1.96 * sqrt(
+  desc_overall$DAT_p_confirmed * (1-desc_overall$DAT_p_confirmed) / 
+    desc_overall$total_residents
+)
+
+desc_overall$DAT_incidence_confirmed <- desc_overall$unique_residents_test_result_positive/sum(residents_overview$susceptible_confirmed_days)
+desc_overall$DAT_incidence_confirmed_SE <- sqrt((1-desc_overall$DAT_incidence_confirmed)/desc_overall$unique_residents_test_result_positive)
+desc_overall$DAT_incidence_confirmed_95low <- exp(
+  log(desc_overall$DAT_incidence_confirmed - 1.96 * desc_overall$DAT_incidence_confirmed_SE)
+)
+desc_overall$DAT_incidence_confirmed_95up <- exp(
+  log(desc_overall$DAT_incidence_confirmed + 1.96 * desc_overall$DAT_incidence_confirmed_SE)
 )
@@
 
-desc_overall$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
-desc_overall$unique_residents_tested <- dplyr::filter(incidents, covid_tested == 1) %>% 
-  dplyr::n_distinct(.$resident_id)
-desc_overall$unique_residents_test_result <- dplyr::filter(incidents, !is.na(covid_test_result)) %>% 
-  dplyr::n_distinct(.$resident_id)
-desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>%   dplyr::n_distinct(.$resident_id)
-
 desc_homes <- bind_rows(desc_homes, desc_national)
@@
   caption =  paste0("Factors associated with the hazard of become a confirmed case: hazard ratios (HR) from a Cox proportional hazards model (n=", 
-  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
   ")"))
@@
   caption =  paste0("Factors associated with all-cause mortality: hazard ratios (HR) from a Cox proportional hazards model (n=", 
-  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
   ")"))
@@
   caption = paste0("Kaplan-Meier estimates of the cumulative probability of residents (n=",
-  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
   ") becoming confirmed cases based on Datix incidents report"))
@@
   caption = paste0("Kaplan-Meier estimates of resident (n=",
-  formatbm(dplyr::n_distinct(desc_residents$resident_id)),
+  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
   ") survival by SARS-COV-2 case type"))
@@
 england_results <- list()
-england_results$n_positive = sum(inner_join(new_cases, filter(homes_country, country == "England"))[c("new_confirmed_home", "new_confirmed_hospital")], na.rm = T)
-england_results$n_negative = n_distinct(inner_join(residents, filter(homes_country, country == "England"))$resident_id) - england_results$n_positive
+england_results$n_positive = sum(inner_join(new_cases, filter(reference_homes, country == "England"))[c("new_confirmed_home", "new_confirmed_hospital")], na.rm = T)
+england_results$n_negative = n_distinct(inner_join(residents, filter(reference_homes, country == "England"))$resident_id) - england_results$n_positive
 
@@
 **Methods:** Retrospective cohort study using individual-level electronic records from residents (incidents reports) and daily care home counts of infection for both residents and staff.
-We estimated the incidence and period prevalence of symptoms, confirmed infection, and death in residents and staff, and compare them with findings from a national community survey.
+We estimated the incidence proportion and incidence rate of symptoms, confirmed infection, and death in residents and staff, and compare them with findings from a national community survey.
 We tested a range of risk factors for resident infections in multivariate regression models.
 
-**Findings:** Records were available for `r formatbm(n_distinct(desc_residents$resident_id))` residents present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
-The period prevalence of COVID-19 symptoms ranged between `r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
-Across the period, care homes registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]) infections confirmed with naropharyngeal RNA tests.
+**Findings:** Records were available for `r formatbm(n_distinct(residents_overview$resident_id))` residents present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
+The incidence proportion of COVID-19 symptoms ranged between `r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
+Across the period, care homes registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]) infections confirmed with nasopharyngeal RNA tests.
 Incident reports accounted just `r formatbm(desc_overall$unique_residents_test_result_positive)` confirmed infections.
@@
 **Interpretation** Despite uncertain case ascertainment in care homes, internal surveillance suggests a greater transmission than in the community.
-Staffing may impact care homes' response to outbreaks and in turn influence rates of confirmed cases.
-The absence of high-quality case ascertainment hinders local infection prevention initiatives which could be better supported by laboratory data surveillance.
+Staffing may impact care homes' response to outbreaks and in turn influence rates of confirmed infections The absence of high-quality case ascertainment hinders local infection prevention initiatives which could be better supported by laboratory data surveillance.
 
@@
 
-An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020] One such investigation estimated that 50% of residents and 20% of staff are infected during an outbreak in London, UK.[@Easter6] Estimates of this nature cannot be generalised to the unknown proportion of care homes that have not reported any outbreaks.
+An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020; @Dora2020] One such investigation estimated that 50% of residents and 20% of staff are infected during an outbreak in London, UK.[@Easter6] Estimates of this nature cannot be generalised to the unknown proportion of care homes that have not reported any outbreaks.
 
@@
 This study obtained data from `r formatbm(dplyr::n_distinct(residents$home_code))` care homes from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care.
-It estimated incidence and period prevalence of COVID-19 events.
+It estimated incidence of SARS-CoV-2 symptoms, diagnoses and all-cause mortality.
 In addition, it tested and identified individual and organisational risk factors associated with the rate of confirmed infections.
@@
 
-We used individual-level incident reports and daily tallies to generate estimates of disease incidence in residents and period prevalence in residents and staff.
-Death status as of `r format(date_max, "%e %B %Y")` was used to estimate the case-fatality rate and the relative risks of all-cause death in confirmed cases .
+We used individual-level incident reports and daily tallies to generate estimates of infection incidence in residents and staff.
 The individual data provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
 A Cox proportional hazards model was used to test the effect of potential risk factors, whether individual- (sex, age, care type) or organisation-level (region, care home size, staff-to-bed ratio).
+Dates of death were used to estimate the case-fatality rate, as well as compare all-cause mortality based on the presence of symptoms and infection confirmation.
+A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
 
-We also investigated whether the number of confirmed infections among staff was associated with subsequent infections amongst residents, independently of time. For this, we regressed.
+We also investigated whether the number of confirmed infections among staff was associated with subsequent infections amongst residents, independently of time.
+For this, a Poisson model regressed the rate of confirmed resident cases against home occupancy levels as well as lagged rates of confirmed staff infections at care home level.
 
-However, a Datix report may not have been filed for every event.
-We therefore also estimate the incidence and period prevalence of events using manager counts, using the total resident-days denominator derived from the individual-level data.
+A Datix report may not have been filed for every event.
+We therefore also estimate the incidence of events using manager counts.
+The resident-days denominator was computed by subtracting past confirmed cases counts from resident-days estimated from the individual-level data.
 
-The incidence of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
+The incidence proportion of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
 
@@
 arsenal::tableby(home_type  ~ gender + ageg + region_UK +
-                                 user_type2 + admission_type + status, 
-                               data = desc_residents) %>% summary()
+   user_type2 + admission_type2 + covid_status + status +
+   resident_days + susceptible_symptomatic_days + susceptible_confirmed_days, 
+   control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
+   data = residents_overview) %>% summary()
 ```
 
-```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 7}
+```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 8}
 
@@
 
+The overall incidence proportion was between `r tbrounding(desc_overall$DAT_p_confirmed*100)`% [`r tbrounding((desc_overall$DAT_p_confirmed - desc_overall$DAT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$DAT_p_confirmed + desc_overall$DAT_p_confirmed_ME)*100)`%] according to incidents reports and  `r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]  according to manager counts.
 For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
@@
 
-```{r km_confirmed, fig.height=4,fig.width=8}
-
-km_confirmed <- tline %>% 
-  left_join(select(reference_homes, home_code, postcode, 
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
-  left_join(select(reference_geography, postcode, region_UK)) %>% 
-  filter(rday == 1 & susceptible_confirmed == 1) %>% 
-  mutate(home_type = factor(home_type, levels = c("Residential", "Nursing"))) %>% 
-  group_by(resident_id, dob, gender, region_UK, home_code,
-           home_type, user_type2, funding_type, admission_type, 
-           staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category) %>% 
-  summarise(
-    start = as.numeric(min(date) - date_min),
-    end = max(date)
-  ) %>% 
-  left_join(select(timelines, resident_id, covid_first_positive)) %>% 
-  mutate(event = if_else(
-    is.na(covid_first_positive),
-    0L,
-    as.integer(covid_first_positive == end)
-  )) %>% 
-  mutate(
-    stop = as.numeric(end +1 - date_min),
-    ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)
-  )
-
-survfit(Surv(time = start, 
-                   time2 = stop,
-                   event = event) ~ region_UK, data = km_confirmed) %>% 
+```{r km_confirmed, fig.height=4,fig.width=8, warning=F}
+survfit(Surv(time = start_int, 
+             time2 = stop_int + 1,
+             event = confirmed) ~ region_UK, data = residents_overview) %>% 
   ggsurvplot(., conf.int = TRUE, 
@@
            break.x.by	= 7,
-           break.y.by = .02,
-           ylim = c(0, .17),
+           break.y.by = .01,
+           ylim = c(0, .08),
            xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
            ylab = "Cum probability of confirmed case",
-           legend.labs = sort(unique(km_confirmed$region_UK)),
+           legend.labs = sort(unique(residents_overview$region_UK)),
            conf.int.alpha = .1) 
-
 ```
@@
 
-```{r KM_CT_residents_infection, fig.height = 4, fig.width = 7}
+```{r KM_CT_residents_infection, fig.height = 4, fig.width = 8}
 
@@
 ```{r CPH_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start, time2 = stop, event = event) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category, data = km_confirmed)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category, data = residents_overview)
 # summary(cph_conf)
 # cox.zph(cph_conf)
-# survfit(Surv(time = start, time2 = stop, event = event)~factor(home_type), data = km_confirmed) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
+# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
 explanatory = c("gender", "ageg", "home_type", "user_type2", "region_UK" ,  
                 "staff_to_bed_ratio", "bed_tot_category")
-dependent = "Surv(time = start, time2 = stop, event = event)"
-CPH_confirmed <-  finalfit(km_confirmed, dependent, 
+dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
+CPH_confirmed <-  finalfit(residents_overview,
+                           dependent, 
                            explanatory, column = T,
@@
 
-```{r km_death}
-km_death <- tline %>% 
-  filter(rday==1) %>% 
-  mutate(home_type = factor(home_type, c("Residential", "Nursing"))) %>% 
-  group_by(resident_id, home_code, dob, gender, 
-           status, home_type, user_type2,
-           susceptible_symptomatic, 
-           susceptible_confirmed) %>% 
-  summarise(entry_date = min(date),
-            exit_date = max(date+1)) %>% 
-  mutate(date1 = as.numeric( entry_date - date_min),
-         date2 = as.numeric(exit_date - date_min),
-         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
-  mutate(symptomatic = -susceptible_symptomatic+1,
-         confirmed = -susceptible_confirmed+1,
-         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
-         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
-  mutate(covid_status = case_when(
-    symptomatic==0 & confirmed == 0 ~ "Negative",
-    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
-    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
-    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
-  )) %>% 
-  mutate(covid_status = factor(covid_status, 
-                               levels = c("Negative", "Symptomatic (not confirmed)",
-                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
-  left_join(select(reference_homes, home_code, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
-  group_by(resident_id) %>% 
-  mutate(death_status = if_else(
-    date1 == max(date1),
-    as.integer(status == "Deceased"),
-    0L
-  )) %>% 
-  ungroup()   
-# km_death <- bind_cols(km_death, data.frame(model.matrix(~-1+covid_status, data = km_death)))
-```
+The crude risk of all-cause death for the entire population of study was 19·5% (`r table_nums("tab_resident_descriptives", display = "cite")`).
+Time-varying covariate Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
+Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
 
-The crude risk of all-cause death for the entire population of study was 19·4% (`r table_nums("tab_resident_descriptives", display = "cite")`).
-Time-varying covariate Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
-Mortality increased with age and men's rate of death was 47% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
-
-After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 3.6 [2·2; 5·9] and 8·3 [6·9; 10·0], confirming that symptomatic cases were more severe.
-As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 5·2 [4·3; 6·3].
+After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 2·6 [1·6; 4·3] and 9·9 [8·3; 11·9], confirming that symptomatic cases were more severe.
+As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 6·2 [5·1; 7·4].
 There was strong evidence that hazards were not proportional across these categories -- a Kaplan-Meier estimator of the unadjusted survival function illustrates this (`r fig_nums("fig_KM_mortality", display="cite")`).
 Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
-The 28% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
+The 27% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
 This effect is likely attributable to a greater burden of comorbidities in nursing home residents.
 Beyond, there was no evidence of an association between care home size and all-cause mortality.
-After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·6; 1.0]).
+After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·7; 1.1]).
 
@@
 ```{r risk_death, results="asis"}
-# cph_death <- coxph(Surv(date1, date2, death_status) ~
-#                       covid_status + gender+ ageg+ home_type +
-#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+
-#                       cluster(resident_id), data = km_death)
+cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+                      gender+ ageg +
+                      staff_to_bed_ratio + bed_tot_category +(region_UK)+ 
+                      home_type*covid_status+ cluster(resident_id), data = residents_episodes)
 # summary(cph_death)
 # cox.zph(cph_death)
+# survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
 # 
-explanatory = c("gender", "ageg", "home_type" , 
+explanatory = c("gender", "ageg", "user_type2", "home_type",
                 "staff_to_bed_ratio", "bed_tot_category", 
                 "covid_status", "region_UK", "cluster(resident_id)")
+explanatory_interaction = c("gender", "ageg", "user_type2",
+                "staff_to_bed_ratio", "bed_tot_category", 
+                "covid_status:home_type", "region_UK", "cluster(resident_id)")
 dependent = "Surv(time = date1, time2 = date2, event = death_status)"
-CPH_death <-  finalfit(km_death, dependent, explanatory, 
+CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
                        column = T, add_dependent_label = F)  
@@
 
-```{r fig_KM_mortality}
-survfit(Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + cluster(resident_id), data = km_death) %>% 
+```{r fig_KM_mortality,  fig.height = 4, fig.width = 8}
+survfit(Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + cluster(resident_id), data = residents_episodes) %>% 
   ggsurvplot(., conf.int = TRUE, 
@@
            break.y.by = .1,
-           legend.labs = levels(km_death$covid_status),
+           legend.labs = levels(residents_episodes$covid_status),
            xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
@@
 ```
-*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevatn Datix incident report.
+
+*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevant Datix incident report.
 
@@
 
-```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 7}
+```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 8}
 
@@
 
-```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 7}
+```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 8}
 km_staff_confirmed <- staff_ct %>% 
@@
 
-Such data have limitations for care and prevention rather than epidemiological surveillance.
-Beyond uncertain on the exact number of cases, the viral antigen test campaign is a likely subject of underascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
+This study was to our knowledge the first to report on the incidence of SARS-CoV-2 infections in UK care homes on a large scale, using individal-level resident records from a large provider network.
+Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r formatbm(desc_overall$CT_p_symptomatic)`% of residents experiencing symptoms at some stage and `r `r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%] having a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+Regional variation was considerable, but the cross-examination of from incident reports and manager counts produces inconsistent findings, suggesting strong variation in recording practices.
+Rates of testing across time and care homes were not yet known at the time of writing and likely distorts measures of incidence.
+
+One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline (***Haydn, please explain***) was a protective factor.
+A 10 percentage point increase was associated with a 13% reduction [9%; 18%] in hazards of confirmed infections.
+Few organisational variables were available to identify potential confounders or mediators of this association.
+One hypothesis is that staffing ratios may be a sign of better-resourced care homes, and of a greater capability to prescribe and observe effective infection control procedures.
+Furthermore, staff can become a vector in the transmission of infections.
+It is possible that higher staffing levels mitigate the transmission of infections since each staff member will care for a lower number of residents.
+Better staffed homes may also be less reliant on bank or agency staff who work in different locations.
+A more detailed survey of care home characteristics, along with protocols and measures adopted during the epidemic may help elucidate mechanisms through which staffing levels are associated with lower infection rates.
+
+staff infection rates
+
+Similar to infection rates, all-cause mortality exhibited strong regional variation. Nursing home mortality was 76% higher, and . I
+
+Uncertainty remains regarding the internal validity of estimates presented in this report.
+Beyond uncertain on the exact number of cases, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
 Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
 According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
-This factor appears to be one of the reasons why nursing homes have a higher rate of confirmed cases; a difference that is no longer significant once the staff-to-bed ratio is present in the model.
+Consequently findings presented in the current study are likely to be underestimates.
+This includes estimates of the incidence ratio comparing care homes with the community in England.
 
-Consequently findings presented in the current study are likely to be underestimates, particularly estimates of the incidence ratio comparing care homes with the community in England.
+Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection.[@Dora2020] Although increasing testing capacity raises the possibility of undertaking frequent, repeat testing in the care home population, this may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
+Understanding which care homes are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention accordingly.
 

### 2020-06-24 covidWarehouse/R/compute.R

@@ -225,12 +225,12 @@ resident_timelines <- function(residents_data = residents,
         dplyr::mutate(
           susceptible_symptomatic = dplyr::case_when(
             rday == 0 ~ 0L,
-            date > covid_first_symptomatic ~ 0L,
+            date >= covid_first_symptomatic ~ 0L,
             TRUE ~ 1L
           ),
           susceptible_confirmed = dplyr::case_when(
             rday == 0 ~ 0L,
-            date > covid_first_positive ~ 0L,
+            date >= covid_first_positive ~ 0L,
             TRUE ~ 1L
           ))
 

### 2020-06-24 covidWarehouse/R/database.R

@@ -227,6 +227,14 @@ extract_incidents <- function() {
         infection_confirmed == "Yes - positive test result: confirmed case" ~ 1L,
         infection_confirmed == "No - negative test result" ~ 0L,
         TRUE ~ NA_integer_
+      ),
+      covid_test_date_corrected = dplyr::case_when(
+        covid_tested == 0 ~ as.Date(NA),
+        lubridate::year(infection_covid_19_test_date) != 2020 ~ as.Date(
+          paste0("2020",
+                 substr(infection_covid_19_test_date, 5, 10))),
+        is.na(infection_covid_19_test_date) ~ incident_date,
+        TRUE ~ infection_covid_19_test_date
       )
     ) %>%
     dplyr::mutate(
@@ -252,12 +260,17 @@ extract_incidents <- function() {
                 as.Date(NA))),
       covid_first_positive = min_non_missing(
         if_else(covid_test_result == 1,
-                incident_date,
+                # incident_date,
+                # 24/06 trying the actual test date
+                covid_test_date_corrected,
                 as.Date(NA))),
       covid_first_tested = min_non_missing(
         if_else(covid_tested == 1,
-                incident_date,
+                # incident_date,
+                # 24/06 trying the actual test date
+                covid_test_date_corrected,
                 as.Date(NA)))
+
     ) %>%
     dplyr::mutate(
       covid_ever_symptomatic = if_else(

### 2020-06-25 covidWarehouse/manuscript/manuscript.Rmd
manuscript version 0.2

@@
 
-```{r descriptives}
+```{r init_data}
 
@@
     ) %>% 
+  ungroup() %>% 
   mutate(covid_status = case_when(
@@
   left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
+
+# CARE HOME AGGREGATES
+home_total_confirmed_DAT <- filter(incidents, !is.na(covid_first_symptomatic)) %>% 
+  group_by(home_code) %>% 
+  summarise(DAT_confirmed = tidyr::replace_na(n_distinct(resident_id), 0))
+home_total_confirmed_CT <- new_cases %>% group_by(home_code) %>% 
+  summarise(CT_confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
+                                 tidyr::replace_na(new_confirmed_hospital, 0)))
 
@@
 
-desc_overall <- list()
-desc_overall$total_residents <- dplyr::n_distinct(residents_overview$resident_id)
-desc_overall$total_staff <- reference_homes %>% filter(home_code %in% unique(residents_overview$home_code)) %>% .$staff_total %>% sum()
-
-#COUNTS
-desc_overall$CT_total_symptomatic <- sum(new_cases$new_symptomatic_home, na.rm=T) + 
-  sum(new_cases$new_symptomatic_hospital, na.rm=T)
-desc_overall$CT_p_symptomatic <- desc_overall$CT_total_symptomatic/desc_overall$total_residents
-desc_overall$CT_p_symptomatic_ME <- 1.96 * sqrt(
-  (desc_overall$CT_p_symptomatic*(1-desc_overall$CT_p_symptomatic)/desc_overall$total_residents)
-)
-desc_overall$CT_total_confirmed <-  sum(new_cases$new_confirmed_home, na.rm = T) +
-  sum(new_cases$new_confirmed_hospital, na.rm = T)
-desc_overall$CT_p_confirmed <- desc_overall$CT_total_confirmed/desc_overall$total_residents
-desc_overall$CT_p_confirmed_ME <- 1.96 * sqrt(
-  desc_overall$CT_p_confirmed * (1-desc_overall$CT_p_confirmed) / 
-    desc_overall$total_residents
-)
-#INCIDENT REPORTS
-desc_overall$unique_residents_incidents <- dplyr::n_distinct(incidents$resident_id)
-
-desc_overall$DAT_total_symptomatic <- n_distinct(filter(incidents, !is.na(covid_first_symptomatic))$resident_id)
-desc_overall$DAT_p_symptomatic <- desc_overall$DAT_total_symptomatic/desc_overall$total_residents
-desc_overall$DAT_p_symptomatic_ME <- 1.96 * sqrt(
-  desc_overall$DAT_p_symptomatic * (1 - desc_overall$DAT_p_symptomatic) / 
-    desc_overall$total_residents
-)
-
-desc_overall$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
-desc_overall$unique_residents_tested <- dplyr::filter(incidents, covid_tested == 1) %>% 
-  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
-desc_overall$unique_residents_test_result <- dplyr::filter(incidents, !is.na(covid_test_result)) %>% 
-  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
-desc_overall$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>%  
-  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
-
-desc_overall$DAT_p_confirmed <- desc_overall$unique_residents_test_result_positive/desc_overall$total_residents
-desc_overall$DAT_p_confirmed_ME <- 1.96 * sqrt(
-  desc_overall$DAT_p_confirmed * (1-desc_overall$DAT_p_confirmed) / 
-    desc_overall$total_residents
-)
-
-desc_overall$DAT_incidence_confirmed <- desc_overall$unique_residents_test_result_positive/sum(residents_overview$susceptible_confirmed_days)
-desc_overall$DAT_incidence_confirmed_SE <- sqrt((1-desc_overall$DAT_incidence_confirmed)/desc_overall$unique_residents_test_result_positive)
-desc_overall$DAT_incidence_confirmed_95low <- exp(
-  log(desc_overall$DAT_incidence_confirmed - 1.96 * desc_overall$DAT_incidence_confirmed_SE)
-)
-desc_overall$DAT_incidence_confirmed_95up <- exp(
-  log(desc_overall$DAT_incidence_confirmed + 1.96 * desc_overall$DAT_incidence_confirmed_SE)
-)
-
-home_total_confirmed_DAT <- filter(incidents, !is.na(covid_first_symptomatic)) %>% 
-  group_by(home_code) %>% 
-  summarise(DAT_confirmed = tidyr::replace_na(n_distinct(resident_id), 0))
-home_total_confirmed_CT <- new_cases %>% group_by(home_code) %>% 
-  summarise(CT_confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
-                                 tidyr::replace_na(new_confirmed_hospital, 0)))
-
 home_total_confirmed <- distinct(reference_homes, home_code, home_name_clickview, postcode) %>% 
@@
 
-desc_overall$total_homes <- n_distinct(home_total_confirmed$home_code)
-desc_overall$CT_zero_case <- sum(home_total_confirmed$CT_confirmed == 0)
-desc_overall$DAT_zero_case <- sum(home_total_confirmed$DAT_confirmed == 0)
-
 desc_homes <- bind_rows(desc_homes, desc_national)
 
+```
+
+```{r km_objects}
+
+residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
+  left_join(distinct(new_cases, date, home_code, 
+                     new_confirmed_home, new_confirmed_hospital)) %>% 
+  filter(!is.na(home_code)) %>% 
+  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
+  left_join(select(reference_geography, postcode, region_UK)) %>% 
+  left_join(approx_occupancy) %>% 
+  arrange(date)
+
+km_ct_residents_confirmed <- residents_ct %>% 
+  group_by(date, region_UK) %>% 
+  summarise(
+    confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
+                      tidyr::replace_na(new_confirmed_hospital, 0)),
+    occupancy = sum(tidyr::replace_na(susceptible_confirmed, 0))
+  ) %>% 
+  group_by(region_UK) %>% 
+  km_ct_estimator(df = ., t = date, d = confirmed, 
+                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
+  ungroup() %>% 
+  mutate(
+    St_max = S_t^exp(1.96*S_t_SElog),
+    St_min = S_t^exp(-1.96*S_t_SElog)
+  )
+
+residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
+  left_join(distinct(new_cases, date, home_code, 
+                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
+  filter(!is.na(home_code)) %>% 
+  left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
+  left_join(select(reference_geography, postcode, region_UK)) %>% 
+  left_join(approx_occupancy) %>% 
+  arrange(date)
+
+km_ct_residents_symp <- residents_ct_symp %>% 
+  group_by(date, region_UK) %>% 
+  summarise(
+    symptomatic = sum(tidyr::replace_na(new_symptomatic_home, 0) + 
+                      tidyr::replace_na(new_symptomatic_hospital, 0)),
+    occupancy = sum(tidyr::replace_na(susceptible_symptomatic, 0))
+  ) %>% 
+  group_by(region_UK) %>% 
+  km_ct_estimator(df = ., t = date, d = symptomatic, 
+                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
+  ungroup() %>% 
+  mutate(
+    St_max = S_t^exp(1.96*S_t_SElog),
+    St_min = S_t^exp(-1.96*S_t_SElog)
+  )
+
+staff_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
+  left_join(distinct(new_cases, date, home_code, 
+                     new_colleague_confirmed, new_colleague_symptomatic)) %>% 
+  filter(!is.na(home_code)) %>% 
+  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
+  left_join(select(reference_geography, postcode, region_UK)) %>% 
+  arrange(date)
+
+km_staff_symptomatic <- staff_ct %>% 
+  group_by(date, region_UK) %>% 
+  summarise(
+    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
+    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
+    staff_total = sum(staff_total)
+  ) %>% 
+  group_by(region_UK) %>% 
+  km_ct_estimator(., t=date, d=symptomatic, pop=staff_total,subtract_cases = TRUE) %>% 
+  ungroup() %>% 
+  mutate(
+    St_max = S_t^exp(1.96*S_t_SElog),
+    St_min = S_t^exp(-1.96*S_t_SElog)
+  )
+
+km_staff_confirmed <- staff_ct %>% 
+  group_by(date, region_UK) %>% 
+  summarise(
+    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
+    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
+    staff_total = sum(staff_total)
+  ) %>% 
+  group_by(region_UK) %>% 
+  km_ct_estimator(., date, confirmed, staff_total) %>% 
+  ungroup() %>% 
+  mutate(
+    St_max = S_t^exp(1.96*S_t_SElog),
+    St_min = S_t^exp(-1.96*S_t_SElog)
+  )
+
+desc_staff <- list()
+desc_staff$total_staff <- reference_homes %>% filter(home_code %in% unique(residents_overview$home_code)) %>% .$staff_total %>% sum()
+
+desc_staff$CT_symptomatic_total <- sum(new_cases$new_colleague_symptomatic, na.rm = T)
+desc_staff$CT_symptomatic_denom <- sum(km_staff_symptomatic$n_t)
+
+desc_staff[paste0(
+  "CT_symptomatic_p", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(
+  desc_staff$CT_symptomatic_total,
+  desc_staff$total_staff)[c("proportion", "lower", "upper")] *100
+
+desc_staff[paste0(
+  "CT_symptomatic_IR", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(
+  desc_staff$CT_symptomatic_total,
+  desc_staff$CT_symptomatic_denom)[c("rate", "lower", "upper")] *100000
+
+desc_staff$CT_confirmed_total <- sum(new_cases$new_colleague_confirmed, na.rm = T)
+desc_staff$CT_confirmed_denom <- sum(km_staff_confirmed$n_t)
+desc_staff[paste0(
+  "CT_confirmed_p", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(
+  desc_staff$CT_confirmed_total,
+  desc_staff$total_staff)[c("proportion", "lower", "upper")] *100
+
+desc_staff[paste0(
+  "CT_confirmed_IR", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(
+  desc_staff$CT_confirmed_total,
+  desc_staff$CT_confirmed_denom)[c("rate", "lower", "upper")] *100000
+
+```
+
+```{r descriptives}
+desc_residents <- list()
+desc_residents$total_residents <- dplyr::n_distinct(residents_overview$resident_id)
+
+#COUNTS
+desc_residents$CT_total_symptomatic <- sum(new_cases$new_symptomatic_home, na.rm=T) + 
+  sum(new_cases$new_symptomatic_hospital, na.rm=T)
+desc_residents$CT_total_symptomatic_denom <- sum(km_ct_residents_symp$n_t)
+desc_residents[paste0(
+  "CT_p_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents$CT_total_symptomatic,
+                            desc_residents$total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+
+desc_residents[paste0(
+  "CT_IR_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents$CT_total_symptomatic,
+                            desc_residents$CT_total_symptomatic_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
+desc_residents$CT_total_confirmed <-  sum(new_cases$new_confirmed_home, na.rm = T) + sum(new_cases$new_confirmed_hospital, na.rm = T)
+desc_residents$CT_total_confirmed_denom <- sum(km_ct_residents_confirmed$n_t)
+
+desc_residents[paste0(
+  "CT_p_confirmed", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents$CT_total_confirmed,
+                            desc_residents$total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+
+desc_residents[paste0(
+  "CT_IR_confirmed", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents$CT_total_confirmed,
+                            desc_residents$CT_total_confirmed_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
+#INCIDENT REPORTS
+desc_residents$unique_residents_incidents <- dplyr::n_distinct(incidents$resident_id)
+
+desc_residents$DAT_total_symptomatic <- n_distinct(filter(incidents, !is.na(covid_first_symptomatic))$resident_id)
+desc_residents$DAT_total_symptomatic_denom <- sum(residents_overview$susceptible_symptomatic_days)
+desc_residents[paste0(
+  "DAT_p_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents$DAT_total_symptomatic,
+                            desc_residents$total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+desc_residents[paste0(
+  "DAT_IR_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents$DAT_total_symptomatic,
+                            desc_residents$DAT_total_symptomatic_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
+desc_residents$total_tests <- nrow(dplyr::filter(incidents, covid_tested == 1))
+desc_residents$unique_residents_tested <- dplyr::filter(incidents, covid_tested == 1) %>% 
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+desc_residents$unique_residents_test_result <- dplyr::filter(incidents, !is.na(covid_test_result)) %>% 
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+desc_residents$unique_residents_test_result_positive <- dplyr::filter(incidents, covid_test_result == 1) %>%  
+  dplyr::summarise(n = n_distinct(resident_id)) %>% .$n
+
+desc_residents$DAT_total_confirmed_denom <- sum(residents_overview$susceptible_confirmed_days)
+desc_residents[paste0(
+  "DAT_p_confirmed", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents$unique_residents_test_result_positive,
+                            desc_residents$total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+
+desc_residents[paste0(
+  "DAT_IR_confirmed", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents$unique_residents_test_result_positive,
+                            desc_residents$DAT_total_confirmed_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
+# CASE FATALITY AND RELATIVE RISK
+
+desc_residents$DAT_total_deaths <- sum(residents_overview$status == "Deceased")
+desc_residents$total_denom <- sum(residents_overview$resident_days)
+desc_residents[paste0(
+  "DAT_p_death",  c("", "_lower", "_upper")
+)] <- epitools::binom.exact(
+  desc_residents$DAT_total_deaths,
+  desc_residents$total_residents
+)[c("proportion", "lower", "upper")] *100
+
+desc_residents[paste0(
+  "DAT_IR_death",  c("", "_lower", "_upper")
+)] <- epitools::pois.exact(
+  desc_residents$DAT_total_deaths,
+  desc_residents$total_denom
+)[c("rate", "lower", "upper")] * 100000
+
+desc_residents$DAT_CFR <- residents_overview %>% 
+  filter(confirmed == 1) %>% 
+  summarise(n = n(),
+            y = sum(status == "Deceased"))
+
+desc_residents$DAT_BFR <- residents_overview %>%
+  filter(confirmed != 1) %>%
+  summarise(n = n(),
+            y = sum(status == "Deceased"))
+
+desc_residents$DAT_CFR_RR <- relative_risk(desc_residents$DAT_CFR$y, desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, desc_residents$DAT_BFR$y, desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
+
+desc_residents$DAT_CFR_RR_SE <- relative_risk_SE(desc_residents$DAT_CFR$y, desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, desc_residents$DAT_BFR$y, desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
+
+desc_residents[paste0(
+  "DAT_CFR", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(
+  desc_residents$DAT_CFR$y,
+  desc_residents$DAT_CFR$n
+)[c("proportion", "lower", "upper")] *100
+
+desc_residents[paste0(
+  "DAT_BFR", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(
+  desc_residents$DAT_BFR$y,
+  desc_residents$DAT_BFR$n
+)[c("proportion", "lower", "upper")] *100
+
+### Homes
+desc_residents$total_homes <- n_distinct(home_total_confirmed$home_code)
+desc_residents$CT_zero_case <- sum(home_total_confirmed$CT_confirmed == 0)
+desc_residents$DAT_zero_case <- sum(home_total_confirmed$DAT_confirmed == 0)
+```
+
+```{r captions}
 ## TABLES
@@
   ")"))
+
+supp_epi_CT <- table_nums(
+  name = "supp_epi_CT",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents and staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+supp_epi_DAT <- table_nums(
+  name = "supp_epi_DAT",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections amongst residents according to Datix incident reports (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+
 ## FIGURES
@@
   name = 'fig_KM_staff_symptomatic',
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_staff$total_staff),
   ") becoming symptomatic cases based on manager-reported daily counts"))
@@
   name = 'fig_KM_staff_confirmed',
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_overall$total_staff),
+  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_staff$total_staff),
   ") becoming confirmed cases based on manager-reported daily counts"))
@@
 
-RR_england <- relative_risk(
+england_results$RR_england <- relative_risk(
   a = england_results$n_positive,
@@
 
-RR_SE <- relative_risk_SE(
+england_results$RR_SE <- relative_risk_SE(
   a = england_results$n_positive,
@@
 **Findings:** Records were available for `r formatbm(n_distinct(residents_overview$resident_id))` residents present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
-The incidence proportion of COVID-19 symptoms ranged between `r tbrounding(desc_overall$DAT_p_symptomatic*100)`% (based on incidents reports) and `r tbrounding(desc_overall$CT_p_symptomatic*100)`% (based on manager-reported counts).
-Across the period, care homes registered a total of `r formatbm(desc_overall$CT_total_confirmed)` (`r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]) infections confirmed with nasopharyngeal RNA tests.
-Incident reports accounted just `r formatbm(desc_overall$unique_residents_test_result_positive)` confirmed infections.
+During this period `r formatbm(desc_residents$CT_total_symptomatic_denom)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
+Across all care homes, the confirmed infection rate of incidence was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
 In England, `r england_results$n_positive` infections were confirmed of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
-Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
+Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
 The staff-to-bed ratio was the only factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
+Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted(
+`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]), corresponding to a rate of incidence of 
+`r tbrounding(desc_staff$CT_confirmed_IR)` per 100,000 staff-days [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`].
 
@@
 
-An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020; @Dora2020] One such investigation estimated that 50% of residents and 20% of staff are infected during an outbreak in London, UK.[@Easter6] Estimates of this nature cannot be generalised to the unknown proportion of care homes that have not reported any outbreaks.
+An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020; @Dora2020] A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long term care facilities.
+Estimates of this nature cannot be generalised due to the unknown proportion of care homes that have not reported any outbreaks.
 
@@
 
-We obtained care home organisational characteristics: staff headcounts, bed counts, ***latest Care Quality Commission rating***, and types of care available (nursing dementia; nursing general; residential dementia; residential general; younger person).
+We obtained care home organisational characteristics: staff headcounts, bed counts, and types of care available (nursing dementia; nursing general; residential dementia; residential general; younger person).
 
@@
 
-Data were analysed in `R`3·5·0[@RCoreTeam2018] and the `survival` package.[@therneau2015] All computer syntax is available on an online repository.[@Dutey2020]
+Data were analysed in `R`3·5·0[@RCoreTeam2018] using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries.
+All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
 
@@
 
-The study population consisted of `r formatbm(desc_overall$total_residents)` residents across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+The study population consisted of `r formatbm(desc_residents$total_residents)` residents across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
 The vast majority (85·4%) were permanent residents.
-A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported on Datix as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_overall$unique_residents_incidents)` unique residents.
+A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported on Datix as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_residents$unique_residents_incidents)` unique residents.
 The reports were returned by `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes  totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
 
-Datix reports and manager counts respectively indicated that `r formatbm(desc_overall$DAT_total_symptomatic)` (`r tbrounding(desc_overall$DAT_p_symptomatic*100)`% [95% confidence interval: `r tbrounding((desc_overall$DAT_p_symptomatic - desc_overall$DAT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$DAT_p_symptomatic + desc_overall$DAT_p_symptomatic_ME)*100)`]) and `r formatbm(desc_overall$CT_total_symptomatic)` (`r tbrounding(desc_overall$CT_p_symptomatic*100)`% [`r tbrounding((desc_overall$CT_p_symptomatic - desc_overall$CT_p_symptomatic_ME)*100)`; `r tbrounding((desc_overall$CT_p_symptomatic + desc_overall$CT_p_symptomatic_ME)*100)`]) residents became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
+Datix reports and manager counts respectively indicated that `r formatbm(desc_residents$DAT_total_symptomatic)` (`r tbrounding(desc_residents$DAT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$DAT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`%]) and `r formatbm(desc_residents$CT_total_symptomatic)` (`r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]) residents became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")` (`table_nums("supp_epi_CT", display = "cite")`).
 Temporal trends broken down by region (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`) indicate wide variation across the UK, with care homes in the South East and Northern Ireland reporting in excess of a third of residents with symptoms.
@@
 ```{r table_resident_descriptives, message = FALSE, results="asis"}
-arsenal::tableby(home_type  ~ gender + ageg + region_UK +
-   user_type2 + admission_type2 + covid_status + status +
-   resident_days + susceptible_symptomatic_days + susceptible_confirmed_days, 
+arsenal::tableby(home_type  ~ gender + ageg  +
+   user_type2 + admission_type2 + covid_status + status + region_UK, 
    control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
@@
 ```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 8}
-
-residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-  left_join(distinct(new_cases, date, home_code, 
-                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
-  filter(!is.na(home_code)) %>% 
-  left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
-  left_join(select(reference_geography, postcode, region_UK)) %>% 
-  left_join(approx_occupancy) %>% 
-  arrange(date)
-
-km_ct_residents_symp <- residents_ct_symp %>% 
-  group_by(date, region_UK) %>% 
-  summarise(
-    symptomatic = sum(tidyr::replace_na(new_symptomatic_home, 0) + 
-                      tidyr::replace_na(new_symptomatic_hospital, 0)),
-    occupancy = sum(tidyr::replace_na(susceptible_symptomatic, 0))
-  ) %>% 
-  group_by(region_UK) %>% 
-  km_ct_estimator(df = ., t = date, d = symptomatic, 
-                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
-  ungroup() %>% 
-  mutate(
-    St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
-  )
-
 order_key <-km_ct_residents_symp %>% 
@@
 
-Datix incidents reports accounted for `r formatbm(desc_overall$unique_residents_tested)` residents tested at least once.
-Test results were known for `r formatbm(desc_overall$unique_residents_test_result)` of them (`r tbrounding(desc_overall$unique_residents_test_result/desc_overall$unique_residents_tested*100)`%), including `r formatbm(desc_overall$unique_residents_test_result_positive)` residents with a positive test result.
+Datix incidents reports accounted for `r formatbm(desc_residents$unique_residents_tested)` residents tested at least once.
+Test results were known for `r formatbm(desc_residents$unique_residents_test_result)` of them (`r tbrounding(desc_residents$unique_residents_test_result/desc_residents$unique_residents_tested*100)`%), including `r formatbm(desc_residents$unique_residents_test_result_positive)` residents with a positive test result.
 This data should be treated as representative, as tests coming back negative are not commonly recorded on Datix.
@@
 
-The overall incidence proportion was between `r tbrounding(desc_overall$DAT_p_confirmed*100)`% [`r tbrounding((desc_overall$DAT_p_confirmed - desc_overall$DAT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$DAT_p_confirmed + desc_overall$DAT_p_confirmed_ME)*100)`%] according to incidents reports and  `r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%]  according to manager counts.
+The overall confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] according to manager counts, with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] (Table 7).
+
 For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
 In comparison, the survey of English community households[@cisbulletin2020] found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
-This implies a relative risk of `r tbrounding(RR_england)` [`r tbrounding(exp(log(RR_england) - 1.96 * RR_SE)) `; `r tbrounding(exp(log(RR_england) + 1.96 * RR_SE))`].
+This implies a relative risk of `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
 
@@
 ```{r KM_CT_residents_infection, fig.height = 4, fig.width = 8}
-
-residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-  left_join(distinct(new_cases, date, home_code, 
-                     new_confirmed_home, new_confirmed_hospital)) %>% 
-  filter(!is.na(home_code)) %>% 
-  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
-  left_join(select(reference_geography, postcode, region_UK)) %>% 
-  left_join(approx_occupancy) %>% 
-  arrange(date)
-
-km_ct_residents_confirmed <- residents_ct %>% 
-  group_by(date, region_UK) %>% 
-  summarise(
-    confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
-                      tidyr::replace_na(new_confirmed_hospital, 0)),
-    occupancy = sum(tidyr::replace_na(susceptible_confirmed, 0))
-  ) %>% 
-  group_by(region_UK) %>% 
-  km_ct_estimator(df = ., t = date, d = confirmed, 
-                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
-  ungroup() %>% 
-  mutate(
-    St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
-  )
-
 order_key <-km_ct_residents_confirmed %>% 
@@
 
-Between `r desc_overall$DAT_zero_case` (`r tbrounding( sum(home_zero_confirmed$zero_datix)/ sum(home_zero_confirmed[["Total homes"]])*100)`%) and  `r desc_overall$CT_zero_case` (`r tbrounding(sum(home_zero_confirmed$zero_CT)/sum(home_zero_confirmed[["Total homes"]])*100)`%) care homes had not accounted any confirmed cases whether in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
+Between `r desc_residents$DAT_zero_case` (`r tbrounding( sum(home_zero_confirmed$zero_datix)/ sum(home_zero_confirmed[["Total homes"]])*100)`%) and  `r desc_residents$CT_zero_case` (`r tbrounding(sum(home_zero_confirmed$zero_CT)/sum(home_zero_confirmed[["Total homes"]])*100)`%) care homes had not accounted any confirmed cases whether in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
 The substantial lack of agreement between the two sources of data makes it difficult to ascertain which care homes had not had any confirmed cases.
@@
 The crude risk of all-cause death for the entire population of study was 19·5% (`r table_nums("tab_resident_descriptives", display = "cite")`).
-Time-varying covariate Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
+The all-cause confirmed case fatality rate was `r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%] based on incident reports.
+This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR)`% [`r tbrounding(desc_residents$DAT_BFR_lower)`%; `r tbrounding(desc_residents$DAT_BFR_upper)`%] in residents that did not test or did not test positive (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(exp(log(desc_residents$DAT_CFR_RR) - 1.96 * desc_residents$DAT_CFR_RR_SE)) `; `r tbrounding(exp(log(desc_residents$DAT_CFR_RR) + 1.96 * desc_residents$DAT_CFR_RR_SE))`]).
+
+The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
 Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
-
 After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 2·6 [1·6; 4·3] and 9·9 [8·3; 11·9], confirming that symptomatic cases were more severe.
 As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 6·2 [5·1; 7·4].
-There was strong evidence that hazards were not proportional across these categories -- a Kaplan-Meier estimator of the unadjusted survival function illustrates this (`r fig_nums("fig_KM_mortality", display="cite")`).
+It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
 Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
@@
 ```{r risk_death, results="asis"}
-cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-                      gender+ ageg +
-                      staff_to_bed_ratio + bed_tot_category +(region_UK)+ 
-                      home_type*covid_status+ cluster(resident_id), data = residents_episodes)
-# summary(cph_death)
-# cox.zph(cph_death)
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+#                       gender+ ageg +
+#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+ 
+#                       home_type*covid_status+ cluster(resident_id), data = residents_episodes)
+# summary(cph_death2)
+# cox.zph(cph_death2)
 # survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
@@
 ```{r fig_KM_mortality,  fig.height = 4, fig.width = 8}
-survfit(Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + cluster(resident_id), data = residents_episodes) %>% 
-  ggsurvplot(., conf.int = TRUE, 
+KM_mortality_by_covid_status <- survfit(
+  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status, 
+        data = residents_episodes)
+ggsurvplot(KM_mortality_by_covid_status, conf.int = TRUE, 
            legend = "right",
@@
 
+Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were recorded as having symptoms by FSHCG managers.
+In total, `r formatbm(desc_staff$CT_confirmed_total)` (	`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff had a confirmed infection by the end of the period.
+Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) indicate a strong progression of infections in the North West, North East and South East of England.
+
 ```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 8}
-
-staff_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-  left_join(distinct(new_cases, date, home_code, 
-                     new_colleague_confirmed, new_colleague_symptomatic)) %>% 
-  filter(!is.na(home_code)) %>% 
-  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
-  left_join(select(reference_geography, postcode, region_UK)) %>% 
-  arrange(date)
-
-km_staff_symptomatic <- staff_ct %>% 
-  group_by(date, region_UK) %>% 
-  summarise(
-    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
-    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
-    staff_total = sum(staff_total)
-  ) %>% 
-  group_by(region_UK) %>% 
-  km_ct_estimator(., t=date, d=symptomatic, pop=staff_total,subtract_cases = TRUE) %>% 
-  ungroup() %>% 
-  mutate(
-    St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
-  )
 
@@
 ```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 8}
-km_staff_confirmed <- staff_ct %>% 
-  group_by(date, region_UK) %>% 
-  summarise(
-    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
-    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
-    staff_total = sum(staff_total)
-  ) %>% 
-  group_by(region_UK) %>% 
-  km_ct_estimator(., date, confirmed, staff_total) %>% 
-  ungroup() %>% 
-  mutate(
-    St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
-  )
 
@@
 
-This study was to our knowledge the first to report on the incidence of SARS-CoV-2 infections in UK care homes on a large scale, using individal-level resident records from a large provider network.
-Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r formatbm(desc_overall$CT_p_symptomatic)`% of residents experiencing symptoms at some stage and `r `r tbrounding(desc_overall$CT_p_confirmed*100)`% [`r tbrounding((desc_overall$CT_p_confirmed - desc_overall$CT_p_confirmed_ME)*100)`%; `r tbrounding((desc_overall$CT_p_confirmed + desc_overall$CT_p_confirmed_ME)*100)`%] having a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-Regional variation was considerable, but the cross-examination of from incident reports and manager counts produces inconsistent findings, suggesting strong variation in recording practices.
+This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large provider network across `r as.integer(date_max-date_min)` days.
+A comparable study was previously conducted on across 623 care homes in Ontario, Canada.[@Stall2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+
+Regional variation was noticeable, but inconsistencies between incident reports and manager counts were indicative of data quality issues, particularly in relation to incident reports.
 Rates of testing across time and care homes were not yet known at the time of writing and likely distorts measures of incidence.
 
-One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline (***Haydn, please explain***) was a protective factor.
+One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline was a protective factor.
 A 10 percentage point increase was associated with a 13% reduction [9%; 18%] in hazards of confirmed infections.
 Few organisational variables were available to identify potential confounders or mediators of this association.
-One hypothesis is that staffing ratios may be a sign of better-resourced care homes, and of a greater capability to prescribe and observe effective infection control procedures.
+One hypothesis is that staffing ratios may be a sign of better-resourced care homes (for instance protective personal equipment), and of a greater capability to prescribe and observe effective infection control procedures.
 Furthermore, staff can become a vector in the transmission of infections.
@@
 
-staff infection rates
-
-Similar to infection rates, all-cause mortality exhibited strong regional variation. Nursing home mortality was 76% higher, and . I
+The case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was higher than previous literature[@Salcher-Konrad2020], which is not exclusively attributable to the long observation period.
+An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
+These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19%.
 
 Uncertainty remains regarding the internal validity of estimates presented in this report.
-Beyond uncertain on the exact number of cases, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
+It was not possible to determine each care home's outbreak status with confidence.
+Consequently, incidence was calculated across all homes, unlike the study by Stall et al.[@Stall2020] Beyond uncertainty in the exact number of cases and outbreaks, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
 Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
@@
 
+# Supplementary tables
+
+`r table_nums("supp_epi_CT")`
+
+|               |  Symptomatic  | Residents  <br> Confirmed  | Symptomatic | Staff <br> Confirmed   |
+| ------------- | -------------:| -------------:|-------------:|-------------:|
+| Cases      | `r formatbm(desc_residents$CT_total_symptomatic)` |	`r formatbm(desc_residents$CT_total_confirmed)` |	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
+|	$N$ exposed  | `r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
+|	Total exposure (days)| `r formatbm(desc_residents$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents$CT_total_confirmed_denom)`	|`r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
+|	Incidence proportion (%)  | `r tbrounding(desc_residents$CT_p_symptomatic)` [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents$CT_p_confirmed)` [`r tbrounding(desc_residents$CT_p_confirmed_lower)`; `r tbrounding(desc_residents$CT_p_confirmed_upper)`] | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
+|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$CT_IR_symptomatic)` [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
+
+`r table_nums("supp_epi_DAT")`
+
+| Residents     |  Symptomatic  |  Confirmed  | All-cause death | 
+| ------------- | -------------:| -------------:|-------------:|
+| Cases      | `r formatbm(desc_residents$DAT_total_symptomatic)` |	`r formatbm(desc_residents$unique_residents_test_result_positive)` | `r formatbm(desc_residents$DAT_total_deaths)`  |
+|	$N$ exposed  | `r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_residents$total_residents)` | `r formatbm(desc_residents$total_residents)` |
+|	Total exposure (days)|`r formatbm(desc_residents$DAT_total_symptomatic_denom)`|	`r formatbm(desc_residents$DAT_total_confirmed_denom)` | `r formatbm(desc_residents$total_denom)` |
+|	Incidence proportion (%)  | `r tbrounding(desc_residents$DAT_p_symptomatic)` [`r tbrounding(desc_residents$DAT_p_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`] |`r tbrounding(desc_residents$DAT_p_confirmed)` [`r tbrounding(desc_residents$DAT_p_confirmed_lower)`; `r tbrounding(desc_residents$DAT_p_confirmed_upper)`] |`r tbrounding(desc_residents$DAT_p_death)` [`r tbrounding(desc_residents$DAT_p_death_lower)`; `r tbrounding(desc_residents$DAT_p_death_upper)`] |
+|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$DAT_IR_symptomatic)` [`r tbrounding(desc_residents$DAT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_IR_symptomatic_upper)`] | `r tbrounding(desc_residents$DAT_IR_confirmed)`  [`r tbrounding(desc_residents$DAT_IR_confirmed_lower)`; `r tbrounding(desc_residents$DAT_IR_confirmed_upper)`] | `r tbrounding(desc_residents$DAT_IR_death)` [`r tbrounding(desc_residents$DAT_IR_death_lower)`; `r tbrounding(desc_residents$DAT_IR_death_upper)` ]|
+
 # References

### 2020-06-25 covidWarehouse/R/stats.R

@@ -26,6 +26,7 @@ funnel_poisson_control_limits <- function(y, denom) {
   )))
 }
 
+
 #' Classify observation as an outlier based on Poisson control limits
 #'
 #' @param o the numeric of the observed count

### 2020-06-26 covidWarehouse/tests/testthat/test-stats.R
test both KM methods

@@ -29,8 +29,13 @@ km_indiv <- data.frame(km_indiv[1:8])
 km_indiv$pop <- km_indiv$n - cumsum(dplyr::lag(km_indiv$n.censor, 1, default = 0))
 km_indiv$n_t <- km_indiv$n
 
-expect_error(km_ct_estimator(km_indiv, time, n.event, pop, subtract_cases = TRUE, overwrite = FALSE))
-output <- expect_warning(km_ct_estimator(km_indiv, time, n.event, pop, overwrite = TRUE))
-expect_equal(output$surv, output$S_t)
-expect_equal(output$std.err, output$S_t_SE)
+output_sub <- expect_warning(km_ct_estimator(df = km_indiv, t = time, d = n.event,
+                              pop = pop, subtract_cases = TRUE, overwrite = TRUE))
+output_nosub<- expect_warning(km_ct_estimator(df = km_indiv, t = time, d = n.event,
+                               pop = n.risk, subtract_cases = FALSE, overwrite = TRUE))
+
+expect_equal(output_nosub$surv, output_nosub$S_t)
+expect_equal(output_nosub$std.err, output_nosub$S_t_SE)
+expect_equal(output_sub$surv, output_sub$S_t)
+expect_equal(output_sub$std.err, output_sub$S_t_SE)
 })

### 2020-06-26 covidWarehouse/R/compute.R
add linear interpolation of occupancy censuses

@@ -240,6 +240,52 @@ resident_timelines <- function(residents_data = residents,
   output
 }
 
+resident_days_linear_approx <- function(min_date = as.Date("2020-01-01"),
+                                        max_date = as.Date("2020-07-01")) {
+  # This function estimates total daily occupancy (including contract beds) using linear
+  # interpolations between weekly censuses
+
+  load(file.path(getOption("fshc_files"), "occupancy.rda"),)
+  occupancy2 <- dplyr::select(occupancy, -resident_days)
+
+  output <- dplyr::tibble(
+    date = seq(min_date, max_date, by = 1),
+    week_ending = get_monday_date(seq(min_date, max_date, by = 1)) + 6
+    )
+  output$day_number <- as.integer( output$date - get_monday_date(output$date))
+
+  output <- tidyr::crossing(output, unique(occupancy2[,c("home_code")]))
+  output <- merge(output, occupancy2, by = c("week_ending", "home_code"),
+                  all.x = T, all.y = F)
+
+  output <- dplyr::rename(output, next_occupancy = occupancy) %>%
+    dplyr::arrange(home_code, date)
+
+  output <- output %>%
+    dplyr::mutate(
+      occupancy = last_occupancy + round(
+        (next_occupancy - last_occupancy) * day_number / 7
+      )
+    )  %>%
+    dplyr::arrange(home_code, date)
+
+  # carry forward last occupancy figure
+
+  output <- output %>%
+    dplyr::mutate(imputed = is.na(occupancy)) %>%
+    dplyr::group_by(home_code) %>%
+    dplyr::mutate(occupancy = dplyr::if_else(
+      imputed,
+      dplyr::last(na.omit(occupancy), order_by = date),
+      occupancy)) %>%
+    ungroup() %>%
+    dplyr::select(home_code, date, occupancy, imputed)
+
+  output
+}
+
+
+
 resident_days_approx <- function(timelines_data = timelines, time_span = get_time_span()) {
   # This function merges individual resident timelines and computes an approximation
   # of daily occupancy for each care home

### 2020-06-27 covidWarehouse/manuscript/manuscript.Rmd
improve life table calculations + add occupancy risk factor

@@
 # author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
-date: "16/06/2020"
 bibliography: references.bib
@@
 date_max <- as.Date("2020-06-07") #as.Date("2020-06-15")
+date_min_tallies = as.Date("2020-03-24")
 date_max_tallies = as.Date("2020-06-07")
@@
            breaks = seq(20,100,25), right = F, 
-           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds")))
+           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
+         )
 
@@
   left_join(approx_occupancy)
+
+occupancy_interpolated <- resident_days_linear_approx() %>% 
+  # TODO: currently putting the individual-level occup for 360 because no occ tally is known!
+  # it's got 57 bed including 5 contract beds, so 
+  left_join(transmute(filter(approx_occupancy, home_code == "360"), 
+                      date, home_code  , occup360 = trunc(occupancy * 57/52)),
+            by = c("home_code", "date")) %>%
+  mutate(occupancy = if_else(home_code == "360", occup360, occupancy)) %>% 
+  select(-occup360)
+
+home_mean_occupancy <- occupancy_interpolated  %>% 
+  filter(between(date, date_min_tallies, date_max_tallies)) %>% 
+  group_by(home_code) %>% 
+  summarise(mean_occupancy = mean(occupancy, na.rm = T))
+
+reference_homes <- reference_homes %>% 
+  left_join(home_mean_occupancy) %>% 
+  mutate(pct_occupancy = mean_occupancy/beds)
 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, beds, pct_occupancy)) %>% 
   group_by(resident_id) %>% 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category)) %>% 
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, beds, beds, pct_occupancy)) %>% 
   left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
@@
 # CARE HOME AGGREGATES
-home_total_confirmed_DAT <- filter(incidents, !is.na(covid_first_symptomatic)) %>% 
+home_total_confirmed_DAT <- filter(residents_overview, !is.na(covid_first_confirmed)) %>% 
+  filter(between(covid_first_confirmed, date_min_tallies, date_max_tallies)) %>% 
   group_by(home_code) %>% 
@@
   left_join(distinct(new_cases, date, home_code, 
-                     new_confirmed_home, new_confirmed_hospital)) %>% 
+                     new_confirmed_home, new_confirmed_hospital,
+                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
   filter(!is.na(home_code)) %>% 
@@
   left_join(select(reference_geography, postcode, region_UK)) %>% 
-  left_join(approx_occupancy) %>% 
-  arrange(date)
+  left_join(occupancy_interpolated) %>% 
+  arrange(home_code, date) %>% 
+  group_by(home_code) %>% 
+  mutate(new_symptomatic = tidyr::replace_na(new_symptomatic_home, 0) +
+                                    tidyr::replace_na(new_symptomatic_hospital, 0),
+         new_confirmed = tidyr::replace_na(new_confirmed_home, 0) +
+                                    tidyr::replace_na(new_confirmed_hospital, 0)) %>% 
+  mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
+         I_confirmed = total_cases_in_home(occupancy, new_confirmed, date)) %>% 
+  ungroup()
 
@@
   summarise(
-    confirmed = sum(tidyr::replace_na(new_confirmed_home, 0) + 
-                      tidyr::replace_na(new_confirmed_hospital, 0)),
-    occupancy = sum(tidyr::replace_na(susceptible_confirmed, 0))
+    confirmed = sum(new_confirmed),
+    I_confirmed = round(sum(I_confirmed)),
+    occupancy = sum(occupancy)
   ) %>% 
@@
   km_ct_estimator(df = ., t = date, d = confirmed, 
-                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
+                  pop = (occupancy - I_confirmed), subtract_cases = FALSE, overwrite = TRUE) %>% 
   ungroup() %>% 
@@
 
-residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-  left_join(distinct(new_cases, date, home_code, 
-                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
-  filter(!is.na(home_code)) %>% 
-  left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
-  left_join(select(reference_geography, postcode, region_UK)) %>% 
-  left_join(approx_occupancy) %>% 
-  arrange(date)
-
-km_ct_residents_symp <- residents_ct_symp %>% 
+# residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
+#   left_join(distinct(new_cases, date, home_code, 
+#                      new_symptomatic_home, new_symptomatic_hospital)) %>% 
+#   filter(!is.na(home_code)) %>% 
+#   left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
+#   left_join(select(reference_geography, postcode, region_UK)) %>% 
+#   left_join(occupancy_interpolated) %>% 
+#   arrange(date)
+
+km_ct_residents_symp <- residents_ct %>% 
   group_by(date, region_UK) %>% 
   summarise(
-    symptomatic = sum(tidyr::replace_na(new_symptomatic_home, 0) + 
-                      tidyr::replace_na(new_symptomatic_hospital, 0)),
-    occupancy = sum(tidyr::replace_na(susceptible_symptomatic, 0))
+    symptomatic = sum(new_symptomatic),
+    I_symptomatic = round(sum(I_symptomatic)),
+    occupancy = sum(occupancy)
   ) %>% 
@@
   km_ct_estimator(df = ., t = date, d = symptomatic, 
-                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
+                  pop = (occupancy - I_symptomatic), subtract_cases = FALSE, overwrite = TRUE) %>% 
   ungroup() %>% 
@@
   left_join(select(reference_geography, postcode, region_UK))  %>% 
-  arrange(date)
+  arrange(date) %>% 
+  group_by(home_code) %>% 
+  mutate(
+    new_colleague_confirmed = tidyr::replace_na(new_colleague_confirmed, 0),
+    new_colleague_symptomatic = tidyr::replace_na(new_colleague_symptomatic, 0)
+    ) %>% 
+  mutate(I_symptomatic = total_cases_in_home(staff_total, new_colleague_symptomatic, date),
+         I_confirmed = total_cases_in_home(staff_total, new_colleague_confirmed, date)) %>% 
+  ungroup()
 
@@
   summarise(
-    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
-    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
+    symptomatic = sum(new_colleague_symptomatic),
+    I_symptomatic = round(sum(I_symptomatic)),
     staff_total = sum(staff_total)
@@
   group_by(region_UK) %>% 
-  km_ct_estimator(., t=date, d=symptomatic, pop=staff_total,subtract_cases = TRUE) %>% 
+  km_ct_estimator(., t=date, d=symptomatic, pop=(staff_total-I_symptomatic),
+                  subtract_cases = FALSE, overwrite = TRUE) %>% 
   ungroup() %>% 
@@
   summarise(
-    confirmed = sum(tidyr::replace_na(new_colleague_confirmed, 0)),
-    symptomatic = sum(tidyr::replace_na(new_colleague_symptomatic, 0)),
+    confirmed = sum(new_colleague_confirmed),
+    I_confirmed = round(sum(I_confirmed)),
     staff_total = sum(staff_total)
@@
   group_by(region_UK) %>% 
-  km_ct_estimator(., date, confirmed, staff_total) %>% 
+  km_ct_estimator(., t=date, d=confirmed, pop=(staff_total-I_confirmed),
+                  subtract_cases = FALSE, overwrite = TRUE) %>% 
   ungroup() %>% 
@@
 desc_residents <- list()
-desc_residents$total_residents <- dplyr::n_distinct(residents_overview$resident_id)
+desc_residents$DAT_total_residents <- dplyr::n_distinct(residents_overview$resident_id)
+desc_residents$DAT_total_residents_england <- dplyr::filter(residents_overview, country == "England") %>% summarise(n = dplyr::n_distinct(resident_id)) %>% .$n
 
 #COUNTS
+# This is estimated in validation.Rmd
+desc_residents$CT_total_residents <- round(desc_residents$DAT_total_residents * 1.117309 + .49)
+desc_residents$CT_total_residents_england <- round(desc_residents$DAT_total_residents_england * 1.0793268 + .49)
 desc_residents$CT_total_symptomatic <- sum(new_cases$new_symptomatic_home, na.rm=T) + 
@@
 )] <- epitools::binom.exact(desc_residents$CT_total_symptomatic,
-                            desc_residents$total_residents)[
+                            desc_residents$CT_total_residents)[
                               c("proportion", "lower", "upper")
@@
 )] <- epitools::binom.exact(desc_residents$CT_total_confirmed,
-                            desc_residents$total_residents)[
+                            desc_residents$CT_total_residents)[
                               c("proportion", "lower", "upper")
@@
 )] <- epitools::binom.exact(desc_residents$DAT_total_symptomatic,
-                            desc_residents$total_residents)[
+                            desc_residents$DAT_total_residents)[
                               c("proportion", "lower", "upper")
@@
 )] <- epitools::binom.exact(desc_residents$unique_residents_test_result_positive,
-                            desc_residents$total_residents)[
+                            desc_residents$DAT_total_residents)[
                               c("proportion", "lower", "upper")
@@
   desc_residents$DAT_total_deaths,
-  desc_residents$total_residents
+  desc_residents$DAT_total_residents
 )[c("proportion", "lower", "upper")] *100
@@
 desc_residents$DAT_zero_case <- sum(home_total_confirmed$DAT_confirmed == 0)
+
+## OUTBREAK HOMES 
+desc_residents_outbreak_only <- list()
+desc_residents_outbreak_only$outbreak_homes <- home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0]
+
+desc_residents_outbreak_only$CT_total_beds <- beds %>% 
+  filter(year == 2020 & home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(beds = sum(beds),
+            contract_beds = sum(contract_beds))
+
+desc_residents_outbreak_only$CT_total_residents <- residents_overview %>% 
+  dplyr::filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  dplyr::summarise(total_residents = round(dplyr::n_distinct(resident_id)* 1.1151324 + .49)) %>% 
+  .$total_residents
+
+desc_residents_outbreak_only$CT_total_confirmed <- new_cases %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(new_confirmed_home, na.rm = T) + sum(new_confirmed_hospital, na.rm = T)) %>% 
+  .$n
+
+desc_residents_outbreak_only$CT_total_symptomatic <- new_cases %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(new_symptomatic_home, na.rm = T) + sum(new_symptomatic_hospital, na.rm = T)) %>% 
+  .$n
+
+desc_residents_outbreak_only$CT_total_confirmed_denom <- residents_ct %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(occupancy - I_confirmed)) %>% 
+  .$n
+
+desc_residents_outbreak_only[paste0(
+  "CT_p_confirmed", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents_outbreak_only$CT_total_confirmed,
+                            desc_residents_outbreak_only$CT_total_residents)[
+                              c("proportion", "lower", "upper")
+                              ] * 100
+
+desc_residents_outbreak_only[paste0(
+  "CT_IR_confirmed", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents_outbreak_only$CT_total_confirmed,
+                           desc_residents_outbreak_only$CT_total_confirmed_denom)[
+                             c("rate", "lower", "upper")
+                             ] * 100000
+
+desc_residents_outbreak_only$CT_total_symptomatic <-  new_cases %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(new_symptomatic_home, na.rm = T) + sum(new_symptomatic_hospital, na.rm = T)) %>% 
+  .$n
+
+desc_residents_outbreak_only$CT_total_symptomatic_denom <- residents_ct %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(occupancy - I_symptomatic)) %>% 
+  .$n
+
+desc_residents_outbreak_only[paste0(
+  "CT_p_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents_outbreak_only$CT_total_symptomatic,
+                            desc_residents_outbreak_only$CT_total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+desc_residents_outbreak_only[paste0(
+  "CT_IR_symptomatic", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents_outbreak_only$CT_total_symptomatic,
+                            desc_residents_outbreak_only$CT_total_symptomatic_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
 ```
@@
   name = "tab_national_descriptives",
-  caption = "Total homes, beds, and residents within FSHCG and nationwide")
+  caption = "Total homes, beds, contract beds, and residents within FSHCG and nationwide")
 tab_resident_descriptives <- table_nums(
   name = "tab_resident_descriptives",
-  caption = paste0("Breakdown of FSHCG residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
-tab_zero_cases <- table_nums(
-  name = "tab_zero_cases",
-  caption = "Number and proportion of homes without confirmed cases by UK region"
-)
+  caption = paste0("Breakdown of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+supp_epi_CT <- table_nums(
+  name = "supp_epi_CT",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents and staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+supp_epi_CT_oubreak_only <- table_nums(
+  name = "supp_epi_CT_oubreak_only",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among residents of outbreak care homes according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
 tab_CoxPH_confirmed <- table_nums(
@@
 
-supp_epi_CT <- table_nums(
-  name = "supp_epi_CT",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents and staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
 supp_epi_DAT <- table_nums(
@@
   caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections amongst residents according to Datix incident reports (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+
+tab_zero_cases <- table_nums(
+  name = "tab_zero_cases",
+  caption = paste0("Number and proportion of homes without confirmed cases by UK region (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
+)
 
@@
 england_results$n_positive = sum(inner_join(new_cases, filter(reference_homes, country == "England"))[c("new_confirmed_home", "new_confirmed_hospital")], na.rm = T)
-england_results$n_negative = n_distinct(inner_join(residents, filter(reference_homes, country == "England"))$resident_id) - england_results$n_positive
+england_results$n_negative = desc_residents$CT_total_residents_england  - england_results$n_positive
 
@@
 
-**Findings:** Records were available for `r formatbm(n_distinct(residents_overview$resident_id))` residents present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")`.
-During this period `r formatbm(desc_residents$CT_total_symptomatic_denom)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
+**Findings:** An estimated `r formatbm(desc_residents$CT_total_residents)` residents were present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+During the period `r formatbm(desc_residents$CT_total_symptomatic_denom)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
 Across all care homes, the confirmed infection rate of incidence was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
@@
 Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
+These estimates include `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) that had not reported a single confirmed case.
+
 The staff-to-bed ratio was the only factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
@@
 
-In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`)).
+In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`).
 As of 22 June 2020, 90% of FSHCG care homes have participated in the whole care home testing programme.
 
-FSHCG collects administrative data on their residents through a variety of electronic systems that are primarily used for billing purposes.
+FSHCG collects administrative data on their private residents through a variety of electronic systems that are primarily used for billing purposes.
 This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as "Datix".
+Such records do not include residents occupying beds contracted by local authorities as part of block agreements.
 
@@
 
-## Individual-level residents data
-
-Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of two datasets:
-
-- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly).
-This dataset does not include any records for occupants of local authority 'block contract' beds  ($n=$ `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))`).
-- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff into FSHCG's Datix incidents management system.
-Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting.
-In addition, reports can include: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.
-
-HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier.
-Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents dataset.
-Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) Datix reports did not contain enough information to generate an identifier and were excluded.
-Identifiers of a further `r desc_before_exclusions$link_fail` reports (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
-A proportion of these reports may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.
-
-## Aggregate data
-
-We obtained care home organisational characteristics: staff headcounts, bed counts, and types of care available (nursing dementia; nursing general; residential dementia; residential general; younger person).
-
-On 24 March 2020, FSHCG introduced an additional system to monitor COVID-19 cases in each care home, requiring care home managers to report daily tallies of suspected and confirmed infection in staff and residents to their regional manager.
+## Aggregate outbreak data (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")` )
+
+On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor COVID-19 cases in each care home.
+This new system requires care home managers to report daily tallies of symptomatic and confirmed infections identified among staff and residents to their FSHCG regional manager.
+
 Daily counts reported by care homes to FSHCG were obtained for every care home:
@@
 
-These counts include events relating to occupants of local authority contract beds as well as occupants receiving a different category of support (eg.
+In addition, the number of beds currently occupied was reported every Sunday. This weekly information was used to approximate daily occupancy using linear interpolation.
+
+These counts include events relating to residents occupying local authority contract beds as well as a minority of residents receiving a different category of support (eg.
 young people) who were excluded from the study population.
 
+## Individual-level residents data (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")` )
+
+Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of two datasets:
+
+- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly).
+This dataset does not include any records for occupants of local authority 'block contract' beds  ($n=$ `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))`).
+- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff into FSHCG's Datix incidents management system.
+Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting.
+In addition, reports can include: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.
+
+HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier.
+Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents dataset.
+Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) Datix reports did not contain enough information to generate an identifier and were excluded.
+Identifiers of a further `r desc_before_exclusions$link_fail` reports (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
+A proportion of these reports may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.
+
 ## Analysis
 
-We used individual-level incident reports and daily tallies to generate estimates of infection incidence in residents and staff.
-The individual data provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
-A Cox proportional hazards model was used to test the effect of potential risk factors, whether individual- (sex, age, care type) or organisation-level (region, care home size, staff-to-bed ratio).
+The two sources of data were employed for separate purposes.
+
+Incidence proportions and rates were calculated using the aggregate outbreak counts, since they were the trusted source of information and encompass all residents.
+The number of unique residents $N$ present in the homes at some point between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
+This approach assumes that the mean stay of local authority residents is identical to that of private residents (supplementary material 1).
+As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrement life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records.
+Admitted cases were accounted for in the daily new case count.
+Based on this $\hat{E_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as
+
+$$\hat{E_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad E_{t_0} = 0$$
+where $d_{t}$ denotes the number of new cases for each day $t = \{1, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate.
+The total exposure time was calculated as $ \sum_{t=1}^T (N_t - \hat{E_t})$ where $N_t$ denotes daily occupancy figures.
+The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
+
+As for individual-level data, they were known to be subject to underreporting of infection cases as well as to exclude residents occupying local authority contracted beds.
+In spite of this, individual-level provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
+By assuming underreporting of cases to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and fatalities.
+A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, mean bed occupancy proportion).
 Dates of death were used to estimate the case-fatality rate, as well as compare all-cause mortality based on the presence of symptoms and infection confirmation.
@@
 
-A Datix report may not have been filed for every event.
-We therefore also estimate the incidence of events using manager counts.
-The resident-days denominator was computed by subtracting past confirmed cases counts from resident-days estimated from the individual-level data.
-
-The incidence proportion of confirmed cases in England is contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
-
 Data were analysed in `R`3·5·0[@RCoreTeam2018] using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries.
@@
 
-The study population consisted of `r formatbm(desc_residents$total_residents)` residents across England, Scotland and Northern Ireland, `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of whom were in nursing homes and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
-The vast majority (85·4%) were permanent residents.
-A total of `r formatbm(nrow(incidents))` COVID-19 incidents were reported on Datix as of `r format(max(incidents$incident_date), "%A %e %b %Y")` for  `r formatbm(desc_residents$unique_residents_incidents)` unique residents.
-The reports were returned by `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes  totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
-
-Datix reports and manager counts respectively indicated that `r formatbm(desc_residents$DAT_total_symptomatic)` (`r tbrounding(desc_residents$DAT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$DAT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`%]) and `r formatbm(desc_residents$CT_total_symptomatic)` (`r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]) residents became symptomatic at some point between `r format(date_min, "%e %B %Y")` and `r format(date_max, "%e %B %Y")` (`table_nums("supp_epi_CT", display = "cite")`).
-Temporal trends broken down by region (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`) indicate wide variation across the UK, with care homes in the South East and Northern Ireland reporting in excess of a third of residents with symptoms.
+The study population consisted of an estimated `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland.
+This included `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
+Characteristics of private residents could be described from individual-level records (`r table_nums("tab_resident_descriptives", display = "cite")`); `r tbrounding(mean(residents$home_type=="Nursing")*100)`% were in nursing homes, and 39·2% received dementia care.
 
@@
 ```
+
+### Resident cases
+
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("supp_epi_CT", display = "cite")`).
+
+`r table_nums("supp_epi_CT")`
+
+|               |  Symptomatic  | Residents  <br> Confirmed  | Symptomatic | Staff <br> Confirmed   |
+| ------------- | -------------:| -------------:|-------------:|-------------:|
+| Cases      | `r formatbm(desc_residents$CT_total_symptomatic)` |	`r formatbm(desc_residents$CT_total_confirmed)` |	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
+|	$N$ exposed  | `r formatbm(desc_residents$CT_total_residents)`|	`r formatbm(desc_residents$CT_total_residents)`|	`r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
+|	Total exposure (days)| `r formatbm(desc_residents$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents$CT_total_confirmed_denom)`	|`r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
+|	Incidence proportion (%)  | `r tbrounding(desc_residents$CT_p_symptomatic)` [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents$CT_p_confirmed)` [`r tbrounding(desc_residents$CT_p_confirmed_lower)`; `r tbrounding(desc_residents$CT_p_confirmed_upper)`] | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
+|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$CT_IR_symptomatic)` [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
+
+FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
+From 11 May 2020, most FSHCG home undertook testing every resident as part of the whole home testing programme.[@dhscTesting2020] Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` residents had an infection confirmed in homes while `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` had an infection confirmed in hospital.
+The overall confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%, with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
+
+Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`, `r fig_nums("fig_KM_CT_residents_confirmed", display = "cite")`) indicate wide variation across the UK.
+We note a strong progression of infections in the North West, North East and South East of England, where more than 1 in 5 residents had a confirmed infection.
+The sharp increase in confirmed infections in the North West between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
 
@@
 
-## Resident confirmed cases
-
-FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
-From 11 May 2020, most FSHCG home undertook testing every resident as part of the whole home testing programme.[@dhscTesting2020]
-
-Datix incidents reports accounted for `r formatbm(desc_residents$unique_residents_tested)` residents tested at least once.
-Test results were known for `r formatbm(desc_residents$unique_residents_test_result)` of them (`r tbrounding(desc_residents$unique_residents_test_result/desc_residents$unique_residents_tested*100)`%), including `r formatbm(desc_residents$unique_residents_test_result_positive)` residents with a positive test result.
-This data should be treated as representative, as tests coming back negative are not commonly recorded on Datix.
-Care home manager daily counts of new confirmed cases added up to `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` cases tested in homes and `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` tested in hospitals.
-The temporal trend of confirmed cases broken down by region is presented using both Datix reports (`r fig_nums("fig_KM_residents_confirmed", display = "cite")`) and manager counts (`r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`).
-
-The overall confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] according to manager counts, with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] (Table 7).
-
-For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
-This implies a relative risk of `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
-
-```{r km_confirmed, fig.height=4,fig.width=8, warning=F}
-survfit(Surv(time = start_int, 
-             time2 = stop_int + 1,
-             event = confirmed) ~ region_UK, data = residents_overview) %>% 
-  ggsurvplot(., conf.int = TRUE, 
-           fun = "event",
-           legend = "right",
-           ggtheme = theme_bw(),
-           # TODO: double check labels at the end
-           break.x.by	= 7,
-           break.y.by = .01,
-           ylim = c(0, .08),
-           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
-           ylab = "Cum probability of confirmed case",
-           legend.labs = sort(unique(residents_overview$region_UK)),
-           conf.int.alpha = .1) 
-```
-
-`r fig_nums("fig_KM_residents_confirmed")`
-
 ```{r KM_CT_residents_infection, fig.height = 4, fig.width = 8}
@@
 
-Between `r desc_residents$DAT_zero_case` (`r tbrounding( sum(home_zero_confirmed$zero_datix)/ sum(home_zero_confirmed[["Total homes"]])*100)`%) and  `r desc_residents$CT_zero_case` (`r tbrounding(sum(home_zero_confirmed$zero_CT)/sum(home_zero_confirmed[["Total homes"]])*100)`%) care homes had not accounted any confirmed cases whether in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
-The substantial lack of agreement between the two sources of data makes it difficult to ascertain which care homes had not had any confirmed cases.
+For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
+This implies a relative risk of `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
+
+Incidence proportions and rates were also estimated for private residents only using the Datix incident reports dataset (supplementary material 1-2).
+Estimates are lower than corresponding estimates based on outbreak counts.
+This is thought to be caused by a combination of underascertainment as well as linkage error.
+
+### Staff cases
+
+Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were recorded as having symptoms by FSHCG managers.
+In total, `r formatbm(desc_staff$CT_confirmed_total)` (	`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff had a confirmed infection by the end of the period.
+Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) indicate a strong progression of infections in the North West, North East and South East of England.
+
+```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 8}
+
+order_key <-km_staff_symptomatic %>% 
+  group_by(region_UK) %>% 
+  summarise(S_t = min(S_t)) %>% 
+  arrange(S_t)
+km_staff_symptomatic$region_UK <- factor(
+  km_staff_symptomatic$region_UK,
+  levels = order_key$region_UK
+)
+km_staff_symptomatic$region_group <- cut(as.integer(km_staff_symptomatic$region_UK), 2, labels = 1:2)
+
+ggplot(aes(x = date, y = 1-S_t, 
+           ymax = 1 - St_max,
+           ymin = 1 - St_min,
+           color = region_UK), data = km_staff_symptomatic) +
+  # facet_grid(region_group ~ .) +
+  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
+  geom_path(lwd=1.2) +
+  scale_color_discrete("Region") +  
+  scale_y_continuous("Cum probability symptomatic case",
+                     breaks = seq(0,1,.05)) +
+  scale_x_date(name = "Date", 
+               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
+               labels = format.Date(
+                 unique(get_monday_date(km_staff_symptomatic$date)),
+                 "%d/%m/%y"
+               )) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
+
+```
+
+`r fig_nums("fig_KM_staff_symptomatic")`
+
+```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 8}
+
+order_key <- km_staff_confirmed %>% 
+  group_by(region_UK) %>% 
+  summarise(S_t = min(S_t)) %>% 
+  arrange(S_t)
+km_staff_confirmed$region_UK <- factor(
+  km_staff_confirmed$region_UK,
+  levels = order_key$region_UK
+)
+km_staff_confirmed$region_group <- cut(as.integer(km_staff_confirmed$region_UK), 2, labels = 1:2)
+
+ggplot(aes(x = date, y = 1-S_t, 
+           ymax = 1 - St_max,#(S_t + 1.96 * S_t_SE),
+           ymin = 1 - St_min, #(S_t - 1.96 * S_t_SE),
+           color = region_UK), data = km_staff_confirmed) +
+  # facet_grid(region_group ~ .) +
+  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
+  geom_path(lwd=1.2) +
+  scale_color_discrete("Region") +
+  scale_y_continuous("Cum probability confirmed case",
+                     breaks = seq(0,1,.01)) +
+  scale_x_date(name = "Date", 
+               breaks = unique(get_monday_date(km_staff_confirmed$date)),
+               labels = format.Date(
+                 unique(get_monday_date(km_staff_confirmed$date)),
+                 "%d/%m/%y"
+               )) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
+
+```
+
+`r fig_nums("fig_KM_staff_confirmed")`
+
+### Outbreak homes
+
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection, whether acquired in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
+
+`r table_nums("supp_epi_CT_oubreak_only")`
+
+|               |  Symptomatic  | Residents  <br> Confirmed  | 
+| ------------- | -------------:| -------------:|
+| Cases      | `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic)` |	`r formatbm(desc_residents_outbreak_only$CT_total_confirmed)` |
+|	$N$ exposed  | `r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	`r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	
+|	Total exposure (days)| `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents_outbreak_only$CT_total_confirmed_denom)`
+|	Incidence proportion (%)  | `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_upper)`] |
+|	Incidence rate (per 100,000) | `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_upper)`] |
+
+## Factors associated with confirmed infections
+
+Using the individual-level data on private bed residents (`table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+According to this, and other than geographical location, the staff-to-bed ratio was the only factor significantly associated with the rate of confirmed cases.
+An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 13% reduction [9%; 18%] in the rate of confirmed cases.
+
+`r table_nums("tab_CoxPH_confirmed")`
+
+```{r CPH_confirmed,   results="asis"}
+cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + scale(pct_occupancy, scale = F), data = residents_overview)
+ summary(cph_conf)
+# cox.zph(cph_conf)
+# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
+explanatory = c("gender", "ageg", "home_type", "user_type2", "region_UK" ,  
+                "staff_to_bed_ratio", "bed_tot_category", "pct_occupancy")
+dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
+CPH_confirmed <-  finalfit(residents_overview,
+                           dependent, 
+                           explanatory, column = T,
+                           add_dependent_label = F)  
+knitr::kable(CPH_confirmed,
+             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
+```
+
+## Factors associated with all-cause mortality
+
+The crude risk of all-cause death for the entire population of study was 19·5% (`r table_nums("tab_resident_descriptives", display = "cite")`).
+The all-cause confirmed case fatality rate was `r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%] based on incident reports.
+This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR)`% [`r tbrounding(desc_residents$DAT_BFR_lower)`%; `r tbrounding(desc_residents$DAT_BFR_upper)`%] in residents that did not test or did not test positive (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(exp(log(desc_residents$DAT_CFR_RR) - 1.96 * desc_residents$DAT_CFR_RR_SE)) `; `r tbrounding(exp(log(desc_residents$DAT_CFR_RR) + 1.96 * desc_residents$DAT_CFR_RR_SE))`]).
+
+The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
+Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
+After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 2·6 [1·6; 4·3] and 9·9 [8·3; 11·9], confirming that symptomatic cases were more severe.
+As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 6·2 [5·1; 7·4].
+It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
+Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
+The 27% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
+This effect is likely attributable to a greater burden of comorbidities in nursing home residents.
+Beyond, there was no evidence of an association between care home size and all-cause mortality.
+After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·7; 1.1]).
+
+`r table_nums("tab_CoxPH_death")`
+
+```{r risk_death, results="asis"}
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+#                       gender+ ageg +
+#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+ 
+#                       home_type*covid_status+ cluster(resident_id), data = residents_episodes)
+# summary(cph_death2)
+# cox.zph(cph_death2)
+# survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
+# 
+explanatory = c("gender", "ageg", "user_type2", "home_type",
+                "staff_to_bed_ratio", "bed_tot_category", 
+                "covid_status", "region_UK", "cluster(resident_id)")
+explanatory_interaction = c("gender", "ageg", "user_type2",
+                "staff_to_bed_ratio", "bed_tot_category", 
+                "covid_status:home_type", "region_UK", "cluster(resident_id)")
+dependent = "Surv(time = date1, time2 = date2, event = death_status)"
+CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
+                       column = T, add_dependent_label = F)  
+knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
+
+```
+
+```{r fig_KM_mortality,  fig.height = 4, fig.width = 8}
+KM_mortality_by_covid_status <- survfit(
+  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status, 
+        data = residents_episodes)
+ggsurvplot(KM_mortality_by_covid_status, conf.int = TRUE, 
+           legend = "right",
+           ggtheme = theme_bw(),  
+           break.x.by	= 7,
+           break.y.by = .1,
+           legend.labs = levels(residents_episodes$covid_status),
+           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
+           ylab = "Survival",
+           conf.int.alpha = .1)  
+```
+
+*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevant Datix incident report.
+
+`r fig_nums("fig_KM_mortality")`
+
+# Interpretation
+
+This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large provider network across `r as.integer(date_max-date_min)` days.
+A comparable study was previously conducted on across 623 care homes in Ontario, Canada.[@Stall2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+
+Regional variation was noticeable, but inconsistencies between incident reports and manager counts were indicative of data quality issues, particularly in relation to incident reports.
+Rates of testing across time and care homes were not yet known at the time of writing and likely distorts measures of incidence.
+
+One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline was a protective factor.
+A 10 percentage point increase was associated with a 13% reduction [9%; 18%] in hazards of confirmed infections.
+Few organisational variables were available to identify potential confounders or mediators of this association.
+One hypothesis is that staffing ratios may be a sign of better-resourced care homes (for instance protective personal equipment), and of a greater capability to prescribe and observe effective infection control procedures.
+Furthermore, staff can become a vector in the transmission of infections.
+It is possible that higher staffing levels mitigate the transmission of infections since each staff member will care for a lower number of residents.
+Better staffed homes may also be less reliant on bank or agency staff who work in different locations.
+A more detailed survey of care home characteristics, along with protocols and measures adopted during the epidemic may help elucidate mechanisms through which staffing levels are associated with lower infection rates.
+
+The case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was higher than previous literature[@Salcher-Konrad2020], which is not exclusively attributable to the long observation period.
+An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
+These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19%.
+
+Uncertainty remains regarding the internal validity of estimates presented in this report.
+It was not possible to determine each care home's outbreak status with confidence.
+Consequently, incidence was calculated across all homes, unlike the study by Stall et al.[@Stall2020] Beyond uncertainty in the exact number of cases and outbreaks, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
+Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
+According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
+Consequently findings presented in the current study are likely to be underestimates.
+This includes estimates of the incidence ratio comparing care homes with the community in England.
+
+Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection.[@Dora2020] Although increasing testing capacity raises the possibility of undertaking frequent, repeat testing in the care home population, this may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
+Understanding which care homes are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention accordingly.
+
+# Declarations
+
+## Funding
+
+This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].
+
+## Research ethics
+
+This study was approved by the University College London Research Ethics Committee (project reference 13355/002).
+
+# Supplementary material
+
+## Supplementary material 1: Quality and methodology report
+
+## Supplementary material 2: Incidence of confirmed and symptomatic cases among private residents according to Datix incident reports.
+
+Datix incidents reports recorded that `r formatbm(desc_residents$unique_residents_tested)` private residents were tested at least once.
+Test results were known for `r formatbm(desc_residents$unique_residents_test_result)` of them (`r tbrounding(desc_residents$unique_residents_test_result/desc_residents$unique_residents_tested*100)`%), including `r formatbm(desc_residents$unique_residents_test_result_positive)` residents with a positive test result.
+This data should be treated as representative, as tests coming back negative were not commonly recorded on Datix.
+Incidence proportions and rates reported in `r  table_nums("supp_epi_DAT", display = "cite")` are lower than corresponding estimates based on outbreak counts.
+This is thought to be caused by a combination of underascertainment as well as linkage error.
+
+`r table_nums("supp_epi_DAT")`
+
+| Residents     |  Symptomatic  |  Confirmed  | All-cause death | 
+| ------------- | -------------:| -------------:|-------------:|
+| Cases      | `r formatbm(desc_residents$DAT_total_symptomatic)` |	`r formatbm(desc_residents$unique_residents_test_result_positive)` | `r formatbm(desc_residents$DAT_total_deaths)`  |
+|	$N$ exposed  | `r formatbm(desc_residents$DAT_total_residents)`|	`r formatbm(desc_residents$DAT_total_residents)` | `r formatbm(desc_residents$DAT_total_residents)` |
+|	Total exposure (days)|`r formatbm(desc_residents$DAT_total_symptomatic_denom)`|	`r formatbm(desc_residents$DAT_total_confirmed_denom)` | `r formatbm(desc_residents$total_denom)` |
+|	Incidence proportion (%)  | `r tbrounding(desc_residents$DAT_p_symptomatic)` [`r tbrounding(desc_residents$DAT_p_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`] |`r tbrounding(desc_residents$DAT_p_confirmed)` [`r tbrounding(desc_residents$DAT_p_confirmed_lower)`; `r tbrounding(desc_residents$DAT_p_confirmed_upper)`] |`r tbrounding(desc_residents$DAT_p_death)` [`r tbrounding(desc_residents$DAT_p_death_lower)`; `r tbrounding(desc_residents$DAT_p_death_upper)`] |
+|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$DAT_IR_symptomatic)` [`r tbrounding(desc_residents$DAT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_IR_symptomatic_upper)`] | `r tbrounding(desc_residents$DAT_IR_confirmed)`  [`r tbrounding(desc_residents$DAT_IR_confirmed_lower)`; `r tbrounding(desc_residents$DAT_IR_confirmed_upper)`] | `r tbrounding(desc_residents$DAT_IR_death)` [`r tbrounding(desc_residents$DAT_IR_death_lower)`; `r tbrounding(desc_residents$DAT_IR_death_upper)` ]|
+
+## Supplementary material 3: Additional tables
 
@@
 
-Using the individual-level data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-According to this, and other than geographical location, the staff-to-bed ratio was the only factor significantly associated with the rate of confirmed cases.
-An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 13% reduction [9%; 18%] in the rate of confirmed cases.
-
-`r table_nums("tab_CoxPH_confirmed")`
-
-```{r CPH_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category, data = residents_overview)
-# summary(cph_conf)
-# cox.zph(cph_conf)
-# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
-explanatory = c("gender", "ageg", "home_type", "user_type2", "region_UK" ,  
-                "staff_to_bed_ratio", "bed_tot_category")
-dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
-CPH_confirmed <-  finalfit(residents_overview,
-                           dependent, 
-                           explanatory, column = T,
-                           add_dependent_label = F)  
-knitr::kable(CPH_confirmed,
-             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
-```
-
-## All-cause mortality
-
-The crude risk of all-cause death for the entire population of study was 19·5% (`r table_nums("tab_resident_descriptives", display = "cite")`).
-The all-cause confirmed case fatality rate was `r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%] based on incident reports.
-This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR)`% [`r tbrounding(desc_residents$DAT_BFR_lower)`%; `r tbrounding(desc_residents$DAT_BFR_upper)`%] in residents that did not test or did not test positive (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(exp(log(desc_residents$DAT_CFR_RR) - 1.96 * desc_residents$DAT_CFR_RR_SE)) `; `r tbrounding(exp(log(desc_residents$DAT_CFR_RR) + 1.96 * desc_residents$DAT_CFR_RR_SE))`]).
-
-The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
-Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
-After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 2·6 [1·6; 4·3] and 9·9 [8·3; 11·9], confirming that symptomatic cases were more severe.
-As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 6·2 [5·1; 7·4].
-It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
-Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
-The 27% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
-This effect is likely attributable to a greater burden of comorbidities in nursing home residents.
-Beyond, there was no evidence of an association between care home size and all-cause mortality.
-After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·7; 1.1]).
-
-`r table_nums("tab_CoxPH_death")`
-
-```{r risk_death, results="asis"}
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg +
-#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+ 
-#                       home_type*covid_status+ cluster(resident_id), data = residents_episodes)
-# summary(cph_death2)
-# cox.zph(cph_death2)
-# survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
-# 
-explanatory = c("gender", "ageg", "user_type2", "home_type",
-                "staff_to_bed_ratio", "bed_tot_category", 
-                "covid_status", "region_UK", "cluster(resident_id)")
-explanatory_interaction = c("gender", "ageg", "user_type2",
-                "staff_to_bed_ratio", "bed_tot_category", 
-                "covid_status:home_type", "region_UK", "cluster(resident_id)")
-dependent = "Surv(time = date1, time2 = date2, event = death_status)"
-CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
-                       column = T, add_dependent_label = F)  
-knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
-
-```
-
-```{r fig_KM_mortality,  fig.height = 4, fig.width = 8}
-KM_mortality_by_covid_status <- survfit(
-  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status, 
-        data = residents_episodes)
-ggsurvplot(KM_mortality_by_covid_status, conf.int = TRUE, 
-           legend = "right",
-           ggtheme = theme_bw(),  
-           break.x.by	= 7,
-           break.y.by = .1,
-           legend.labs = levels(residents_episodes$covid_status),
-           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
-           ylab = "Survival",
-           conf.int.alpha = .1)  
-```
-
-*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevant Datix incident report.
-
-`r fig_nums("fig_KM_mortality")`
-
-## Staff symptomatic and confirmed cases
-
-Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were recorded as having symptoms by FSHCG managers.
-In total, `r formatbm(desc_staff$CT_confirmed_total)` (	`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff had a confirmed infection by the end of the period.
-Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) indicate a strong progression of infections in the North West, North East and South East of England.
-
-```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 8}
-
-order_key <-km_staff_symptomatic %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_staff_symptomatic$region_UK <- factor(
-  km_staff_symptomatic$region_UK,
-  levels = order_key$region_UK
-)
-km_staff_symptomatic$region_group <- cut(as.integer(km_staff_symptomatic$region_UK), 2, labels = 1:2)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,
-           ymin = 1 - St_min,
-           color = region_UK), data = km_staff_symptomatic) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  scale_color_discrete("Region") +  
-  scale_y_continuous("Cum probability symptomatic case",
-                     breaks = seq(0,1,.05)) +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_staff_symptomatic$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_staff_symptomatic")`
-
-```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 8}
-
-order_key <- km_staff_confirmed %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_staff_confirmed$region_UK <- factor(
-  km_staff_confirmed$region_UK,
-  levels = order_key$region_UK
-)
-km_staff_confirmed$region_group <- cut(as.integer(km_staff_confirmed$region_UK), 2, labels = 1:2)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,#(S_t + 1.96 * S_t_SE),
-           ymin = 1 - St_min, #(S_t - 1.96 * S_t_SE),
-           color = region_UK), data = km_staff_confirmed) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  scale_color_discrete("Region") +
-  scale_y_continuous("Cum probability confirmed case",
-                     breaks = seq(0,1,.01)) +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_staff_confirmed$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_staff_confirmed$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_staff_confirmed")`
-
-# Interpretation
-
-This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large provider network across `r as.integer(date_max-date_min)` days.
-A comparable study was previously conducted on across 623 care homes in Ontario, Canada.[@Stall2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-
-Regional variation was noticeable, but inconsistencies between incident reports and manager counts were indicative of data quality issues, particularly in relation to incident reports.
-Rates of testing across time and care homes were not yet known at the time of writing and likely distorts measures of incidence.
-
-One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline was a protective factor.
-A 10 percentage point increase was associated with a 13% reduction [9%; 18%] in hazards of confirmed infections.
-Few organisational variables were available to identify potential confounders or mediators of this association.
-One hypothesis is that staffing ratios may be a sign of better-resourced care homes (for instance protective personal equipment), and of a greater capability to prescribe and observe effective infection control procedures.
-Furthermore, staff can become a vector in the transmission of infections.
-It is possible that higher staffing levels mitigate the transmission of infections since each staff member will care for a lower number of residents.
-Better staffed homes may also be less reliant on bank or agency staff who work in different locations.
-A more detailed survey of care home characteristics, along with protocols and measures adopted during the epidemic may help elucidate mechanisms through which staffing levels are associated with lower infection rates.
-
-The case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was higher than previous literature[@Salcher-Konrad2020], which is not exclusively attributable to the long observation period.
-An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
-These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19%.
-
-Uncertainty remains regarding the internal validity of estimates presented in this report.
-It was not possible to determine each care home's outbreak status with confidence.
-Consequently, incidence was calculated across all homes, unlike the study by Stall et al.[@Stall2020] Beyond uncertainty in the exact number of cases and outbreaks, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
-Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
-According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
-Consequently findings presented in the current study are likely to be underestimates.
-This includes estimates of the incidence ratio comparing care homes with the community in England.
-
-Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection.[@Dora2020] Although increasing testing capacity raises the possibility of undertaking frequent, repeat testing in the care home population, this may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
-Understanding which care homes are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention accordingly.
-
-# Declarations
-
-## Funding
-
-This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].
-
-## Research ethics
-
-This study was approved by the University College London Research Ethics Committee (project reference 13355/002).
-
-# Supplementary tables
-
-`r table_nums("supp_epi_CT")`
-
-|               |  Symptomatic  | Residents  <br> Confirmed  | Symptomatic | Staff <br> Confirmed   |
-| ------------- | -------------:| -------------:|-------------:|-------------:|
-| Cases      | `r formatbm(desc_residents$CT_total_symptomatic)` |	`r formatbm(desc_residents$CT_total_confirmed)` |	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
-|	$N$ exposed  | `r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
-|	Total exposure (days)| `r formatbm(desc_residents$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents$CT_total_confirmed_denom)`	|`r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
-|	Incidence proportion (%)  | `r tbrounding(desc_residents$CT_p_symptomatic)` [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents$CT_p_confirmed)` [`r tbrounding(desc_residents$CT_p_confirmed_lower)`; `r tbrounding(desc_residents$CT_p_confirmed_upper)`] | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
-|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$CT_IR_symptomatic)` [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
-
-`r table_nums("supp_epi_DAT")`
-
-| Residents     |  Symptomatic  |  Confirmed  | All-cause death | 
-| ------------- | -------------:| -------------:|-------------:|
-| Cases      | `r formatbm(desc_residents$DAT_total_symptomatic)` |	`r formatbm(desc_residents$unique_residents_test_result_positive)` | `r formatbm(desc_residents$DAT_total_deaths)`  |
-|	$N$ exposed  | `r formatbm(desc_residents$total_residents)`|	`r formatbm(desc_residents$total_residents)` | `r formatbm(desc_residents$total_residents)` |
-|	Total exposure (days)|`r formatbm(desc_residents$DAT_total_symptomatic_denom)`|	`r formatbm(desc_residents$DAT_total_confirmed_denom)` | `r formatbm(desc_residents$total_denom)` |
-|	Incidence proportion (%)  | `r tbrounding(desc_residents$DAT_p_symptomatic)` [`r tbrounding(desc_residents$DAT_p_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`] |`r tbrounding(desc_residents$DAT_p_confirmed)` [`r tbrounding(desc_residents$DAT_p_confirmed_lower)`; `r tbrounding(desc_residents$DAT_p_confirmed_upper)`] |`r tbrounding(desc_residents$DAT_p_death)` [`r tbrounding(desc_residents$DAT_p_death_lower)`; `r tbrounding(desc_residents$DAT_p_death_upper)`] |
-|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$DAT_IR_symptomatic)` [`r tbrounding(desc_residents$DAT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_IR_symptomatic_upper)`] | `r tbrounding(desc_residents$DAT_IR_confirmed)`  [`r tbrounding(desc_residents$DAT_IR_confirmed_lower)`; `r tbrounding(desc_residents$DAT_IR_confirmed_upper)`] | `r tbrounding(desc_residents$DAT_IR_death)` [`r tbrounding(desc_residents$DAT_IR_death_lower)`; `r tbrounding(desc_residents$DAT_IR_death_upper)` ]|
-
 # References

### 2020-06-27 covidWarehouse/R/compute.R

@@ -422,7 +422,39 @@ table_weekly_rate_tests <- function() {
   output
 }
 
+#' Compute the total number of cases in the home
+#'
+#' @param occupancy a numeric vector of total number of residents present
+#' in the home for every date
+#' @param new_cases a numeric vector of new cases recorded in the home at
+#' every date
+#' @param order_by a vector of dates or day numbers
+#'
+#' @return a numeric vector of total new and previous cases currently in
+#' the home
+#' @export
+total_cases_in_home <- function(occupancy, new_cases, order_by) {
 
+  if(any(is.na(order_by) | duplicated(order_by))) {
+    stop("order_by must be unique and non missing")
+  }
+
+  sort_vector <- order(order_by)
+  reverse_sort_vector <- order(sort_vector)
+
+  occupancy <- occupancy[sort_vector]
+  new_cases <- new_cases[sort_vector]
+
+  occupancy_change_rate <- pmin(1, tidyr::replace_na(occupancy/lag(occupancy), 1))
+
+  stopifnot(all(occupancy_change_rate <= 1 & occupancy_change_rate >= 0))
+  I_t <- 0
+  for(i in 2:length(occupancy)){
+    I_t[i] <- (I_t[i-1] + new_cases[i-1]) * occupancy_change_rate[i]
+  }
+
+  I_t[reverse_sort_vector]
+}
 
 #' Deduplicated cases and events
 #'

### 2020-06-27 covidWarehouse/tests/testthat/test-data.R

@@ -47,3 +47,20 @@ test_that("resident bed day approximation", {
                c(0, 1, 1, 0, 1, 0, 1, 0))
 })
 
+test_that("total_cases_in_home", {
+  data <- dplyr::tibble(
+    date = 1:5,
+    occupancy = c(3000,2990,2980,2975,2971),
+    new_cases = c(10,8,9,12,8)
+    )
+  truncated_result <- c(0, 9, 17, 26, 38)
+
+  expect_equal(trunc(total_cases_in_home(data$occupancy, data$new_cases, data$date)),
+               truncated_result)
+
+  data <- rev(data)
+  truncated_result <- rev(truncated_result)
+  expect_equal(trunc(total_cases_in_home(data$occupancy, data$new_cases, data$date)),
+               truncated_result)
+
+})

### 2020-06-30 covidWarehouse/manuscript/manuscript.Rmd
update data till 14 June

@@
 date_min <- as.Date("2020-03-02")
-date_max <- as.Date("2020-06-07") #as.Date("2020-06-15")
+date_max <- as.Date("2020-06-14")
 date_min_tallies = as.Date("2020-03-24")
-date_max_tallies = as.Date("2020-06-07")
+date_max_tallies = as.Date("2020-06-14")
 
@@
 
-occupancy_interpolated <- resident_days_linear_approx() %>% 
-  # TODO: currently putting the individual-level occup for 360 because no occ tally is known!
-  # it's got 57 bed including 5 contract beds, so 
-  left_join(transmute(filter(approx_occupancy, home_code == "360"), 
-                      date, home_code  , occup360 = trunc(occupancy * 57/52)),
-            by = c("home_code", "date")) %>%
-  mutate(occupancy = if_else(home_code == "360", occup360, occupancy)) %>% 
-  select(-occup360)
+occupancy_interpolated <- resident_days_linear_approx() #%>% 
+  # # TODO: currently putting the individual-level occup for 360 because no occ tally is known!
+  # # it's got 57 bed including 5 contract beds, so 
+  # left_join(transmute(filter(approx_occupancy, home_code == "360"), 
+  #                     date, home_code  , occup360 = trunc(occupancy * 57/52)),
+  #           by = c("home_code", "date")) %>%
+  # mutate(occupancy = if_else(home_code == "360", occup360, occupancy)) %>% 
+  # select(-occup360)
 
@@
   left_join(home_mean_occupancy) %>% 
-  mutate(pct_occupancy = mean_occupancy/beds)
+  mutate(pct_occupancy = mean_occupancy/beds,
+         occ_bedroom_ratio = mean_occupancy/no_bedrooms)
 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, beds, pct_occupancy)) %>% 
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, 
+                   beds, pct_occupancy, occ_bedroom_ratio, no_bedrooms)) %>% 
   group_by(resident_id) %>% 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, beds, beds, pct_occupancy)) %>% 
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
+                   beds,  pct_occupancy, occ_bedroom_ratio, no_bedrooms)) %>% 
   left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
@@
 
-home_total_confirmed <- distinct(reference_homes, home_code, home_name_clickview, postcode) %>% 
-  left_join(select(reference_geography, postcode, country, region_UK), by = "postcode") %>% 
+home_total_confirmed <- distinct(reference_homes, home_code, home_name_clickview, 
+                                 country, region_UK, staff_to_bed_ratio, carer_to_bed_ratio, 
+                                 bed_tot_category, beds, pct_occupancy, no_bedrooms, occ_bedroom_ratio,
+                                 younger_person, residential_general, residential_dementia,
+                                 nursing_general,nursing_dementia) %>% 
   left_join(res_counts, by = "home_code") %>% 
@@
 residents_ct <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-  left_join(distinct(new_cases, date, home_code, 
+  left_join(select(new_cases, date, home_code, 
                      new_confirmed_home, new_confirmed_hospital,
-                     new_symptomatic_home, new_symptomatic_hospital)) %>% 
+                     new_symptomatic_home, new_symptomatic_hospital,
+                     new_deaths_home, new_deaths_hospital)) %>% 
   filter(!is.na(home_code)) %>% 
-  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, staff_total, nursing_dementia, nursing_general, residential_dementia, residential_general, younger_person)) %>% 
+  left_join(distinct(reference_homes, home_code, home_name_datix, postcode, 
+                     staff_total, nursing_dementia, nursing_general, 
+                     residential_dementia, residential_general, younger_person)) %>% 
   left_join(select(reference_geography, postcode, region_UK)) %>% 
@@
          new_confirmed = tidyr::replace_na(new_confirmed_home, 0) +
-                                    tidyr::replace_na(new_confirmed_hospital, 0)) %>% 
+                                    tidyr::replace_na(new_confirmed_hospital, 0),
+         new_deaths = tidyr::replace_na(new_deaths_home, 0) +
+                                    tidyr::replace_na(new_deaths_hospital, 0)) %>% 
   mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
@@
 
+km_ct_residents_deaths <- residents_ct %>% 
+  group_by(date) %>% 
+  summarise(
+    deaths = sum(new_deaths),
+    occupancy = sum(occupancy)
+  ) %>% 
+  km_ct_estimator(df = ., t = date, d = deaths, 
+                  pop = occupancy, subtract_cases = FALSE, overwrite = TRUE) %>% 
+  ungroup() %>% 
+  mutate(
+    St_max = S_t^exp(1.96*S_t_SElog),
+    St_min = S_t^exp(-1.96*S_t_SElog),
+    event = "Death"
+  )
+
 km_ct_residents_confirmed <- residents_ct %>% 
-  group_by(date, region_UK) %>% 
+  group_by(date) %>% 
   summarise(
@@
   ) %>% 
-  group_by(region_UK) %>% 
   km_ct_estimator(df = ., t = date, d = confirmed, 
@@
     St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
-  )
-
-# residents_ct_symp <- tibble(date = seq(date_min, date_max_tallies, 1)) %>% 
-#   left_join(distinct(new_cases, date, home_code, 
-#                      new_symptomatic_home, new_symptomatic_hospital)) %>% 
-#   filter(!is.na(home_code)) %>% 
-#   left_join(distinct(reference_homes, home_code, home_name_datix, postcode)) %>% 
-#   left_join(select(reference_geography, postcode, region_UK)) %>% 
-#   left_join(occupancy_interpolated) %>% 
-#   arrange(date)
+    St_min = S_t^exp(-1.96*S_t_SElog),
+    event = "Confirmed infection"
+  )
 
 km_ct_residents_symp <- residents_ct %>% 
-  group_by(date, region_UK) %>% 
+  group_by(date) %>% 
   summarise(
@@
   ) %>% 
-  group_by(region_UK) %>% 
   km_ct_estimator(df = ., t = date, d = symptomatic, 
@@
     St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
+    St_min = S_t^exp(-1.96*S_t_SElog),
+    event = "Symptomatic case"
   )
@@
 km_staff_symptomatic <- staff_ct %>% 
-  group_by(date, region_UK) %>% 
+  group_by(date) %>% 
   summarise(
@@
   ) %>% 
-  group_by(region_UK) %>% 
   km_ct_estimator(., t=date, d=symptomatic, pop=(staff_total-I_symptomatic),
@@
 km_staff_confirmed <- staff_ct %>% 
-  group_by(date, region_UK) %>% 
+  group_by(date) %>% 
   summarise(
@@
   ) %>% 
-  group_by(region_UK) %>% 
   km_ct_estimator(., t=date, d=confirmed, pop=(staff_total-I_confirmed),
@@
     St_max = S_t^exp(1.96*S_t_SElog),
-    St_min = S_t^exp(-1.96*S_t_SElog)
+    St_min = S_t^exp(-1.96*S_t_SElog),
+    event = "Confirmed infection"
   )
@@
 desc_residents$DAT_total_residents <- dplyr::n_distinct(residents_overview$resident_id)
-desc_residents$DAT_total_residents_england <- dplyr::filter(residents_overview, country == "England") %>% summarise(n = dplyr::n_distinct(resident_id)) %>% .$n
 
@@
 # This is estimated in validation.Rmd
-desc_residents$CT_total_residents <- round(desc_residents$DAT_total_residents * 1.117309 + .49)
-desc_residents$CT_total_residents_england <- round(desc_residents$DAT_total_residents_england * 1.0793268 + .49)
+desc_residents$CT_total_residents <- tline %>% 
+  dplyr::filter(between(date, date_min_tallies, date_max_tallies) & rday == 1) %>% summarise(n = round(dplyr::n_distinct(resident_id) * 1.123865 + .49)) %>% .$n
+
+desc_residents$CT_total_residents_england <- tline %>% 
+  dplyr::inner_join(select(filter(reference_homes, country == "England"), home_code)) %>%
+  dplyr::filter(between(date, date_min_tallies, date_max_tallies) & rday == 1) %>%
+  summarise(n = round(dplyr::n_distinct(resident_id) * 1.087621 + .49)) %>% .$n
+
 desc_residents$CT_total_symptomatic <- sum(new_cases$new_symptomatic_home, na.rm=T) + 
@@
   dplyr::filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
-  dplyr::summarise(total_residents = round(dplyr::n_distinct(resident_id)* 1.1151324 + .49)) %>% 
+  dplyr::summarise(total_residents = round(dplyr::n_distinct(resident_id)* 1.122205 + .49)) %>% 
   .$total_residents
@@
 
-fig_KM_CT_residents_symptomatic <- fig_nums(
-  name = "fig_KM_CT_residents_symptomatic",
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of residents becoming symptomatic based on care home manager daily counts")
-)
-fig_KM_residents_confirmed <- fig_nums(
-  name = 'fig_KM_residents_confirmed',
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of residents (n=",
-  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
-  ") becoming confirmed cases based on Datix incidents report"))
-fig_KM_CT_confirmed_residents <- fig_nums(
-  name = 'fig_KM_CT_confirmed_residents',
-  caption = "Kaplan-Meier estimates of the cumulative probability of residents becoming confirmed cases based on care home manager daily counts")
+fig_KM_CT_residents <- fig_nums(
+  name = "fig_KM_CT_residents",
+  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in residents according to FSHCG outbreak surveillance reports (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
+)
+
+fig_KM_staff <- fig_nums(
+  name = 'fig_KM_staff',
+  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infection in care home staff (n=",   formatbm(desc_staff$total_staff),
+  ") according to FSHCG outbreak surveillance reports (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")"))
+
 fig_KM_mortality <- fig_nums(
@@
   ") survival by SARS-COV-2 case type"))
-fig_KM_staff_symptomatic <- fig_nums(
-  name = 'fig_KM_staff_symptomatic',
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_staff$total_staff),
-  ") becoming symptomatic cases based on manager-reported daily counts"))
-fig_KM_staff_confirmed <- fig_nums(
-  name = 'fig_KM_staff_confirmed',
-  caption = paste0("Kaplan-Meier estimates of the cumulative probability of care home staff (n=",   formatbm(desc_staff$total_staff),
-  ") becoming confirmed cases based on manager-reported daily counts"))
-
-```
-
-```{r period_prevalence}
-cis_data <- list(
-  n_positive = 66,
-  n_negative = 60618
-)
-
-england_results <- list()
-england_results$n_positive = sum(inner_join(new_cases, filter(reference_homes, country == "England"))[c("new_confirmed_home", "new_confirmed_hospital")], na.rm = T)
-england_results$n_negative = desc_residents$CT_total_residents_england  - england_results$n_positive
-
-england_results$RR_england <- relative_risk(
-  a = england_results$n_positive,
-  b = england_results$n_negative,
-  c = cis_data$n_positive,
-  d = cis_data$n_negative)
-
-england_results$RR_SE <- relative_risk_SE(
-  a = england_results$n_positive,
-  b = england_results$n_negative,
-  c = cis_data$n_positive,
-  d = cis_data$n_negative)
+
+```
+
+```{r period_prevalence, message=F}
+# Cumulative sample counts, directly standardised for
+# 11 May to 24 May
+# 25 May to 7 June
+
+england_results <- list(
+  date_min = as.Date("2020-05-11"),
+  date_max = as.Date("2020-06-07"),
+  cis_confirmed = 35,
+  cis_person_days = 483259
+)
+
+england_results$FSHC_confirmed <- new_cases %>% 
+  filter(between(date, england_results$date_min, england_results$date_max)) %>% 
+  inner_join(filter(reference_homes, country == "England")) %>% 
+  summarise(
+    n = sum(new_confirmed_home, na.rm = T) + sum(new_confirmed_hospital, na.rm = T)
+  ) %>% 
+  .$n
+
+england_results$FSHC_resident_days <- filter(km_ct_residents_confirmed,
+  between(date, england_results$date_min, england_results$date_max)) %>% 
+  summarise(n = sum(n_t, na.rm = T) ) %>% 
+  .$n
+
+england_results$results <- epitools::rateratio.wald(
+  c(england_results$cis_confirmed,
+    england_results$FSHC_confirmed,
+    england_results$cis_person_days,
+    england_results$FSHC_resident_days))
 ```
@@
 Across all care homes, the confirmed infection rate of incidence was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-In England, `r england_results$n_positive` infections were confirmed of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
-Comparing with the community, the relative risk of confirmed cases in England was `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
 These estimates include `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) that had not reported a single confirmed case.
-
-The staff-to-bed ratio was the only factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
+In a comparison with results from an English household survey, we estimated a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`] in care homes compared with the community in England between `r format(england_results$date_min, "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
+The staff-to-bed ratio was the main organisational factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
 Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted(
@@
 
-## Aggregate outbreak data (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")` )
+## Aggregate outbreak data (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")`)
 
@@
 
-## Individual-level residents data (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")` )
+## Individual-level residents data (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")`)
 
@@
 Admitted cases were accounted for in the daily new case count.
-Based on this $\hat{E_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as
-
-$$\hat{E_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad E_{t_0} = 0$$
-where $d_{t}$ denotes the number of new cases for each day $t = \{1, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate.
-The total exposure time was calculated as $ \sum_{t=1}^T (N_t - \hat{E_t})$ where $N_t$ denotes daily occupancy figures.
+Based on this $\hat{I_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as
+
+$$\hat{I_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad I_{t_0} = 0$$
+where $d_{t}$ denotes the number of new cases for each day $t = \{0, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate.
+The total exposure time was calculated as $\sum_{t=0}^{T} (N_t-\hat{I}_t)$ where $N_t$ denotes daily occupancy figures.
 The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
@@
 
-FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from ***XXXXX***.
+FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from 6 April 2020.
 From 11 May 2020, most FSHCG home undertook testing every resident as part of the whole home testing programme.[@dhscTesting2020] Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` residents had an infection confirmed in homes while `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` had an infection confirmed in hospital.
@@
 
-Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`, `r fig_nums("fig_KM_CT_residents_confirmed", display = "cite")`) indicate wide variation across the UK.
+Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`, `r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`) indicate wide variation across the UK.
 We note a strong progression of infections in the North West, North East and South East of England, where more than 1 in 5 residents had a confirmed infection.
@@
 ```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 8}
-order_key <-km_ct_residents_symp %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_ct_residents_symp$region_UK <- factor(
-  km_ct_residents_symp$region_UK,
-  levels = order_key$region_UK
-)
-km_ct_residents_symp$region_group <- cut(as.integer(km_ct_residents_symp$region_UK), 2, labels = 1:2)
+all_km_residents <- bind_rows(
+  km_ct_residents_symp,
+  km_ct_residents_confirmed,
+  km_ct_residents_deaths
+)
+all_km_residents$Event <- factor(
+  all_km_residents$event,
+  levels = c(
+    "Symptomatic case",
+    "Confirmed infection",
+    "Death"
+  )
+)
 
@@
            ymin = 1 - St_min,
-           color = region_UK), data = km_ct_residents_symp) +
+           color = Event), data = all_km_residents) +
   # facet_grid(region_group ~ .) +
@@
   geom_path(lwd=1.2) +
-  scale_y_continuous("Cum probability symptomatic case",
+  scale_y_continuous("Cum probability",
                      breaks = seq(0,1,.05)) +
-  scale_color_discrete("Region") +
+  ggsci::scale_color_lancet() +
   scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_ct_residents_symp$date)),
+               breaks = unique(get_monday_date(all_km_residents$date)),
                labels = format.Date(
-                 unique(get_monday_date(km_ct_residents_symp$date)),
+                 unique(get_monday_date(all_km_residents$date)),
                  "%d/%m/%y"
@@
 
-`r fig_nums("fig_KM_CT_residents_symptomatic")`
-
-```{r KM_CT_residents_infection, fig.height = 4, fig.width = 8}
-order_key <-km_ct_residents_confirmed %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_ct_residents_confirmed$region_UK <- factor(
-  km_ct_residents_confirmed$region_UK,
-  levels = order_key$region_UK
-)
-km_ct_residents_confirmed$region_group <- cut(as.integer(km_ct_residents_confirmed$region_UK), 2, labels = 1:2)
+`r fig_nums("fig_KM_CT_residents")`
+
+For England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format((england_results$date_man), "%e %B %Y")`.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$FSHC_confirmed` confirmed cases out of `r formatbm(england_results$FSHC_confirmed)` person-days during this period.
+This implies a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
+
+Incidence proportions and rates were also estimated separately for private residents using the Datix incident reports dataset (supplementary material 1-2).
+Estimates are lower than estimates in `r table_nums("supp_epi_CT", display = "cite")`.
+This is thought to be caused by a combination of underascertainment as well as linkage error affecting incidents reports.
+
+### Staff cases
+
+Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were reported to have symptoms by FSHCG managers between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+During the same period, `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff were reported to have had a confirmed infection.
+Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) once again indicate a strong progression of infections in the North West, North East and South East of England.
+
+```{r KM_Staff, fig.height = 4, fig.width = 8}
+all_km_staff <- bind_rows(
+  km_staff_symptomatic,
+  km_staff_confirmed
+)
+all_km_staff$Event <- factor(
+  all_km_staff$event,
+  levels = c(
+    "Symptomatic case",
+    "Confirmed infection"
+  )
+)
 
@@
            ymin = 1 - St_min,
-           color = region_UK), data = km_ct_residents_confirmed) +
+           color = Event), data = all_km_staff) +
   # facet_grid(region_group ~ .) +
@@
   geom_path(lwd=1.2) +
-  scale_y_continuous("Cum probability confirmed case",
+  ggsci::scale_color_lancet()+
+  scale_y_continuous("Cum probability symptomatic case",
                      breaks = seq(0,1,.02)) +
-  scale_color_discrete("Region") +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_ct_residents_confirmed$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_ct_residents_confirmed$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_CT_confirmed_residents")`
-
-For England only, a total of `r england_results$n_positive` confirmed cases occurred out of a total `r formatbm(england_results$n_positive + england_results$n_negative)` residents.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r cis_data$n_positive` confirmed cases out of `r formatbm(cis_data$n_positive + cis_data$n_negative)` sampled individuals.
-This implies a relative risk of `r tbrounding(england_results$RR_england)` [`r tbrounding(exp(log(england_results$RR_england) - 1.96 * england_results$RR_SE)) `; `r tbrounding(exp(log(england_results$RR_england) + 1.96 * england_results$RR_SE))`].
-
-Incidence proportions and rates were also estimated for private residents only using the Datix incident reports dataset (supplementary material 1-2).
-Estimates are lower than corresponding estimates based on outbreak counts.
-This is thought to be caused by a combination of underascertainment as well as linkage error.
-
-### Staff cases
-
-Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were recorded as having symptoms by FSHCG managers.
-In total, `r formatbm(desc_staff$CT_confirmed_total)` (	`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff had a confirmed infection by the end of the period.
-Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) indicate a strong progression of infections in the North West, North East and South East of England.
-
-```{r KM_Staff_Symptomatic, fig.height = 4, fig.width = 8}
-
-order_key <-km_staff_symptomatic %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_staff_symptomatic$region_UK <- factor(
-  km_staff_symptomatic$region_UK,
-  levels = order_key$region_UK
-)
-km_staff_symptomatic$region_group <- cut(as.integer(km_staff_symptomatic$region_UK), 2, labels = 1:2)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,
-           ymin = 1 - St_min,
-           color = region_UK), data = km_staff_symptomatic) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  scale_color_discrete("Region") +  
-  scale_y_continuous("Cum probability symptomatic case",
-                     breaks = seq(0,1,.05)) +
   scale_x_date(name = "Date", 
@@
 
-`r fig_nums("fig_KM_staff_symptomatic")`
-
-```{r KM_Staff_Confirmed, fig.height = 4, fig.width = 8}
-
-order_key <- km_staff_confirmed %>% 
-  group_by(region_UK) %>% 
-  summarise(S_t = min(S_t)) %>% 
-  arrange(S_t)
-km_staff_confirmed$region_UK <- factor(
-  km_staff_confirmed$region_UK,
-  levels = order_key$region_UK
-)
-km_staff_confirmed$region_group <- cut(as.integer(km_staff_confirmed$region_UK), 2, labels = 1:2)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,#(S_t + 1.96 * S_t_SE),
-           ymin = 1 - St_min, #(S_t - 1.96 * S_t_SE),
-           color = region_UK), data = km_staff_confirmed) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  scale_color_discrete("Region") +
-  scale_y_continuous("Cum probability confirmed case",
-                     breaks = seq(0,1,.01)) +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_staff_confirmed$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_staff_confirmed$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_staff_confirmed")`
+`r fig_nums("fig_KM_staff")`
 
@@
 
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection, whether acquired in the home or in hospital (`r table_nums("tab_zero_cases", display="cite")`).
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection, whether acquired in the home or in hospital (supplementary material 3).
+Incidence estimates are reported for those homes separately in `r table_nums("tab_zero_cases", display="cite")`.
 
@@
 
+```{r factors_outbreak}
+home_total_confirmed$outbreak <-  as.factor(home_total_confirmed$CT_confirmed>0)
+# # factors_outbreak <- glm( 
+# #   outbreak ~ staff_to_bed_ratio + bed_tot_category + 
+# #     pct_occupancy +occ_bedroom_ratio + region_UK , 
+# #   family = binomial(), data = home_total_confirmed)
+# # summary(factors_outbreak)
+# explanatory = c( "bed_tot_category", "pct_occupancy", "staff_to_bed_ratio",
+#                  "occ_bedroom_ratio",
+#                  #"younger_person", "residential_general", "factor(residential_dementia)",
+#                  # "nursing_general","nursing_dementia", 
+#                  "region_UK")
+# dependent = "outbreak"
+# odds_outbreak <-  finalfit(home_total_confirmed,
+#                            dependent, 
+#                            explanatory, column = T,
+#                            add_dependent_label = F)  
+# knitr::kable(odds_outbreak,
+#              row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
+#  
+# ggplot(home_total_confirmed, aes(occ_bedroom_ratio, pct_occupancy, colour = outbreak)) +
+#   geom_point()
+```
+
 ## Factors associated with confirmed infections
 
-Using the individual-level data on private bed residents (`table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+Using the individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
 According to this, and other than geographical location, the staff-to-bed ratio was the only factor significantly associated with the rate of confirmed cases.
@@
 ```{r CPH_confirmed,   results="asis"}
-cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + scale(pct_occupancy, scale = F), data = residents_overview)
- summary(cph_conf)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + pct_occupancy+occ_bedroom_ratio, data = residents_overview)
+#  summary(cph_conf)
 # cox.zph(cph_conf)
@@
 #                       staff_to_bed_ratio + bed_tot_category +(region_UK)+
-#                       home_type*covid_status+ cluster(resident_id), data = residents_episodes)
+#                       pct_occupancy + occ_bedroom_ratio + cluster(resident_id), data = residents_episodes)
 # summary(cph_death2)
@@
 # survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
-# 
-explanatory = c("gender", "ageg", "user_type2", "home_type",
-                "staff_to_bed_ratio", "bed_tot_category", 
+#  pct_occupancy, occ_bedroom_ratio,no_bedroom 
+explanatory = c("gender", "ageg", "home_type", #"user_type2", 
+                "staff_to_bed_ratio", "bed_tot_category", "pct_occupancy", # "occ_bedroom_ratio",
                 "covid_status", "region_UK", "cluster(resident_id)")
-explanatory_interaction = c("gender", "ageg", "user_type2",
-                "staff_to_bed_ratio", "bed_tot_category", 
-                "covid_status:home_type", "region_UK", "cluster(resident_id)")
+# explanatory_interaction = c("gender", "ageg", "user_type2",
+#                 "staff_to_bed_ratio", "bed_tot_category", 
+#                 "covid_status:home_type", "region_UK", "cluster(resident_id)")
 dependent = "Surv(time = date1, time2 = date2, event = death_status)"
@@
 
-## Supplementary material 2: Incidence of confirmed and symptomatic cases among private residents according to Datix incident reports.
+## Supplementary material 2: Incidence of confirmed and symptomatic cases among private residents according to Datix incident reports
 

### 2020-07-01 covidWarehouse/R/database.R
overall refactoring

@@ -1,61 +1,6 @@
 
 
 
-
-get_db_conn <- function() {
-  getOption("fshc_warehouse")
-}
-
-#' Create or overwrite the \code{reference_homes} table
-#'
-#' @param conn a database connection
-#' @param overwrite whether to overwrite the table (default is TRUE)
-#' @return the name of the table if data was successfully loaded
-#' @importFrom dplyr tbl
-#' @export
-create_table_reference_homes <- function(homes = reference_homes,
-                                         conn = get_db_conn(), overwrite = TRUE){
-
-  dbplyr::db_copy_to(con = conn,
-                     table = "reference_homes",
-                     values = homes,
-                     overwrite = overwrite,
-                     temporary = FALSE,
-                     indexes = list(
-                       "home_code",
-                       "postcode"
-                     ))
-
-}
-
-
-#' Create or overwrite \code{reference_geography} table
-#'
-#' @param conn a database connection
-#' @param data a data frame, set to  `reference_geography` by default.
-#' @param overwrite whether to overwrite the `reference_geography` table.
-#' Default is `TRUE`.
-#' @importFrom magrittr %>%
-#' @importFrom dplyr group_by summarise select filter
-#' @return
-#' @export
-#' @importFrom dplyr tbl
-create_table_reference_geography <- function(conn = get_db_conn(),
-                                       data = reference_geography,
-                                       overwrite = TRUE){
-
-  dbplyr::db_copy_to(con = conn,
-                     table = "reference_geography",
-                     values = data,
-                     overwrite = overwrite,
-                     temporary = FALSE,
-                     indexes = list(
-                       "postcode"
-                     ))
-
-}
-
-
 check_homes <- function(home_codes, call = NULL) {
 
   if(!file.exists(file.path(getOption("fshc_files"), "reference_homes.rda"))) {

### 2020-07-01 manuscript/manuscript.Rmd
general refactoring 2

--- 2020-06-30 covidWarehouse/manuscript/manuscript.Rmd
@@
 home_mean_occupancy <- occupancy_interpolated  %>% 
-  filter(between(date, date_min_tallies, date_max_tallies)) %>% 
+  filter(between(date, as.Date("2020-01-01"), as.Date("2020-03-14"))) %>%
   group_by(home_code) %>% 
@@
          occ_bedroom_ratio = mean_occupancy/no_bedrooms)
+
+reference_homes <- reference_homes %>% 
+  mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
+         occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)),
+         region_UK = as.factor(region_UK)) %>% 
+  mutate(occ_bedroom_ratio_g6 = relevel(occ_bedroom_ratio_g6, '(0·7,0·85]'),
+         occ_bedroom_ratio_g5 = relevel(occ_bedroom_ratio_g5, '(0·8,1]'),
+         region_UK <- relevel(region_UK, "Scotland"))
 
@@
                    staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, 
-                   beds, pct_occupancy, occ_bedroom_ratio, no_bedrooms)) %>% 
+                   beds, pct_occupancy, no_bedrooms, 
+                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
   group_by(resident_id) %>% 
@@
                    staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
-                   beds,  pct_occupancy, occ_bedroom_ratio, no_bedrooms)) %>% 
+                   beds,  pct_occupancy, no_bedrooms, 
+                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
   left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
@@
                                  country, region_UK, staff_to_bed_ratio, carer_to_bed_ratio, 
-                                 bed_tot_category, beds, pct_occupancy, no_bedrooms, occ_bedroom_ratio,
+                                 bed_tot_category, beds, pct_occupancy, no_bedrooms, 
+                                 occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6,
                                  younger_person, residential_general, residential_dementia,
@@
            tidyr::replace_na(new_confirmed_hospital, 0),
-         new_deaths = tidyr::replace_na(new_deaths_home, 0) +
-                                    tidyr::replace_na(new_deaths_hospital, 0)) %>% 
+         new_deaths = tidyr::replace_na(new_deaths_home, 0) # +                                    tidyr::replace_na(new_deaths_hospital, 0)
+         ) %>% 
   mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
@@
 # ggplot(home_total_confirmed, aes(occ_bedroom_ratio, pct_occupancy, colour = outbreak)) +
-#   geom_point()
+#   geom_point() +
+#   ggsci::scale_color_lancet() +
+#   xlab("Ratio mean occupancy:number bedroom") + ylab("Ratio mean occupancy:beds")
 ```
@@
 # survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
-explanatory = c("gender", "ageg", "home_type", "user_type2", "region_UK" ,  
-                "staff_to_bed_ratio", "bed_tot_category", "pct_occupancy")
+explanatory = c("gender", "ageg", "home_type", "user_type2",  
+                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6", "region_UK" )
 dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
@@
 # cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg +
+#                       gender+ ageg +covid_status+
 #                       staff_to_bed_ratio + bed_tot_category +(region_UK)+
@@
 explanatory = c("gender", "ageg", "home_type", #"user_type2", 
-                "staff_to_bed_ratio", "bed_tot_category", "pct_occupancy", # "occ_bedroom_ratio",
+                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
                 "covid_status", "region_UK", "cluster(resident_id)")

### 2020-07-01 assets/js/testmap.js

@@ -1,39 +1,0 @@
-
-Highcharts.getJSON('https://ons-inspire.esriuk.com/arcgis/rest/services/Eurostat_Boundaries/NUTS_Level_1_January_2018_Boundaries/MapServer/3/query?where=1%3D1&outFields=*&outSR=3857&f=geojson', function (geojson) {
-    chart = Highcharts.mapChart('covid-map', {
-        chart: {
-            map: geojson
-        },
-        title: {
-            text: 'GeoJSON in Highmaps'
-        },
-
-        mapNavigation: {
-            enabled: true,
-            buttonOptions: {
-                verticalAlign: 'bottom'
-            }
-        },
-        colorAxis: {
-            tickPixelInterval: 100
-        },
-        data: {
-            csvURL:  window.location.origin + '/assets/data/testdata.csv',
-            enablePolling: true,
-            seriesMapping: [{
-                nuts118cd: 0,
-                value: 1
-            }]
-        },
-        series: [{
-            name: 'Random data',
-            joinBy: 'nuts118cd',
-            borderColor: '#606060',
-            nullColor: '#bfbfbf',
-            dataLabels: {
-                enabled: true,
-                format: '{point.properties.nuts118nm}'
-            }
-        }]
-    });
-});

### 2020-07-01 manuscript/manuscript.Rmd
add maps

@@
 **Background** COVID-19 has caused substantial morbidity and mortality in care home residents and staff.
-National surveys suggest infection is declining in the general population in the UK, but outbreaks persist in care homes.
+National surveys suggest infections are declining in the general population in the UK, but outbreaks persist in care homes.
 Epidemiological data on disease trends in care home staff and residents are scarce.
@@
 We estimated the incidence proportion and incidence rate of symptoms, confirmed infection, and death in residents and staff, and compare them with findings from a national community survey.
-We tested a range of risk factors for resident infections in multivariate regression models.
+We tested a range of risk factors for resident infection and mortality in multivariate Cox proportional hazards models.
 
 **Findings:** An estimated `r formatbm(desc_residents$CT_total_residents)` residents were present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-During the period `r formatbm(desc_residents$CT_total_symptomatic_denom)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
-Across all care homes, the confirmed infection rate of incidence was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-These estimates include `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) that had not reported a single confirmed case.
-In a comparison with results from an English household survey, we estimated a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`] in care homes compared with the community in England between `r format(england_results$date_min, "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
-The staff-to-bed ratio was the main organisational factor to be independently associated with the rate of confirmed cases: an increase by 10 percentage point in the number of staff relative to the bed total was associated with a 13% reduction [9%, 18%] in the rate of confirmed cases.
+During the period `r formatbm(desc_residents$CT_total_symptomatic)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
+The confirmed infection incidence rate was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] across all care homes, including `r desc_residents$CT_zero_case` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) that had not reported a single confirmed case.
+Within England, the confirmed infection rate was `r tbrounding(england_results$results$measure[2,"estimate"])` times [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`] greater in care homes than in community households.
 Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted(
@@
 
+Several risk factors such as being male or aged 85 years and over were independently associated with residents' confirmed infection rate.
+An increase by 10 percentage points in the staff-to-bed ratio was associated with a 25% reduction [19%, 31%] in the rate of confirmed cases.
+The ratio of occupants to bedrooms measured at the beginning of the epidemic had a strong positive association with the infection rate.
+Hazards of confirmed infection in homes with a ratio greater than 1.00 were 4·9 times [2·8; 8·4] greater than in homes with a ratio of 0.70 to 0.85.
+Mortality was strongly associated with both being male and age.
+An increase by 10 percentage points in the staff-to-bed ratio was associated with a 25% reduction [19%, 31%] in the rate of confirmed cases.
+The ratio of occupants to bedrooms measured at the beginning of the epidemic had a strong positive association with the infection rate.
+Hazards of confirmed infection in homes with a ratio greater than 1.00 were 4·9 times [2·8; 8·4] greater than in homes with a ratio of 0.70 to 0.85.
+All-cause mortality was strongly associated with being male, age, but not rates of staffing or occupancy.
+Adjusted mortality hazard ratios for asymptomatic and symptomatic confirmed cases respectively were 2.4 [1.4; 4.0] and 10 [8.3; 12] compared with residents that were asymptomatic and never tested positive.
+
 **Interpretation** Despite uncertain case ascertainment in care homes, internal surveillance suggests a greater transmission than in the community.
-Staffing may impact care homes' response to outbreaks and in turn influence rates of confirmed infections The absence of high-quality case ascertainment hinders local infection prevention initiatives which could be better supported by laboratory data surveillance.
+Organisational factors such as staffing and occupancy are strongly associated with rates of confirmed infections and can consequently place residents at increased risks of death.
+The absence of high-quality case ascertainment across the social care sector hinders local infection prevention initiatives.
 
@@
 ```
+
+```{r FSHCG_map, fig.width=3, fig.height=5, warning=F}
+reference_homes_lat_long <- reference_homes %>% 
+  dplyr::filter(home_code %in% home_inclusion) %>% 
+  dplyr::select(home_code, home_name_datix, postcode) %>% 
+  dplyr::left_join(reference_geography, by = "postcode")
+
+care_homes_map <- ggplot(UK_country_region_sf)  +
+  geom_sf(size = 0.1) +
+  geom_sf_text(aes(label = ctry_reg_abb), fontface = "bold") +
+  theme_void() +
+  geom_point(data = reference_homes_lat_long, aes(y = latitude, x = longitude),
+             color = "#0099B4FF") +
+  ggtitle(paste0(length(home_inclusion), " FSHCG care homes"))
+# pdf("map.pdf", width = 3, height = 5)
+care_homes_map
+# dev.off()
+```
+
+*Note:* `r paste(paste0(UK_country_region_sf$ctry_reg_abb, ": ", UK_country_region_sf$ctry_reg), collapse = "; ")`.
 
@@
 ```{r CPH_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + pct_occupancy+occ_bedroom_ratio, data = residents_overview)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + pct_occupancy+occ_bedroom_ratio_g6, data = residents_overview)
 #  summary(cph_conf)
@@
 # cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg +covid_status+
-#                       staff_to_bed_ratio + bed_tot_category +(region_UK)+
-#                       pct_occupancy + occ_bedroom_ratio + cluster(resident_id), data = residents_episodes)
+#                       gender+ ageg + home_type +covid_status+
+#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
+#                       region_UK + cluster(resident_id), data = residents_episodes)
 # summary(cph_death2)

### 2020-07-01 R/UK_country_region_sf.R

@@ -1,0 +1,17 @@
+
+#' Simple feature map of UK countries and English regions
+#'
+#' A dataset containing the ultra generalised clip boundaries of UK countries
+#' and regions of England as of 2019
+#'
+#' @format An sf object with two polygon name variables:
+#' \describe{
+#'   \item{ctry_reg}{name of the country or region (in the case of England)}
+#'   \item{ctry_reg_abb}{three-letter abbreviation of the country/region name}
+#' }
+#' @details Source: Office for National Statistics licensed under the
+#' Open Government Licence v.3.0
+#' Contains OS data (c) Crown copyright and database right [2020]
+#' @source UK countries: \url{https://geoportal.statistics.gov.uk/datasets/countries-december-2019-boundaries-uk-buc}
+#' English regions: \url{https://geoportal.statistics.gov.uk/datasets/regions-december-2019-boundaries-en-buc}
+"UK_country_region_sf"

### 2020-07-01 data-raw/UK_country_region_sf.R

@@ -1,0 +1,26 @@
+
+download_temp <- tempdir()
+library(magrittr)
+ENG_url <- "https://ons-inspire.esriuk.com/arcgis/rest/services/Administrative_Boundaries/Regions_December_2019_Boundaries_EN_BUC/MapServer/0/query?where=1%3D1&text=&objectIds=&time=&geometry=&geometryType=esriGeometryEnvelope&inSR=&spatialRel=esriSpatialRelIntersects&relationParam=&outFields=*&returnGeometry=true&returnTrueCurves=false&maxAllowableOffset=&geometryPrecision=&outSR=4326&having=&returnIdsOnly=false&returnCountOnly=false&orderByFields=&groupByFieldsForStatistics=&outStatistics=&returnZ=false&returnM=false&gdbVersion=&historicMoment=&returnDistinctValues=false&resultOffset=&resultRecordCount=&queryByDistance=&returnExtentOnly=false&datumTransformation=&parameterValues=&rangeValues=&quantizationParameters=&featureEncoding=esriDefault&f=geojson"
+UK_url <- "https://ons-inspire.esriuk.com/arcgis/rest/services/Administrative_Boundaries/Countries_December_2019_Boundaries_UK_BUC/MapServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=geojson"
+download.file(ENG_url, file.path(download_temp, "ENG.geojson"))
+download.file(UK_url, file.path(download_temp, "UK.geojson"))
+ENG_data <- sf::st_read(file.path(download_temp, "ENG.geojson")) %>%
+  dplyr::mutate(ctry_reg = rgn19nm) %>%
+  dplyr::select(-objectid, -rgn19cd, -rgn19nm)
+UK_data <- sf::st_read(file.path(download_temp, "UK.geojson")) %>%
+  dplyr::filter(ctry19cd != "E92000001") %>%
+  dplyr::mutate(ctry_reg = ctry19nm) %>%
+  dplyr::select(-objectid, -ctry19cd, -ctry19nm, -ctry19nmw)
+UK_country_region_sf <- rbind(UK_data, ENG_data)
+
+UK_country_region_sf <- dplyr::mutate(
+  UK_country_region_sf,
+  ctry_reg_abb = sapply(regmatches(ctry_reg, gregexpr("[A-Z]", ctry_reg)), paste, collapse = ""))
+
+
+usethis::use_data(UK_country_region_sf, overwrite = TRUE)
+rm(download_temp)
+
+
+

### 2020-07-02 manuscript/manuscript.Rmd
manuscript v0.3

@@
 output:
+  word_document:
+    toc: yes
+    reference_docx: template.docx
   html_document:
@@
     code_folding: hide
-  word_document:
-    toc: yes
-    reference_docx: template.docx
 link-citations: yes
@@
   mutate(
-    user_type2 = case_when(
+    user_type2 = as.factor(case_when(
       user_type %in% c("General", "Elderly") ~ "General/elderly",
@@
       TRUE ~ NA_character_ 
-      ),
+      )),
     admission_type2 = case_when(
@@
       TRUE ~ admission_type
-    ))
+    )) %>% 
+  mutate(
+    user_type2 = relevel(user_type2, "General/elderly")
+  )
 
@@
 occupancy_interpolated <- resident_days_linear_approx() #%>% 
-  # # TODO: currently putting the individual-level occup for 360 because no occ tally is known!
-  # # it's got 57 bed including 5 contract beds, so 
-  # left_join(transmute(filter(approx_occupancy, home_code == "360"), 
-  #                     date, home_code  , occup360 = trunc(occupancy * 57/52)),
-  #           by = c("home_code", "date")) %>%
-  # mutate(occupancy = if_else(home_code == "360", occup360, occupancy)) %>% 
-  # select(-occup360)
-
 home_mean_occupancy <- occupancy_interpolated  %>% 
@@
 desc_homes <- bind_rows(desc_homes, desc_national) 
-
 ```
@@
            tidyr::replace_na(new_confirmed_hospital, 0),
-         new_deaths = tidyr::replace_na(new_deaths_home, 0) # +                                    tidyr::replace_na(new_deaths_hospital, 0)
+         new_deaths = tidyr::replace_na(new_deaths_home, 0)  +
+           tidyr::replace_na(new_deaths_hospital, 0)
          ) %>% 
@@
     St_min = S_t^exp(-1.96*S_t_SElog),
-    event = "Death"
+    event = "COVID-19-related death"
   )
@@
 
+desc_residents$CT_total_cdeath <-  sum(new_cases$new_deaths_home, na.rm = T) + sum(new_cases$new_deaths_hospital, na.rm = T)
+desc_residents$CT_total_cdeath_denom <- sum(km_ct_residents_deaths$n_t)
+
+desc_residents[paste0(
+  "CT_p_cdeath", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents$CT_total_cdeath,
+                            desc_residents$CT_total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+
+desc_residents[paste0(
+  "CT_IR_cdeath", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents$CT_total_cdeath,
+                            desc_residents$CT_total_cdeath_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
 #INCIDENT REPORTS
@@
 desc_residents_outbreak_only <- list()
-desc_residents_outbreak_only$outbreak_homes <- home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0]
+desc_residents_outbreak_only$outbreak_homes <- unique(
+  home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0],
+  distinct(filter(new_cases, tidyr::replace_na(new_deaths_hospital, 0) > 0 |
+                    tidyr::replace_na(new_deaths_home, 0)>0), home_code)$home_code
+)
 
@@
 
+desc_residents_outbreak_only$CT_total_cdeath <-  sum(new_cases$new_deaths_home, na.rm = T) + sum(new_cases$new_deaths_hospital, na.rm = T)
+desc_residents_outbreak_only$CT_total_cdeath_denom <- residents_ct %>% 
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes) %>% 
+  summarise(n = sum(occupancy)) %>% 
+  .$n
+
+desc_residents_outbreak_only[paste0(
+  "CT_p_cdeath", c("", "_lower", "_upper")
+)] <- epitools::binom.exact(desc_residents_outbreak_only$CT_total_cdeath,
+                            desc_residents_outbreak_only$CT_total_residents)[
+                              c("proportion", "lower", "upper")
+                            ] * 100
+
+desc_residents_outbreak_only[paste0(
+  "CT_IR_cdeath", c("", "_lower", "_upper")
+)] <- epitools::pois.exact(desc_residents_outbreak_only$CT_total_cdeath,
+                            desc_residents_outbreak_only$CT_total_cdeath_denom)[
+                              c("rate", "lower", "upper")
+                            ] * 100000
+
 ```
@@
   caption = paste0("Breakdown of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
-supp_epi_CT <- table_nums(
-  name = "supp_epi_CT",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents and staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+incidence_CT_residents <- table_nums(
+  name = "incidence_CT_residents",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
 supp_epi_CT_oubreak_only <- table_nums(
-  name = "supp_epi_CT_oubreak_only",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among residents of outbreak care homes according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+  name = "incidence_CT_staff",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
 tab_CoxPH_confirmed <- table_nums(
@@
 
+fig_study_diagram <- fig_nums(
+  name = "fig_study_diagram",
+  caption = paste0("Study overview: location of FSHCG care homes and diagram of data sources")
+)
+
 fig_KM_CT_residents <- fig_nums(
   name = "fig_KM_CT_residents",
-  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in residents according to FSHCG outbreak surveillance reports (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
+  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in residents according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
 )
@@
   caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infection in care home staff (n=",   formatbm(desc_staff$total_staff),
-  ") according to FSHCG outbreak surveillance reports (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")"))
+  ") according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")"))
 
@@
 National surveys suggest infections are declining in the general population in the UK, but outbreaks persist in care homes.
-Epidemiological data on disease trends in care home staff and residents are scarce.
-We used administrative data from the Four Seasons Health Care Group (FSHCG), a national care home chain representing 9% of all beds in England, Scotland and Northern Ireland.
+We measured the incidence of infections among care home staff and residents, using administrative data from the Four Seasons Health Care Group (FSHCG), a national care home chain representing 9% of all beds in England, Scotland and Northern Ireland.
 
 **Methods:** Retrospective cohort study using individual-level electronic records from residents (incidents reports) and daily care home counts of infection for both residents and staff.
-We estimated the incidence proportion and incidence rate of symptoms, confirmed infection, and death in residents and staff, and compare them with findings from a national community survey.
+We estimated the incidence proportion and incidence rate of symptoms, confirmed infections, and deaths in residents and staff.
+The rate of confirmed infections was compared with data from a national community survey.
 We tested a range of risk factors for resident infection and mortality in multivariate Cox proportional hazards models.
@@
 During the period `r formatbm(desc_residents$CT_total_symptomatic)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
-The confirmed infection incidence rate was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] across all care homes, including `r desc_residents$CT_zero_case` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) that had not reported a single confirmed case.
+The confirmed infection incidence rate was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] across all care homes, including `r desc_residents$CT_zero_case` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) which had not reported a single confirmed infection or death.
 Within England, the confirmed infection rate was `r tbrounding(england_results$results$measure[2,"estimate"])` times [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`] greater in care homes than in community households.
-Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted(
-`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]), corresponding to a rate of incidence of 
+Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]), corresponding to a rate of incidence of
 `r tbrounding(desc_staff$CT_confirmed_IR)` per 100,000 staff-days [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`].
 
-Several risk factors such as being male or aged 85 years and over were independently associated with residents' confirmed infection rate.
-An increase by 10 percentage points in the staff-to-bed ratio was associated with a 25% reduction [19%, 31%] in the rate of confirmed cases.
-The ratio of occupants to bedrooms measured at the beginning of the epidemic had a strong positive association with the infection rate.
-Hazards of confirmed infection in homes with a ratio greater than 1.00 were 4·9 times [2·8; 8·4] greater than in homes with a ratio of 0.70 to 0.85.
-Mortality was strongly associated with both being male and age.
-An increase by 10 percentage points in the staff-to-bed ratio was associated with a 25% reduction [19%, 31%] in the rate of confirmed cases.
-The ratio of occupants to bedrooms measured at the beginning of the epidemic had a strong positive association with the infection rate.
-Hazards of confirmed infection in homes with a ratio greater than 1.00 were 4·9 times [2·8; 8·4] greater than in homes with a ratio of 0.70 to 0.85.
-All-cause mortality was strongly associated with being male, age, but not rates of staffing or occupancy.
-Adjusted mortality hazard ratios for asymptomatic and symptomatic confirmed cases respectively were 2.4 [1.4; 4.0] and 10 [8.3; 12] compared with residents that were asymptomatic and never tested positive.
+Several risk factors were independently associated with residents' confirmed infection hazards.
+Males and those aged 85 years had significantly greater hazards of confirmed infection.
+Homes with a ratio of occupants to bedrooms (measured at the beginning of the epidemic) in excess of 1.00 were found to have confirmed infections hazards 10 times [7·0; 15] greater than homes with a ratio of 0.70 to 0.85.
+Conversely, an increase by 10 percentage points in the staff-to-bed ratio was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infections.
+
+All-cause deaths both within homes and in hospitals were recorded for occupants of private beds. 
+The all-cause mortality proportion was 
+`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%], with a confirmed infection case-fatality rate of
+`r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%].
+All-cause mortality was strongly associated with both older age and being male, but not with rates of staffing or occupancy.
+Adjusted mortality hazard ratios for asymptomatic and symptomatic confirmed cases respectively were 2·1 [1·3; 3·6] and 8·5 [7·1; 10] compared with asymptomatic residents who never tested positive for SARS-CoV-2.
 
@@
 
-Care home residents are extremely vulnerable to infection due to their advanced age and high prevalence of comorbidity, and this is compounded by frequent exposure to infection, for example through contact with infected staff members.
-At the peak of the pandemic the number of deaths in UK care home residents was threefold higher than the equivalent period in 2019,[@onsDR2020] but estimates of infection incidence and prevalence over this time period are lacking due to delays in establishing universal testing for care home residents and staff.
+Care home residents are vulnerable to infections due to their advanced age and high prevalence of comorbidity.
+This is compounded by frequent exposure to infections, for example through contact with infected staff members.
+At the peak of the pandemic the number of deaths in UK care home residents was threefold higher than the equivalent period in 2019.[@onsDR2020]
+
+Information on disease incidence and prevalence, and how they vary across care homes is essential to inform the public health response to future waves of infection, and in particular decisions around the need for widespread universal testing in care home staff and residents.
+Yet, estimates of infection incidence and prevalence over this time period are lacking due to delays in establishing universal testing for care home residents and staff.
 
@@
 Testing was also performed in care homes that notified an outbreak to Public Health England, but this was limited to a maximum of five test per care home.
-Widespread testing for residents first became available on 11 May 2020 through the whole care home testing programme.
-[@dhscTesting2020] Consequently national estimates of incidence and prevalence based on test results obtained prior to 11 May will substantially underestimate the burden of infection in care home residents.
-
-An important source of epidemiological data on the burden of disease in care homes are outbreak investigations.[@McMichael2020; @Dora2020] A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long term care facilities.
-Estimates of this nature cannot be generalised due to the unknown proportion of care homes that have not reported any outbreaks.
-
-Information on disease incidence and prevalence, and how this varies by care home is essential to inform the public health response to future waves of infection, and in particular decisions around the need for widespread universal testing in care home staff and residents.
+Widespread testing for residents only became available on 11 May 2020 through the whole care home testing programme.[@dhscTesting2020] As a result, national estimates of incidence and prevalence based on test results obtained prior to 11 May substantially underestimate the burden of infection in care home residents.
+
+Outbreak investigations constitute an important source of epidemiological data on the burden of disease in care homes.[@McMichael2020; @Dora2020] A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long term care facilities.
+Despite this, estimates of this nature cannot be generalised due to the unknown proportion of care homes that have not reported any outbreaks.
 
 This study obtained data from `r formatbm(dplyr::n_distinct(residents$home_code))` care homes from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care.
-It estimated incidence of SARS-CoV-2 symptoms, diagnoses and all-cause mortality.
-In addition, it tested and identified individual and organisational risk factors associated with the rate of confirmed infections.
+It estimated incidence of SARS-CoV-2 symptoms, diagnoses and all-cause mortality across all homes.
+In addition, it tested and identified individual and organisational risk factors associated with the rate of confirmed infections and deaths.
 
@@
 
-In 2020, FSHCG represented approximately 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`).
-As of 22 June 2020, 90% of FSHCG care homes have participated in the whole care home testing programme.
+In 2020, FSHCG counted 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`).
+Under 9% of these beds were contracted to local authorities for their statutory social care responsibilities.
 
@@
 This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as "Datix".
-Such records do not include residents occupying beds contracted by local authorities as part of block agreements.
+Such records do not include residents occupying beds contracted by local authorities as part of block agreements, but home-level aggregate information was nevertheless available on all residents (`r fig_nums("fig_study_diagram", display = "cite")`).
 
@@
 ```{r FSHCG_map, fig.width=3, fig.height=5, warning=F}
-reference_homes_lat_long <- reference_homes %>% 
-  dplyr::filter(home_code %in% home_inclusion) %>% 
-  dplyr::select(home_code, home_name_datix, postcode) %>% 
-  dplyr::left_join(reference_geography, by = "postcode")
-
-care_homes_map <- ggplot(UK_country_region_sf)  +
-  geom_sf(size = 0.1) +
-  geom_sf_text(aes(label = ctry_reg_abb), fontface = "bold") +
-  theme_void() +
-  geom_point(data = reference_homes_lat_long, aes(y = latitude, x = longitude),
-             color = "#0099B4FF") +
-  ggtitle(paste0(length(home_inclusion), " FSHCG care homes"))
-# pdf("map.pdf", width = 3, height = 5)
-care_homes_map
+# reference_homes_lat_long <- reference_homes %>% 
+#   dplyr::filter(home_code %in% home_inclusion) %>% 
+#   dplyr::select(home_code, home_name_datix, postcode) %>% 
+#   dplyr::left_join(reference_geography, by = "postcode")
+#  
+# care_homes_map <- ggplot(UK_country_region_sf)  +
+#   geom_sf(size = 0.1) +
+#   geom_sf_text(aes(label = ctry_reg_abb), fontface = "bold") +
+#   theme_void() +
+#   geom_point(data = reference_homes_lat_long, aes(y = latitude, x = longitude),
+#              color = "#0099B4FF", size = .5) + # shape = 3
+#   ggtitle(paste0(length(home_inclusion), " FSHCG care homes"))
+# cairo_ps("map.ps", width = 3, height = 5)
+# care_homes_map
 # dev.off()
@@
 
+```{r, fig.width=10, fig.height=5}
+knitr::include_graphics("study_diagram.svg")
+```
+
 *Note:* `r paste(paste0(UK_country_region_sf$ctry_reg_abb, ": ", UK_country_region_sf$ctry_reg), collapse = "; ")`.
 
-## Aggregate outbreak data (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")`)
+`r fig_nums("fig_study_diagram")`
+
+## Aggregate outbreak surveillance counts (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")`)
 
 On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor COVID-19 cases in each care home.
-This new system requires care home managers to report daily tallies of symptomatic and confirmed infections identified among staff and residents to their FSHCG regional manager.
-
-Daily counts reported by care homes to FSHCG were obtained for every care home:
+This new system requires care homes to report daily tallies of new cases identified among staff and residents to their FSHCG regional manager.
+Daily counts were obtained for every care home:
 
 - new symptomatic cases (residents)
-- new confirmed cases in home (residents)
-- new confirmed cases in hospital (residents)
-- deaths related to COVID-19 (Haydn please confirm what these correspond to: suspected/coroner report/tested positive before death?)
-- new symptomatic cases (staff)
-- new confirmed cases (staff)
-- deaths related to COVID-19 (staff)
+- newly confirmed infection in home (residents)
+- newly confirmed infection in hospital (residents)
+- deaths related to COVID-19 (residents previously confirmed or with a death attributed to SARS-CoV-2 by coroner)
+- symptomatic cases (staff)
+- newly confirmed cases (staff)
 
@@
 
-These counts include events relating to residents occupying local authority contract beds as well as a minority of residents receiving a different category of support (eg.
-young people) who were excluded from the study population.
-
-## Individual-level residents data (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")`)
-
-Secondary use individual-level data were extracted and pseudonymised by HW. They consisted of two datasets:
+These aggregate data related to private residents, residents occupying local authority contract beds, as well as a minority of residents receiving a different category of support (eg.
+young people) whose individual-level records were not extracted.
+
+## Individual-level residents records (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")`)
+
+Secondary use individual-level data for private residents receiving general and elderly care were extracted and pseudonymised by HW. They consisted of two datasets:
 
@@
 Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting.
-In addition, reports can include: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.
+In addition, reports could record: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.
 
@@
 Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents dataset.
-Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) Datix reports did not contain enough information to generate an identifier and were excluded.
-Identifiers of a further `r desc_before_exclusions$link_fail` reports (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
-A proportion of these reports may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.
+Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) incident records did not contain enough information to generate an identifier and were excluded.
+Identifiers of a further `r desc_before_exclusions$link_fail` records (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
+A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.
 
@@
 
-Incidence proportions and rates were calculated using the aggregate outbreak counts, since they were the trusted source of information and encompass all residents.
-The number of unique residents $N$ present in the homes at some point between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
-This approach assumes that the mean stay of local authority residents is identical to that of private residents (supplementary material 1).
-As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrement life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records.
+On the one hand, incidence proportions and rates were calculated using the aggregate outbreak counts, as these were the trusted source of information and encompassed all residents.
+The number of unique residents present in the homes at some point between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
+This approach assumes that the mean length of stay of local authority residents is identical to that of private residents (supplementary material 1).
+As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records.
 Admitted cases were accounted for in the daily new case count.
@@
 
-As for individual-level data, they were known to be subject to underreporting of infection cases as well as to exclude residents occupying local authority contracted beds.
-In spite of this, individual-level provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicate when a resident was discharged or died.
-By assuming underreporting of cases to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and fatalities.
-A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, mean bed occupancy proportion).
-Dates of death were used to estimate the case-fatality rate, as well as compare all-cause mortality based on the presence of symptoms and infection confirmation.
+On the other hand, individual-level data were affected by some underreporting of infection cases and did not extend to residents occupying local authority contract beds.
+In spite of this, they provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicated when a resident was discharged.
+By assuming infection underreporting to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and fatalities.
+A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, occupant-to-bedroom ratio).
+Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and infection confirmation.
 A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
-
-We also investigated whether the number of confirmed infections among staff was associated with subsequent infections amongst residents, independently of time.
-For this, a Poisson model regressed the rate of confirmed resident cases against home occupancy levels as well as lagged rates of confirmed staff infections at care home level.
 
@@
 
+Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab testing across the study period.
+Whole care home testing[@dhscTesting2020] began on 11 May 2020, meaning some confirmed infection are likely to have been missed before this date.
+By 22 June 2020, 90% of FSHCG care homes had participated in the whole home testing programme.
+
 # Findings
 
-## Descriptives
+## Resident cases
 
@@
 
-### Resident cases
-
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("supp_epi_CT", display = "cite")`).
-
-`r table_nums("supp_epi_CT")`
-
-|               |  Symptomatic  | Residents  <br> Confirmed  | Symptomatic | Staff <br> Confirmed   |
-| ------------- | -------------:| -------------:|-------------:|-------------:|
-| Cases      | `r formatbm(desc_residents$CT_total_symptomatic)` |	`r formatbm(desc_residents$CT_total_confirmed)` |	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
-|	$N$ exposed  | `r formatbm(desc_residents$CT_total_residents)`|	`r formatbm(desc_residents$CT_total_residents)`|	`r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
-|	Total exposure (days)| `r formatbm(desc_residents$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents$CT_total_confirmed_denom)`	|`r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
-|	Incidence proportion (%)  | `r tbrounding(desc_residents$CT_p_symptomatic)` [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents$CT_p_confirmed)` [`r tbrounding(desc_residents$CT_p_confirmed_lower)`; `r tbrounding(desc_residents$CT_p_confirmed_upper)`] | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
-|	Incidence rate (per 100,000) | `r tbrounding(desc_residents$CT_IR_symptomatic)` [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+
+`r table_nums("incidence_CT_residents")`
+
+| Residents     |  Symptomatic  | Confirmed  |    All homes <br>  COVID-19-related deaths  | Symptomatic  |  Confirmed  | Outbreak homes <br>  COVID-19-related deaths  |
+| ------------- | -------------:| -------------:|-------------:|-------------:|-------------:|-------------:|
+| Cases      | `r formatbm(desc_residents$CT_total_symptomatic)` |	`r formatbm(desc_residents$CT_total_confirmed)` |	`r formatbm(desc_residents$CT_total_cdeath)` | `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic)` |	`r formatbm(desc_residents_outbreak_only$CT_total_confirmed)` | 	`r formatbm(desc_residents_outbreak_only$CT_total_cdeath)` |
+|	$N$ exposed  | `r formatbm(desc_residents$CT_total_residents)`|	`r formatbm(desc_residents$CT_total_residents)`|	 	`r formatbm(desc_residents$CT_total_residents)`| `r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	`r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	`r formatbm(desc_residents_outbreak_only$CT_total_residents)`|
+|	Total exposure (days)| `r formatbm(desc_residents$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents$CT_total_confirmed_denom)`	| `r formatbm(desc_residents$CT_total_cdeath_denom)`	|  `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents_outbreak_only$CT_total_confirmed_denom)` |`r formatbm(desc_residents_outbreak_only$CT_total_cdeath_denom)` |
+|	Incidence proportion (%)  | `r tbrounding(desc_residents$CT_p_symptomatic)` [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents$CT_p_confirmed)` [`r tbrounding(desc_residents$CT_p_confirmed_lower)`; `r tbrounding(desc_residents$CT_p_confirmed_upper)`] | `r tbrounding(desc_residents$CT_p_cdeath)` [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] | `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_upper)`] | `r tbrounding(desc_residents_outbreak_only$CT_p_cdeath)` [`r tbrounding(desc_residents_outbreak_only$CT_p_cdeath_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_cdeath_upper)`] |
+|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_residents$CT_IR_symptomatic)` [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] | `r tbrounding(desc_residents$CT_IR_cdeath)` [`r tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r tbrounding(desc_residents$CT_IR_cdeath_upper)`] | `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_upper)`] |`r tbrounding(desc_residents_outbreak_only$CT_IR_cdeath)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_cdeath_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_cdeath_upper)`] |
 
@@
 The overall confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%, with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-
-Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases (`r fig_nums("fig_KM_CT_residents_symptomatic", display = "cite")`, `r fig_nums("fig_KM_CT_confirmed_residents", display = "cite")`) indicate wide variation across the UK.
-We note a strong progression of infections in the North West, North East and South East of England, where more than 1 in 5 residents had a confirmed infection.
-The sharp increase in confirmed infections in the North West between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
+Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown on `r fig_nums("fig_KM_CT_residents", display = "cite")`.
+The sharp increase in confirmed infections between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
+
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection or COVID-19-related death, whether in the home or in hospital (supplementary material 3).
+Incidence estimates are reported for those homes only alongside overall incidence in `r table_nums("incidence_CT_residents", display = "cite")`.
 
@@
     "Confirmed infection",
-    "Death"
+    "COVID-19-related death"
   )
@@
 
-For England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format((england_results$date_man), "%e %B %Y")`.
+For England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
 In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$FSHC_confirmed` confirmed cases out of `r formatbm(england_results$FSHC_confirmed)` person-days during this period.
@@
 Incidence proportions and rates were also estimated separately for private residents using the Datix incident reports dataset (supplementary material 1-2).
-Estimates are lower than estimates in `r table_nums("supp_epi_CT", display = "cite")`.
+Estimates are lower than estimates in `r table_nums("incidence_CT_residents", display = "cite")`.
 This is thought to be caused by a combination of underascertainment as well as linkage error affecting incidents reports.
 
-### Staff cases
-
-Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were reported to have symptoms by FSHCG managers between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-During the same period, `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff were reported to have had a confirmed infection.
-Regional trends (`r fig_nums("fig_KM_staff_symptomatic", display="cite")`, `r fig_nums("fig_KM_staff_confirmed", display="cite")`) once again indicate a strong progression of infections in the North West, North East and South East of England.
-
-```{r KM_Staff, fig.height = 4, fig.width = 8}
-all_km_staff <- bind_rows(
-  km_staff_symptomatic,
-  km_staff_confirmed
-)
-all_km_staff$Event <- factor(
-  all_km_staff$event,
-  levels = c(
-    "Symptomatic case",
-    "Confirmed infection"
-  )
-)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,
-           ymin = 1 - St_min,
-           color = Event), data = all_km_staff) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  ggsci::scale_color_lancet()+
-  scale_y_continuous("Cum probability symptomatic case",
-                     breaks = seq(0,1,.02)) +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_staff_symptomatic$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_staff")`
-
-### Outbreak homes
-
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection, whether acquired in the home or in hospital (supplementary material 3).
-Incidence estimates are reported for those homes separately in `r table_nums("tab_zero_cases", display="cite")`.
-
-`r table_nums("supp_epi_CT_oubreak_only")`
-
-|               |  Symptomatic  | Residents  <br> Confirmed  | 
-| ------------- | -------------:| -------------:|
-| Cases      | `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic)` |	`r formatbm(desc_residents_outbreak_only$CT_total_confirmed)` |
-|	$N$ exposed  | `r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	`r formatbm(desc_residents_outbreak_only$CT_total_residents)`|	
-|	Total exposure (days)| `r formatbm(desc_residents_outbreak_only$CT_total_symptomatic_denom)`	|`r formatbm(desc_residents_outbreak_only$CT_total_confirmed_denom)`
-|	Incidence proportion (%)  | `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_symptomatic_upper)`]| `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_p_confirmed_upper)`] |
-|	Incidence rate (per 100,000) | `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_symptomatic_upper)`] |`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed)` [`r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents_outbreak_only$CT_IR_confirmed_upper)`] |
-
 ```{r factors_outbreak}
-home_total_confirmed$outbreak <-  as.factor(home_total_confirmed$CT_confirmed>0)
+# home_total_confirmed$outbreak <-  as.factor(home_total_confirmed$CT_confirmed>0)
 # # factors_outbreak <- glm( 
@@
 
+## Staff cases
+
+Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were reported to have symptoms by FSHCG managers between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+During the same period, `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff were reported to have had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`).
+
+`r table_nums("incidence_CT_staff")`
+
+|   Staff       |  Symptomatic  |    Confirmed  | 
+| ------------- | -------------:| -------------:|
+| Cases      | 	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
+|	$N$ exposed  | `r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
+|	Total exposure (days)| `r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
+|	Incidence proportion (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
+|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
+
+```{r KM_Staff, fig.height = 4, fig.width = 8}
+all_km_staff <- bind_rows(
+  km_staff_symptomatic,
+  km_staff_confirmed
+)
+all_km_staff$Event <- factor(
+  all_km_staff$event,
+  levels = c(
+    "Symptomatic case",
+    "Confirmed infection"
+  )
+)
+
+ggplot(aes(x = date, y = 1-S_t, 
+           ymax = 1 - St_max,
+           ymin = 1 - St_min,
+           color = Event), data = all_km_staff) +
+  # facet_grid(region_group ~ .) +
+  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
+  geom_path(lwd=1.2) +
+  ggsci::scale_color_lancet()+
+  scale_y_continuous("Cum probability symptomatic case",
+                     breaks = seq(0,1,.02)) +
+  scale_x_date(name = "Date", 
+               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
+               labels = format.Date(
+                 unique(get_monday_date(km_staff_symptomatic$date)),
+                 "%d/%m/%y"
+               )) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
+
+```
+
+`r fig_nums("fig_KM_staff")`
+
 ## Factors associated with confirmed infections
@@
 Using the individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-According to this, and other than geographical location, the staff-to-bed ratio was the only factor significantly associated with the rate of confirmed cases.
-An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 13% reduction [9%; 18%] in the rate of confirmed cases.
+According to this, men were at significant greater risk of confirmed infections (adjusted HR = 1·3 [1·1; 1·6]), as were residents aged 85 years and over (adjusted HRs $\geq$ 1·4) and nursing home residents (adjusted HR = 1·5 [1·2; 1·8]).
+
+Large care homes had greater rates of infection, but this was partly attributable to other factors.
+Care homes with 70 beds or more had an adjusted HR of 1·8 [1·4; 2·4] compared with homes counting under 35 beds.
+Occupancy had an even greater effect: compared to homes between 7 and 8·5 residents for 10 bedrooms, confirmed infection hazards in homes with between 8·5 and 10 residents for 10 bedrooms were 2·5 times greater [1·9; 3·3].
+In  homes with more residents than bedrooms, hazards were 10 times [7·0; 15] greater.
+
+This was based on occupancy measured before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
+As occupancy reduced more steeply in outbreak care homes during the study period due to hospitalisations and deaths, the average occupancy during the epidemic was negatively associated with infection rates (not reported).
+Finally, staffing acted as an important protective factor.
+An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infections, once adjusting for other factors.
 
@@
 ```{r CPH_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+region_UK+ staff_to_bed_ratio+bed_tot_category + pct_occupancy+occ_bedroom_ratio_g6, data = residents_overview)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
 #  summary(cph_conf)
@@
 explanatory = c("gender", "ageg", "home_type", "user_type2",  
-                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6", "region_UK" )
+                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6"  )
 dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
@@
 
-The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by COVID-19 status and other factors.
+The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by the infection status along other factors.
 Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
-After controlling for other risk factors the hazard ratios of being asymptomatic confirmed and symptomatic confirmed cases respectively were 2·6 [1·6; 4·3] and 9·9 [8·3; 11·9], confirming that symptomatic cases were more severe.
-As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 6·2 [5·1; 7·4].
+After controlling for other risk factors, the hazard ratio for asymptomatic confirmed and symptomatic confirmed cases respectively were 2·1 [1·3; 3·6] and 8·5 [7·1; 10·2], confirming that symptomatic cases were more severe.
+As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 5·3 [4·4; 6·4].
 It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
-Mortality was significantly higher in nursing care homes (adjusted HR = 1·8 [1·6; 2·0]).
-The 27% increase in the hazard ratio after covariate adjustment reinforces the finding that this effect does not appear to be confounded by the greater proportion of symptomatic cases or common risk factors.
-This effect is likely attributable to a greater burden of comorbidities in nursing home residents.
+Mortality was significantly higher in nursing care homes (adjusted HR = 1·4 [1·3; 1.6]).
+This effect is likely to be at least partly attributable to a greater burden of comorbidities in nursing home residents.
 Beyond, there was no evidence of an association between care home size and all-cause mortality.
-After adjusting for other variables, the staff-to-bed ratio did not exhibit a statistically significant association with all-cause mortality (adjusted HR = 0·8 [0·7; 1.1]).
+After adjusting for other variables, an increase by 10 percentage points in staffing was associated with a 4.3% reduction [1.8%; 6.9%] in mortality after adjustment for other variables.
+Mortality did not appear to be associated with care home size or occupancy, at least not in a linear fashion: mortality was lower in homes with a ratio of occupants to bedrooms in excess of 1 compared with a ratio comprised between .85 and 1; while also being lower in homes with a ratio between 0·55 and 0·7.
 
@@
 #                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       region_UK + cluster(resident_id), data = residents_episodes)
+#                       cluster(resident_id), data = residents_episodes)
 # summary(cph_death2)
@@
 #  pct_occupancy, occ_bedroom_ratio,no_bedroom 
-explanatory = c("gender", "ageg", "home_type", #"user_type2", 
+explanatory = c("gender", "ageg", "home_type", "user_type2", 
                 "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
-                "covid_status", "region_UK", "cluster(resident_id)")
+                "covid_status", "cluster(resident_id)")
 # explanatory_interaction = c("gender", "ageg", "user_type2",
@@
            ggtheme = theme_bw(),  
+           palette = "lancet",
            break.x.by	= 7,
@@
 
-This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large provider network across `r as.integer(date_max-date_min)` days.
-A comparable study was previously conducted on across 623 care homes in Ontario, Canada.[@Stall2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-
-Regional variation was noticeable, but inconsistencies between incident reports and manager counts were indicative of data quality issues, particularly in relation to incident reports.
-Rates of testing across time and care homes were not yet known at the time of writing and likely distorts measures of incidence.
-
-One prominent organisational factor was significantly and independently associated with the risks of confirmed infection: the staff-to-bed ratio at baseline was a protective factor.
-A 10 percentage point increase was associated with a 13% reduction [9%; 18%] in hazards of confirmed infections.
-Few organisational variables were available to identify potential confounders or mediators of this association.
+This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large care provider across `r as.integer(date_max-date_min)` days.
+Comparable studies were conducted in Ontario, Canada and Spain.[@Brown2020; @Stall2020; @Moraga2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+
+As is the case with most investigations, the confirmed infection incidence parameters are likely to be severely underestimated.
+Despite the very rapid development of viral antigen tests, testing capacity was limited in the early weeks of the epidemic, resulting in many infections being missed before 11 May 2020.
+This comes in addition to the low sensitivity of such tests, even when administered in a timely manner.
+According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
+Other than underestimation of incidence, this may mean some underestimation in the incidence ratio comparing care homes with the community in England.
+Impact on case-fatality estimates is unclear, as this may affect both numerators and denominators.
+
+We adopted a very broad fatality case definition, by including deaths of all causes in the individual-level data analysis, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all COVID-19-related deaths reported by outbreak surveillance counts).
+The resulting case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was far higher than in previous literature,[@Salcher-Konrad2020] which often had both a narrower fatality case definition, and shorter observation periods.
+An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
+These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19% among our private residents.
+Infections were also less common and more concentrated in Stall et al.[@Stall2020]
+
+Like a large study conducted in Ontario, Canada,[@Brown2020] we note a strong association between the occupancy rate and the incidence of infection.
+Although the effect was highest where residents had to share bedrooms, we observed that this association remained noticeable at levels well below 1 resident per bedroom.
+The limited range of variables measured did not allow to investigate potential counfounding extensively, particularly in relation to resident casemix (eg: comorbidities).
+
+We found strong associations between the incidence of infections and organisational factors: staff-to-bed ratio, and occupancy.
+These are connected to important healthcare infection control procedures, such as resident isolation and workplace hygiene.
 One hypothesis is that staffing ratios may be a sign of better-resourced care homes (for instance protective personal equipment), and of a greater capability to prescribe and observe effective infection control procedures.
@@
 A more detailed survey of care home characteristics, along with protocols and measures adopted during the epidemic may help elucidate mechanisms through which staffing levels are associated with lower infection rates.
-
-The case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was higher than previous literature[@Salcher-Konrad2020], which is not exclusively attributable to the long observation period.
-An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
-These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19%.
-
-Uncertainty remains regarding the internal validity of estimates presented in this report.
-It was not possible to determine each care home's outbreak status with confidence.
-Consequently, incidence was calculated across all homes, unlike the study by Stall et al.[@Stall2020] Beyond uncertainty in the exact number of cases and outbreaks, the viral antigen test campaign is a likely subject of under-ascertainment, on account of both the diagnostics' intrinsically low sensitivity, and of the timing of the systematic test campaign initiated on 11 May 2020.
-Residents may not have provided sample at their peak viral load, thereby reducing the rate of detection.
-According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
-Consequently findings presented in the current study are likely to be underestimates.
-This includes estimates of the incidence ratio comparing care homes with the community in England.
-
+As for occupancy, our finding echoes the conclusions of Brown et al.,[@Brown2020] although the association did not extend to mortality.
+
+This study was not able to examine some other important factors such as the presence of co-morbidities, the importing of cases from hospitals, or the rate of testing.
+This last factor is a major determinant of confirmed infection rates.
 Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection.[@Dora2020] Although increasing testing capacity raises the possibility of undertaking frequent, repeat testing in the care home population, this may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
 Understanding which care homes are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention accordingly.
+The development of saliva antigen tests should also provide a less invasive alternative.
+
+We note that a large proportion of the analysis in this paper was made possible by the centralised reporting of individual-level data, the quality of which could be improved with support from national health organisations.
+Although we made good use of home-level aggregate outbreak surveillance counts using a life table approach, this involves some strong assumptions on all residents being susceptible at the beginning of the study period.
+Such assumptions would not hold in the same way in the event of a second wave of the epidemic.
+Anticipation of a possible resurgence should constitute a further motive for rapidly strengthening surveillance systems.
 

### 2020-07-05 manuscript/manuscript.Rmd
manuscript version 0.4

@@
 ---
-title: "Incidence of confirmed COVID-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
+title: "Incidence of confirmed Covid-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
 # author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
@@
 
+residents_overview <- tline %>% 
+  filter(rday==1) %>%
+  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
+  group_by(resident_id, home_code, ageg, gender, 
+           admission_type2, home_type, user_type2, status) %>% 
+  summarise(resident_days = sum(rday),
+            susceptible_symptomatic_days = sum(susceptible_symptomatic),
+            susceptible_confirmed_days = sum(susceptible_confirmed),
+            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
+            confirmed = max(-susceptible_confirmed+1, na.rm = T),
+            start_date = min(date, na.rm = T),
+            stop_date = max(date, na.rm = T),
+            start_int = as.numeric(min(date, na.rm = T) - date_min),
+            stop_int = as.numeric(max(date, na.rm = T)- date_min)
+    ) %>% 
+  ungroup() %>% 
+  mutate(covid_status = case_when(
+    symptomatic==0 & confirmed == 0 ~ "Negative",
+    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
+    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
+    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
+  )) %>% 
+  mutate(covid_status = factor(covid_status, 
+                               levels = c("Negative", "Symptomatic (not confirmed)",
+                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
+  left_join(select(reference_homes, home_code, country, region_UK,
+                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
+                   beds,  pct_occupancy, no_bedrooms, 
+                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
+  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
+
+home_first_case <- bind_rows(
+  new_cases %>% 
+    dplyr::filter(tidyr::replace_na(new_deaths_hospital, 0) > 0 |
+                    tidyr::replace_na(new_deaths_home, 0)>0) %>% 
+    dplyr::group_by(home_code) %>% 
+    dplyr::summarise(first_case_date = min(date),
+                     source = "CT"),
+  residents_overview %>% 
+    dplyr::filter(confirmed == 1) %>% 
+    dplyr::group_by(home_code) %>% 
+    dplyr::summarise(first_case_date = min(covid_first_confirmed, na.rm = T), source="DAT")
+) %>% 
+  dplyr::group_by(home_code) %>% 
+    dplyr::summarise(first_case_date = min(first_case_date)) %>% 
+  tidyr::crossing(tibble(outbreak = c(0, 1)))
+
 residents_episodes <- tline %>% 
   filter(rday==1)  %>% 
+  left_join(home_first_case) %>%
+  mutate(outbreak = dplyr::case_when(
+    date >= first_case_date ~ 1L,
+    date < first_case_date ~ 0L,
+    TRUE ~ 0L
+  )) %>% 
   group_by(resident_id, home_code, dob, gender, 
@@
            susceptible_symptomatic, 
-           susceptible_confirmed) %>% 
+           susceptible_confirmed,
+           outbreak) %>% 
   summarise(entry_date = min(date),
@@
 
-residents_overview <- tline %>% 
-  filter(rday==1) %>%
-  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
-  group_by(resident_id, home_code, ageg, gender, 
-           admission_type2, home_type, user_type2, status) %>% 
-  summarise(resident_days = sum(rday),
-            susceptible_symptomatic_days = sum(susceptible_symptomatic),
-            susceptible_confirmed_days = sum(susceptible_confirmed),
-            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
-            confirmed = max(-susceptible_confirmed+1, na.rm = T),
-            start_date = min(date, na.rm = T),
-            stop_date = max(date, na.rm = T),
-            start_int = as.numeric(min(date, na.rm = T) - date_min),
-            stop_int = as.numeric(max(date, na.rm = T)- date_min)
-    ) %>% 
-  ungroup() %>% 
-  mutate(covid_status = case_when(
-    symptomatic==0 & confirmed == 0 ~ "Negative",
-    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
-    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
-    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
-  )) %>% 
-  mutate(covid_status = factor(covid_status, 
-                               levels = c("Negative", "Symptomatic (not confirmed)",
-                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
-  left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
-                   beds,  pct_occupancy, no_bedrooms, 
-                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
-  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
-
 # CARE HOME AGGREGATES
@@
     St_min = S_t^exp(-1.96*S_t_SElog),
-    event = "COVID-19-related death"
+    event = "Covid-19-related death"
   )
@@
 
-desc_residents$DAT_CFR_RR <- relative_risk(desc_residents$DAT_CFR$y, desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, desc_residents$DAT_BFR$y, desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
-
-desc_residents$DAT_CFR_RR_SE <- relative_risk_SE(desc_residents$DAT_CFR$y, desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, desc_residents$DAT_BFR$y, desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
+desc_residents$DAT_CFR_RR <- relative_risk(
+  desc_residents$DAT_CFR$y, 
+  desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, 
+  desc_residents$DAT_BFR$y, 
+  desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
+
+desc_residents$DAT_CFR_RR_SE <- relative_risk_SE(
+  desc_residents$DAT_CFR$y, 
+  desc_residents$DAT_CFR$n - desc_residents$DAT_CFR$y, 
+  desc_residents$DAT_BFR$y, 
+  desc_residents$DAT_BFR$n - desc_residents$DAT_BFR$y)
+
+desc_residents$DAT_CFR_RR_lower <- exp(log(desc_residents$DAT_CFR_RR) - 1.96 * desc_residents$DAT_CFR_RR_SE)
+desc_residents$DAT_CFR_RR_upper <- exp(log(desc_residents$DAT_CFR_RR) + 1.96 * desc_residents$DAT_CFR_RR_SE)
 
 desc_residents[paste0(
-  "DAT_CFR", c("", "_lower", "_upper")
+  "DAT_CFR_rate", c("", "_lower", "_upper")
 )] <- epitools::binom.exact(
@@
 desc_residents[paste0(
-  "DAT_BFR", c("", "_lower", "_upper")
+  "DAT_BFR_rate", c("", "_lower", "_upper")
 )] <- epitools::binom.exact(
@@
 desc_residents_outbreak_only$outbreak_homes <- unique(
-  home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0],
+  c(home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0],
   distinct(filter(new_cases, tidyr::replace_na(new_deaths_hospital, 0) > 0 |
-                    tidyr::replace_na(new_deaths_home, 0)>0), home_code)$home_code
+                    tidyr::replace_na(new_deaths_home, 0)>0), home_code)$home_code,
+  distinct(filter(residents_overview, confirmed == 1), home_code)$home_code)
+)
+
+desc_residents$DAT_BFR_outbreak_homes <- residents_overview %>%
+  filter(home_code %in% desc_residents_outbreak_only$outbreak_homes & confirmed != 1) %>%
+  summarise(n = n(),
+            y = sum(status == "Deceased"))
+
+desc_residents$DAT_BFR_other_homes <- residents_overview %>%
+  filter((!home_code %in% desc_residents_outbreak_only$outbreak_homes) & confirmed != 1) %>%
+  summarise(n = n(),
+            y = sum(status == "Deceased"))
+
+desc_residents$DAT_BFR_outbreak_homes$RR <- relative_risk(
+  desc_residents$DAT_BFR_outbreak_homes$y,
+  desc_residents$DAT_BFR_outbreak_homes$n - desc_residents$DAT_BFR_outbreak_homes$y,
+  desc_residents$DAT_BFR_other_homes$y, 
+  desc_residents$DAT_BFR_other_homes$n - desc_residents$DAT_BFR_other_homes$y)
+
+desc_residents$DAT_BFR_outbreak_homes$RR_SE <- relative_risk_SE(
+  desc_residents$DAT_BFR_outbreak_homes$y, 
+  desc_residents$DAT_BFR_outbreak_homes$n - desc_residents$DAT_BFR_outbreak_homes$y,
+  desc_residents$DAT_BFR_other_homes$y, 
+  desc_residents$DAT_BFR_other_homes$n - desc_residents$DAT_BFR_other_homes$y)
+
+desc_residents$DAT_BFR_outbreak_homes$RR_lower <- exp(
+  log(desc_residents$DAT_BFR_outbreak_homes$RR) - 
+    1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE
+)
+desc_residents$DAT_BFR_outbreak_homes$RR_upper <- exp(
+  log(desc_residents$DAT_BFR_outbreak_homes$RR) + 
+    1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE
 )
@@
 table_nums <- captioner::captioner(prefix = "Table")
-tab_national_descriptives <- table_nums(
-  name = "tab_national_descriptives",
-  caption = "Total homes, beds, contract beds, and residents within FSHCG and nationwide")
+
 tab_resident_descriptives <- table_nums(
@@
 
+tab_national_descriptives <- table_nums(
+  name = "tab_national_descriptives",
+  caption = "Total homes, beds, contract beds, and residents within FSHCG and nationwide")
+
 tab_zero_cases <- table_nums(
@@
 
-**Background** COVID-19 has caused substantial morbidity and mortality in care home residents and staff.
-National surveys suggest infections are declining in the general population in the UK, but outbreaks persist in care homes.
-We measured the incidence of infections among care home staff and residents, using administrative data from the Four Seasons Health Care Group (FSHCG), a national care home chain representing 9% of all beds in England, Scotland and Northern Ireland.
-
-**Methods:** Retrospective cohort study using individual-level electronic records from residents (incidents reports) and daily care home counts of infection for both residents and staff.
-We estimated the incidence proportion and incidence rate of symptoms, confirmed infections, and deaths in residents and staff.
-The rate of confirmed infections was compared with data from a national community survey.
-We tested a range of risk factors for resident infection and mortality in multivariate Cox proportional hazards models.
-
-**Findings:** An estimated `r formatbm(desc_residents$CT_total_residents)` residents were present in `r dplyr::n_distinct(residents$home_code)` FSHCG homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-During the period `r formatbm(desc_residents$CT_total_symptomatic)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), and `r formatbm(desc_residents$CT_total_confirmed)` (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) had an infection confirmed with nasopharyngeal RNA tests.
-The confirmed infection incidence rate was `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] across all care homes, including `r desc_residents$CT_zero_case` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) which had not reported a single confirmed infection or death.
-Within England, the confirmed infection rate was `r tbrounding(england_results$results$measure[2,"estimate"])` times [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`] greater in care homes than in community households.
-Out of `r formatbm(desc_staff$total_staff)` staff, `r formatbm(desc_staff$CT_confirmed_total)` confirmed infections were counted (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]), corresponding to a rate of incidence of
-`r tbrounding(desc_staff$CT_confirmed_IR)` per 100,000 staff-days [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`].
-
-Several risk factors were independently associated with residents' confirmed infection hazards.
-Males and those aged 85 years had significantly greater hazards of confirmed infection.
-Homes with a ratio of occupants to bedrooms (measured at the beginning of the epidemic) in excess of 1.00 were found to have confirmed infections hazards 10 times [7·0; 15] greater than homes with a ratio of 0.70 to 0.85.
-Conversely, an increase by 10 percentage points in the staff-to-bed ratio was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infections.
-
-All-cause deaths both within homes and in hospitals were recorded for occupants of private beds. 
-The all-cause mortality proportion was 
-`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%], with a confirmed infection case-fatality rate of
-`r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%].
-All-cause mortality was strongly associated with both older age and being male, but not with rates of staffing or occupancy.
-Adjusted mortality hazard ratios for asymptomatic and symptomatic confirmed cases respectively were 2·1 [1·3; 3·6] and 8·5 [7·1; 10] compared with asymptomatic residents who never tested positive for SARS-CoV-2.
-
-**Interpretation** Despite uncertain case ascertainment in care homes, internal surveillance suggests a greater transmission than in the community.
-Organisational factors such as staffing and occupancy are strongly associated with rates of confirmed infections and can consequently place residents at increased risks of death.
-The absence of high-quality case ascertainment across the social care sector hinders local infection prevention initiatives.
+**Background** Covid-19 has led to a 45% increase in mortality in care home residents in England and Wales, but the proportion of deaths directly attributable to Covid-19 is uncertain due to limited testing for infection, and lack of syndromic surveillance.
+
+**Methods:**  Retrospective cohort study using individual-level electronic records from residents and daily care home counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` care homes from the Four Seasons Health Care Group (FSHCG, `r format(date_min, "%e %B %Y")`-`r format(date_max, "%e %B %Y")`).
+We estimated the incidence of symptoms, confirmed infections, and deaths in residents and staff and modelled risk factors for infection and mortality in residents using Cox proportional hazards.
+
+**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had confirmed infections.
+The incidence of confirmed infection in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days across all care homes.
+This includes `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) which had not reported a single confirmed infection or Covid-related death.
+
+Among `r formatbm(desc_residents$DAT_total_residents)` private residents, `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred either in homes or in hospital (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]), including `r desc_residents$DAT_CFR$y` deaths of residents who had had a confirmed infection.
+This suggests an all-cause case-fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%], of which an estimated `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)` [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`] is attributable to confirmed SARS-CoV-2 infections.
+Independently, care home reports of Covid-related deaths suggest that `r desc_residents$CT_total_cdeath`/`r formatbm(desc_residents$CT_total_residents)` residents (`r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`]) died of a cause directly attributable to Covid-19.
+In addition to male gender and older age, confirmed asymptomatic (HR = 3·3 [1·9; 5·6]) and confirmed symptomatic (HR = 12 [10; 15]) infection were independently associated with all-cause mortality, by comparison with asymptomatic residents living in homes with no record of confirmed infections.
+All-cause mortality for uninfected residents was 93% higher in care homes with outbreaks compared to those without outbreaks (adjusted HR = 1·9 [1·7; 2·2]).
+
+**Interpretation:** Covid-19 infection directly increases residents’ risk of death, but in outbreak care homes mortality is also elevated in individuals with no record of infection compared to homes without outbreaks.
+Preparations for future waves of the pandemic should prioritise regular testing in care homes to address case under-ascertainment, and consider the indirect effect of Covid-19 on mortality.
 
@@
 
-Care home residents are vulnerable to infections due to their advanced age and high prevalence of comorbidity.
-This is compounded by frequent exposure to infections, for example through contact with infected staff members.
-At the peak of the pandemic the number of deaths in UK care home residents was threefold higher than the equivalent period in 2019.[@onsDR2020]
-
-Information on disease incidence and prevalence, and how they vary across care homes is essential to inform the public health response to future waves of infection, and in particular decisions around the need for widespread universal testing in care home staff and residents.
-Yet, estimates of infection incidence and prevalence over this time period are lacking due to delays in establishing universal testing for care home residents and staff.
-
-For example, at the start of the pandemic, testing for SARS-CoV-2 infection testing using reverse transcriptase polymerase chain reaction (RT-PCR) was only routinely available for care home residents or staff who were admitted to hospital.
-Testing was also performed in care homes that notified an outbreak to Public Health England, but this was limited to a maximum of five test per care home.
-Widespread testing for residents only became available on 11 May 2020 through the whole care home testing programme.[@dhscTesting2020] As a result, national estimates of incidence and prevalence based on test results obtained prior to 11 May substantially underestimate the burden of infection in care home residents.
-
-Outbreak investigations constitute an important source of epidemiological data on the burden of disease in care homes.[@McMichael2020; @Dora2020] A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long term care facilities.
-Despite this, estimates of this nature cannot be generalised due to the unknown proportion of care homes that have not reported any outbreaks.
-
-This study obtained data from `r formatbm(dplyr::n_distinct(residents$home_code))` care homes from the Four Seasons Health Care Group (FSHCG), one of the UK's largest independent provider of long-term and respite residential and nursing care.
-It estimated incidence of SARS-CoV-2 symptoms, diagnoses and all-cause mortality across all homes.
-In addition, it tested and identified individual and organisational risk factors associated with the rate of confirmed infections and deaths.
+Care home residents are particularly vulnerable to Covid-19 infection due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and this is compounded by frequent exposure to infection through close contact with infected staff members, other residents and contaminated surfaces in the care home setting.
+In the UK, an estimated 420,000 residents lived in approximately 11,000 homes for the elderly.[@cqcdata2020] At the peak of the pandemic the number of deaths in care home residents increased threefold over the equivalent period in 2019.[@onsDR2020]
+
+However, this is likely to underestimate of the true number of deaths in care home residents that were caused by the pandemic, since it only includes deaths in which Covid-19 was directly implicated.
+Death certifications in England and Wales between 28th December 2019 and 12th June 2020 suggest that two thirds of the 29,393 excess deaths recorded in care home residents (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] As part of efforts to mitigate the anticipated second wave of infection, it is critical to understand the ratio of direct to indirect deaths in care home residents.
+This will inform decisions around instigating lockdown measures to protect residents and staff from infection, whilst recognising the potential unintended consequences of interrupting usual care.
+
+Disentangling direct and indirect deaths from Covid-19 requires accurate information on the incidence and prevalence of infection among staff and residents linked to individual-level clinical outcomes.
+There is no syndromic surveillance for infection in care homes in England, and universal testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established in care homes in England until 11th May 2020, when the government launched the whole care home testing programme.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations and within a maximum of five tests per care home.
+Consequently national estimates of incidence and prevalence based on test results obtained prior to 11th May will substantially underestimate the burden of infection in care home residents and staff.
+
+Outbreak investigations are useful to estimate the proportion of residents and staff who become infected during an outbreak,[@McMichael2020; @Dora2020] but they provide limited insight into the relationship between infection and mortality across the entire pandemic period, and cannot be generalised to care homes that have not reported any outbreaks.
+A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long-term care facilities.
+Mortality rates in residents with confirmed Covid-19 infection ranged from 0 to 10%, with case fatality rates between 0 and 34%, however this review did not include estimates indirect mortality.
+
+We used administrative data from the Four Seasons Health Care Group (FSHCG), one of the UK’s largest independent provider of long-term and respite residential and nursing care, to establish syndromic surveillance for Covid-19 across `r formatbm(dplyr::n_distinct(residents$home_code))` FSHCG homes.
+We used this system to estimate the incidence and prevalence of suspected and confirmed Covid-19 across a national network of care homes, and to investigate the fraction of all-cause deaths attributable to confirmed infections.
 
@@
 
-In 2020, FSHCG counted 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (`r table_nums("tab_national_descriptives", display = "cite")`).
-Under 9% of these beds were contracted to local authorities for their statutory social care responsibilities.
+In 2020, the FSHCG included 9,568 residential or nursing care beds for elderly residents, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 3).
+Approximately 9% of these beds were contracted to local authorities, with the remainder for privately funded residents.
 
 FSHCG collects administrative data on their private residents through a variety of electronic systems that are primarily used for billing purposes.
-This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as "Datix".
-Such records do not include residents occupying beds contracted by local authorities as part of block agreements, but home-level aggregate information was nevertheless available on all residents (`r fig_nums("fig_study_diagram", display = "cite")`).
-
-`r table_nums("tab_national_descriptives")` [@cqcdata2020; @hscni2020; @isd2020]
-
-```{r table_national_descriptives, fig.cap = tab_national_descriptives, message=F}
-mutate_all(desc_homes,  .funs = formatbm) %>% arrange(gsub("FSHC ", "", Scope)) %>% t %>% kable
-```
+This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as “Datix”.
+Such records do not include residents occupying beds contracted by local authorities, but home-level aggregate information was nevertheless available on all residents (`r fig_nums("fig_study_diagram", display = "cite")`).
 
@@
 
-On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor COVID-19 cases in each care home.
-This new system requires care homes to report daily tallies of new cases identified among staff and residents to their FSHCG regional manager.
-Daily counts were obtained for every care home:
+On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor Covid-19 cases in each care home for all care home residents and staff (including residents in contract beds).
+This new system requires care homes to report daily tallies of new cases identified among staff and residents to their FSHCG regional manager, and was used to report Covid-19 deaths to the Care Quality Commission (CQC).[@CQC2020stats] Daily counts were obtained for every care home:
 
@@
 - newly confirmed infection in hospital (residents)
-- deaths related to COVID-19 (residents previously confirmed or with a death attributed to SARS-CoV-2 by coroner)
+- deaths related to Covid-19 (residents previously confirmed or with a death attributed to SARS-CoV-2 by coroner)
 - symptomatic cases (staff)
@@
 
-In addition, the number of beds currently occupied was reported every Sunday. This weekly information was used to approximate daily occupancy using linear interpolation.
-
-These aggregate data related to private residents, residents occupying local authority contract beds, as well as a minority of residents receiving a different category of support (eg.
-young people) whose individual-level records were not extracted.
+In addition, the number of beds currently occupied was reported every Sunday.
+This weekly information was used to approximate daily occupancy using linear interpolation.
+Aggregate data were only extracted for residents in care homes for the elderly.
+
+Care home characteristics were inferred from organisational data.
+Baseline care home occupancy was computed by averaging weekly occupancy before the first Covid-19 cases (between 1 January and 1 March 2020).
+This was used to calculate a ratio of the baseline occupancy to the number of bedroom ratio.
+We also estimated a ratio of the headcount of staff employed to the number of beds.
 
@@
 - **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly).
-This dataset does not include any records for occupants of local authority 'block contract' beds  ($n=$ `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))`).
-- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff into FSHCG's Datix incidents management system.
+This dataset does not include any records for `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))` occupants of local authority 'block contract' beds.
+- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff.
 Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting.
-In addition, reports could record: date of birth, gender, information on COVID-19 symptoms, tests and test results, resident current location (home/hospital), and death.
+In addition, reports could record: date of birth, gender, information on Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death.
+Incidents were classified to indicate whether the resident was symptomatic, and whether an infection was confirmed on the basis of multiple variables (supplementary material 1).
 
 HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier.
-Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents dataset.
+Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents' dataset.
 Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) incident records did not contain enough information to generate an identifier and were excluded.
 Identifiers of a further `r desc_before_exclusions$link_fail` records (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
-A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents dataset.
+A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents' dataset.
 
@@
 
-The two sources of data were employed for separate purposes.
-
-On the one hand, incidence proportions and rates were calculated using the aggregate outbreak counts, as these were the trusted source of information and encompassed all residents.
-The number of unique residents present in the homes at some point between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
+Incidence proportions and rates were calculated using the aggregate outbreak counts, as these were the trusted source of information used for reporting deaths to the CQC and encompassed all residents.
+The number of unique residents present in the homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")` was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
 This approach assumes that the mean length of stay of local authority residents is identical to that of private residents (supplementary material 1).
-As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records.
-Admitted cases were accounted for in the daily new case count.
-Based on this $\hat{I_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as
-
-$$\hat{I_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad I_{t_0} = 0$$
-where $d_{t}$ denotes the number of new cases for each day $t = \{0, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate.
-The total exposure time was calculated as $\sum_{t=0}^{T} (N_t-\hat{I}_t)$ where $N_t$ denotes daily occupancy figures.
+Aggregate outbreak counts were also used to estimate incidence and cumulative incidence across time using a life table approach.
+Daily occupancy was inferred from weekly occupied bed censuses using linear interpolation.
+As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents (supplementary material 1).
+A Kaplan-Meier product limit estimator was applied to this life table to estimate the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths every day between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
 The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
 
-On the other hand, individual-level data were affected by some underreporting of infection cases and did not extend to residents occupying local authority contract beds.
-In spite of this, they provided an accurate of the duration of exposure in resident-days as well as right-censoring, since individual records indicated when a resident was discharged.
-By assuming infection underreporting to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and fatalities.
+Individual-level data (including dates of care home entry and exit) were available for private residents only.
+By assuming infection underreporting to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and mortality.
 A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, occupant-to-bedroom ratio).
-Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and infection confirmation.
+Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and/or confirmed infection.
 A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
+
+Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab (PCR) testing across the study period following the introduction of whole care home testing on 11 May 2020.[@dhscTesting2020].
+By 22 June 2020, 90% of FSHCG care homes had participated in the whole home testing programme.
 
@@
 
-Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab testing across the study period.
-Whole care home testing[@dhscTesting2020] began on 11 May 2020, meaning some confirmed infection are likely to have been missed before this date.
-By 22 June 2020, 90% of FSHCG care homes had participated in the whole home testing programme.
-
 # Findings
@@
 The study population consisted of an estimated `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland.
-This included `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
-Characteristics of private residents could be described from individual-level records (`r table_nums("tab_resident_descriptives", display = "cite")`); `r tbrounding(mean(residents$home_type=="Nursing")*100)`% were in nursing homes, and 39·2% received dementia care.
+Two thirds (`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes`) of homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, reported at least one confirmed infection or Covid-related death, either through Datix or outbreak surveillance counts.
+
+Individual-date were available for a subset of `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
+`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of private residents were in nursing homes, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
 
@@
 ```{r table_resident_descriptives, message = FALSE, results="asis"}
-arsenal::tableby(home_type  ~ gender + ageg  +
-   user_type2 + admission_type2 + covid_status + status + region_UK, 
+residents_overview$outbreak <- relevel(as.factor(dplyr::case_when(
+  residents_overview$home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
+  TRUE ~ "Other homes"
+)), "Outbreak homes")
+
+arsenal::tableby(outbreak  ~ gender + ageg  +
+                   user_type2 + admission_type2 + home_type +  covid_status + status + region_UK, 
                  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
@@
 
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r filter(new_cases, new_symptomatic_home>0 | new_symptomatic_hospital>0) %>% summarise(n=n_distinct(home_code)) %>% .$n`/`r desc_residents$total_homes` care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+Infection was confirmed in `r desc_residents$CT_total_confirmed` residents, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
+The confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
+
+In England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$FSHC_confirmed` confirmed cases out of `r formatbm(england_results$FSHC_confirmed)` person-days during this period.
+This implies a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
 
@@
 
-| Residents     |  Symptomatic  | Confirmed  |    All homes <br>  COVID-19-related deaths  | Symptomatic  |  Confirmed  | Outbreak homes <br>  COVID-19-related deaths  |
+| Residents     |  Symptomatic  | Confirmed  |    All homes <br>  Covid-19-related deaths  | Symptomatic  |  Confirmed  | Outbreak homes <br>  Covid-19-related deaths  |
 | ------------- | -------------:| -------------:|-------------:|-------------:|-------------:|-------------:|
@@
 
-FSHCG residents underwent reverse transcriptase polymerase chain reaction (RT-PCR) viral antigen diagnostics based on nose and throat swab antigen tests during hospitalisations in the first instance, but such tests later became available in care homes from 6 April 2020.
-From 11 May 2020, most FSHCG home undertook testing every resident as part of the whole home testing programme.[@dhscTesting2020] Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`  `r formatbm(sum(new_cases$new_confirmed_home, na.rm = T))` residents had an infection confirmed in homes while `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` had an infection confirmed in hospital.
-The overall confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%, with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
 Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown on `r fig_nums("fig_KM_CT_residents", display = "cite")`.
@@
 
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) care homes had not reported a single confirmed infection or COVID-19-related death, whether in the home or in hospital (supplementary material 3).
-Incidence estimates are reported for those homes only alongside overall incidence in `r table_nums("incidence_CT_residents", display = "cite")`.
+Estimates of incidence within outbreak homes are reported alongside the overall estimates in the right handside of `r table_nums("incidence_CT_residents", display = "cite")`.
+Furthermore, incidence estimates for `r formatbm(desc_residents$DAT_total_residents)` private residents (using Datix incident records) are reported in supplementary material 2.
+Lower estimates were obtained which is likely due to reflect incomplete recording of cases in Datix, and the inability to link all incident records to an individual.
 
 ```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 8}
+write.csv(km_ct_residents_symp, file = "supp_table_kaplan_meier_residents_symptomatic.csv", row.names = F)
+write.csv(km_ct_residents_confirmed, file = "supp_table_kaplan_meier_residents_confirmed.csv", row.names = F)
+write.csv(km_ct_residents_deaths, file = "supp_table_kaplan_meier_residents_deaths.csv", row.names = F)
+
 all_km_residents <- bind_rows(
@@
     "Confirmed infection",
-    "COVID-19-related death"
+    "Covid-19-related death"
   )
@@
 
-For England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$FSHC_confirmed` confirmed cases out of `r formatbm(england_results$FSHC_confirmed)` person-days during this period.
-This implies a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
-
-Incidence proportions and rates were also estimated separately for private residents using the Datix incident reports dataset (supplementary material 1-2).
-Estimates are lower than estimates in `r table_nums("incidence_CT_residents", display = "cite")`.
-This is thought to be caused by a combination of underascertainment as well as linkage error affecting incidents reports.
+Mortality could be described from two sources of data:
+
+- Covid-related mortality in all residents as reported to the CQC (outbreak surveillance counts, `r table_nums("incidence_CT_residents", display = "cite")`)
+- all-cause mortality in private residents (individual-level data, `r table_nums("tab_resident_descriptives", display = "cite")`).
+
+`r desc_residents$CT_total_cdeath` Covid-related resident deaths were reported across all care homes, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
+
+In comparison, the incidence of all-cause death in private residents was `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%], with an all-cause confirmed case fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] among private residents.
+This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR_rate)`% [`r tbrounding(desc_residents$DAT_BFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_BFR_rate_upper)`%] in private residents that did not test positive for Covid-19 (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(desc_residents$DAT_CFR_RR_lower) `; `r tbrounding(desc_residents$DAT_CFR_RR_upper)`]).
+The excess death fraction attributable to confirmed infection is thus `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)` [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`].
+
+We compared outbreak care homes with other care homes, focusing on private residents who had not tested positive for Covid-19: `r formatbm(desc_residents$DAT_BFR_outbreak_homes$y)`/`r formatbm(desc_residents$DAT_BFR_outbreak_homes$n)` died in homes reporting at least one infection or Covid-related death, compared with `r formatbm(desc_residents$DAT_BFR_other_homes$y)`/`r formatbm(desc_residents$DAT_BFR_other_homes$n)` in other homes.
+The risk of all-cause death was therefore twice as large in outbreak homes for uninfected residents (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) - 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) + 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
+This means `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR-1)/desc_residents$DAT_BFR_outbreak_homes$RR*100)`% [`r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_lower-1)/desc_residents$DAT_BFR_outbreak_homes$RR_lower*100)`; `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_upper-1)/desc_residents$DAT_BFR_outbreak_homes$RR_upper*100)`] of deaths in residents not testing positive in outbreak care homes is in some way attributable to the Covid-19 outbreak.
+Some of this difference could be an artefact of low case ascertainment: many residents who were infected are likely to be counted as uninfected due to not being tested, or receiving a false negative test result.
+In particular, because of delays in provided enough tests for all residents, many residents may have been tested late, at a point where their viral load was no longer detectable.
+
+***ADD PROPORTION of hospital deaths*** 
+24·7% of all COVID-19-related deaths reported by outbreak surveillance counts
 
@@
 
-Out of a total `r formatbm(desc_staff$total_staff)` staff,  `r formatbm(desc_staff$CT_symptomatic_total)`	(`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) were reported to have symptoms by FSHCG managers between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-During the same period, `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) staff were reported to have had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`).
+Symptoms of infection were reported by `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)`  staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_staff", display = "cite")`).
 
@@
 |	Incidence proportion (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
-|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`] |
+|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] |
 
@@
 )
+write.csv(km_staff_symptomatic, file = "supp_table_kaplan_meier_staff_symptomatic.csv", row.names = F)
+write.csv(km_staff_confirmed, file = "supp_table_kaplan_meier_staff_confirmed.csv", row.names = F)
 
@@
 
-## Factors associated with confirmed infections
-
-Using the individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-According to this, men were at significant greater risk of confirmed infections (adjusted HR = 1·3 [1·1; 1·6]), as were residents aged 85 years and over (adjusted HRs $\geq$ 1·4) and nursing home residents (adjusted HR = 1·5 [1·2; 1·8]).
-
-Large care homes had greater rates of infection, but this was partly attributable to other factors.
-Care homes with 70 beds or more had an adjusted HR of 1·8 [1·4; 2·4] compared with homes counting under 35 beds.
-Occupancy had an even greater effect: compared to homes between 7 and 8·5 residents for 10 bedrooms, confirmed infection hazards in homes with between 8·5 and 10 residents for 10 bedrooms were 2·5 times greater [1·9; 3·3].
-In  homes with more residents than bedrooms, hazards were 10 times [7·0; 15] greater.
-
-This was based on occupancy measured before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
+## Factors associated with confirmed infections in residents
+
+Using the individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
+Male gender, age $\geq$ 85 years (adjusted HRs $\geq$ 1·4) and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+Large care homes had greater rates of infection (adjusted HR = 1.8 [1·4; 2·4] for care homes with $\geq$ 70 beds versus <35 beds).
+Care home baseline occupancy and staffing ratios had the greatest effect on residents risk of infection.
+For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1·9; 3·3] greater in homes with 0·85-1 resident per room versus homes with 0·7-0·85 resident per room.
+In homes with more residents than bedrooms, hazards were 10 times [6·9; 15] greater.
+
+Higher staff to resident ratios were associated with lower risk of infection: a 10 percentage point increase in the number of staff relative to the total number of care home beds was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infection, once adjusting for other factors.
+
+It is important to note that bedroom occupancy and staffing ratios were estimated on the basis of baseline occupancy averaged before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
 As occupancy reduced more steeply in outbreak care homes during the study period due to hospitalisations and deaths, the average occupancy during the epidemic was negatively associated with infection rates (not reported).
-Finally, staffing acted as an important protective factor.
-An increase by 10 percentage point in the staff headcount relative to the total bed count was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infections, once adjusting for other factors.
 
@@
 
-The crude risk of all-cause death for the entire population of study was 19·5% (`r table_nums("tab_resident_descriptives", display = "cite")`).
-The all-cause confirmed case fatality rate was `r tbrounding(desc_residents$DAT_CFR)`% [`r tbrounding(desc_residents$DAT_CFR_lower)`%; `r tbrounding(desc_residents$DAT_CFR_upper)`%] based on incident reports.
-This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR)`% [`r tbrounding(desc_residents$DAT_BFR_lower)`%; `r tbrounding(desc_residents$DAT_BFR_upper)`%] in residents that did not test or did not test positive (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(exp(log(desc_residents$DAT_CFR_RR) - 1.96 * desc_residents$DAT_CFR_RR_SE)) `; `r tbrounding(exp(log(desc_residents$DAT_CFR_RR) + 1.96 * desc_residents$DAT_CFR_RR_SE))`]).
-
 The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by the infection status along other factors.
-Mortality increased with age and men's rate of death was 46% greater than women's (adjusted HR = 1·5 [1·3; 1·6]) all other things equals.
-After controlling for other risk factors, the hazard ratio for asymptomatic confirmed and symptomatic confirmed cases respectively were 2·1 [1·3; 3·6] and 8·5 [7·1; 10·2], confirming that symptomatic cases were more severe.
-As for symptomatic residents without a confirmed diagnosis, the adjusted hazard ratio of all-cause mortality was intermediary at 5·3 [4·4; 6·4].
+After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·5 [1·3; 1·6]), and receiving nursing rather than residential care (adjusted HR = 1·5 [1·3; 1.7]).
+All-cause death was strongly strongly associated with asymptomatic confirmed infection (adjusted HR = 3·3 [1·9; 5·6]) and symptomatic confirmed infection (adjusted HR = 12 [10; 15]), compared to residents with no evidence of Covid-19 infection.
+For all other residents, we studied mortality separately depending on whether their care home had recorded at least one confirmed infection or death yet (time-variant covariate):
+
+- all-cause death hazards of residents with no evidence of infection were 1·9 times greater [1·7; 2·2] in outbreak homes than in other homes
+- all-cause death hazards of residents with symptoms were 4·4 times greater [3·0; 6·2] than uninfected residents in homes without any outbreak
+- all-cause death hazards of residents with symptoms in outbreak homes were 9·3 times greater [7·5; 12] than uninfected residents in homes without any outbreak
+
+An increase by 10 percentage points in staffing was associated with a 4·4% reduction [1·9%; 7·0%] in mortality after adjustment for other variables.
+
 It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
-Mortality was significantly higher in nursing care homes (adjusted HR = 1·4 [1·3; 1.6]).
+
 This effect is likely to be at least partly attributable to a greater burden of comorbidities in nursing home residents.
@@
 ```{r risk_death, results="asis"}
+residents_episodes$Uninfected_other <- as.factor(residents_episodes$covid_status == "Negative" &
+  residents_episodes$outbreak==0)
+residents_episodes$Uninfected_outbreak <- as.factor(residents_episodes$covid_status == "Negative" &
+  residents_episodes$outbreak==1)
+residents_episodes$Symptomatic_not_confirmed_other <- as.factor(
+  residents_episodes$covid_status == "Symptomatic (not confirmed)" &
+  residents_episodes$outbreak==0)
+residents_episodes$Symptomatic_not_confirmed_outbreak <- as.factor(
+  residents_episodes$covid_status == "Symptomatic (not confirmed)" &
+    residents_episodes$outbreak==1)
+residents_episodes$Confirmed_Asymptomatic <- as.factor(
+  residents_episodes$covid_status == "Asymptomatic confirmed")
+residents_episodes$Confirmed_Symptomatic <- as.factor(
+  residents_episodes$covid_status == "Symptomatic confirmed")
+# 
 # cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type +covid_status+
+#                       gender+ ageg + home_type +covid_status * outbreak+
+#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
+#                       cluster(resident_id), data = residents_episodes)
+# summary(cph_death2)
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+#                       gender+ ageg + home_type + Uninfected_other +Uninfected_outbreak +
+#                       Symptomatic_not_confirmed_other +
+#                       Symptomatic_not_confirmed_outbreak +
+#                       Confirmed_Asymptomatic + Confirmed_Symptomatic +
 #                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
@@
                 "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
-                "covid_status", "cluster(resident_id)")
+                "Uninfected_outbreak",
+                "Symptomatic_not_confirmed_other", "Symptomatic_not_confirmed_outbreak",
+                "Confirmed_Asymptomatic", "Confirmed_Symptomatic",  "cluster(resident_id)")
 # explanatory_interaction = c("gender", "ageg", "user_type2",
@@
 knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
-
 ```
@@
 
-This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 infections in UK care homes independently of outbreaks, using individual-level resident records from a large care provider across `r as.integer(date_max-date_min)` days.
-Comparable studies were conducted in Ontario, Canada and Spain.[@Brown2020; @Stall2020; @Moraga2020] Findings confirm a very rapid progression of the epidemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms at some stage and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-
-As is the case with most investigations, the confirmed infection incidence parameters are likely to be severely underestimated.
-Despite the very rapid development of viral antigen tests, testing capacity was limited in the early weeks of the epidemic, resulting in many infections being missed before 11 May 2020.
-This comes in addition to the low sensitivity of such tests, even when administered in a timely manner.
-According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] on day 16.
-Other than underestimation of incidence, this may mean some underestimation in the incidence ratio comparing care homes with the community in England.
-Impact on case-fatality estimates is unclear, as this may affect both numerators and denominators.
-
-We adopted a very broad fatality case definition, by including deaths of all causes in the individual-level data analysis, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all COVID-19-related deaths reported by outbreak surveillance counts).
-The resulting case-fatality rate (`r tbrounding(desc_residents$DAT_CFR)`%) was far higher than in previous literature,[@Salcher-Konrad2020] which often had both a narrower fatality case definition, and shorter observation periods.
+## Main findings
+
+This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 in UK care homes during the pandemic period.
+Findings confirm a very rapid progression of the pandemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
+The case-fatality rate was `r tbrounding(desc_residents$DAT_CFR_rate)`%, `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% of which was attributable to Covid-19.
+Surprisingly, similar excess mortality is noticeable in residents with no record of infection in outbreak care homes, compared to those living in homes without outbreak (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
+Whilst substantial case under-ascertainment is likely, our findings suggest that both direct and indirect effects of the pandemic may contribute to excess deaths in care home residents.
+
+Our estimates of the prevalence of infection and deaths in residents are very similar to those reported in a recent large scale survey of care home managers in England,[@onsVivaldi2020] but it remains likely that we have underestimated the proportion of residents and staff who became infected during the pandemic.
+Testing capacity was limited in the early weeks of the epidemic, and PCR testing has been shown to have low sensitivity even when administered in a timely manner.
+According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] by day 16.
+Evidence from a systematic review[@Salcher-Konrad2020] suggests that up to 75% of care home residents have asymptomatic infection, increasing the likelihood that infected cases were not detected before universal testing was introduced in May.
+
+We adopted a broad definition of fatality, by including all-cause deaths for any individual with a confirmed infection, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts).
+Our case-fatality estimate (`r tbrounding(desc_residents$DAT_CFR_rate)`% on a period of `r as.integer(date_max - date_min)` days) thus represents an upper estimate of the case-fatality rate and was higher than in previous literature,[@Salcher-Konrad2020] which often had both a narrower fatality case definition, and shorter observation periods.
 An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
-These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19% among our private residents.
-Infections were also less common and more concentrated in Stall et al.[@Stall2020]
-
-Like a large study conducted in Ontario, Canada,[@Brown2020] we note a strong association between the occupancy rate and the incidence of infection.
-Although the effect was highest where residents had to share bedrooms, we observed that this association remained noticeable at levels well below 1 resident per bedroom.
-The limited range of variables measured did not allow to investigate potential counfounding extensively, particularly in relation to resident casemix (eg: comorbidities).
-
-We found strong associations between the incidence of infections and organisational factors: staff-to-bed ratio, and occupancy.
-These are connected to important healthcare infection control procedures, such as resident isolation and workplace hygiene.
-One hypothesis is that staffing ratios may be a sign of better-resourced care homes (for instance protective personal equipment), and of a greater capability to prescribe and observe effective infection control procedures.
-Furthermore, staff can become a vector in the transmission of infections.
-It is possible that higher staffing levels mitigate the transmission of infections since each staff member will care for a lower number of residents.
-Better staffed homes may also be less reliant on bank or agency staff who work in different locations.
-A more detailed survey of care home characteristics, along with protocols and measures adopted during the epidemic may help elucidate mechanisms through which staffing levels are associated with lower infection rates.
-As for occupancy, our finding echoes the conclusions of Brown et al.,[@Brown2020] although the association did not extend to mortality.
-
-This study was not able to examine some other important factors such as the presence of co-morbidities, the importing of cases from hospitals, or the rate of testing.
-This last factor is a major determinant of confirmed infection rates.
-Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection.[@Dora2020] Although increasing testing capacity raises the possibility of undertaking frequent, repeat testing in the care home population, this may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
-Understanding which care homes are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention accordingly.
+These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19% among FSHCG private residents, and as they focus on the outbreak period are less likely to identify indirect deaths associated with the pandemic.
+Infections were also less frequent and concentrated in a smaller proportion of care homes in Stall et al.[@Stall2020]
+
+`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/ desc_residents$total_homes * 100)`% of care homes in our study reported at least one case of resident infection or Covid-related death.
+This proportion is higher than PHE estimates based on outbreak detection, but is similar to the 56% of care homes that reported at least one case of infection in a national survey of care home managers.[@onsVivaldi2020]
+
+In common with a large study from Canada,[@Brown2020] we found strong associations between the incidence of infections and increased care home occupancy.
+Care homes with higher staff to resident ratios also had lower rates of infection and mortality in residents.
+These organisational factors are likely to moderate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, training staff, and regular deep cleaning of the care home environment.
+Higher staffing ratios may also be a sign of better-resourced care homes, with greater access to  protective personal equipment, a reduced risk of transmission of infection from staff to residents (since staff work with a smaller number of residents) and less reliance on temporary staff who are likely to work across multiple care settings.
+
+## Strengths and limitations
+
+The surveillance system established by FSHCG enabled us to establish a unique cohort study of staff and residents across a large care home chain in three countries throughout the pandemic period.
+Daily reporting of suspected and confirmed cases to care home managers ensured high quality data capture, since this system was used for mandatory reporting to the care home regulator.
+Importantly, it allowed us to estimate the burden of undiagnosed infection, since widespread testing for SARS-CoV-2 was not widely available to care homes until May.
+The incident reporting (Datix) system allowed detailed analysis of the relationship between suspected and confirmed infection and death in private residents.
+
+An important limitation of this study is lack of access to information on comorbidity and ethnicity, both of which have been shown in previous studies to be an important risk factors for Covid-19 infection and adverse outcomes.
+Our finding that residents receiving nursing rather than residential care have higher rates of infection and mortality is likely to be partly explained by higher rates of comorbidity in those receiving nursing care, however future studies should aim to capture individual level data on both of these risk factors.
+We were unable to investigate the relationship between residents being discharged to the care home from hospital and subsequent risk of infection in care home residents, since we did not have access to reliable information on the dates of hospital admission or discharge.
+Finally, we lacked information on the overall rate of care home testing, since incident reporting recorded positive test results and symptoms rather than whether a resident has been tested, and this is clearly a major determinant of confirmed infection rates.
+However, by 22nd June 2020, at least 90% of FSHCG homes had participated in the whole care home testing programme (when all residents and staff are offered testing for infection).
+Whilst the testing programme does not address the issue of case under-ascertainment at the start of the pandemic, it provides some evidence that around one-third of care homes did not experience any cases during the pandemic at all.
+It seems likely that these homes remain vulnerable to future waves of infection.
+
+## Clinical, research and policy implications
+
+Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection,[@Dora2020] since various studies have shown effective transmission of SARS-CoV-2 from people who are infected but not yet symptomatic.[@He2020; @Kimball2020] The UK government has recently announced plans[@dhscRegularTesting2020] to increase the frequency of testing in care home staff and residents.
+Although our findings strongly support increased used of testing to improve case ascertainment, frequent testing in care home residents may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
+Focussing testing on staff and/or developing strategies to identify care homes that are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention activities.
 The development of saliva antigen tests should also provide a less invasive alternative.
+
+Although we have largely attributed our finding of increased mortality in residents with no record of infection (in outbreak homes) to poor case ascertainment, the difference between outbreak and non-outbreak homes raises the possibility that Covid-19 may have indirectly increased mortality in care home residents.
+There are many ways in which delivery of usual care may have been impacted by the pandemic, increasing residents’ risk of preventable adverse events such as falls or pressure ulcers.
+Detailed analysis of cause of death and reasons for hospital admission in care home residents during the pandemic will be important to further explore this issue.
 
@@
 Such assumptions would not hold in the same way in the event of a second wave of the epidemic.
-Anticipation of a possible resurgence should constitute a further motive for rapidly strengthening surveillance systems.
+Anticipation of a possible resurgence should constitute a further motive for rapidly strengthening care home infection surveillance systems.
 
@@
 
-`r table_nums("tab_zero_cases")`
-
-```{r tab_zero_cases, results="asis"}
-home_zero_confirmed %>% 
-  transmute(`UK region/nation` = region_UK,
-            `Total homes`,
-            `Homes with no Datix confirmed case` = paste0(zero_datix, " (", 
-                                          tbrounding( pct_datix*100), "%)"),
-            `Homes with no manager-reported case` = paste0(zero_CT, " (", 
-                                          tbrounding( pct_CT*100), "%)")) %>% 
-  bind_rows(
-    tibble(`UK region/nation` = "Total",
-           `Total homes` = sum(home_zero_confirmed$`Total homes`),
-           `Homes with no Datix confirmed case` = paste0(
-             sum(home_zero_confirmed$zero_datix), " (", 
-             tbrounding( sum(home_zero_confirmed$zero_datix) / 
-                           sum(home_zero_confirmed$`Total homes`)*100), "%)"),
-            `Homes with no manager-reported case` = paste0(
-             sum(home_zero_confirmed$zero_CT), " (", 
-             tbrounding( sum(home_zero_confirmed$zero_CT) / 
-                           sum(home_zero_confirmed$`Total homes`)*100), "%)"))
-  ) %>% kable(align = c("l", "r", "r", "r", "r"))
-
+`r table_nums("tab_national_descriptives")` [@cqcdata2020; @hscni2020; @isd2020]
+
+```{r table_national_descriptives, fig.cap = tab_national_descriptives, message=F}
+mutate_all(desc_homes,  .funs = formatbm) %>% arrange(gsub("FSHC ", "", Scope)) %>% t %>% kable
 ```

### 2020-07-05 R/stats.R

@@ -186,7 +186,7 @@ km_ct_estimator <- function(df, t, d, pop, subtract_cases = TRUE, overwrite = TR
 
 
 relative_risk <- function(a, b, c, d) {
-  (a/b)/(c/d)
+  (a/(a + b))/(c/(c + d))
 }
 
 relative_risk_SE <- function(a, b, c, d){

### 2020-07-08 manuscript/manuscript.Rmd
manuscript v0.5

@@
 ---
-title: "Incidence of confirmed Covid-19 cases in care home staff and residents: Cohort study using electronic records in a large UK care home provider (March-June 2020)"
+title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
 # author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
@@
 library(finalfit)
+# override the finalfit cox PH modelling which defaults to Efron ties
+# using Breslow ties instead
+# coxphmulti <- function (.data, dependent, explanatory) {
+#     requireNamespace("survival")
+#     coxph(as.formula(paste0(dependent, "~", paste(explanatory, 
+#         collapse = "+"))), data = .data, ties = "breslow")
+# }
+# assignInNamespace("coxphmulti",coxphmulti,ns="finalfit")
 
@@
 date_max_tallies = as.Date("2020-06-14")
+date_min_cox_mortality <- as.Date("2020-04-01")
 
@@
   mutate(covid_status = case_when(
-    symptomatic==0 & confirmed == 0 ~ "Negative",
+    symptomatic==0 & confirmed == 0 ~ "Uninfected",
     symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
@@
   mutate(covid_status = factor(covid_status, 
-                               levels = c("Negative", "Symptomatic (not confirmed)",
+                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                           "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
@@
   dplyr::group_by(home_code) %>% 
-    dplyr::summarise(first_case_date = min(first_case_date)) %>% 
-  tidyr::crossing(tibble(outbreak = c(0, 1)))
+  dplyr::summarise(first_case_date = min(first_case_date)) 
+  # tidyr::crossing(tibble(outbreak = c(0, 1)))
 
@@
   mutate(covid_status = factor(covid_status, 
-                               levels = c("Negative", "Symptomatic (not confirmed)",
+                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                           "Asymptomatic confirmed", "Symptomatic confirmed")),
@@
   ungroup()
+
+residents_episodes <- dplyr::mutate(
+  residents_episodes,
+  covid_status_outbreak = as.factor(case_when(
+    covid_status == "Uninfected" & outbreak == 0 ~ "0 Uninfected (other home)",
+    covid_status == "Symptomatic (not confirmed)" &
+      outbreak==0 ~ "1 Symptomatic not confirmed  (other home)",
+    covid_status == "Uninfected" & outbreak == 1 ~ "2 Uninfected (outbreak home)",
+    covid_status == "Symptomatic (not confirmed)" &
+      outbreak==1 ~ "3 Symptomatic not confirmed  (outbreak home)",
+    covid_status == "Asymptomatic confirmed" ~ "4 Confirmed asymptomatic",
+    covid_status == "Symptomatic confirmed" ~ "5 Confirmed symptomatic"
+  )))
+
+residents_episodes$covid_status_outbreak_digit <- as.factor(
+  as.numeric(residents_episodes$covid_status_outbreak)-1)
+residents_episodes <- bind_cols(
+  residents_episodes,
+  as_tibble(model.matrix(~covid_status_outbreak_digit, residents_episodes)[ ,-1])
+)
+
+residents_episodes_short <- residents_episodes %>% 
+  filter(exit_date > date_min_cox_mortality) %>% 
+  mutate(date1 = if_else(
+    entry_date > date_min_cox_mortality,
+    date1,
+    as.numeric( date_min_cox_mortality - date_min),
+  ))
 
@@
 
+incidence_CT_residents <- table_nums(
+  name = "incidence_CT_residents",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+supp_epi_CT_oubreak_only <- table_nums(
+  name = "incidence_CT_staff",
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_resident_descriptives <- table_nums(
   name = "tab_resident_descriptives",
-  caption = paste0("Breakdown of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
-incidence_CT_residents <- table_nums(
-  name = "incidence_CT_residents",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
-supp_epi_CT_oubreak_only <- table_nums(
-  name = "incidence_CT_staff",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+tab_CFR_age_sex <- table_nums(
+  name = "tab_CFR_age_sex",
+  caption =  paste0("All-cause case-fatality rates and relative risks by age and sex among private residents (n=", 
+  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
+  "; ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_CoxPH_confirmed <- table_nums(
   name = "tab_CoxPH_confirmed",
-  caption =  paste0("Factors associated with the hazard of become a confirmed case: hazard ratios (HR) from a Cox proportional hazards model (n=", 
+  caption =  paste0("Risk factors for confirmed infection in private residents: hazard ratios (HR) from a Cox proportional hazards model (n=", 
   formatbm(dplyr::n_distinct(residents_overview$resident_id)),
@@
   name = "tab_CoxPH_death",
-  caption =  paste0("Factors associated with all-cause mortality: hazard ratios (HR) from a Cox proportional hazards model (n=", 
-  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
-  ")"))
+  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
+  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
+  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+tab_attributable_deaths <- table_nums(
+  name = "tab_attributable_deaths",
+  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
+  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
+  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 
@@
   name = "supp_epi_DAT",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections amongst residents according to Datix incident reports (", format(date_min, "%d/%m/%y"), "-", format(date_max, "%d/%m/%y"), ")"))
+  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections amongst residents according to Datix incident reports (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 
@@
   name = "tab_zero_cases",
-  caption = paste0("Number and proportion of homes without confirmed cases by UK region (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
+  caption = paste0("Number and proportion of homes without confirmed cases by UK region (", format(date_min_tallies, "%e %b %Y"), "-", format(date_max_tallies, "%e %b %Y"), ")")
 )
@@
 
-fig_KM_CT_residents <- fig_nums(
-  name = "fig_KM_CT_residents",
-  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in residents according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")")
-)
-
-fig_KM_staff <- fig_nums(
-  name = 'fig_KM_staff',
-  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infection in care home staff (n=",   formatbm(desc_staff$total_staff),
-  ") according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%d/%m/%y"), "-", format(date_max_tallies, "%d/%m/%y"), ")"))
+fig_KM_CT_plots <- fig_nums(
+  name = "fig_KM_CT_plots",
+  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in (A) residents (n=",   formatbm(desc_residents$CT_total_residents),
+  ") and (B) care home staff (n=",   formatbm(desc_staff$total_staff),
+  ") according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%e %b %Y"), "-", format(date_max_tallies, "%e %b %Y"), ")"))
 
@@
 **Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had confirmed infections.
-The incidence of confirmed infection in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days across all care homes.
-This includes `r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) which had not reported a single confirmed infection or Covid-related death.
-
-Among `r formatbm(desc_residents$DAT_total_residents)` private residents, `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred either in homes or in hospital (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]), including `r desc_residents$DAT_CFR$y` deaths of residents who had had a confirmed infection.
-This suggests an all-cause case-fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%], of which an estimated `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)` [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`] is attributable to confirmed SARS-CoV-2 infections.
-Independently, care home reports of Covid-related deaths suggest that `r desc_residents$CT_total_cdeath`/`r formatbm(desc_residents$CT_total_residents)` residents (`r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`]) died of a cause directly attributable to Covid-19.
-In addition to male gender and older age, confirmed asymptomatic (HR = 3·3 [1·9; 5·6]) and confirmed symptomatic (HR = 12 [10; 15]) infection were independently associated with all-cause mortality, by comparison with asymptomatic residents living in homes with no record of confirmed infections.
+The incidence of confirmed infection in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days.
+`r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) had neither confirmed infections nor Covid-related deaths.
+
+Among `r formatbm(desc_residents$DAT_total_residents)` private residents, there were `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]), `r desc_residents$DAT_CFR$y` of which occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%].
+`r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`%; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`%] of these deaths were attributable to SARS-CoV-2.
+Symptomatic and asymptomatic infection were independent risk factors for mortality.
 All-cause mortality for uninfected residents was 93% higher in care homes with outbreaks compared to those without outbreaks (adjusted HR = 1·9 [1·7; 2·2]).
@@
 Care home residents are particularly vulnerable to Covid-19 infection due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and this is compounded by frequent exposure to infection through close contact with infected staff members, other residents and contaminated surfaces in the care home setting.
-In the UK, an estimated 420,000 residents lived in approximately 11,000 homes for the elderly.[@cqcdata2020] At the peak of the pandemic the number of deaths in care home residents increased threefold over the equivalent period in 2019.[@onsDR2020]
-
-However, this is likely to underestimate of the true number of deaths in care home residents that were caused by the pandemic, since it only includes deaths in which Covid-19 was directly implicated.
-Death certifications in England and Wales between 28th December 2019 and 12th June 2020 suggest that two thirds of the 29,393 excess deaths recorded in care home residents (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] As part of efforts to mitigate the anticipated second wave of infection, it is critical to understand the ratio of direct to indirect deaths in care home residents.
-This will inform decisions around instigating lockdown measures to protect residents and staff from infection, whilst recognising the potential unintended consequences of interrupting usual care.
+In the UK, an estimated 420,000 residents lived in approximately 11,000 homes for the elderly.[@cqcdata2020]
+
+At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic.
+Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures that prevent infection may have a negative impact on residents by disrupting usual care.
 
 Disentangling direct and indirect deaths from Covid-19 requires accurate information on the incidence and prevalence of infection among staff and residents linked to individual-level clinical outcomes.
-There is no syndromic surveillance for infection in care homes in England, and universal testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established in care homes in England until 11th May 2020, when the government launched the whole care home testing programme.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations and within a maximum of five tests per care home.
+There is no syndromic surveillance for infection in care homes in England, and universal testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established in care homes in England until 11th May 2020, when the government launched the whole care home testing programme.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations with a maximum of five tests per care home.
 Consequently national estimates of incidence and prevalence based on test results obtained prior to 11th May will substantially underestimate the burden of infection in care home residents and staff.
 
-Outbreak investigations are useful to estimate the proportion of residents and staff who become infected during an outbreak,[@McMichael2020; @Dora2020] but they provide limited insight into the relationship between infection and mortality across the entire pandemic period, and cannot be generalised to care homes that have not reported any outbreaks.
+Outbreak investigations provide estimates of the proportion of residents and staff who become infected during an outbreak and case-fatality rates,[@McMichael2020; @Dora2020] but they provide limited insight into the relationship between infection and mortality across the entire pandemic period, and cannot be generalised to care homes that have not reported any outbreaks.
 A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long-term care facilities.
-Mortality rates in residents with confirmed Covid-19 infection ranged from 0 to 10%, with case fatality rates between 0 and 34%, however this review did not include estimates indirect mortality.
+Mortality rates in residents with confirmed Covid-19 infection ranged from 0 to 10%, with case fatality rates between 0 and 34%, however this review did not include estimates of indirect mortality.
 
@@
 
-## Aggregate outbreak surveillance counts (`r format(date_min_tallies, "%d/%m/%Y")`--`r format(date_max_tallies, "%d/%m/%Y")`)
+## Aggregate outbreak surveillance counts (`r format(date_min_tallies, "%e %b %Y")`--`r format(date_max_tallies, "%e %b %Y")`)
 
@@
 
-## Individual-level residents records (`r format(date_min, "%d/%m/%Y")`--`r format(date_max, "%d/%m/%Y")`)
+## Individual-level residents records (`r format(date_min, "%e %b %Y")`--`r format(date_max, "%e %b %Y")`)
 
@@
 A Kaplan-Meier product limit estimator was applied to this life table to estimate the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths every day between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between 26 April and 13 June 2020.[@cisbulletin2020; @cisdata2020]
+The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between `r format(england_results$date_min, "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
 
@@
 A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, occupant-to-bedroom ratio).
+
+).
+We also compared care ratesDates of infectiondischarge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality in care homes withbased on the presence of symptoms and without outbreaks, defined as at least one /or confirmed case or Covid-19 related death during the study period.
+infection.
+A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
+We also compared rates of infection and mortality in care homes with and without outbreaks, defined as at least one confirmed infection or Covid-19 related death during the study period.
+A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (uninfected; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed) as well as the time-dependent home outbreak status (no Covid-related death or confirmed infection; at least one Covid-related death or confirmed infection).
+The associated hazard ratios enable us to compare the mortality in each group as compared to the reference category (uninfected in non-outbreak care homes).
+Interpretation is facilitated by computing the associated attributable risk $ AR = (HR - 1)/HR$, which
+
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5259851/pdf/12874_2016_Article_285.pdf
+
+TO DOOOOO
+
+Bedroom occupancy and staffing ratios were estimated on the basis of baseline occupancy averaged before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
 Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and/or confirmed infection.
-A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
-
 Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab (PCR) testing across the study period following the introduction of whole care home testing on 11 May 2020.[@dhscTesting2020].
@@
 
-## Resident cases
+## Cases of infection in residents and staff
 
@@
 
-Individual-date were available for a subset of `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
-`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of private residents were in nursing homes, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
-
-`r table_nums("tab_resident_descriptives")`
-
-```{r table_resident_descriptives, message = FALSE, results="asis"}
-residents_overview$outbreak <- relevel(as.factor(dplyr::case_when(
-  residents_overview$home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
-  TRUE ~ "Other homes"
-)), "Outbreak homes")
-
-arsenal::tableby(outbreak  ~ gender + ageg  +
-                   user_type2 + admission_type2 + home_type +  covid_status + status + region_UK, 
-                 control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
-                 data = residents_overview) %>% summary()
-```
-
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r filter(new_cases, new_symptomatic_home>0 | new_symptomatic_hospital>0) %>% summarise(n=n_distinct(home_code)) %>% .$n`/`r desc_residents$total_homes` care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` cases of residents developing symptoms in homes, alongside `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents readmitted from hospitals presenting symptoms, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r filter(new_cases, new_symptomatic_home>0 | new_symptomatic_hospital>0) %>% summarise(n=n_distinct(home_code)) %>% .$n`/`r desc_residents$total_homes` care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` residents who developed symptoms in the care home and `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents who developed symptoms following readmission from hospital, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
 Infection was confirmed in `r desc_residents$CT_total_confirmed` residents, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
@@
 
+Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown on `r fig_nums("fig_KM_CT_plots", display = "cite")`.
+The sharp increase in confirmed infections between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
+
 In England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$FSHC_confirmed` confirmed cases out of `r formatbm(england_results$FSHC_confirmed)` person-days during this period.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$cis_confirmed` confirmed cases out of `r formatbm(england_results$cis_person_days)` person-days during this period.
 This implies a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
@@
 
-Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown on `r fig_nums("fig_KM_CT_residents", display = "cite")`.
-The sharp increase in confirmed infections between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
-
-Estimates of incidence within outbreak homes are reported alongside the overall estimates in the right handside of `r table_nums("incidence_CT_residents", display = "cite")`.
-Furthermore, incidence estimates for `r formatbm(desc_residents$DAT_total_residents)` private residents (using Datix incident records) are reported in supplementary material 2.
-Lower estimates were obtained which is likely due to reflect incomplete recording of cases in Datix, and the inability to link all incident records to an individual.
-
-```{r KM_CT_residents_symptomatic, fig.height = 4, fig.width = 8}
+```{r fig_KM_CT_plots, fig.height = 4.5, fig.width = 14, message=FALSE}
 write.csv(km_ct_residents_symp, file = "supp_table_kaplan_meier_residents_symptomatic.csv", row.names = F)
@@
 
-ggplot(aes(x = date, y = 1-S_t, 
+km_residents_graph <- ggplot(aes(x = date, y = 1-S_t, 
            ymax = 1 - St_max,
@@
   # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
+  geom_ribbon(alpha=0.1, linetype = 2 ) +
   geom_path(lwd=1.2) +
-  scale_y_continuous("Cum probability",
-                     breaks = seq(0,1,.05)) +
+  scale_y_continuous("Cumulative incidence",
+                     breaks = seq(0,1,.02)) +
   ggsci::scale_color_lancet() +
@@
                  unique(get_monday_date(all_km_residents$date)),
-                 "%d/%m/%y"
+                 "%e %b %Y"
                )) +
   theme_bw()  +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_CT_residents")`
-
-Mortality could be described from two sources of data:
-
-- Covid-related mortality in all residents as reported to the CQC (outbreak surveillance counts, `r table_nums("incidence_CT_residents", display = "cite")`)
-- all-cause mortality in private residents (individual-level data, `r table_nums("tab_resident_descriptives", display = "cite")`).
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
+  coord_cartesian(ylim = c(0,0.27)) + ggtitle("(A) Residents")
+
+all_km_staff <- bind_rows(
+  km_staff_symptomatic,
+  km_staff_confirmed
+)
+all_km_staff$Event <- factor(
+  all_km_staff$event,
+  levels = c(
+    "Symptomatic case",
+    "Confirmed infection"
+  )
+)
+write.csv(km_staff_symptomatic, file = "supp_table_kaplan_meier_staff_symptomatic.csv", row.names = F)
+write.csv(km_staff_confirmed, file = "supp_table_kaplan_meier_staff_confirmed.csv", row.names = F)
+
+km_staff_graph <- ggplot(aes(x = date, y = 1-S_t, 
+           ymax = 1 - St_max,
+           ymin = 1 - St_min,
+           color = Event), data = all_km_staff) +
+  # facet_grid(region_group ~ .) +
+  geom_ribbon(alpha=0.1, linetype = 2,stat = ) +
+  geom_path(lwd=1.2) +
+  ggsci::scale_color_lancet()+
+  scale_y_continuous("Cumulative incidence",
+                     breaks = seq(0,1,.02)) +
+  scale_x_date(name = "Date", 
+               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
+               labels = format.Date(
+                 unique(get_monday_date(km_staff_symptomatic$date)),
+                 "%e %b %Y"
+               )) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
+  coord_cartesian(ylim = c(0,0.27)) +  ggtitle("(B) Staff")
+
+km_ct_plots <- ggpubr::ggarrange(
+  km_residents_graph,
+  km_staff_graph
+)
+
+pdf("figure2.pdf", 14,4)
+km_ct_plots
+dev.off()
+km_ct_plots
+```
+
+`r fig_nums("fig_KM_CT_plots")`
+
+Symptoms of infection were reported by `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)`  staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+
+`r table_nums("incidence_CT_staff")`
+
+|   Staff       |  Symptomatic  |    Confirmed  | 
+| ------------- | -------------:| -------------:|
+| Cases      | 	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
+|	$N$ exposed  | `r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
+|	Total exposure (days)| `r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
+|	Incidence proportion (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
+|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] |
+
+Incidence estimates for `r formatbm(desc_residents$DAT_total_residents)` private residents derived from Datix incident records are reported in supplementary material 2.
+Lower estimates were obtained which is likely due to reflect incomplete recording of cases in Datix, and the inability to link all incident records to an individual.
+
+## Mortality in residents
 
 `r desc_residents$CT_total_cdeath` Covid-related resident deaths were reported across all care homes, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
-
-In comparison, the incidence of all-cause death in private residents was `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%], with an all-cause confirmed case fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] among private residents.
+`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts took place in hospital.
+
+Individual-level data were available for a subset of `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
+`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of private residents were in nursing homes, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+The crude incidence of all-cause death in private residents was `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
+All cause mortality in residents with confirmed infection was `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%].
 This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR_rate)`% [`r tbrounding(desc_residents$DAT_BFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_BFR_rate_upper)`%] in private residents that did not test positive for Covid-19 (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(desc_residents$DAT_CFR_RR_lower) `; `r tbrounding(desc_residents$DAT_CFR_RR_upper)`]).
@@
 
-We compared outbreak care homes with other care homes, focusing on private residents who had not tested positive for Covid-19: `r formatbm(desc_residents$DAT_BFR_outbreak_homes$y)`/`r formatbm(desc_residents$DAT_BFR_outbreak_homes$n)` died in homes reporting at least one infection or Covid-related death, compared with `r formatbm(desc_residents$DAT_BFR_other_homes$y)`/`r formatbm(desc_residents$DAT_BFR_other_homes$n)` in other homes.
+`r table_nums("tab_resident_descriptives")`
+
+```{r table_resident_descriptives, message = FALSE, results="asis"}
+residents_overview$outbreak <- relevel(as.factor(dplyr::case_when(
+  residents_overview$home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
+  TRUE ~ "Other homes"
+)), "Outbreak homes")
+
+arsenal::tableby(
+  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Home type` +  
+    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
+  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
+  data = mutate(residents_overview,
+                Sex = gender,
+                Age = ageg,
+                `Resident type` = user_type2,
+                `Admission type` = admission_type2,
+                `Home type` = home_type,
+                `Infection status by 14 June` = covid_status, 
+                `Status as of 14 June` = status,
+                `Region/nation` = region_UK 
+  )) %>% summary()
+```
+
+To investigate whether increased mortality was only seen in individuals who were infected with Covid-19, we compared mortality in residents with no record of Covid-19 from care homes with outbreaks and those without.
+`r formatbm(desc_residents$DAT_BFR_outbreak_homes$y)`/`r formatbm(desc_residents$DAT_BFR_outbreak_homes$n)` died in outbreak homes, compared with `r formatbm(desc_residents$DAT_BFR_other_homes$y)`/`r formatbm(desc_residents$DAT_BFR_other_homes$n)` in non-outbreak homes.
 The risk of all-cause death was therefore twice as large in outbreak homes for uninfected residents (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) - 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) + 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
-This means `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR-1)/desc_residents$DAT_BFR_outbreak_homes$RR*100)`% [`r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_lower-1)/desc_residents$DAT_BFR_outbreak_homes$RR_lower*100)`; `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_upper-1)/desc_residents$DAT_BFR_outbreak_homes$RR_upper*100)`] of deaths in residents not testing positive in outbreak care homes is in some way attributable to the Covid-19 outbreak.
-Some of this difference could be an artefact of low case ascertainment: many residents who were infected are likely to be counted as uninfected due to not being tested, or receiving a false negative test result.
-In particular, because of delays in provided enough tests for all residents, many residents may have been tested late, at a point where their viral load was no longer detectable.
-
-***ADD PROPORTION of hospital deaths*** 
-24·7% of all COVID-19-related deaths reported by outbreak surveillance counts
+This implies that `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR-1)/desc_residents$DAT_BFR_outbreak_homes$RR*100)`% [`r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_lower-1)/desc_residents$DAT_BFR_outbreak_homes$RR_lower*100)`%; `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_upper-1)/desc_residents$DAT_BFR_outbreak_homes$RR_upper*100)`%] of deaths in residents who did not test positive for infection in outbreak homes were in some way attributable to Covid-19.
 
@@
 
-## Staff cases
-
-Symptoms of infection were reported by `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)`  staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_staff", display = "cite")`).
-
-`r table_nums("incidence_CT_staff")`
-
-|   Staff       |  Symptomatic  |    Confirmed  | 
-| ------------- | -------------:| -------------:|
-| Cases      | 	`r formatbm(desc_staff$CT_symptomatic_total)`|	`r formatbm(desc_staff$CT_confirmed_total)` |
-|	$N$ exposed  | `r formatbm(desc_staff$total_staff)`|	`r formatbm(desc_staff$total_staff)`|
-|	Total exposure (days)| `r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
-|	Incidence proportion (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
-|	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] |
-
-```{r KM_Staff, fig.height = 4, fig.width = 8}
-all_km_staff <- bind_rows(
-  km_staff_symptomatic,
-  km_staff_confirmed
-)
-all_km_staff$Event <- factor(
-  all_km_staff$event,
-  levels = c(
-    "Symptomatic case",
-    "Confirmed infection"
-  )
-)
-write.csv(km_staff_symptomatic, file = "supp_table_kaplan_meier_staff_symptomatic.csv", row.names = F)
-write.csv(km_staff_confirmed, file = "supp_table_kaplan_meier_staff_confirmed.csv", row.names = F)
-
-ggplot(aes(x = date, y = 1-S_t, 
-           ymax = 1 - St_max,
-           ymin = 1 - St_min,
-           color = Event), data = all_km_staff) +
-  # facet_grid(region_group ~ .) +
-  geom_ribbon(alpha=0.05, linetype = 2,stat = ) +
-  geom_path(lwd=1.2) +
-  ggsci::scale_color_lancet()+
-  scale_y_continuous("Cum probability symptomatic case",
-                     breaks = seq(0,1,.02)) +
-  scale_x_date(name = "Date", 
-               breaks = unique(get_monday_date(km_staff_symptomatic$date)),
-               labels = format.Date(
-                 unique(get_monday_date(km_staff_symptomatic$date)),
-                 "%d/%m/%y"
-               )) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
-
-```
-
-`r fig_nums("fig_KM_staff")`
+`r table_nums("tab_CFR_age_sex")`
+
+```{r tab_CFR_age_sex, results="asis"}
+residents_overview %>%
+  mutate(Age = ageg, Sex = gender) %>% 
+  group_by(Age, Sex) %>% 
+  summarise(
+    n = n_distinct(resident_id),
+    resident_days = sum(resident_days),
+    confirmed_infections = sum(confirmed == 1),
+    deaths = sum(status == "Deceased"),
+    deaths_confirmed = sum(status == "Deceased" & confirmed == 1)) %>% 
+  mutate(
+    cfr = tbrounding(deaths_confirmed/confirmed_infections*100),
+    cfr_lower = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$lower*100),
+    cfr_upper = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$upper*100),
+    RR = relative_risk(deaths_confirmed, confirmed_infections - deaths_confirmed,
+                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed)),
+    RR_SE = relative_risk_SE(deaths_confirmed, confirmed_infections - deaths_confirmed,
+                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed))
+  ) %>%  ungroup() %>% 
+  transmute(
+    Age,
+    Sex,
+    N = formatbm(n),
+    `Confirmed infections` = confirmed_infections,
+    `Total deaths` = deaths,
+    `Deaths in confirmed infections` = deaths_confirmed,
+    `Case-fatality rate (%)` = paste0(cfr, " [", cfr_lower, "; ", cfr_upper, "]"),
+    `Relative risk` = paste0(
+      tbrounding(RR, 2), " [",
+      tbrounding(exp(log(RR) - 1.96 * RR_SE), 2), "; ",
+      tbrounding(exp(log(RR) + 1.96 * RR_SE), 2), "]"
+    )
+  ) %>% kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
+```
 
@@
 
-Using the individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
+Using individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
 Male gender, age $\geq$ 85 years (adjusted HRs $\geq$ 1·4) and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
@@
 Higher staff to resident ratios were associated with lower risk of infection: a 10 percentage point increase in the number of staff relative to the total number of care home beds was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infection, once adjusting for other factors.
-
-It is important to note that bedroom occupancy and staffing ratios were estimated on the basis of baseline occupancy averaged before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
-As occupancy reduced more steeply in outbreak care homes during the study period due to hospitalisations and deaths, the average occupancy during the epidemic was negatively associated with infection rates (not reported).
 
@@
 
-The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine how mortality is affected by the infection status along other factors.
+The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status and mortality.
 After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·5 [1·3; 1·6]), and receiving nursing rather than residential care (adjusted HR = 1·5 [1·3; 1.7]).
-All-cause death was strongly strongly associated with asymptomatic confirmed infection (adjusted HR = 3·3 [1·9; 5·6]) and symptomatic confirmed infection (adjusted HR = 12 [10; 15]), compared to residents with no evidence of Covid-19 infection.
-For all other residents, we studied mortality separately depending on whether their care home had recorded at least one confirmed infection or death yet (time-variant covariate):
-
-- all-cause death hazards of residents with no evidence of infection were 1·9 times greater [1·7; 2·2] in outbreak homes than in other homes
-- all-cause death hazards of residents with symptoms were 4·4 times greater [3·0; 6·2] than uninfected residents in homes without any outbreak
-- all-cause death hazards of residents with symptoms in outbreak homes were 9·3 times greater [7·5; 12] than uninfected residents in homes without any outbreak
-
-An increase by 10 percentage points in staffing was associated with a 4·4% reduction [1·9%; 7·0%] in mortality after adjustment for other variables.
+All-cause death was strongly associated with asymptomatic confirmed infection (adjusted HR = 3·3 [1·9; 5·6]) and symptomatic confirmed infection (adjusted HR = 12 [10; 15]), compared to residents with no evidence of Covid-19 infection.
+
+We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection in non-outbreak homes as the reference group.
+Risk of all-cause mortality was almost two-fold higher in residents with no evidence of infection in outbreak versus non-outbreak homes (adjusted HR = 1·9 [1·7; 2·2]) in outbreak homes than in other homes.
+Risk of death was also higher in individuals with symptoms but unconfirmed infection in outbreak homes compared to non-outbreak homes (adjusted HR = 4·4 [3·0; 6·2] versus 9·3 [7·5; 12]).
 
@@
 
-This effect is likely to be at least partly attributable to a greater burden of comorbidities in nursing home residents.
-Beyond, there was no evidence of an association between care home size and all-cause mortality.
-After adjusting for other variables, an increase by 10 percentage points in staffing was associated with a 4.3% reduction [1.8%; 6.9%] in mortality after adjustment for other variables.
-Mortality did not appear to be associated with care home size or occupancy, at least not in a linear fashion: mortality was lower in homes with a ratio of occupants to bedrooms in excess of 1 compared with a ratio comprised between .85 and 1; while also being lower in homes with a ratio between 0·55 and 0·7.
+After adjusting for other variables, an increase by 10 percentage points in staffing was associated with a 3·1% reduction [0·6%; 5·7%] in mortality after adjustment for other  variables.
+
+```{r residents_daily}
+
+residents_daily <- tline %>% 
+  filter(rday==1)  %>% 
+  left_join(home_first_case) %>%
+  mutate(outbreak = dplyr::case_when(
+    date >= first_case_date ~ 1L,
+    date < first_case_date ~ 0L,
+    TRUE ~ 0L
+  )) %>% 
+  group_by(resident_id) %>% 
+  mutate(death = case_when(
+    status == "Deceased" & date == max(date) ~ 1L,
+    TRUE ~ 0L
+  )) %>% 
+  ungroup() %>% 
+  mutate(symptomatic = -susceptible_symptomatic+1,
+         confirmed = -susceptible_confirmed+1,
+         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1)) %>% 
+  mutate(covid_status_outbreak = as.factor(case_when(
+    symptomatic==0 & confirmed == 0 & outbreak == 0 ~ "0 Uninfected (other home)",
+    symptomatic==1 & confirmed == 0 & outbreak == 0 ~ "1 Symptomatic not confirmed  (other home)",
+    symptomatic==0 & confirmed == 0 & outbreak == 1 ~ "2 Uninfected (outbreak home)",
+    symptomatic==1 & confirmed == 0 & outbreak == 1 ~ "3 Symptomatic not confirmed  (outbreak home)",
+    symptomatic==0 & confirmed == 1 ~ "4 Confirmed asymptomatic",
+    symptomatic==1 & confirmed == 1 ~ "5 Confirmed symptomatic"
+  )))
+residents_daily$covid_status_outbreak_digit <- as.factor(
+  as.numeric(residents_daily$covid_status_outbreak)-1)
+residents_daily <- bind_cols(
+  residents_daily,
+  as_tibble(model.matrix(~covid_status_outbreak_digit, residents_daily)[,-1])
+)
+
+# residents_daily <- residents_daily %>% 
+#   group_by(date) %>% 
+#   summarise(n = sum(rday),
+#             d = sum(death),
+#             covid_status_outbreak_digit0 = 1 - 
+#               mean(covid_status_outbreak_digit1) -
+#               mean(covid_status_outbreak_digit2) -
+#               mean(covid_status_outbreak_digit3) -
+#               mean(covid_status_outbreak_digit4) -
+#               mean(covid_status_outbreak_digit5),
+#             covid_status_outbreak_digit1 = mean(covid_status_outbreak_digit1),
+#             covid_status_outbreak_digit2 = mean(covid_status_outbreak_digit2),
+#             covid_status_outbreak_digit3 = mean(covid_status_outbreak_digit3),
+#             covid_status_outbreak_digit4 = mean(covid_status_outbreak_digit4),
+#             covid_status_outbreak_digit5 = mean(covid_status_outbreak_digit5)) %>% 
+#   mutate(date_int = as.integer(date - date_min))
+
+```
 
@@
 ```{r risk_death, results="asis"}
-residents_episodes$Uninfected_other <- as.factor(residents_episodes$covid_status == "Negative" &
-  residents_episodes$outbreak==0)
-residents_episodes$Uninfected_outbreak <- as.factor(residents_episodes$covid_status == "Negative" &
-  residents_episodes$outbreak==1)
-residents_episodes$Symptomatic_not_confirmed_other <- as.factor(
-  residents_episodes$covid_status == "Symptomatic (not confirmed)" &
-  residents_episodes$outbreak==0)
-residents_episodes$Symptomatic_not_confirmed_outbreak <- as.factor(
-  residents_episodes$covid_status == "Symptomatic (not confirmed)" &
-    residents_episodes$outbreak==1)
-residents_episodes$Confirmed_Asymptomatic <- as.factor(
-  residents_episodes$covid_status == "Asymptomatic confirmed")
-residents_episodes$Confirmed_Symptomatic <- as.factor(
-  residents_episodes$covid_status == "Symptomatic confirmed")
-# 
 # cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type +covid_status * outbreak+
+#                       gender+ ageg + home_type + covid_status_outbreak +
 #                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       cluster(resident_id), data = residents_episodes)
-# summary(cph_death2)
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type + Uninfected_other +Uninfected_outbreak +
-#                       Symptomatic_not_confirmed_other +
-#                       Symptomatic_not_confirmed_outbreak +
-#                       Confirmed_Asymptomatic + Confirmed_Symptomatic +
-#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       cluster(resident_id), data = residents_episodes)
+#                       cluster(resident_id), data = residents_episodes_short, ties = "breslow")
 # summary(cph_death2)
 # cox.zph(cph_death2)
-# survfit(Surv(time = date1, time2 = date2, event = death_status)~ gender, data = residents_episodes)
-#  pct_occupancy, occ_bedroom_ratio,no_bedroom 
+
 explanatory = c("gender", "ageg", "home_type", "user_type2",
                 "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
-                "Uninfected_outbreak",
-                "Symptomatic_not_confirmed_other", "Symptomatic_not_confirmed_outbreak",
-                "Confirmed_Asymptomatic", "Confirmed_Symptomatic",  "cluster(resident_id)")
-# explanatory_interaction = c("gender", "ageg", "user_type2",
-#                 "staff_to_bed_ratio", "bed_tot_category", 
-#                 "covid_status:home_type", "region_UK", "cluster(resident_id)")
+                "covid_status_outbreak",  "cluster(resident_id)")
 dependent = "Surv(time = date1, time2 = date2, event = death_status)"
-CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
+CPH_death <-  finalfit(residents_episodes_short, dependent, explanatory, 
                        column = T, add_dependent_label = F)  
@@
 
-```{r fig_KM_mortality,  fig.height = 4, fig.width = 8}
+*Note:* Baseline group = uninfected residents in non-outbreak care home
+
+```{r fig_KM_mortality,  fig.height = 6, fig.width = 8, message=F}
 KM_mortality_by_covid_status <- survfit(
-  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status, 
-  data = residents_episodes)
-ggsurvplot(KM_mortality_by_covid_status, conf.int = TRUE, 
-           legend = "right",
-           ggtheme = theme_bw(),  
-           palette = "lancet",
-           break.x.by	= 7,
+  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + outbreak, 
+  data = residents_episodes_short)
+KM_mortality_plot <- ggsurvplot_facet(KM_mortality_by_covid_status,
+                 residents_episodes_short,
+                 facet.by  = "covid_status",
+                 legend.labs = c("Non-outbreak home", "Outbreak home"),
+                 short.panel.labs	 = T, break.x.by	= 7, 
                  break.y.by = .1,
-           legend.labs = levels(residents_episodes$covid_status),
-           xlab  = paste0("Days since ", format(date_min, "%d %B %Y")),
-           ylab = "Survival",
-           conf.int.alpha = .1)  
+                 palette = "aaas", conf.int = T,
+                 scales = "free", ggtheme = theme_bw(), xlim = c(32, 104),
+                 xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
+                 conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)
+pdf("figure5.pdf",8, 6)
+KM_mortality_plot
+dev.off()
+KM_mortality_plot
 ```
@@
 `r fig_nums("fig_KM_mortality")`
+
+`r table_nums("tab_attributable_deaths")`
+
+```{r, warning=F, results = "asis"}
+cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+                      gender+ ageg + home_type + covid_status_outbreak +
+                      staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
+                      cluster(resident_id), data = residents_episodes_short)
+HRs_covid <- tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak", 
+                                                          names(cph_death2$coefficients))]) %>% 
+  transmute(covid_status_outbreak = gsub("covid_status_outbreak", "", name),
+            HR = exp(value))
+HRs_CI_covid <- as_tibble(exp(confint(cph_death2))[grep("covid_status_outbreak", 
+                                                          names(cph_death2$coefficients)),]) %>% 
+  rename(HR_lower =  `2·5 %`,
+         HR_upper = `97·5 %`)
+HRs_CI_covid$covid_status_outbreak <- gsub(
+  "covid_status_outbreak", "",
+  grep("covid_status_outbreak", names(cph_death2$coefficients), value = T)
+) 
+HRs_covid <- merge(HRs_covid, HRs_CI_covid)
+
+attributable_deaths <- residents_episodes_short %>% 
+  filter(death_status == 1) %>% 
+  group_by(covid_status_outbreak) %>% 
+  summarise(Deaths = n()) %>% 
+    left_join(HRs_covid) %>% 
+    mutate(HR = tidyr::replace_na(HR, 1),
+           HR_lower = tidyr::replace_na(HR_lower, 1),
+           HR_upper = tidyr::replace_na(HR_upper, 1)) %>% 
+    mutate(AH = (HR - 1)/HR,
+           AH_lower = (HR_lower - 1)/HR_lower,
+           AH_upper = (HR_upper - 1)/HR_upper) %>% 
+    mutate(`Attributable deaths` = AH * Deaths) %>% 
+    mutate(`AD (%)` = tbrounding(`Attributable deaths`/ sum(Deaths)*100)) %>% 
+  transmute(`Infection/outbreak status` = covid_status_outbreak,
+            Deaths,
+            HR = paste0(tbrounding(HR,2), " [",
+                        tbrounding(HR_lower,2),"; ",
+                        tbrounding(HR_upper,2), "]"),
+            `Attibutable hazard (%)`= paste0(tbrounding(AH*100,1), " [",
+                        tbrounding(AH_lower*100,1),"; ",
+                        tbrounding(AH_upper*100,1), "]"),
+           `Attributable deaths` = round(`Attributable deaths`),
+           `AD (%)`)
+attributable_deaths <- bind_rows(
+  attributable_deaths,
+  tibble(
+    `Infection/outbreak status` = "TOTAL",
+            Deaths = (sum(attributable_deaths$Deaths)),
+            HR = "",
+            `Attibutable hazard (%)`= "",
+           `Attributable deaths` = sum(attributable_deaths$`Attributable deaths`),
+           `AD (%)` = tbrounding(
+             sum(attributable_deaths$`Attributable deaths`)/sum(attributable_deaths$Deaths)*100)
+  )
+)
+
+kable(attributable_deaths, align =c("l", "r", "r", "r", "r", "r", "r"))
+```
+
+```{r}
+
+# residents_episodes$resident_id_int <- as.numeric(as.factor(residents_episodes$resident_id))
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+#                       gender+ ageg + home_type + covid_status_outbreak_digit1 +
+#                       covid_status_outbreak_digit2 + covid_status_outbreak_digit3+
+#                       covid_status_outbreak_digit4 +covid_status_outbreak_digit5 +
+#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
+#                       cluster(resident_id_int), data = data.frame(residents_episodes), 
+#                     ties = "breslow")
+# 
+# trucmuche <- survfit(Surv(date1, date2, death_status) ~
+#                       gender+ ageg + home_type + covid_status_outbreak_digit1 +
+#                       covid_status_outbreak_digit2 + covid_status_outbreak_digit3+
+#                       covid_status_outbreak_digit4 +covid_status_outbreak_digit5 +
+#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6,
+#                      residents_episodes_short, id = "resident_id")
+# 
+# AH_fractions <- residents_daily %>% 
+#   tidyr::pivot_longer(cols=starts_with("covid_status_outbreak_")) %>% 
+#   dplyr::rename(prevalence = value) %>% 
+#   left_join(tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak_digit", names(cph_death2$coefficients))])) %>% 
+#   mutate(AH = (prevalence * tidyr::replace_na(exp(value) - 1, 0)) / 
+#            (prevalence * tidyr::replace_na(exp(value) - 1, 0) +1))
+# 
+# 
+# AH_coefs <- tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak_digit",
+#                                                          names(cph_death2$coefficients))]) %>% 
+#   rename(covid_status_outbreak_digit = name) %>% 
+#   mutate(covid_status_outbreak_digit = gsub("covid_status_outbreak_digit", "", covid_status_outbreak_digit))
+#                               
+# AH_fractions <- residents_daily %>%
+#   group_by(date, covid_status_outbreak_digit) %>%
+#   summarise(n = sum(rday),
+#             d = sum(death)) %>% 
+#   left_join(AH_coefs)  %>% 
+#   filter(date >= as.Date("2020-04-01")) %>% 
+#   mutate(AH =  (tidyr::replace_na(exp(value), 1) -1 ) / tidyr::replace_na(exp(value), 1)) %>% 
+#   group_by(covid_status_outbreak_digit) %>% 
+#   summarise(AH = sum(AH*d),
+#             d = sum(d),
+#             n = sum(n))  
+#             
+# 
+# 
+#  
+# 
+# ggplot(AH_fractions)+
+#   geom_line(aes(x = date, y = AH*d, color = name ))
+#
+
+# 
+# # summary(cph_death2)
+# AF_estimates <- list()
+# for(i in 1:5) {
+#   AF_estimates[[levels(residents_episodes$covid_status_outbreak)[i]]] <- 
+#     AF::AFcoxph(object = cph_death2, data = data.frame(residents_episodes), 
+#             exposure = paste0("covid_status_outbreak_digit", i), clusterid = "resident_id_int")
+#   AF_estimates[[i]]$se <- sqrt(AF_estimates[[i]]$AF.var)
+#   AF_estimates[[i]]$confidence.interval <- AF:::CI.AF(
+#     AF = AF_estimates[[i]]$AF.est, 
+#     Std.Error = AF_estimates[[i]]$se, 
+#     confidence.level = 0.95, CI.transform = "log")
+#   colnames(AF_estimates[[i]]$confidence.interval) <- c("Lower limit", "Upper limit")
+#   AF_estimates[[i]]$covid_status_outbreak <- levels(
+#     residents_episodes$covid_status_outbreak)[i+1]
+#   AF_estimates[[i]] <- bind_cols(
+#     as_tibble(
+#       AF_estimates[[i]][c('times', 'covid_status_outbreak', 'AF.est', 'AF.var', 
+#                          'S.est', 'S0.est', 'S.var', 'S0.var')]
+#     ),
+#     as_tibble(AF_estimates[[i]]$confidence.interval)
+#   )
+# }
+# 
+# 
+# ggplot(bind_rows(AF_estimates[2:5]),
+#        aes(x = times,  color =  covid_status_outbreak )) +
+#   geom_ribbon(aes(ymin = `Lower limit`*100, ymax = `Upper limit`*100), 
+#               alpha=0.05, linetype = 2) +
+#   geom_line(aes(y=AF.est*100), size = 1) +
+#   coord_cartesian(ylim=c(0, 30)) +
+#   ggsci::scale_color_lancet() +
+#   xlab("Excess mortality (%)") +
+#   ggtitle("% excess mortality compared with uninfected in non-outbreak homes")
+# 
+# # 
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
+#                         covid_status_outbreak_digit  +
+#                       cluster(resident_id_int), data = residents_episodes, 
+#                     ties = "breslow")
+# cph_death2 <- flexsurv::flexsurvreg(Surv(date1, date2, death_status) ~
+#                         covid_status_outbreak_digit  +
+#                       cluster(resident_id), 
+#                       data = residents_episodes_short,dist = "weibull")
+# 
+# truc <- survfit(cph_death2, tibble(date1 = 30, covid_status_outbreak_digit = factor(1:5, levels = 1:5) )) 
+#   ggsurvplot(truc,
+#              data = tibble(date1 = 30, covid_status_outbreak_digit = factor(5, levels = 1:5) ))
+#     
+```
 
@@
 The case-fatality rate was `r tbrounding(desc_residents$DAT_CFR_rate)`%, `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% of which was attributable to Covid-19.
-Surprisingly, similar excess mortality is noticeable in residents with no record of infection in outbreak care homes, compared to those living in homes without outbreak (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
+Similar excess mortality is noticeable in residents with no record of infection in outbreak care homes, compared to those living in homes without outbreak (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
 Whilst substantial case under-ascertainment is likely, our findings suggest that both direct and indirect effects of the pandemic may contribute to excess deaths in care home residents.
@@
 Our estimates of the prevalence of infection and deaths in residents are very similar to those reported in a recent large scale survey of care home managers in England,[@onsVivaldi2020] but it remains likely that we have underestimated the proportion of residents and staff who became infected during the pandemic.
-Testing capacity was limited in the early weeks of the epidemic, and PCR testing has been shown to have low sensitivity even when administered in a timely manner.
+Testing capacity was limited in the early weeks of the epidemic, and PCR testing has been shown to have low sensitivity even when administered early in the duration of illness.
 According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] by day 16.
@@
 
-We adopted a broad definition of fatality, by including all-cause deaths for any individual with a confirmed infection, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts).
+We adopted a broad definition of mortality, by including all-cause deaths for any individual with a confirmed infection, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts).
 Our case-fatality estimate (`r tbrounding(desc_residents$DAT_CFR_rate)`% on a period of `r as.integer(date_max - date_min)` days) thus represents an upper estimate of the case-fatality rate and was higher than in previous literature,[@Salcher-Konrad2020] which often had both a narrower fatality case definition, and shorter observation periods.
@@
 
+**Need to add a paragraph about the limitations of estimating attributable mortality.**
+
 `r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/ desc_residents$total_homes * 100)`% of care homes in our study reported at least one case of resident infection or Covid-related death.
-This proportion is higher than PHE estimates based on outbreak detection, but is similar to the 56% of care homes that reported at least one case of infection in a national survey of care home managers.[@onsVivaldi2020]
+This proportion is higher than PHE estimates based on outbreak detection, and higher than the 56% of care homes that reported at least one case of infection in an English national survey of care home manager.[@onsVivaldi2020] Reasons for this discrepancy may relate to systematic differences between FSHCG homes and homes that were surveyed, better detection of cases though FSHCG's daily reporting system, or comparatively high rates of participation in the national whole care home testing programme.
 
@@
 Care homes with higher staff to resident ratios also had lower rates of infection and mortality in residents.
-These organisational factors are likely to moderate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, training staff, and regular deep cleaning of the care home environment.
+These organisational factors are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, training staff, and regular deep cleaning of the care home environment.
 Higher staffing ratios may also be a sign of better-resourced care homes, with greater access to  protective personal equipment, a reduced risk of transmission of infection from staff to residents (since staff work with a smaller number of residents) and less reliance on temporary staff who are likely to work across multiple care settings.
@@
 
-An important limitation of this study is lack of access to information on comorbidity and ethnicity, both of which have been shown in previous studies to be an important risk factors for Covid-19 infection and adverse outcomes.
-Our finding that residents receiving nursing rather than residential care have higher rates of infection and mortality is likely to be partly explained by higher rates of comorbidity in those receiving nursing care, however future studies should aim to capture individual level data on both of these risk factors.
+An important limitation of this study is lack of access to information on comorbidity and ethnicity, both of which have been shown in previous studies to be an important risk factors for Covid-19 infection and adverse outcomes.[@Zhou2020] Our finding that residents receiving nursing rather than residential care have higher rates of infection and mortality is likely to be partly explained by higher rates of comorbidity in those receiving nursing care, however future studies should aim to capture individual level data on both of these risk factors.
 We were unable to investigate the relationship between residents being discharged to the care home from hospital and subsequent risk of infection in care home residents, since we did not have access to reliable information on the dates of hospital admission or discharge.
@@
 However, by 22nd June 2020, at least 90% of FSHCG homes had participated in the whole care home testing programme (when all residents and staff are offered testing for infection).
-Whilst the testing programme does not address the issue of case under-ascertainment at the start of the pandemic, it provides some evidence that around one-third of care homes did not experience any cases during the pandemic at all.
+Whilst results from the testing programme cannot address the issue of case under-ascertainment at the start of the pandemic, the fact that the majority of FSHCG homes participated in the whole home testing programme makes it likely that one third of care homes have genuinely not experienced any Covid-19 cases.
 It seems likely that these homes remain vulnerable to future waves of infection.
@@
 Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection,[@Dora2020] since various studies have shown effective transmission of SARS-CoV-2 from people who are infected but not yet symptomatic.[@He2020; @Kimball2020] The UK government has recently announced plans[@dhscRegularTesting2020] to increase the frequency of testing in care home staff and residents.
-Although our findings strongly support increased used of testing to improve case ascertainment, frequent testing in care home residents may not be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
+Although our findings strongly support increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
 Focussing testing on staff and/or developing strategies to identify care homes that are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention activities.
-The development of saliva antigen tests should also provide a less invasive alternative.
-
-Although we have largely attributed our finding of increased mortality in residents with no record of infection (in outbreak homes) to poor case ascertainment, the difference between outbreak and non-outbreak homes raises the possibility that Covid-19 may have indirectly increased mortality in care home residents.
+Alternatively the development of saliva antigen tests could provide a less invasive solution to testing.
+
+Although we have largely attributed our findings of increased mortality in residents with no record of infection (in outbreak homes) to poor case ascertainment, the difference between outbreak and non-outbreak homes raises the possibility that Covid-19 may have indirectly increased mortality in care home residents.
 There are many ways in which delivery of usual care may have been impacted by the pandemic, increasing residents’ risk of preventable adverse events such as falls or pressure ulcers.

### 2020-07-08 manuscript/manuscript.Rmd
excess mortality

@@
 output:
-  word_document:
-    toc: yes
-    reference_docx: template.docx
   html_document:
@@
 
-```{r, fig.width=10, fig.height=5}
+```{r study_diagram, fig.width=10, fig.height=5}
 knitr::include_graphics("study_diagram.svg")
@@
 The associated hazard ratios enable us to compare the mortality in each group as compared to the reference category (uninfected in non-outbreak care homes).
-Interpretation is facilitated by computing the associated attributable risk $ AR = (HR - 1)/HR$, which
-
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5259851/pdf/12874_2016_Article_285.pdf
-
-TO DOOOOO
+The corresponding attributable fraction is obtained by using the model to predict the counterfactual mortality, then computing AFW, the attributable fraction within study:[@Samuelsen2008]
+
+$$ AFW = \dfrac{\sum_{i=1}^{n}(\hat{F}_i - \hat{F}_i^\star)}{\sum_{i=1}^{n}\hat{F}_i} $$
+
+where $\hat{F}_i$ designates the predicted probability of death of individual $i = \{1, \dots, n\}$ based on the observed infection/outbreak classification, and  $\hat{F}_i^\star$ the same probability had the individual been uninfected and living in a non-outbreak care home.
+Observed censoring times are used.
 
@@
 
-```{r fig_KM_CT_plots, fig.height = 4.5, fig.width = 14, message=FALSE}
+```{r fig_KM_CT_plots, fig.height = 4, fig.width = 12, message=FALSE}
 write.csv(km_ct_residents_symp, file = "supp_table_kaplan_meier_residents_symptomatic.csv", row.names = F)
@@
            color = Event), data = all_km_residents) +
-  # facet_grid(region_group ~ .) +
   geom_ribbon(alpha=0.1, linetype = 2 ) +
   geom_path(lwd=1.2) +
+  geom_text(aes(label = if_else(
+    date == get_monday_date(date),
+    formatbm(round((1-S_t) * desc_residents$CT_total_residents)),
+    ""
+  )), position = position_nudge(y = .02)) +
   scale_y_continuous("Cumulative incidence",
@@
            color = Event), data = all_km_staff) +
-  # facet_grid(region_group ~ .) +
   geom_ribbon(alpha=0.1, linetype = 2,stat = ) +
   geom_path(lwd=1.2) +
+  geom_text(aes(label = if_else(
+    date == get_monday_date(date),
+    formatbm(round((1-S_t) * desc_staff$total_staff)),
+    ""
+  )), position = position_nudge(y = .02)) +
   ggsci::scale_color_lancet()+
@@
   theme_bw() +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
+  theme(legend.position = "none", 
+        axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
   coord_cartesian(ylim = c(0,0.27)) +  ggtitle("(B) Staff")
@@
   km_residents_graph,
-  km_staff_graph
-)
-
-pdf("figure2.pdf", 14,4)
+  km_staff_graph,
+  widths = c(1.4, 1)
+)
+
+pdf("figure2.pdf", 12,4)
 km_ct_plots
@@
   mutate(Age = ageg, Sex = gender) %>% 
-  group_by(Age, Sex) %>% 
+  group_by(Sex, Age) %>% 
   summarise(
@@
     `Deaths in confirmed infections` = deaths_confirmed,
-    `Case-fatality rate (%)` = paste0(cfr, " [", cfr_lower, "; ", cfr_upper, "]"),
-    `Relative risk` = paste0(
-      tbrounding(RR, 2), " [",
-      tbrounding(exp(log(RR) - 1.96 * RR_SE), 2), "; ",
-      tbrounding(exp(log(RR) + 1.96 * RR_SE), 2), "]"
-    )
+    `Case-fatality rate (%)` = paste0(cfr, " [", cfr_lower, "; ", cfr_upper, "]")
+    # `Relative risk` = paste0(
+    #   tbrounding(RR, 2), " [",
+    #   tbrounding(exp(log(RR) - 1.96 * RR_SE), 2), "; ",
+    #   tbrounding(exp(log(RR) + 1.96 * RR_SE), 2), "]"
+    # )
   ) %>% kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
@@
 
-```{r CPH_confirmed,   results="asis"}
+```{r hazard_ratios_confirmed,   results="asis"}
 # cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
@@
 # survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
+
 explanatory = c("gender", "ageg", "home_type", "user_type2",  
@@
 ```{r residents_daily}
-
 residents_daily <- tline %>% 
@@
 
-# residents_daily <- residents_daily %>% 
-#   group_by(date) %>% 
-#   summarise(n = sum(rday),
-#             d = sum(death),
-#             covid_status_outbreak_digit0 = 1 - 
-#               mean(covid_status_outbreak_digit1) -
-#               mean(covid_status_outbreak_digit2) -
-#               mean(covid_status_outbreak_digit3) -
-#               mean(covid_status_outbreak_digit4) -
-#               mean(covid_status_outbreak_digit5),
-#             covid_status_outbreak_digit1 = mean(covid_status_outbreak_digit1),
-#             covid_status_outbreak_digit2 = mean(covid_status_outbreak_digit2),
-#             covid_status_outbreak_digit3 = mean(covid_status_outbreak_digit3),
-#             covid_status_outbreak_digit4 = mean(covid_status_outbreak_digit4),
-#             covid_status_outbreak_digit5 = mean(covid_status_outbreak_digit5)) %>% 
-#   mutate(date_int = as.integer(date - date_min))
-
 ```
@@
 
-```{r risk_death, results="asis"}
+```{r hazard_ratios_death, results="asis"}
 # cph_death2 <- coxph(Surv(date1, date2, death_status) ~
@@
 
-```{r, warning=F, results = "asis"}
-cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-                      gender+ ageg + home_type + covid_status_outbreak +
+```{r attributable_deaths, warning=F, message=F, results = "asis"}
+cph_death3 <- coxph(Surv(date1, date2, death_status) ~
+                    gender+ ageg + home_type + covid_status_outbreak_digit +
                     staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-                      cluster(resident_id), data = residents_episodes_short)
-HRs_covid <- tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak", 
-                                                          names(cph_death2$coefficients))]) %>% 
-  transmute(covid_status_outbreak = gsub("covid_status_outbreak", "", name),
+                    cluster(resident_id),
+                    data =  residents_episodes_short)
+
+HRs_covid <- tibble::enframe(cph_death3$coefficients[grep("covid_status_outbreak", 
+                                                          names(cph_death3$coefficients))]) %>% 
+  transmute(covid_status_outbreak_digit = gsub("covid_status_outbreak_digit", "", name),
             HR = exp(value))
-HRs_CI_covid <- as_tibble(exp(confint(cph_death2))[grep("covid_status_outbreak", 
-                                                          names(cph_death2$coefficients)),]) %>% 
+HRs_CI_covid <- as_tibble(exp(confint(cph_death3))[grep("covid_status_outbreak", 
+                                                          names(cph_death3$coefficients)),]) %>% 
   rename(HR_lower =  `2·5 %`,
          HR_upper = `97·5 %`)
-HRs_CI_covid$covid_status_outbreak <- gsub(
-  "covid_status_outbreak", "",
-  grep("covid_status_outbreak", names(cph_death2$coefficients), value = T)
+HRs_CI_covid$covid_status_outbreak_digit <- gsub(
+  "covid_status_outbreak_digit", "",
+  grep("covid_status_outbreak", names(cph_death3$coefficients), value = T)
 ) 
@@
 attributable_deaths <- residents_episodes_short %>%
-  filter(death_status == 1) %>% 
-  group_by(covid_status_outbreak) %>% 
-  summarise(Deaths = n()) %>% 
+  group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
+  summarise(
+    rdays = sum(as.integer(date2-date1)),
+    deaths = sum(death_status == 1)
+    ) %>% 
   left_join(HRs_covid) %>% 
@@
            HR_lower = tidyr::replace_na(HR_lower, 1),
-           HR_upper = tidyr::replace_na(HR_upper, 1)) %>% 
-    mutate(AH = (HR - 1)/HR,
-           AH_lower = (HR_lower - 1)/HR_lower,
-           AH_upper = (HR_upper - 1)/HR_upper) %>% 
-    mutate(`Attributable deaths` = AH * Deaths) %>% 
-    mutate(`AD (%)` = tbrounding(`Attributable deaths`/ sum(Deaths)*100)) %>% 
-  transmute(`Infection/outbreak status` = covid_status_outbreak,
-            Deaths,
-            HR = paste0(tbrounding(HR,2), " [",
+           HR_upper = tidyr::replace_na(HR_upper, 1))
+
+death_predictions <- 1-predict(cph_death3, residents_episodes_short, type = "survival")
+
+att_fractions <- 0
+for(i in 1:5) {
+  residents_episodes_short_counterfact <- mutate(
+    residents_episodes_short, 
+    # date2 = if_else(death_status == 1 & covid_status_outbreak_digit == as.character(i), 105, date2),
+    # death_status = if_else(death_status == 1 & 
+    #                          covid_status_outbreak_digit == as.character(i),
+    #                        0L, death_status),
+    covid_status_outbreak_digit = case_when(
+      covid_status_outbreak_digit == as.character(i) ~ factor("0", levels = 0:5),
+      TRUE ~ covid_status_outbreak_digit)
+  )
+  death_predictions_counterfactual <- 1 - predict(cph_death3, 
+                                                  residents_episodes_short_counterfact, 
+                                                  type = "survival")
+  att_fractions[i+1] <- sum(death_predictions - death_predictions_counterfactual) / sum(death_predictions)
+}
+
+rm(death_predictions, death_predictions_counterfactual)
+
+attributable_deaths$AF <- att_fractions
+attributable_deaths <- bind_rows(attributable_deaths, 
+                                 colSums(attributable_deaths[,c("rdays", "deaths", "AF")]))
+attributable_deaths %>% ungroup() %>% transmute(
+  `Infection/outbreak status` = tidyr::replace_na(as.character(covid_status_outbreak), "TOTAL"),
+  `Ajusted HR` = paste0(tbrounding(HR,2), " [",
                         tbrounding(HR_lower,2),"; ",
                         tbrounding(HR_upper,2), "]"),
-            `Attibutable hazard (%)`= paste0(tbrounding(AH*100,1), " [",
-                        tbrounding(AH_lower*100,1),"; ",
-                        tbrounding(AH_upper*100,1), "]"),
-           `Attributable deaths` = round(`Attributable deaths`),
-           `AD (%)`)
-attributable_deaths <- bind_rows(
-  attributable_deaths,
-  tibble(
-    `Infection/outbreak status` = "TOTAL",
-            Deaths = (sum(attributable_deaths$Deaths)),
-            HR = "",
-            `Attibutable hazard (%)`= "",
-           `Attributable deaths` = sum(attributable_deaths$`Attributable deaths`),
-           `AD (%)` = tbrounding(
-             sum(attributable_deaths$`Attributable deaths`)/sum(attributable_deaths$Deaths)*100)
-  )
-)
-
-kable(attributable_deaths, align =c("l", "r", "r", "r", "r", "r", "r"))
-```
-
-```{r}
-
-# residents_episodes$resident_id_int <- as.numeric(as.factor(residents_episodes$resident_id))
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type + covid_status_outbreak_digit1 +
-#                       covid_status_outbreak_digit2 + covid_status_outbreak_digit3+
-#                       covid_status_outbreak_digit4 +covid_status_outbreak_digit5 +
-#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       cluster(resident_id_int), data = data.frame(residents_episodes), 
-#                     ties = "breslow")
-# 
-# trucmuche <- survfit(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type + covid_status_outbreak_digit1 +
-#                       covid_status_outbreak_digit2 + covid_status_outbreak_digit3+
-#                       covid_status_outbreak_digit4 +covid_status_outbreak_digit5 +
-#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6,
-#                      residents_episodes_short, id = "resident_id")
-# 
-# AH_fractions <- residents_daily %>% 
-#   tidyr::pivot_longer(cols=starts_with("covid_status_outbreak_")) %>% 
-#   dplyr::rename(prevalence = value) %>% 
-#   left_join(tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak_digit", names(cph_death2$coefficients))])) %>% 
-#   mutate(AH = (prevalence * tidyr::replace_na(exp(value) - 1, 0)) / 
-#            (prevalence * tidyr::replace_na(exp(value) - 1, 0) +1))
-# 
-# 
-# AH_coefs <- tibble::enframe(cph_death2$coefficients[grep("covid_status_outbreak_digit",
-#                                                          names(cph_death2$coefficients))]) %>% 
-#   rename(covid_status_outbreak_digit = name) %>% 
-#   mutate(covid_status_outbreak_digit = gsub("covid_status_outbreak_digit", "", covid_status_outbreak_digit))
-#                               
-# AH_fractions <- residents_daily %>%
-#   group_by(date, covid_status_outbreak_digit) %>%
-#   summarise(n = sum(rday),
-#             d = sum(death)) %>% 
-#   left_join(AH_coefs)  %>% 
-#   filter(date >= as.Date("2020-04-01")) %>% 
-#   mutate(AH =  (tidyr::replace_na(exp(value), 1) -1 ) / tidyr::replace_na(exp(value), 1)) %>% 
-#   group_by(covid_status_outbreak_digit) %>% 
-#   summarise(AH = sum(AH*d),
-#             d = sum(d),
-#             n = sum(n))  
-#             
-# 
-# 
-#  
-# 
-# ggplot(AH_fractions)+
-#   geom_line(aes(x = date, y = AH*d, color = name ))
-#
-
-# 
-# # summary(cph_death2)
-# AF_estimates <- list()
-# for(i in 1:5) {
-#   AF_estimates[[levels(residents_episodes$covid_status_outbreak)[i]]] <- 
-#     AF::AFcoxph(object = cph_death2, data = data.frame(residents_episodes), 
-#             exposure = paste0("covid_status_outbreak_digit", i), clusterid = "resident_id_int")
-#   AF_estimates[[i]]$se <- sqrt(AF_estimates[[i]]$AF.var)
-#   AF_estimates[[i]]$confidence.interval <- AF:::CI.AF(
-#     AF = AF_estimates[[i]]$AF.est, 
-#     Std.Error = AF_estimates[[i]]$se, 
-#     confidence.level = 0.95, CI.transform = "log")
-#   colnames(AF_estimates[[i]]$confidence.interval) <- c("Lower limit", "Upper limit")
-#   AF_estimates[[i]]$covid_status_outbreak <- levels(
-#     residents_episodes$covid_status_outbreak)[i+1]
-#   AF_estimates[[i]] <- bind_cols(
-#     as_tibble(
-#       AF_estimates[[i]][c('times', 'covid_status_outbreak', 'AF.est', 'AF.var', 
-#                          'S.est', 'S0.est', 'S.var', 'S0.var')]
-#     ),
-#     as_tibble(AF_estimates[[i]]$confidence.interval)
-#   )
-# }
-# 
-# 
-# ggplot(bind_rows(AF_estimates[2:5]),
-#        aes(x = times,  color =  covid_status_outbreak )) +
-#   geom_ribbon(aes(ymin = `Lower limit`*100, ymax = `Upper limit`*100), 
-#               alpha=0.05, linetype = 2) +
-#   geom_line(aes(y=AF.est*100), size = 1) +
-#   coord_cartesian(ylim=c(0, 30)) +
-#   ggsci::scale_color_lancet() +
-#   xlab("Excess mortality (%)") +
-#   ggtitle("% excess mortality compared with uninfected in non-outbreak homes")
-# 
-# # 
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                         covid_status_outbreak_digit  +
-#                       cluster(resident_id_int), data = residents_episodes, 
-#                     ties = "breslow")
-# cph_death2 <- flexsurv::flexsurvreg(Surv(date1, date2, death_status) ~
-#                         covid_status_outbreak_digit  +
-#                       cluster(resident_id), 
-#                       data = residents_episodes_short,dist = "weibull")
-# 
-# truc <- survfit(cph_death2, tibble(date1 = 30, covid_status_outbreak_digit = factor(1:5, levels = 1:5) )) 
-#   ggsurvplot(truc,
-#              data = tibble(date1 = 30, covid_status_outbreak_digit = factor(5, levels = 1:5) ))
-#     
+  `Resident-days` = formatbm(rdays),
+  `Total deaths` = formatbm(deaths),
+  `Deaths attributable to Covid-19 (% all-cause deaths)` = paste0(
+    round(AF * 1330), " (",
+    tbrounding(AF*100, 2), ")")
+) %>% kable(align=c("l", rep("r", 4)))
+
 ```

### 2020-07-09 manuscript/manuscript.Rmd
manuscript v0.6

@@
            paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
-         )
+         ) %>% 
+  mutate(staff_to_bed_ratio_g5 = cut(
+    staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
+    quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
+    max(staff_to_bed_ratio))))
 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
+                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
                    beds,  pct_occupancy, no_bedrooms, 
@@
   left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, 
+                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
                    beds, pct_occupancy, no_bedrooms, 
@@
   mutate(date1 = if_else(
-    entry_date > date_min_cox_mortality,
+    entry_date >= date_min_cox_mortality,
     date1,
@@
   name = "tab_CFR_age_sex",
-  caption =  paste0("All-cause case-fatality rates and relative risks by age and sex among private residents (n=", 
+  caption =  paste0("All-cause case-fatality rates by age and sex among private residents (n=", 
   formatbm(dplyr::n_distinct(residents_overview$resident_id)),
@@
   caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
-  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
-  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
+  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_attributable_deaths <- table_nums(
@@
   caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
-  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
-  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
+  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 
@@
 ```{r table_resident_descriptives, message = FALSE, results="asis"}
-residents_overview$outbreak <- relevel(as.factor(dplyr::case_when(
-  residents_overview$home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
+
+residents_overview <- residents_overview %>% 
+  mutate(outbreak = as.integer(home_code %in% desc_residents_outbreak_only$outbreak_homes) ) %>% 
+  dplyr::mutate(
+    covid_status_outbreak = as.factor(case_when(
+    covid_status == "Uninfected" & outbreak == 0 ~ "0 Uninfected (other home)",
+    covid_status == "Symptomatic (not confirmed)" &
+      outbreak==0 ~ "1 Symptomatic not confirmed  (other home)",
+    covid_status == "Uninfected" & outbreak == 1 ~ "2 Uninfected (outbreak home)",
+    covid_status == "Symptomatic (not confirmed)" &
+      outbreak==1 ~ "3 Symptomatic not confirmed  (outbreak home)",
+    covid_status == "Asymptomatic confirmed" ~ "4 Confirmed asymptomatic",
+    covid_status == "Symptomatic confirmed" ~ "5 Confirmed symptomatic"
+  ))) %>% 
+  mutate(outbreak = relevel(as.factor(dplyr::case_when(
+    home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
     TRUE ~ "Other homes"
-)), "Outbreak homes")
+    )), "Outbreak homes"))
 
 arsenal::tableby(
-  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Home type` +  
+  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Funding type` +  
     `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
@@
                 `Admission type` = admission_type2,
-                `Home type` = home_type,
+                `Funding type` = home_type,
                 `Infection status by 14 June` = covid_status, 
@@
 ```{r tab_CFR_age_sex, results="asis"}
-residents_overview %>%
+CFR_age_sex <- residents_overview %>%
   mutate(Age = ageg, Sex = gender) %>% 
@@
     deaths_confirmed = sum(status == "Deceased" & confirmed == 1)) %>% 
+  ungroup()
+
+CFR_age_sex <- bind_rows(
+  CFR_age_sex,
+  colSums(CFR_age_sex[3:7])
+) %>% mutate(
+  Sex = tidyr::replace_na(Sex, "All"),
+  Age = tidyr::replace_na(Sex, "")
+)
+
+CFR_age_sex <- CFR_age_sex %>% 
   mutate(
@@
     `Confirmed infections` = confirmed_infections,
-    `Total deaths` = deaths,
+    `Total deaths` = formatbm(deaths),
     `Deaths in confirmed infections` = deaths_confirmed,
@@
     # )
-  ) %>% kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
+  ) 
+CFR_age_sex %>%
+  kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
 ```
@@
 
-Higher staff to resident ratios were associated with lower risk of infection: a 10 percentage point increase in the number of staff relative to the total number of care home beds was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infection, once adjusting for other factors.
+Higher staff to resident ratios were associated with lower risk of infection: care homes with less than 1 staff per bed had a risk of infection four times higher than those with >=1.4 staff per bed, once adjusting for other factors.
 
@@
 ```{r hazard_ratios_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio_g5+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
 #  summary(cph_conf)
@@
 explanatory = c("gender", "ageg", "home_type", "user_type2",  
-                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6"  )
+                "bed_tot_category", "staff_to_bed_ratio_g5",  "occ_bedroom_ratio_g6")
 dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
@@
   group_by(resident_id) %>% 
-  mutate(death = case_when(
+  mutate(
+    death = case_when(
     status == "Deceased" & date == max(date) ~ 1L,
+    TRUE ~ 0L
+  ),death_Apr = case_when(
+    status == "Deceased" & date == max(date) & date >= as.Date("2020-04-01") ~ 1L,
     TRUE ~ 0L
@@
 #                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       cluster(resident_id), data = residents_episodes_short, ties = "breslow")
+#                       cluster(resident_id), data = residents_episodes)
 # summary(cph_death2)
@@
 explanatory = c("gender", "ageg", "home_type", "user_type2",
-                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
+                "bed_tot_category", "staff_to_bed_ratio_g5", "occ_bedroom_ratio_g6",
                 "covid_status_outbreak",  "cluster(resident_id)")
 dependent = "Surv(time = date1, time2 = date2, event = death_status)"
-CPH_death <-  finalfit(residents_episodes_short, dependent, explanatory, 
+CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
                        column = T, add_dependent_label = F)  
@@
 KM_mortality_by_covid_status <- survfit(
-  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + outbreak, 
-  data = residents_episodes_short)
-KM_mortality_plot <- ggsurvplot_facet(KM_mortality_by_covid_status,
-                 residents_episodes_short,
+  Surv(time = start_int, time2 = stop_int, event = death_status) ~ covid_status + outbreak, 
+  data = mutate(
+    residents_overview, 
+    death_status = as.integer(status == "Deceased")
+  ))
+KM_mortality_plot <- ggsurvplot_facet(
+  KM_mortality_by_covid_status,
+  mutate(residents_overview, 
+         death_status = as.integer(status == "Deceased")  ),
   facet.by  = "covid_status",
@@
   palette = "aaas", conf.int = T,
-                 scales = "free", ggtheme = theme_bw(), xlim = c(32, 104),
+  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), ylim = c(.4, 1),
   xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
   conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)
-pdf("figure5.pdf",8, 6)
+
+KM_mortality_plot <- KM_mortality_plot +  
+  scale_x_continuous(name = "Date", 
+                     breaks = c(seq(0, 7*50, 7)),
+                     labels = format.Date(date_min + seq(0, 7*50, 7), "%e %b %Y" )) +
+  theme(axis.text.x = element_text(size = 8, angle = 35, vjust = 1, hjust=1)) 
+pdf("figure3.pdf", 8, 6)
 KM_mortality_plot
@@
 ```
-
-*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevant Datix incident report.
 
@@
                     staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-                    cluster(resident_id),
-                    data =  residents_episodes_short)
+                    cluster(resident_id), data =  residents_episodes)
 
@@
 
-attributable_deaths <- residents_episodes_short %>%
+attributable_deaths <- residents_episodes %>%
   group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
@@
     rdays = sum(as.integer(date2-date1)),
-    deaths = sum(death_status == 1)
+    deaths = sum(death_status == 1),
+    deaths_apr = sum(death_status == 1 & date2 >= 30)
     ) %>% 
@@
 
-death_predictions <- 1-predict(cph_death3, residents_episodes_short, type = "survival")
-
+# All deaths
+
+death_predictions <- 1-predict(cph_death3, residents_episodes, type = "survival")
 att_fractions <- 0
 for(i in 1:5) {
-  residents_episodes_short_counterfact <- mutate(
-    residents_episodes_short, 
-    # date2 = if_else(death_status == 1 & covid_status_outbreak_digit == as.character(i), 105, date2),
-    # death_status = if_else(death_status == 1 & 
-    #                          covid_status_outbreak_digit == as.character(i),
-    #                        0L, death_status),
+  residents_episodes_counterfact <- mutate(
+    residents_episodes, 
     covid_status_outbreak_digit = case_when(
@@
   death_predictions_counterfactual <- 1 - predict(cph_death3, 
-                                                  residents_episodes_short_counterfact, 
+                                                  residents_episodes_counterfact, 
                                                   type = "survival")
@@
 attributable_deaths$AF <- att_fractions
+
+# Just for April-June
+death_predictions <- 1-predict(cph_death3, residents_episodes_short, type = "survival")
+att_fractions <- 0
+for(i in 1:5) {
+  residents_episodes_counterfact <- mutate(
+    residents_episodes_short, 
+    covid_status_outbreak_digit = case_when(
+      covid_status_outbreak_digit == as.character(i) ~ factor("0", levels = 0:5),
+      TRUE ~ covid_status_outbreak_digit)
+  )
+  death_predictions_counterfactual <- 1 - predict(cph_death3, 
+                                                  residents_episodes_counterfact, 
+                                                  type = "survival")
+  att_fractions[i+1] <- sum(death_predictions - death_predictions_counterfactual) / sum(death_predictions)
+}
+rm(death_predictions, death_predictions_counterfactual)
+
+attributable_deaths$AF_apr <- att_fractions
+
 attributable_deaths <- bind_rows(attributable_deaths, 
-                                 colSums(attributable_deaths[,c("rdays", "deaths", "AF")]))
+                                 colSums(attributable_deaths[,c("rdays", "deaths", "deaths_apr",
+                                                                "AF", "AF_apr")]))
 attributable_deaths %>% ungroup() %>% transmute(
   `Infection/outbreak status` = tidyr::replace_na(as.character(covid_status_outbreak), "TOTAL"),
-  `Ajusted HR` = paste0(tbrounding(HR,2), " [",
+  `Adjusted HR` = paste0(tbrounding(HR,2), " [",
                         tbrounding(HR_lower,2),"; ",
@@
   `Total deaths` = formatbm(deaths),
+  `Total deaths (April-June)` = formatbm(deaths_apr),
   `Deaths attributable to Covid-19 (% all-cause deaths)` = paste0(
-    round(AF * 1330), " (",
-    tbrounding(AF*100, 2), ")")
+    round(AF * sum(deaths/2)), " (",
+    tbrounding(AF*100, 2), ")"),
+  `Deaths attributable to Covid-19 (% all-cause deaths April-June)` = paste0(
+    round(AF_apr * sum(deaths_apr/2)), " (",
+    tbrounding(AF_apr*100, 2), ")")
 ) %>% kable(align=c("l", rep("r", 4)))

### 2020-07-10 manuscript/manuscript.Rmd
edits

@@
 ---
+
+Peter Dutey Magni,^{1} Haydn Williams,^{2} Arnoupe Jhass,^{1} Greta Rait, Harry Hemingway,^{134} Andrew Hayward,^{1} Laura Shallcross^{1}
+
+1 Institute of Health Informatics, University College London, London, UK
+2 Four Seasons Health Care Group
+3 Health Data Research UK, University College London, London, UK
+4 University College London Hospitals NIHR Biomedical Research Centre, London, UK
+5 Institute of Epidemiology & Health Care, University College London, London, UK
 
@@
 At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic.
-Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures that prevent infection may have a negative impact on residents by disrupting usual care.
+Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures[@Hale2020] that prevent infection may have a negative impact on residents by disrupting usual care.
 
@@
 
-Data were analysed in `R`3·5·0[@RCoreTeam2018] using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries.
+Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries.
 All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
@@
   palette = "aaas", conf.int = T,
-  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), ylim = c(.4, 1),
+  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), 
   xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
@@
 
-This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].
+This work was supported by the Economic and Social Research Council [ES/V003887/1].
+This work was also supported by a National Institute for Health Research (NIHR) Clinician Scientist award (CS-2016-007 to L.S.); In-Practice fellowship (NIHR300293 to A.J.); a NIHR School of Primary Care Research fellowship (SPCR-118 to A.J.); and a NIHR In-Practice Fellowship (to M.R.).
+AH and HH are NIHR Senior Investigators .
+AH and HH are supported by Health Data Research UK (grant No.
+LOND1), which is funded by the UK Medical Research Council, Engineering and Physical Sciences Research Council, Economic and Social Research Council, Department of Health and Social Care (England), Chief Scientist Office of the Scottish Government Health and Social Care Directorates, Health and Social Care Research and Development Division (Welsh Government), Public Health Agency (Northern Ireland), British Heart Foundation, Wellcome Trust.
+HH is funded by the NIHR University College London Hospitals Biomedical Research Centre, The BigData@Heart Consortium, funded by the Innovative Medicines Initiative-2 Joint Undertaking under grant agreement No.
+116074.
 

### 2020-07-12 manuscript/manuscript.Rmd
manuscript v0.9

@@
 bibliography: references.bib
-csl: elsevier-vancouver.csl
+csl: sage-vancouver.csl
 output:
+  word_document:
+    toc: yes
+    reference_docx: template.docx
   html_document:
@@
 
-Peter Dutey Magni,^{1} Haydn Williams,^{2} Arnoupe Jhass,^{1} Greta Rait, Harry Hemingway,^{134} Andrew Hayward,^{1} Laura Shallcross^{1}
-
-1 Institute of Health Informatics, University College London, London, UK
-2 Four Seasons Health Care Group
-3 Health Data Research UK, University College London, London, UK
-4 University College London Hospitals NIHR Biomedical Research Centre, London, UK
-5 Institute of Epidemiology & Health Care, University College London, London, UK
+Peter Dutey Magni,$^{12}$ Haydn Williams,$^{3}$ Arnoupe Jhass,$^{4}$ Greta Rait,$^4$ Harry Hemingway,$^{125}$ Andrew Hayward,$^{6}$ Laura Shallcross$^{12*}$
+
+$^1$ Institute of Health Informatics, University College London, London, UK
+
+$^2$ Health Data Research UK, University College London, London, UK
+
+$^3$ Four Seasons Health Care Group, Norcliffe House, Station Road, Wilmslow, Cheshire, UK
+
+$^4$ Primary Care & Population Health, University College London, London, UK
+
+$^5$ University College London Hospitals NIHR Biomedical Research Centre, London, UK
+
+$^6$ Institute of Epidemiology & Health Care, University College London, London, UK
+
+$*$ Corresponding author
 
@@
   name = "incidence_CT_residents",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 supp_epi_CT_oubreak_only <- table_nums(
   name = "incidence_CT_staff",
-  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_resident_descriptives <- table_nums(
@@
 
-**Background** Covid-19 has led to a 45% increase in mortality in care home residents in England and Wales, but the proportion of deaths directly attributable to Covid-19 is uncertain due to limited testing for infection, and lack of syndromic surveillance.
-
-**Methods:**  Retrospective cohort study using individual-level electronic records from residents and daily care home counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` care homes from the Four Seasons Health Care Group (FSHCG, `r format(date_min, "%e %B %Y")`-`r format(date_max, "%e %B %Y")`).
-We estimated the incidence of symptoms, confirmed infections, and deaths in residents and staff and modelled risk factors for infection and mortality in residents using Cox proportional hazards.
-
-**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had confirmed infections.
-The incidence of confirmed infection in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days.
-`r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) had neither confirmed infections nor Covid-related deaths.
-
-Among `r formatbm(desc_residents$DAT_total_residents)` private residents, there were `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]), `r desc_residents$DAT_CFR$y` of which occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%].
-`r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`%; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`%] of these deaths were attributable to SARS-CoV-2.
-Symptomatic and asymptomatic infection were independent risk factors for mortality.
-All-cause mortality for uninfected residents was 93% higher in care homes with outbreaks compared to those without outbreaks (adjusted HR = 1·9 [1·7; 2·2]).
-
-**Interpretation:** Covid-19 infection directly increases residents’ risk of death, but in outbreak care homes mortality is also elevated in individuals with no record of infection compared to homes without outbreaks.
-Preparations for future waves of the pandemic should prioritise regular testing in care homes to address case under-ascertainment, and consider the indirect effect of Covid-19 on mortality.
-
-**Funding:** This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].
+**Background** Rates of Covid-19 infection have declined in many countries, but outbreaks persist in care home residents who are at high risk of severe outcomes.
+Epidemiological data from care homes are scarce.
+We used population-level active surveillance to estimate incidence of, and risk factors for Covid-19, and attributable mortality in care home residents.
+
+**Methods:** Cohort study using individual-level electronic health records from 8,713 residents and daily counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` UK care homes.
+We modelled risk factors for infection and mortality using Cox proportional hazards and estimated attributable fractions.
+
+**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections.
+Confirmed infection incidence in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one Covid-19 infection or death.
+Lower staffing ratios and higher occupancy rates were independent risk factors for infection.
+
+`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]) residents.
+`r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
+Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1·9 [1·7; 2·2]).
+33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.
+
+**Interpretation:** 1 in 5 residents had symptoms of infection during the pandemic, but many cases were not tested.
+Higher occupancy and lower staffing levels increase infection risk.
+Disease control measures should integrate active surveillance and testing with fundamental changes in staffing and care home occupancy to protect staff and residents from infection.
+
+**Funding:** Economic and Social Research Council [ES/V003887/1].
 
@@
 
-Care home residents are particularly vulnerable to Covid-19 infection due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and this is compounded by frequent exposure to infection through close contact with infected staff members, other residents and contaminated surfaces in the care home setting.
-In the UK, an estimated 420,000 residents lived in approximately 11,000 homes for the elderly.[@cqcdata2020]
-
-At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic.
-Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures[@Hale2020] that prevent infection may have a negative impact on residents by disrupting usual care.
-
-Disentangling direct and indirect deaths from Covid-19 requires accurate information on the incidence and prevalence of infection among staff and residents linked to individual-level clinical outcomes.
-There is no syndromic surveillance for infection in care homes in England, and universal testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established in care homes in England until 11th May 2020, when the government launched the whole care home testing programme.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations with a maximum of five tests per care home.
-Consequently national estimates of incidence and prevalence based on test results obtained prior to 11th May will substantially underestimate the burden of infection in care home residents and staff.
-
-Outbreak investigations provide estimates of the proportion of residents and staff who become infected during an outbreak and case-fatality rates,[@McMichael2020; @Dora2020] but they provide limited insight into the relationship between infection and mortality across the entire pandemic period, and cannot be generalised to care homes that have not reported any outbreaks.
-A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long-term care facilities.
-Mortality rates in residents with confirmed Covid-19 infection ranged from 0 to 10%, with case fatality rates between 0 and 34%, however this review did not include estimates of indirect mortality.
-
-We used administrative data from the Four Seasons Health Care Group (FSHCG), one of the UK’s largest independent provider of long-term and respite residential and nursing care, to establish syndromic surveillance for Covid-19 across `r formatbm(dplyr::n_distinct(residents$home_code))` FSHCG homes.
-We used this system to estimate the incidence and prevalence of suspected and confirmed Covid-19 across a national network of care homes, and to investigate the fraction of all-cause deaths attributable to confirmed infections.
+Globally the number of cases of Covid-19 continues to increase, but cases in Europe have declined since April 2020,[@ECDC2020] following the introduction of lockdown measures.[@Hale2020] Although the incidence of infection in the general population in England is low (0.04%),[@cisbulletin2020] new infections persist, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] This raises the possibility that these settings represent a reservoir for transmission of infection back to the community.
+
+In the UK, there are an estimated 400,000 residents living in approximately 11,000 homes for the elderly.[@LaingBuisson2019] Care home residents are particularly vulnerable to Covid-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care home setting.
+At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Care home staff also have higher aged-standardised rates of Covid-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved Covid-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
+Understanding the proportion of excess deaths that can be directly and indirectly attributed to Covid-19 infection is important, to fully assess the impact of the pandemic on care homes.
+
+The development of public health strategies to protect the public, residents and staff from Covid-19 requires knowledge of the burden of and risk factors for infection in care home residents and staff, linked to outcomes.
+However, there is no syndromic surveillance for infection in care homes in England, and widespread one-off testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for care home staff and residents until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England’s (PHE) outbreak investigations which permitted a maximum of five tests per care home.
+Consequently national estimates of incidence and prevalence will substantially underestimate the burden of infection in care home residents and staff.
+
+In the absence of cohort studies or active surveillance, outbreak investigations provide the most reliable estimates of the burden of infection and case-fatality.[@McMichael2020; @Dora2020] An estimated 44% of English care homes have had at least one outbreak, with a living systematic review[@Salcher-Konrad2020] reporting substantial variation in cumulative incidence of infection (0%-72%) and case fatality (0-34%) in care home residents.
+A major limitation of outbreak investigations is that follow-up is usually less than 30 days,[@Salcher-Konrad2020] and investigations are only conducted in settings in which outbreaks have been detected.
+Understanding the proportion of care homes with undetected cases is crucial for policy decisions around the frequency of and justification for regular testing in care homes.
+
+To our knowledge, there are no studies which have employed population-level active surveillance in care homes throughout this pandemic to measure outcomes of both suspected and confirmed infection in residents and staff.
+We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest independent provider of residential and nursing care, with the aim of identifying strategies to protect care home staff and residents from future waves of infection.
+Our objectives were to estimate incidence of and risk factors for infection and mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
+We also estimated mortality attributable to Covid-19.
 
@@
 
-In 2020, the FSHCG included 9,568 residential or nursing care beds for elderly residents, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 3).
-Approximately 9% of these beds were contracted to local authorities, with the remainder for privately funded residents.
-
-FSHCG collects administrative data on their private residents through a variety of electronic systems that are primarily used for billing purposes.
-This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as “Datix”.
-Such records do not include residents occupying beds contracted by local authorities, but home-level aggregate information was nevertheless available on all residents (`r fig_nums("fig_study_diagram", display = "cite")`).
+## Study population and setting
+
+Staff or residents living and working in care homes for the elderly between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`, which were run by the Four Seasons Healthcare Group (FSHCG) were eligible for study inclusion.
+The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is primarily state funded.
+Most residents are permanent, but a small proportion receive temporary (respite) care.
+
+In 2020, there were 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 1).
+90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for Covid-19 at least once between 11 May and 22 June 2020.
+
+## Data sources
+
+Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in research.[@Smith2020]
+
+*Individual-level data*
+
+The care home chain collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
+The dataset includes: dates of care home entry and exit, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).
+Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, care home identifier, incident date/time, date of birth, sex, Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death.
+Information in Datix was used categorise residents’ infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1).
+The term 'confirmed' denoted a positive PCR test.
+Datix was also used to differentiate in-hospital and in care home deaths, and to identify Covid-19 related deaths.
+Individual level data on residents was linked to Datix reports by creating an index file based on a combination of forename, surname and the care home identifier (supplementary material 2).
+1492/1880 (79%) of Datix reports were successfully linked.
+Individual-level records were available between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`.
+
+*Aggregate data*
+
+On 24 March the care home chain introduced a new system to report Covid-19 cases and deaths.
+This required care home managers to report daily tallies in residents (new symptomatic cases, new confirmed infection in home, new confirmed infection in hospital, deaths related to Covid-19) and staff (symptomatic cases, new confirmed cases).
+Data on staff deaths were not extracted due to the small number of cases.
+Covid-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to Covid-19 by the coroner.
+The number of occupied beds in each care home was reported weekly.
+
+Care home characteristics (number of beds, region, nursing versus residential care) were collected from organisational data.
+We therefore extracted organisational data, individual-level data for 8713 residents and aggregate data for all care home staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).
 
@@
 
-## Aggregate outbreak surveillance counts (`r format(date_min_tallies, "%e %b %Y")`--`r format(date_max_tallies, "%e %b %Y")`)
-
-On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor Covid-19 cases in each care home for all care home residents and staff (including residents in contract beds).
-This new system requires care homes to report daily tallies of new cases identified among staff and residents to their FSHCG regional manager, and was used to report Covid-19 deaths to the Care Quality Commission (CQC).[@CQC2020stats] Daily counts were obtained for every care home:
-
-- new symptomatic cases (residents)
-- newly confirmed infection in home (residents)
-- newly confirmed infection in hospital (residents)
-- deaths related to Covid-19 (residents previously confirmed or with a death attributed to SARS-CoV-2 by coroner)
-- symptomatic cases (staff)
-- newly confirmed cases (staff)
-
-In addition, the number of beds currently occupied was reported every Sunday.
-This weekly information was used to approximate daily occupancy using linear interpolation.
-Aggregate data were only extracted for residents in care homes for the elderly.
-
-Care home characteristics were inferred from organisational data.
-Baseline care home occupancy was computed by averaging weekly occupancy before the first Covid-19 cases (between 1 January and 1 March 2020).
-This was used to calculate a ratio of the baseline occupancy to the number of bedroom ratio.
-We also estimated a ratio of the headcount of staff employed to the number of beds.
-
-## Individual-level residents records (`r format(date_min, "%e %b %Y")`--`r format(date_max, "%e %b %Y")`)
-
-Secondary use individual-level data for private residents receiving general and elderly care were extracted and pseudonymised by HW. They consisted of two datasets:
-
-- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly).
-This dataset does not include any records for `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))` occupants of local authority 'block contract' beds.
-- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff.
-Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting.
-In addition, reports could record: date of birth, gender, information on Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death.
-Incidents were classified to indicate whether the resident was symptomatic, and whether an infection was confirmed on the basis of multiple variables (supplementary material 1).
-
-HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier.
-Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents' dataset.
-Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) incident records did not contain enough information to generate an identifier and were excluded.
-Identifiers of a further `r desc_before_exclusions$link_fail` records (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident.
-A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents' dataset.
-
-## Analysis
-
-Incidence proportions and rates were calculated using the aggregate outbreak counts, as these were the trusted source of information used for reporting deaths to the CQC and encompassed all residents.
-The number of unique residents present in the homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")` was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data).
-This approach assumes that the mean length of stay of local authority residents is identical to that of private residents (supplementary material 1).
-Aggregate outbreak counts were also used to estimate incidence and cumulative incidence across time using a life table approach.
-Daily occupancy was inferred from weekly occupied bed censuses using linear interpolation.
-As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents (supplementary material 1).
-A Kaplan-Meier product limit estimator was applied to this life table to estimate the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths every day between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between `r format(england_results$date_min, "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
-
-Individual-level data (including dates of care home entry and exit) were available for private residents only.
-By assuming infection underreporting to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and mortality.
-A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, occupant-to-bedroom ratio).
-
-).
-We also compared care ratesDates of infectiondischarge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality in care homes withbased on the presence of symptoms and without outbreaks, defined as at least one /or confirmed case or Covid-19 related death during the study period.
-infection.
-A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed).
-We also compared rates of infection and mortality in care homes with and without outbreaks, defined as at least one confirmed infection or Covid-19 related death during the study period.
-A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (uninfected; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed) as well as the time-dependent home outbreak status (no Covid-related death or confirmed infection; at least one Covid-related death or confirmed infection).
-The associated hazard ratios enable us to compare the mortality in each group as compared to the reference category (uninfected in non-outbreak care homes).
-The corresponding attributable fraction is obtained by using the model to predict the counterfactual mortality, then computing AFW, the attributable fraction within study:[@Samuelsen2008]
-
-$$ AFW = \dfrac{\sum_{i=1}^{n}(\hat{F}_i - \hat{F}_i^\star)}{\sum_{i=1}^{n}\hat{F}_i} $$
-
-where $\hat{F}_i$ designates the predicted probability of death of individual $i = \{1, \dots, n\}$ based on the observed infection/outbreak classification, and  $\hat{F}_i^\star$ the same probability had the individual been uninfected and living in a non-outbreak care home.
-Observed censoring times are used.
-
-Bedroom occupancy and staffing ratios were estimated on the basis of baseline occupancy averaged before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality.
-Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and/or confirmed infection.
-Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab (PCR) testing across the study period following the introduction of whole care home testing on 11 May 2020.[@dhscTesting2020].
-By 22 June 2020, 90% of FSHCG care homes had participated in the whole home testing programme.
-
-Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries.
-All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
+## Risk factors for infection
+
+Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, care home occupancy and staff to bed ratio).
+Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first Covid-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms.
+We also estimated the ratio of staff to beds.
+A care home outbreak was defined as at least one confirmed infection or Covid-19 related death.
+
+## Statistical analysis
+
+*Infection in care home staff and residents*
+
+Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies.
+These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff.[@CQC2020stats] Infection incidence was also estimated from Datix, but was subject to under-reporting (supplementary material 2).
+In order to calculate infection and death incidence, we estimated the total number of residents in each care home by extrapolating estimates of care home occupancy from the individual-level dataset (because dates of care home entry and exit were not available for 855 beds, supplementary material 1).
+Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation.
+The total number of residents at risk of infection was unknown, so it was approximated in a multiple decrements life table (supplementary material 1).
+The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths by day.
+The rate ratio for care home versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
+
+*Mortality, attributable mortality and risk factors*
+
+Individual-level data were used to estimate rates of infection, all-cause mortality and case-fatality by age and gender in residents.
+Aggregate data were also used to estimate the crude rate of Covid-19 related mortality.
+Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and confirmed infection.
+In order to investigate the relationship between Covid-19 infection and excess mortality, we assumed that residents in non-outbreak care homes had not been exposed to infection, and would therefore not experience excess Covid-19 related mortality.
+We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
+
+A Cox proportional hazards regression model tested the effect of individual and care home level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.
+We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes.
+This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
+
+Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]
 
@@
 
-The study population consisted of an estimated `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland.
-Two thirds (`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes`) of homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, reported at least one confirmed infection or Covid-related death, either through Datix or outbreak surveillance counts.
-
-Between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`, `r filter(new_cases, new_symptomatic_home>0 | new_symptomatic_hospital>0) %>% summarise(n=n_distinct(home_code)) %>% .$n`/`r desc_residents$total_homes` care homes reported `r formatbm(sum(new_cases$new_symptomatic_home, na.rm = T))` residents who developed symptoms in the care home and `r formatbm(sum(new_cases$new_symptomatic_hospital, na.rm = T))` residents who developed symptoms following readmission from hospital, contributing to an overall incidence proportion of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
-Infection was confirmed in `r desc_residents$CT_total_confirmed` residents, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
+The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one Covid-19 outbreak (including unconfirmed outbreaks) in either the individual-level or aggregate datasets.
+The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset.
+Mean and median duration of resident follow-up was `r round(mean(residents_overview$resident_days))` and `r median(residents_overview$resident_days)` days respectively in the individual-level dataset.
+
+Symptoms of infection were recorded in  `r formatbm(desc_residents$CT_total_symptomatic)` residents based on the aggregate dataset, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
 The confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
 
-Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown on `r fig_nums("fig_KM_CT_plots", display = "cite")`.
-The sharp increase in confirmed infections between 30 April and 2 May (67 new cases) can largely be attributed to the piloting of whole-home testing in one care home, leading to 42 cases being confirmed.
-
-In England only, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$cis_confirmed` confirmed cases out of `r formatbm(england_results$cis_person_days)` person-days during this period.
-This implies a confirmed infection rate ratio of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
+Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown in `r fig_nums("fig_KM_CT_plots", display = "cite")`.
+
+In England, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
+In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$cis_confirmed` confirmed cases out of `r formatbm(england_results$cis_person_days)` person-days during the same period.
+This implies a confirmed infection rate ratio comparing care homes to the community of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
 
@@
 
+*Note*: underlying data available from supplementary material 3.
+
 `r fig_nums("fig_KM_CT_plots")`
 
-Symptoms of infection were reported by `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)`  staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+`r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) reported symptoms of infection during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
 
@@
 |	Total exposure (days)| `r formatbm(desc_staff$CT_symptomatic_denom)`	|`r formatbm(desc_staff$CT_confirmed_denom)`|
-|	Incidence proportion (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
+|	Cumulative incidence (%)  | 	`r tbrounding(desc_staff$CT_symptomatic_p)` [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`]	| `r tbrounding(desc_staff$CT_confirmed_p)` [`r tbrounding(desc_staff$CT_confirmed_p_lower)`; `r tbrounding(desc_staff$CT_confirmed_p_upper)`] |
 |	Incidence rate (per 100,000 person-days) | `r tbrounding(desc_staff$CT_symptomatic_IR)` [`r tbrounding(desc_staff$CT_symptomatic_IR_lower)`; `r tbrounding(desc_staff$CT_symptomatic_IR_upper)`]| `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] |
 
-Incidence estimates for `r formatbm(desc_residents$DAT_total_residents)` private residents derived from Datix incident records are reported in supplementary material 2.
-Lower estimates were obtained which is likely due to reflect incomplete recording of cases in Datix, and the inability to link all incident records to an individual.
+Estimates of incidence for private residents derived from Datix are reported in supplementary material 2.
 
@@
 
-`r desc_residents$CT_total_cdeath` Covid-related resident deaths were reported across all care homes, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
-`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts took place in hospital.
-
-Individual-level data were available for a subset of `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents, the vast majority (85·4%) of whom were permanent residents.
-`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of private residents were in nursing homes, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
-The crude incidence of all-cause death in private residents was `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
-All cause mortality in residents with confirmed infection was `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%].
-This is to be contrasted with a mortality of `r tbrounding(desc_residents$DAT_BFR_rate)`% [`r tbrounding(desc_residents$DAT_BFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_BFR_rate_upper)`%] in private residents that did not test positive for Covid-19 (RR = `r tbrounding(desc_residents$DAT_CFR_RR)` [`r tbrounding(desc_residents$DAT_CFR_RR_lower) `; `r tbrounding(desc_residents$DAT_CFR_RR_upper)`]).
-The excess death fraction attributable to confirmed infection is thus `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)` [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`].
+`r desc_residents$CT_total_cdeath` Covid-related resident deaths were reported in the aggregate dataset, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
+`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).
+
+Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.
+`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in these residents, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
+The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22·6% versus 11·2%).
+
+`r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (`r table_nums("tab_CFR_age_sex", display = "cite")`).
+The case-fatality rate increased with age and was higher in men compared to women.
 
@@
 ```
-
-To investigate whether increased mortality was only seen in individuals who were infected with Covid-19, we compared mortality in residents with no record of Covid-19 from care homes with outbreaks and those without.
-`r formatbm(desc_residents$DAT_BFR_outbreak_homes$y)`/`r formatbm(desc_residents$DAT_BFR_outbreak_homes$n)` died in outbreak homes, compared with `r formatbm(desc_residents$DAT_BFR_other_homes$y)`/`r formatbm(desc_residents$DAT_BFR_other_homes$n)` in non-outbreak homes.
-The risk of all-cause death was therefore twice as large in outbreak homes for uninfected residents (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) - 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR) + 1.96 * desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
-This implies that `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR-1)/desc_residents$DAT_BFR_outbreak_homes$RR*100)`% [`r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_lower-1)/desc_residents$DAT_BFR_outbreak_homes$RR_lower*100)`%; `r tbrounding((desc_residents$DAT_BFR_outbreak_homes$RR_upper-1)/desc_residents$DAT_BFR_outbreak_homes$RR_upper*100)`%] of deaths in residents who did not test positive for infection in outbreak homes were in some way attributable to Covid-19.
 
@@
   Sex = tidyr::replace_na(Sex, "All"),
-  Age = tidyr::replace_na(Sex, "")
+  Age = tidyr::replace_na(Age, "")
 )
@@
 
-Using individual-level data on private bed residents (`r table_nums("tab_resident_descriptives", display = "cite")`), factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
-Male gender, age $\geq$ 85 years (adjusted HRs $\geq$ 1·4) and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
+Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
 Large care homes had greater rates of infection (adjusted HR = 1.8 [1·4; 2·4] for care homes with $\geq$ 70 beds versus <35 beds).
@@
 
-Higher staff to resident ratios were associated with lower risk of infection: care homes with less than 1 staff per bed had a risk of infection four times higher than those with >=1.4 staff per bed, once adjusting for other factors.
+Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1·23 [1·17; 1·31]).
 
@@
 ```{r hazard_ratios_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio_g5+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
+# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+
+#                     home_type+user_type2+  bed_tot_category + bed_to_staff_ratio +occ_bedroom_ratio_g6, data = residents_overview)
 #  summary(cph_conf)
@@
 # survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
-
+residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
 explanatory = c("gender", "ageg", "home_type", "user_type2",  
-                "bed_tot_category", "staff_to_bed_ratio_g5",  "occ_bedroom_ratio_g6")
+                "bed_tot_category", "bed_to_staff_ratio",  "occ_bedroom_ratio_g6")
 dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
@@
                            explanatory, column = T,
-                           add_dependent_label = F)  
+                           add_dependent_label = F)  %>%     ff_remove_p() 
 knitr::kable(CPH_confirmed,
@@
 
+```{r tablefiller1, results="asis"}
+arsenal::tableby(
+  confirmed  ~ gender + ageg  +  home_type + user_type2  + staff_to_bed_ratio_g5 +  occ_bedroom_ratio_g6, 
+  control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
+  data = residents_overview) %>%  summary()
+
+# residents_overview %>% 
+#   mutate(confirmed = as.factor(confirmed)) %>% 
+#   summary_factorlist("confirmed", explanatory) %>% kable
+```
+
 ## Factors associated with all-cause mortality
 
-The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status and mortality.
-After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·5 [1·3; 1·6]), and receiving nursing rather than residential care (adjusted HR = 1·5 [1·3; 1.7]).
-All-cause death was strongly associated with asymptomatic confirmed infection (adjusted HR = 3·3 [1·9; 5·6]) and symptomatic confirmed infection (adjusted HR = 12 [10; 15]), compared to residents with no evidence of Covid-19 infection.
-
-We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection in non-outbreak homes as the reference group.
-Risk of all-cause mortality was almost two-fold higher in residents with no evidence of infection in outbreak versus non-outbreak homes (adjusted HR = 1·9 [1·7; 2·2]) in outbreak homes than in other homes.
-Risk of death was also higher in individuals with symptoms but unconfirmed infection in outbreak homes compared to non-outbreak homes (adjusted HR = 4·4 [3·0; 6·2] versus 9·3 [7·5; 12]).
+The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality.
+After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·4 [1·3; 1·6]), and receiving nursing care (adjusted HR = 1·4 [1·2; 1.5]).
+
+We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection (group A) in non-outbreak homes as the reference group.
+Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2·0 [1·7; 2·2]).
+Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4·3 [3·0; 6·2] versus 9·4 [7·6; 12]).
+All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3·3 [2·0; 5·7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.
 
 It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
-
-After adjusting for other variables, an increase by 10 percentage points in staffing was associated with a 3·1% reduction [0·6%; 5·7%] in mortality after adjustment for other  variables.
 
@@
 ```{r hazard_ratios_death, results="asis"}
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
-#                       gender+ ageg + home_type + covid_status_outbreak +
-#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-#                       cluster(resident_id), data = residents_episodes)
+# cph_death2 <- coxph(Surv(date1, date2, death_status) ~ gender+ ageg + home_type + user_type2 + 
+#                       bed_tot_category + bed_to_staff_ratio + occ_bedroom_ratio_g6 + 
+#                       covid_status_outbreak  + cluster(resident_id), data = residents_episodes)
 # summary(cph_death2)
 # cox.zph(cph_death2)
-
+residents_episodes$bed_to_staff_ratio <- 1/residents_episodes$staff_to_bed_ratio
 explanatory = c("gender", "ageg", "home_type", "user_type2",
-                "bed_tot_category", "staff_to_bed_ratio_g5", "occ_bedroom_ratio_g6",
+                "bed_tot_category", "bed_to_staff_ratio", "occ_bedroom_ratio_g6",
                 "covid_status_outbreak",  "cluster(resident_id)")
@@
 CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
-                       column = T, add_dependent_label = F)  
+                       column = T, add_dependent_label = F)  %>%   ff_remove_p() 
 knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
@@
 *Note:* Baseline group = uninfected residents in non-outbreak care home
+
+```{r tablefiller2, results="asis"}
+residents_overview$death_status <- as.integer(residents_overview$status == "Deceased")
+arsenal::tableby(
+  death_status  ~ gender + ageg  +  home_type + user_type2  + staff_to_bed_ratio_g5 +  occ_bedroom_ratio_g6, 
+  control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
+  data = residents_overview) %>%  summary()
+
+# residents_overview %>% 
+#   mutate(confirmed = as.factor(confirmed)) %>% 
+#   summary_factorlist("confirmed", explanatory) %>% kable
+```
 
@@
 
+### Attributable mortality
+
+Model-based estimates of attributable mortality were derived from the individual-level data.
+Overall, 567/1,694 (33%) deaths were attributed to Covid-19.
+In homes with outbreaks only 28% (159 residents) of the mortality attributable to COVID-19 occurred in people with confirmed infection (Groups C and D), (`r table_nums("tab_attributable_deaths", display = "cite")`).
+Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41·7%).
+Model-based estimates of deaths based on individual-level were slightly higher (8%) than counts from the aggregate data.
+
 `r table_nums("tab_attributable_deaths")`
@@
   `Total deaths` = formatbm(deaths),
-  `Total deaths (April-June)` = formatbm(deaths_apr),
+  # `Total deaths (April-June)` = formatbm(deaths_apr),
   `Deaths attributable to Covid-19 (% all-cause deaths)` = paste0(
@@
     tbrounding(AF*100, 2), ")"),
-  `Deaths attributable to Covid-19 (% all-cause deaths April-June)` = paste0(
-    round(AF_apr * sum(deaths_apr/2)), " (",
-    tbrounding(AF_apr*100, 2), ")")
+  # `Deaths attributable to Covid-19 (% all-cause deaths April-June)` = paste0(
+  #   round(AF_apr * sum(deaths_apr/2)), " (",
+  #   tbrounding(AF_apr*100, 2), ")")
 ) %>% kable(align=c("l", rep("r", 4)))
@@
 
-This study is to our knowledge the first large scale investigation of the incidence of SARS-CoV-2 in UK care homes during the pandemic period.
-Findings confirm a very rapid progression of the pandemic across care home residents, with up to `r  tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%] of residents experiencing symptoms and `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%] receiving a positive nasopharyngeal antigen test result between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`.
-The case-fatality rate was `r tbrounding(desc_residents$DAT_CFR_rate)`%, `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% of which was attributable to Covid-19.
-Similar excess mortality is noticeable in residents with no record of infection in outbreak care homes, compared to those living in homes without outbreak (RR = `r tbrounding(desc_residents$DAT_BFR_outbreak_homes$RR)` [`r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`; `r tbrounding(exp(log(desc_residents$DAT_BFR_outbreak_homes$RR)-1.96*desc_residents$DAT_BFR_outbreak_homes$RR_SE))`]).
-Whilst substantial case under-ascertainment is likely, our findings suggest that both direct and indirect effects of the pandemic may contribute to excess deaths in care home residents.
-
-Our estimates of the prevalence of infection and deaths in residents are very similar to those reported in a recent large scale survey of care home managers in England,[@onsVivaldi2020] but it remains likely that we have underestimated the proportion of residents and staff who became infected during the pandemic.
-Testing capacity was limited in the early weeks of the epidemic, and PCR testing has been shown to have low sensitivity even when administered early in the duration of illness.
-According to a meta-analysis of 7 studies,[@Kucirka2020] the probability of a false-negative result in an infected person on the day of symptom onset (day 0) was 38% [18%; 65%], decreasing to 20% [12%; 30%] on day 3, subsequently rising to 66% [54%; 77%] by day 16.
-Evidence from a systematic review[@Salcher-Konrad2020] suggests that up to 75% of care home residents have asymptomatic infection, increasing the likelihood that infected cases were not detected before universal testing was introduced in May.
-
-We adopted a broad definition of mortality, by including all-cause deaths for any individual with a confirmed infection, as well as deaths occurring in hospitals (`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of all Covid-19-related deaths reported by outbreak surveillance counts).
-Our case-fatality estimate (`r tbrounding(desc_residents$DAT_CFR_rate)`% on a period of `r as.integer(date_max - date_min)` days) thus represents an upper estimate of the case-fatality rate and was higher than in previous literature,[@Salcher-Konrad2020] which often had both a narrower fatality case definition, and shorter observation periods.
-An outbreak investigation[@Graham2020] on 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days.
-These two studies had a much lower overall mortality incidence proportion at 7% and 6% respectively compared to 19% among FSHCG private residents, and as they focus on the outbreak period are less likely to identify indirect deaths associated with the pandemic.
-Infections were also less frequent and concentrated in a smaller proportion of care homes in Stall et al.[@Stall2020]
-
-**Need to add a paragraph about the limitations of estimating attributable mortality.**
-
-`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/ desc_residents$total_homes * 100)`% of care homes in our study reported at least one case of resident infection or Covid-related death.
-This proportion is higher than PHE estimates based on outbreak detection, and higher than the 56% of care homes that reported at least one case of infection in an English national survey of care home manager.[@onsVivaldi2020] Reasons for this discrepancy may relate to systematic differences between FSHCG homes and homes that were surveyed, better detection of cases though FSHCG's daily reporting system, or comparatively high rates of participation in the national whole care home testing programme.
-
-In common with a large study from Canada,[@Brown2020] we found strong associations between the incidence of infections and increased care home occupancy.
-Care homes with higher staff to resident ratios also had lower rates of infection and mortality in residents.
-These organisational factors are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, training staff, and regular deep cleaning of the care home environment.
-Higher staffing ratios may also be a sign of better-resourced care homes, with greater access to  protective personal equipment, a reduced risk of transmission of infection from staff to residents (since staff work with a smaller number of residents) and less reliance on temporary staff who are likely to work across multiple care settings.
+This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35·7%.
+Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks, implying substantial case under-ascertainment.
+Less than one-third of deaths attributable to COVID-19 in outbreak homes were confirmed which is likely due to poor availability of testing until late in the pandemic.
+In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy are critically important to reducing the spread of infection.
+
+Our estimates of the prevalence of confirmed Covid-19 infections and deaths in residents are comparable to a large survey of care home managers in England.[@onsVivaldi2020] However, both studies are likely to have underestimated the proportion of residents who became infected due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is higher than previous literature,[@Salcher-Konrad2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
+For example, an outbreak investigation[@Easter6] in 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days in a Canadian study.
+
+Whereas two-thirds of care homes in our study reported at least one case of infection or death, just 44% of care homes have notified an outbreak to PHE.[@PHE12020]  This suggests that nationally, local health protection teams may be unaware of Covid-19 infections in up to 1 in 5 care homes.
+Integration of data systems, so that test results can be accessed and acted upon by local public health teams is fundamental to the pandemic response.
+
+In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home occupancy.
+We also identified lower staff to resident ratios as a risk factor for infection.
+These organisational factors, linked to chronic underfunding of the care home sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
+When staff care for fewer residents they also have reduced likelihood of spreading infection between patients.
+Higher staff to resident ratios may also decrease reliance on agency staff who may spread infection between homes, and indicate better resourced care homes with greater access to protective personal equipment.
 
@@
 
-The surveillance system established by FSHCG enabled us to establish a unique cohort study of staff and residents across a large care home chain in three countries throughout the pandemic period.
-Daily reporting of suspected and confirmed cases to care home managers ensured high quality data capture, since this system was used for mandatory reporting to the care home regulator.
-Importantly, it allowed us to estimate the burden of undiagnosed infection, since widespread testing for SARS-CoV-2 was not widely available to care homes until May.
-The incident reporting (Datix) system allowed detailed analysis of the relationship between suspected and confirmed infection and death in private residents.
-
-An important limitation of this study is lack of access to information on comorbidity and ethnicity, both of which have been shown in previous studies to be an important risk factors for Covid-19 infection and adverse outcomes.[@Zhou2020] Our finding that residents receiving nursing rather than residential care have higher rates of infection and mortality is likely to be partly explained by higher rates of comorbidity in those receiving nursing care, however future studies should aim to capture individual level data on both of these risk factors.
-We were unable to investigate the relationship between residents being discharged to the care home from hospital and subsequent risk of infection in care home residents, since we did not have access to reliable information on the dates of hospital admission or discharge.
-Finally, we lacked information on the overall rate of care home testing, since incident reporting recorded positive test results and symptoms rather than whether a resident has been tested, and this is clearly a major determinant of confirmed infection rates.
-However, by 22nd June 2020, at least 90% of FSHCG homes had participated in the whole care home testing programme (when all residents and staff are offered testing for infection).
-Whilst results from the testing programme cannot address the issue of case under-ascertainment at the start of the pandemic, the fact that the majority of FSHCG homes participated in the whole home testing programme makes it likely that one third of care homes have genuinely not experienced any Covid-19 cases.
-It seems likely that these homes remain vulnerable to future waves of infection.
+#The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of care homes, and identify symptomatic as well as confirmed and asymptomatic cases.
+To our knowledge, this is the most compete reporting system for Covid-19 infections in care homes published to date.
+It is possible that care homes that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.
+
+A limitation is lack of access to information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in Covid-19.[@Zhou2020] However, we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
+We also lacked information on the overall rate of care home testing.
 
@@
 
-Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection,[@Dora2020] since various studies have shown effective transmission of SARS-CoV-2 from people who are infected but not yet symptomatic.[@He2020; @Kimball2020] The UK government has recently announced plans[@dhscRegularTesting2020] to increase the frequency of testing in care home staff and residents.
+In the UK the number of infected residents and staff has been underestimated, due to limited availability of testing until late in the pandemic.
+High levels of asymptomatic infection will also lead to under-ascertainment.
+It is important to note however, that mortality rates were also increased in those recorded as asymptomatic infections suggesting that in an elderly cohort patients may have atypical presentations that do not conform to standard case definitions.
+Although our findings support increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because  the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
+Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short.
+Rapid early diagnosis of symptomatic cases in residents and staff and expansion of more widespread testing after a case is identified may also be effective strategies to prevent transmission.
+Such approaches depend on strengthened surveillance in care homes and would be greatly facilitated by the availability of near patient testing platforms, which may be achievable in larger homes.
+
+Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of Covid-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
+Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which Covid-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected .
+Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes.
+Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.
+
+Covid-19 cases are declining in Europe, but there is an opportunity to mitigate the impact of future waves of infection on care home staff and residents.
+Our findings suggest that countries can achieve this by adopting a holistic approach, which integrates surveillance and focused testing for Covid-19 with increased investment to reduce care home occupancy and increase staffing.
+
+<!-- Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection,[@Dora2020] since various studies have shown effective transmission of SARS-CoV-2 from people who are infected but not yet symptomatic.[@He2020; @Kimball2020] The UK government has recently announced plans[@dhscRegularTesting2020] to increase the frequency of testing in care home staff and residents.
 Although our findings strongly support increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
@@
 Alternatively the development of saliva antigen tests could provide a less invasive solution to testing.
-
-Although we have largely attributed our findings of increased mortality in residents with no record of infection (in outbreak homes) to poor case ascertainment, the difference between outbreak and non-outbreak homes raises the possibility that Covid-19 may have indirectly increased mortality in care home residents.
-There are many ways in which delivery of usual care may have been impacted by the pandemic, increasing residents’ risk of preventable adverse events such as falls or pressure ulcers.
-Detailed analysis of cause of death and reasons for hospital admission in care home residents during the pandemic will be important to further explore this issue.
-
-We note that a large proportion of the analysis in this paper was made possible by the centralised reporting of individual-level data, the quality of which could be improved with support from national health organisations.
-Although we made good use of home-level aggregate outbreak surveillance counts using a life table approach, this involves some strong assumptions on all residents being susceptible at the beginning of the study period.
-Such assumptions would not hold in the same way in the event of a second wave of the epidemic.
-Anticipation of a possible resurgence should constitute a further motive for rapidly strengthening care home infection surveillance systems.
+-->
 
@@
 
-## Supplementary material 3: Additional tables
-
-`r table_nums("tab_national_descriptives")` [@cqcdata2020; @hscni2020; @isd2020]
+## Supplementary material 3: Kaplan-Meier estimators data
 
 ```{r table_national_descriptives, fig.cap = tab_national_descriptives, message=F}
+#
 mutate_all(desc_homes,  .funs = formatbm) %>% arrange(gsub("FSHC ", "", Scope)) %>% t %>% kable

### 2020-07-12 manuscript/manuscript.Rmd
minor fixes

@@
 arsenal::tableby(
-  confirmed  ~ gender + ageg  +  home_type + user_type2  + staff_to_bed_ratio_g5 +  occ_bedroom_ratio_g6, 
+  confirmed  ~ gender + ageg  +  home_type + user_type2 +bed_tot_category + bed_to_staff_ratio +  occ_bedroom_ratio_g6, 
   control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
@@
 arsenal::tableby(
-  death_status  ~ gender + ageg  +  home_type + user_type2  + staff_to_bed_ratio_g5 +  occ_bedroom_ratio_g6, 
+  death_status  ~ gender + ageg  +  home_type + user_type2  + bed_to_staff_ratio + bed_tot_category +  occ_bedroom_ratio_g6 + covid_status_outbreak, 
   control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
@@
 
-### Attributable mortality
+## Attributable mortality
 
@@
 
-#The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of care homes, and identify symptomatic as well as confirmed and asymptomatic cases.
+The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of care homes, and identify symptomatic as well as confirmed and asymptomatic cases.
 To our knowledge, this is the most compete reporting system for Covid-19 infections in care homes published to date.
@@
 |	Total exposure (days)|`r formatbm(desc_residents$DAT_total_symptomatic_denom)`|	`r formatbm(desc_residents$DAT_total_confirmed_denom)` | `r formatbm(desc_residents$total_denom)` |
-|	Incidence proportion (%)  | `r tbrounding(desc_residents$DAT_p_symptomatic)` [`r tbrounding(desc_residents$DAT_p_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`] |`r tbrounding(desc_residents$DAT_p_confirmed)` [`r tbrounding(desc_residents$DAT_p_confirmed_lower)`; `r tbrounding(desc_residents$DAT_p_confirmed_upper)`] |`r tbrounding(desc_residents$DAT_p_death)` [`r tbrounding(desc_residents$DAT_p_death_lower)`; `r tbrounding(desc_residents$DAT_p_death_upper)`] |
+|	Cumulative incidence (%)  | `r tbrounding(desc_residents$DAT_p_symptomatic)` [`r tbrounding(desc_residents$DAT_p_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_p_symptomatic_upper)`] |`r tbrounding(desc_residents$DAT_p_confirmed)` [`r tbrounding(desc_residents$DAT_p_confirmed_lower)`; `r tbrounding(desc_residents$DAT_p_confirmed_upper)`] |`r tbrounding(desc_residents$DAT_p_death)` [`r tbrounding(desc_residents$DAT_p_death_lower)`; `r tbrounding(desc_residents$DAT_p_death_upper)`] |
 |	Incidence rate (per 100,000) | `r tbrounding(desc_residents$DAT_IR_symptomatic)` [`r tbrounding(desc_residents$DAT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$DAT_IR_symptomatic_upper)`] | `r tbrounding(desc_residents$DAT_IR_confirmed)`  [`r tbrounding(desc_residents$DAT_IR_confirmed_lower)`; `r tbrounding(desc_residents$DAT_IR_confirmed_upper)`] | `r tbrounding(desc_residents$DAT_IR_death)` [`r tbrounding(desc_residents$DAT_IR_death_lower)`; `r tbrounding(desc_residents$DAT_IR_death_upper)` ]|

### 2020-07-12 manuscript/manuscript.Rmd
manuscript v1.0

@@
 title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
-# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
 bibliography: references.bib
@@
 ```{r fig_KM_CT_plots, fig.height = 4, fig.width = 12, message=FALSE}
+
+km_ct_residents_symp$y <- (1-km_ct_residents_symp$S_t) * desc_residents$CT_total_residents
+km_ct_residents_confirmed$y <- (1-km_ct_residents_confirmed$S_t) * desc_residents$CT_total_residents
+km_ct_residents_deaths$y <- (1-km_ct_residents_deaths$S_t) * desc_residents$CT_total_residents
+
 write.csv(km_ct_residents_symp, file = "supp_table_kaplan_meier_residents_symptomatic.csv", row.names = F)
@@
     ""
-  )), position = position_nudge(y = .02)) +
+  )), size = 3, position = position_nudge(y = .02)) +
   scale_y_continuous("Cumulative incidence",
@@
 )
+
+km_staff_symptomatic$y <- (1-km_staff_symptomatic$S_t) * desc_staff$total_staff
+km_staff_confirmed$y <- (1-km_staff_confirmed$S_t) * desc_staff$total_staff
 write.csv(km_staff_symptomatic, file = "supp_table_kaplan_meier_staff_symptomatic.csv", row.names = F)
@@
     ""
-  )), position = position_nudge(y = .02)) +
+  )), size = 3, position = position_nudge(y = .02)) +
   ggsci::scale_color_lancet()+
@@
 ```{r tablefiller1, results="asis"}
-arsenal::tableby(
-  confirmed  ~ gender + ageg  +  home_type + user_type2 +bed_tot_category + bed_to_staff_ratio +  occ_bedroom_ratio_g6, 
-  control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
-  data = residents_overview) %>%  summary()
+# arsenal::tableby(
+#   confirmed  ~ gender + ageg  +  home_type + user_type2 +bed_tot_category + bed_to_staff_ratio +  occ_bedroom_ratio_g6, 
+#   control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
+#   data = residents_overview) %>%  summary()
 
@@
 ```{r tablefiller2, results="asis"}
-residents_overview$death_status <- as.integer(residents_overview$status == "Deceased")
-arsenal::tableby(
-  death_status  ~ gender + ageg  +  home_type + user_type2  + bed_to_staff_ratio + bed_tot_category +  occ_bedroom_ratio_g6 + covid_status_outbreak, 
-  control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
-  data = residents_overview) %>%  summary()
+# residents_overview$death_status <- as.integer(residents_overview$status == "Deceased")
+# arsenal::tableby(
+#   death_status  ~ gender + ageg  +  home_type + user_type2  + bed_to_staff_ratio + bed_tot_category +  occ_bedroom_ratio_g6 + covid_status_outbreak, 
+#   control = arsenal::tableby.control(test = FALSE, cat.stats = "countrowpct", digits = 1L),
+#   data = residents_overview) %>%  summary()
 
@@
 KM_mortality_plot
+
+summary(KM_mortality_by_covid_status)[c("time",  "n.risk", "n.event", "n.censor", "surv", "strata")] %>% 
+  as_tibble %>% 
+  mutate(infection = gsub("covid_status=", "", gsub(", outbreak=.*", "", strata)),
+         outbreak = gsub(".*, outbreak=", "", strata))  %>% 
+  select(-strata) %>% 
+  write.csv(file = "supp_table_kaplan_meier_indiv_alldeaths.csv", row.names = F)
 ```

### 2020-07-12 manuscript_appendix/manuscript_appendix.Rmd
manuscript v1.0

---
title: 'Incidence of confirmed COVID-19 cases in care home staff and residents: Data
  quality and methodology report'
bibliography: ../manuscript/references.bib
output:
  html_document:
    code_folding: hide
    df_print: kable
    keep_md: yes
    toc: yes
    toc_depth: 1
  word_document:
    toc: yes
    toc_depth: 1
    reference_docx: ../manuscript/template.docx
always_allow_html: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(OutDec = ".")

library(summarytools)
st_options(
  headings = FALSE,
  footnote = NA
)
source("../.Rprofile")
devtools::load_all("..")
library(dplyr)
library(ggplot2)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
load_data()
date_min <- as.Date("2020-03-02")
date_max <- as.Date("2020-06-07") #as.Date("2020-06-15")
date_min_tallies <- as.Date("2020-03-24")
date_max_tallies = as.Date("2020-06-07")

tline <- dplyr::bind_rows(timelines$timeline)
stopifnot(unique(tline$date_end)>=date_max)
tline <- dplyr::filter(tline, between(date, date_min, date_max)) %>% 
  dplyr::filter(resident_id %in% unique(residents$resident_id))

home_inclusion <- unique(tline$home_code)

reference_homes <- reference_homes %>% 
  dplyr::filter(home_code %in% home_inclusion) %>% 
  dplyr::left_join(dplyr::select(reference_geography, postcode, country, region_UK), by = "postcode")

incidents <- incidents %>% 
  filter(dplyr::between(incident_date, date_min, date_max)) %>% 
  filter(home_code %in% home_inclusion)
beds <- beds %>% 
  filter(home_code %in% home_inclusion)

desc_before_exclusions <- list()
desc_before_exclusions$nrow_incidents <- nrow(incidents)
desc_before_exclusions$missing_id <- sum(grepl("^MISSING", incidents$resident_id))
desc_before_exclusions$missing_id_pct <- mean(grepl("^MISSING", incidents$resident_id))*100
desc_before_exclusions$link_fail <- sum(!incidents$resident_id[
  which(!grepl("^MISSING", incidents$resident_id))] %in% unique(residents$resident_id))
desc_before_exclusions$link_fail_pct <- tbrounding(
  sum(!incidents$resident_id[
    which(!grepl("^MISSING", incidents$resident_id))] %in% unique(residents$resident_id)) / 
    nrow(incidents)*100)

# removing erroneous incident reports (no identifier or not linking to residents file) 
incidents <- filter(incidents, !grepl("^MISSING", resident_id)) %>% 
  filter(resident_id %in% unique(residents$resident_id))

new_cases <- filter(new_cases, between(date, date_min, date_max_tallies)) %>% 
    filter(home_code %in% home_inclusion)

```

# Overview

This report provides additional information on the validation of:

- procedures for classifying the SARS-CoV-2 status (symptomatic, tested, confirmed)
- procedures for computing the at-risk resident-days 
- overall numbers of symptomatic and confirmed cases

# Classification of SARS-CoV-2 cases from incident reports

The SARS-CoV-2 status is determined by a combination of fields from the Datix incidents management system:

- `infection_covid_19_type` is a mandatory field one of two possible values:
  - "Confirmed formal clinical diagnosis of COVID-19"   
  - "Symptoms, but no definite clinical diagnosis of COVID-19"
- `infection_confirmed` contains one of 4 possible values (if not missing):
  - "Yes - positive test result: confirmed case"
  - "Test result not yet received"
  - "No - negative test result"
  - "Not tested"   
- `infection_result` is a free-text field containing, for the most part, the words 'positive' or 'negative'

Incident reports are dynamic: they are designed to be updated as more information becomes available. If all fields are not accurately kept updated, some values may become inconsistent. Inconsistencies are observed in incident reports  as illustrated in the tabulations below. Due to this, a reclassification algorithm is used which prioritises any information on a positive test result.

```{r , results="asis"}
ctable(incidents$infection_confirmed, incidents$infection_covid_19_type, useNA = "always", style = "rmarkdown")
```

## Algorithm

The following algorithm is used to classify symptomatic, tested, and test-confirmed cases from the Datix reports of COVID-19 infections.

```{r echo=FALSE}
DiagrammeR::grViz("../man/figure/case_classification_1.gv", width=700, height= 500) %>% 
  export_svg %>% charToRaw %>% rsvg::rsvg_png("case_classification_1.png", )
knitr::include_graphics("case_classification_1.png")
```

```{r echo = FALSE}
grViz("../man/figure/case_classification_2.gv", width=700, height= 400) %>% 
  export_svg %>% charToRaw %>% rsvg::rsvg_png("case_classification_2.png")
knitr::include_graphics("case_classification_2.png")
```

```{r echo = FALSE}
DiagrammeR::grViz("../man/figure/case_classification_3.gv", width=400, height= 400)  %>% 
  export_svg %>% charToRaw %>% rsvg::rsvg_png("case_classification_3.png")
knitr::include_graphics("case_classification_3.png")
```

## Validation

Data are tabulated below to evaluate the internal consistency of the rules set out in the previous section.

The `infection_result` field provides information on test results which broadly agrees yet sometimes conflicts with the `infection_confirmed` classification of incidents. It is potentially more up to date although it is very sparsely populated.

```{r, results="asis"}
ctable(incidents$infection_covid_19_type, incidents$covid_confirmed, useNA = "always", style = "rmarkdown")
```

```{r, results="asis"}
ctable(incidents$infection_covid_19_type, incidents$covid_test_result, useNA = "always", style = "rmarkdown")
```

```{r, results="asis"}
ctable(incidents$infection_confirmed, incidents$covid_test_result, useNA = "always", style = "rmarkdown")
```

# Data quality

## Data linkage

Secondary use individual-level data for private residents receiving general and elderly care were extracted and pseudonymised by HW. They consisted of two datasets:

- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly). This dataset does not include any records for local authority 'block contract' beds.
- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff. Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting. In addition, reports could record: date of birth, gender, information on Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death. Incidents were classified to indicate whether the resident was symptomatic, and whether an infection was confirmed on the basis of multiple variables (supplementary material 1).

HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier. Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents' dataset. Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) incident records did not contain enough information to generate an identifier and were excluded. Identifiers of a further `r desc_before_exclusions$link_fail` records (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident. A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents' dataset.

## Exclusions

A total of `r desc_before_exclusions$nrow_incidents` incident reports were obtained. Resident pseudoidentifiers were absent in `r formatbm(desc_before_exclusions$missing_id)` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) of Datix incident reports. The pseudoidentifiers of a further `r desc_before_exclusions$link_fail ` reports (`r tbrounding(desc_before_exclusions$link_fail /desc_before_exclusions$nrow_incidents*100)`%) were could not be successfully linked to the residents database. All were excluded, leaving a total of `r nrow(incidents)` reports dated between `r format(date_min, "%A %e %b %Y")` and  `r format(date_max, "%A %e %b %Y")` relating to `r formatbm(dplyr::n_distinct(incidents$resident_id))` unique residents.
These reports belongued to `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds. For the period of study, the Four Seasons Health Care Group (FSHCG) comprises `r formatbm(dplyr::n_distinct(beds[beds$year=="2020", "home_code"]))` care homes with a total of `r formatbm(sum(beds[beds$year=="2020", "beds"]))` beds.

## Occupancy

```{r occupancy_appx}
approx_occupancy <- tline %>% 
  filter(date %in% unique(occupancy$week_ending)) %>% 
  mutate(week_ending = date) %>% 
  group_by(home_code, week_ending) %>% 
  summarise(approx_occupancy = sum(rday))

occ_home <- merge(
  approx_occupancy,
  #remove Warmley House Nursing which has no occupancy report
  na.omit(select(occupancy, #filter(occupancy, home_code != "360"),
                 home_code, week_ending, occupancy)),
  all = FALSE
)

bed2020 <- dplyr::filter(beds, year == "2020") %>% 
  dplyr::select(-year) %>% 
  mutate(pct_contract = contract_beds/beds,
         contract_10pc = dplyr::case_when(
    contract_beds/beds >= .1 ~ ">=10% contract beds",
    TRUE ~ "<10% contract beds"
  ))

occ_home <- merge(
  occ_home,
  bed2020,
  all.x = TRUE
) %>% 
  dplyr::mutate(contract_10pc = dplyr::case_when(
    contract_beds/beds >= .1 ~ ">=10% contract beds",
    TRUE ~ "<10% contract beds"
  )) %>% 
  dplyr::left_join(select(reference_homes, home_code, country)) %>% 
 dplyr::filter(between(week_ending, date_min_tallies, date_max_tallies))

ratio_occupant_estimates <- (sum(occ_home$occupancy)/sum(occ_home$approx_occupancy))
ratio_occupant_estimates_england <- dplyr::filter(occ_home, country == "England") %>% 
  dplyr::summarise(ratio = sum(occupancy)/sum(approx_occupancy)) %>% 
  .$ratio

outbreak_homes <- dplyr::mutate(new_cases, new_confirmed = tidyr::replace_na(new_confirmed_home, 0) + tidyr::replace_na(new_confirmed_hospital, 0)) %>% 
  dplyr::group_by(home_code) %>% 
  dplyr::summarise(new_confirmed = sum(new_confirmed)) %>% 
  dplyr::filter(new_confirmed != 0) %>% 
  dplyr::distinct(home_code) %>% .$home_code

ratio_occupant_estimates_outbreak_only <- occ_home %>% 
  dplyr::filter(home_code %in% outbreak_homes) %>% 
  dplyr::summarise(ratio = sum(occupancy)/sum(approx_occupancy)) %>% 
  .$ratio

occ_global <- occ_home %>% 
  dplyr::group_by(home_code, contract_10pc, pct_contract) %>% 
  dplyr::summarise(
    approx_occupancy = sum(approx_occupancy),
    occupancy = sum(occupancy)
  ) %>% 
  dplyr::mutate(
    outliers = dplyr::if_else(
      abs(1-approx_occupancy/occupancy) > .10 &
        contract_10pc != ">=10% contract beds",
      home_code,
      "")
  )
```

Care home resident-days are available from two sources:

- reports to FSHCG by every home on their occupancy every Sunday, which include residents occupying contract beds 
- individual-level resident records of first and last admission and discharge dates, to he exclusion of occupants of contract beds. These records are used to compute occupancy on a daily basis.

In total, `r sum(filter(bed2020, home_code %in% home_inclusion)$contract_beds)` out of `r sum(filter(bed2020, home_code %in% home_inclusion)$beds)` beds were contracted by the local authority, with `r nrow(filter(bed2020, pct_contract>.5 & home_code %in% home_inclusion))` homes having more than half of their beds contracted.

Plots below compare both measures for every Sunday (Figure 1) and in total across the period (Figure 2) at the level of care homes, drawing attention to care homes with a large proportion of contract beds. The Pearson's correlation between the weekly measures was $r=$ `r tbrounding(cor(occ_home$occupancy, occ_home$approx_occupancy), 2)` overall, rising to $r=$ `r with(filter(occ_home, contract_beds/beds <.1), tbrounding(cor(occupancy, approx_occupancy), 2))` once excluding care homes with $\geq$ 10% contract beds. These findings confirm that occupancy could be approximated reliably using just residents' records of first and last admission and discharge dates.

This enabled us to estimate exposure (resident-days) excluding local authority contract beds, since no Datix incident reports were returned for occupants of such beds.

In addition, this validation enabled us to approximate the total number of unique residents (including contract beds) during the period of observation for aggregate count data (`r date_min_tallies` -- `r date_max_tallies`). The ratio of occupants recorded by FSHCG to the number approximated by us was `r ratio_occupant_estimates `. The number of unique residents in non-contract beds was `r formatbm(length(unique(tline$resident_id[tline$rday ==1])))` during the study period. By assuming that the resident turnover in local authority beds was the same as the rate in other beds, we infer the number of unique residents inclusive of contract beds to be `r formatbm(round(length(unique((tline$resident_id[tline$rday ==1])))*ratio_occupant_estimates+.5))`.

In England, this ratio was `r ratio_occupant_estimates_england`, and for outbreak homes across the UK, `r ratio_occupant_estimates_outbreak_only`

```{r,  fig.cap = "Weekly total resident-days per care home", message = F}
ggplot(occ_home, aes(x = occupancy, y = approx_occupancy)) +
  geom_point(aes(shape = contract_10pc)) +
  geom_abline(aes(intercept = 0, slope = 1,  col = "Identity", linetype = "Identity")) +
  geom_smooth(aes(col="OLS", linetype = "OLS"), method = "lm", se = T) +
  xlab("Occupancy provided by FSHC") +
  ylab("Occupancy approximated\nfrom resident records") +
  ggtitle(paste0("Figure 1: Weekly total resident-days by care home (L=", dplyr::n_distinct(occ_home$home_code), ")")) +
  scale_colour_manual(name="Line of fit",
    values = c("Identity" = "black", "OLS" = "blue", "Spline" = "red")) +
  scale_linetype_manual(name="Line of fit",
    values = c("Identity" = "solid", "OLS" = "longdash", "Spline" = "dotted")) +
  scale_shape_manual(name="Bed composition",
    values = c(1, 3)) 
```

```{r, fig.cap = "Weekly total resident-days", message = F}

ggplot(occ_global, aes(x = occupancy, y = approx_occupancy)) +
  geom_point(aes(shape = contract_10pc)) +
  # geom_text(aes(label = outliers), 
  #           nudge_x = .1,  nudge_y = .15, vjust = "outward", hjust = "outward") +
   geom_abline(aes(intercept = 0, slope = 1,  col = "Identity", linetype = "Identity")) +
  geom_smooth(aes(col="OLS", linetype = "OLS"), method = "lm", se = T) +
  xlab("Occupancy provided by FSHC") +
  ylab("Occupancy approximated from resident records") +
  ggtitle(paste0("Figure 2: Total resident-days across FSHC during study period \n(", format(date_min, "%d %b %Y"), " - ", format(date_max, "%d %b %Y"),")")) +
  scale_colour_manual(name="Line of fit",
    values = c("Identity" = "black", "OLS" = "blue", "Spline" = "red")) +
  scale_linetype_manual(name="Line of fit",
    values = c("Identity" = "solid", "OLS" = "longdash", "Spline" = "dotted")) +
  scale_shape_manual(name="Bed composition",
    values = c(1, 3)) 
```

We nevertheless identified 11 care homes in which the total resident-days approximated differed by more than 10 percent despite not belonging (marked on Figure 2).

```{r, message=F}
# merge(dplyr::select(reference_homes, home_code, home_name_datix),
#       dplyr::filter(occ_global, outliers != "")) %>%
#   dplyr::transmute(home_code, home_name_datix, approx_occupancy, occupancy,
#                    `% contract beds` = round(pct_contract*100, 1))
```

## Comparison of manager-reported tally counts with Datix-derived counts

```{r case_validation, message=F}
residents_overview <- tline %>% 
  dplyr::filter(between(date, date_min_tallies, date_max_tallies)) %>% 
  dplyr::filter(rday==1) %>%
  dplyr::group_by(resident_id, home_code) %>% 
  dplyr::summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed),
            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
            confirmed = max(-susceptible_confirmed+1, na.rm = T),
            start_date = min(date, na.rm = T),
            stop_date = max(date, na.rm = T),
            start_int = as.numeric(min(date, na.rm = T) - date_min),
            stop_int = as.numeric(max(date, na.rm = T)- date_min)
    ) %>% 
  dplyr::left_join(dplyr::select(reference_homes, home_code, home_name_datix)) %>% 
  dplyr::left_join(dplyr::distinct(dplyr::select(incidents, resident_id, dplyr::starts_with("covid_first"))) )

datix_counts <- residents_overview %>% 
  dplyr::group_by(home_code, home_name_datix) %>% 
  dplyr::summarise(
    datix_symptomatic = sum(symptomatic, na.rm = T),
    datix_confirmed = sum(confirmed, na.rm = T) ) %>% 
  dplyr::left_join(bed2020)

tally_counts <- new_cases %>% 
  filter(home_code %in% home_inclusion) %>% 
  filter(between(date, date_min, date_max)) %>% 
  dplyr::group_by(home_code) %>% 
  dplyr::summarise(
    tally_symptomatic = sum(new_symptomatic_home, na.rm = T) + 
      sum(new_symptomatic_hospital, na.rm = T),
    tally_confirmed = sum(new_confirmed_home, na.rm = T) + 
      sum(new_confirmed_hospital, na.rm = T)
  )

ch_data_all <- merge(datix_counts, tally_counts, all = T) %>% 
  dplyr::mutate(
    home_code_lbl = if_else(tally_symptomatic>20, home_code, ""),
    home_code_lbl2 = if_else(tally_confirmed>15 | datix_confirmed>15, home_code, ""))
```

During the period, manager counted `r sum(ch_data_all$tally_confirmed)` confirmed infection while Datix recorded `r sum(ch_data_all$datix_confirmed)` such infections. The `r round(sum(ch_data_all$tally_confirmed)/sum(ch_data_all$datix_confirmed)*100-100 )`% excess could be due to a range of explanations:

1. contract beds occupants, who may represent at least `r tbrounding((1-1/ratio_occupant_estimates)*100)`% of residents (assuming that their turnover is identical to that of occupants of other beds, in reality turnover is likely to be higher, and so is this proportion), may have a higher infection rate than other residents 
2. underreporting on Datix
3. linkage error
4. double-counting of cases in manager counts

It is not possible to rely on explanations (1-3) exclusively: the number of confirmed infections on Datix was greater than manager counts in `r sum(ch_data_all$datix_confirmed > ch_data_all$tally_confirmed)` care homes. This suggest double counting (4) is a possibility.

```{r comparison_symptomatic, message = F}
ggplot(data = ch_data_all, 
       aes(x = tally_symptomatic, y = datix_symptomatic)) + 
  geom_point(aes(shape = contract_10pc)) +
  # geom_text(aes(label = home_code_lbl),
  #           nudge_x = 2,  nudge_y = 1) + 
  geom_abline(aes(intercept = 0, slope = 1,  col = "Identity", linetype = "Identity")) +
  geom_smooth(aes(col="OLS", linetype = "OLS"), method = "lm", se = T) +
  xlab("Manager counts") + ylab("Datix-derived counts") + labs(title="Figure 3: Comparison of Datix and manager counts of symptomatic cases") +
  scale_colour_manual(name="Line of fit",
    values = c("Identity" = "black", "OLS" = "blue", "Spline" = "red")) +
  scale_linetype_manual(name="Line of fit",
    values = c("Identity" = "solid", "OLS" = "longdash", "Spline" = "dotted")) +
  scale_shape_manual(name="Bed composition",
    values = c(1, 3)) 
```

```{r comparison_confirmed, message=F}

ggplot(data = ch_data_all, aes(x = tally_confirmed, y = datix_confirmed)) + 
  geom_point(aes(shape = contract_10pc)) +
  # geom_text(aes(label = home_code_lbl2), 
  #           nudge_x = 1,  nudge_y = 1.2) + 
  geom_abline(aes(intercept = 0, slope = 1,  col = "Identity", linetype = "Identity")) +
  geom_smooth(aes(col="OLS", linetype = "OLS"), method = "lm", se = T) +
  xlab("Manager counts") + ylab("Datix-derived counts") + labs(title="Figure 4: Comparison of Datix and manager counts of confirmed cases") +
  scale_colour_manual(name="Line of fit",
    values = c("Identity" = "black", "OLS" = "blue", "Spline" = "red")) +
  scale_linetype_manual(name="Line of fit",
    values = c("Identity" = "solid", "OLS" = "longdash", "Spline" = "dotted")) +
  scale_shape_manual(name="Bed composition",
    values = c(1, 3))

```

# Estimating the cumulative incidence of cases and resident-days at risk

Manager-reported outbreak surveillance counts were used together with occupancy estimated from weekly occupied bed censuses in order to estimate the incidence of cases across time using a Kaplan-Meier life table product limit estimator. As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records. Admitted cases were accounted for in the daily new case count. Based on this $\hat{I_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as

$$\hat{I_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad I_{t_0} = 0$$
where $d_{t}$ denotes the number of new cases for each day $t = \{0, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate. The total exposure time was calculated as $\sum_{t=0}^{T} (N_t-\hat{I}_t)$ where $N_t$ denotes daily occupancy figures.

# Total homes, beds, contract beds, and residents within FSHCG and nationwide

Data sources: [@cqcdata2020; @hscni2020; @isd2020]

# References

### 2020-07-16 manuscript/manuscript.Rmd
version 1.0

@@
 ---
-title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
+title: "Covid-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
 bibliography: references.bib
@@
 
+## Availability of data and materials
+
+Figure 2 and 3 data are available on request to authors.
+This work uses data provided by residents and collected by the Four Seasons Health Care Group as part of their care and support.
+These are available from the Four Seasons Health Care Group but restrictions apply to the availability of these data, which were used under license for the current study, and so are not publicly available.
+Data are however available from the authors upon reasonable request and with permission of the Four Seasons Health Care Group.
+
 # Supplementary material

### 2020-07-16 manuscript_appendix/manuscript_appendix.Rmd
version 1.0

@@
 ---
-title: 'Incidence of confirmed COVID-19 cases in care home staff and residents: Data
-  quality and methodology report'
+title: "Covid-19 infection and attributable mortality in UK long term care facilities: Data quality and methodology report"
 bibliography: ../manuscript/references.bib

### 2020-07-21 manuscript_appendix/manuscript_appendix.Rmd
realign dates to manuscript + fix typos

--- 2020-07-12 manuscript_appendix/manuscript_appendix.Rmd
@@
 ---
-title: 'Incidence of confirmed COVID-19 cases in care home staff and residents: Data
-  quality and methodology report'
+title: "Covid-19 infection and attributable mortality in UK long term care facilities: Data quality and methodology report"
 bibliography: ../manuscript/references.bib
 output:
+  word_document:
+    toc: yes
+    toc_depth: 1
+    reference_docx: ../manuscript/template.docx
   html_document:
@@
     toc_depth: 1
-  word_document:
-    toc: yes
-    toc_depth: 1
-    reference_docx: ../manuscript/template.docx
 always_allow_html: yes
@@
 date_min <- as.Date("2020-03-02")
-date_max <- as.Date("2020-06-07") #as.Date("2020-06-15")
+date_max <- as.Date("2020-06-14") #as.Date("2020-06-15")
 date_min_tallies <- as.Date("2020-03-24")
-date_max_tallies = as.Date("2020-06-07")
+date_max_tallies = as.Date("2020-06-14")
 
@@
 
-A total of `r desc_before_exclusions$nrow_incidents` incident reports were obtained.
+A total of `r formatbm(desc_before_exclusions$nrow_incidents)` incident reports were obtained.
 Resident pseudoidentifiers were absent in `r formatbm(desc_before_exclusions$missing_id)` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) of Datix incident reports.
 The pseudoidentifiers of a further `r desc_before_exclusions$link_fail ` reports (`r tbrounding(desc_before_exclusions$link_fail /desc_before_exclusions$nrow_incidents*100)`%) were could not be successfully linked to the residents database.
-All were excluded, leaving a total of `r nrow(incidents)` reports dated between `r format(date_min, "%A %e %b %Y")` and  `r format(date_max, "%A %e %b %Y")` relating to `r formatbm(dplyr::n_distinct(incidents$resident_id))` unique residents.
-These reports belongued to `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
+All were excluded, leaving a total of `r nrow(incidents)` reports dated between `r format(date_min, "%A %e %B %Y")` and  `r format(date_max, "%A %e %B %Y")` relating to `r formatbm(dplyr::n_distinct(incidents$resident_id))` unique residents.
+These reports belonged to `r formatbm(dplyr::n_distinct(incidents$home_code))` care homes totalling `r formatbm(sum(beds[beds$year=="2020" & beds$home_code %in% unique(incidents$home_code), "beds"]))` beds.
 For the period of study, the Four Seasons Health Care Group (FSHCG) comprises `r formatbm(dplyr::n_distinct(beds[beds$year=="2020", "home_code"]))` care homes with a total of `r formatbm(sum(beds[beds$year=="2020", "beds"]))` beds.

### 2020-10-14 manuscript/manuscript.Rmd
add index of multiple deprivation

@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE)
-options(OutDec = "·")
 knit_output_pdf <- F
@@
   left_join(transmute(reference_geography, postcode, 
-                      country, region_england = region), by = "postcode") %>% 
+                      country, region_england = region,
+                      uk_imd_england_quintile), by = "postcode") %>% 
   mutate(region_UK = if_else(country == "England", region_england, country)) %>% 
@@
          region_UK = as.factor(region_UK)) %>% 
-  mutate(occ_bedroom_ratio_g6 = relevel(occ_bedroom_ratio_g6, '(0·7,0·85]'),
-         occ_bedroom_ratio_g5 = relevel(occ_bedroom_ratio_g5, '(0·8,1]'),
+  mutate(occ_bedroom_ratio_g6 = relevel(occ_bedroom_ratio_g6, '(0.7,0.85]'),
+         occ_bedroom_ratio_g5 = relevel(occ_bedroom_ratio_g5, '(0.8,1]'),
          region_UK <- relevel(region_UK, "Scotland"))
@@
                    staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
-                   beds,  pct_occupancy, no_bedrooms, 
+                   beds,  pct_occupancy, no_bedrooms, uk_imd_england_quintile,
                    occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
@@
                    staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
-                   beds, pct_occupancy, no_bedrooms, 
+                   beds, pct_occupancy, no_bedrooms, uk_imd_england_quintile,
                    occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
@@
   group_by(home_code) %>% 
-  summarise(unique_residents = n_distinct(resident_id))
+  summarise(unique_residents = n_distinct(resident_id),
+            total_rdays = sum(resident_days))
 
@@
   left_join(occupancy_interpolated) %>% 
-  arrange(home_code, date) %>% 
-  group_by(home_code) %>% 
   mutate(new_symptomatic = tidyr::replace_na(new_symptomatic_home, 0) + 
@@
          ) %>%   
+  arrange(home_code, date) %>% 
+  group_by(home_code) %>% 
   mutate(I_symptomatic = total_cases_in_home(occupancy, new_symptomatic, date),
@@
 `r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
-Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1·9 [1·7; 2·2]).
+Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1.9 [1.7; 2.2]).
 33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.
@@
 
-Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]
+Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]
 
@@
 Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.
-`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39·2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
 `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in these residents, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
-The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22·6% versus 11·2%).
+The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
 
@@
   outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Funding type` +  
-    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
+    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation` + `Index of Multiple Deprivation`, 
   control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
@@
                 `Status as of 14 June` = status,
-                `Region/nation` = region_UK 
+                `Region/nation` = region_UK,
+                `Index of Multiple Deprivation` = uk_imd_england_quintile
   )) %>% summary()
@@
 Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
-Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1·5 [1·2; 1·8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-Large care homes had greater rates of infection (adjusted HR = 1.8 [1·4; 2·4] for care homes with $\geq$ 70 beds versus <35 beds).
+Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1.5 [1.2; 1.8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+Large care homes had greater rates of infection (adjusted HR = 1.8 [1.4; 2.4] for care homes with $\geq$ 70 beds versus <35 beds).
 Care home baseline occupancy and staffing ratios had the greatest effect on residents risk of infection.
-For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1·9; 3·3] greater in homes with 0·85-1 resident per room versus homes with 0·7-0·85 resident per room.
-In homes with more residents than bedrooms, hazards were 10 times [6·9; 15] greater.
-
-Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1·23 [1·17; 1·31]).
+For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1.9; 3.3] greater in homes with 0.85-1 resident per room versus homes with 0.7-0.85 resident per room.
+In homes with more residents than bedrooms, hazards were 10 times [6.9; 15] greater.
+
+Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1.23 [1.17; 1.31]).
 
@@
 # survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
+residents_overview$uk_imd_england_quintile <- factor(residents_overview$uk_imd_england_quintile,
+                                                     levels = c(
+                                                       '3', 
+                                                       '1 - least deprived', '2', '4', '5 - most deprived'
+                                                     ))
 residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
-explanatory = c("gender", "ageg", "home_type", "user_type2",  
+explanatory = c("gender", "ageg", "home_type", "user_type2",  "uk_imd_england_quintile",
                 "bed_tot_category", "bed_to_staff_ratio",  "occ_bedroom_ratio_g6")
@@
 The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality.
-After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1·4 [1·3; 1·6]), and receiving nursing care (adjusted HR = 1·4 [1·2; 1.5]).
+After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.4 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.2; 1.5]).
 
 We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection (group A) in non-outbreak homes as the reference group.
-Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2·0 [1·7; 2·2]).
-Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4·3 [3·0; 6·2] versus 9·4 [7·6; 12]).
-All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3·3 [2·0; 5·7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.
+Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2.0 [1.7; 2.2]).
+Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4.3 [3.0; 6.2] versus 9.4 [7.6; 12]).
+All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3.3 [2.0; 5.7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.
 
@@
 # cox.zph(cph_death2)
+residents_episodes$uk_imd_england_quintile <- factor(residents_episodes$uk_imd_england_quintile,
+                                                     levels = c(
+                                                       '3', 
+                                                       '1 - least deprived', '2', '4', '5 - most deprived'
+                                                     ))
+
 residents_episodes$bed_to_staff_ratio <- 1/residents_episodes$staff_to_bed_ratio
-explanatory = c("gender", "ageg", "home_type", "user_type2",
+explanatory = c("gender", "ageg", "home_type", "user_type2", "uk_imd_england_quintile",
                 "bed_tot_category", "bed_to_staff_ratio", "occ_bedroom_ratio_g6",
@@
 In homes with outbreaks only 28% (159 residents) of the mortality attributable to COVID-19 occurred in people with confirmed infection (Groups C and D), (`r table_nums("tab_attributable_deaths", display = "cite")`).
-Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41·7%).
+Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41.7%).
 Model-based estimates of deaths based on individual-level were slightly higher (8%) than counts from the aggregate data.
@@
                                                           names(cph_death3$coefficients)),]) %>% 
-  rename(HR_lower =  `2·5 %`,
-         HR_upper = `97·5 %`)
+  rename(HR_lower =  `2.5 %`,
+         HR_upper = `97.5 %`)
 HRs_CI_covid$covid_status_outbreak_digit <- gsub(
@@
 
-This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35·7%.
+This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35.7%.
 Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks, implying substantial case under-ascertainment.
@@
 
+## Supplementary analysis
+
+```{r effect_organisational_variables}
+# adhoc1 <- residents_ct %>% 
+#   filter(home_code %in% home_inclusion) %>% 
+#   group_by(home_code) %>% 
+#   summarise(confirmed_cases = sum(new_confirmed),
+#             symptomatic_cases = sum(new_symptomatic),
+#             denominator = sum(occupancy - I_confirmed)) %>% 
+#   left_join(reference_homes) %>% 
+#   left_join(select(filter(beds, year=="2020"), -year)) %>% 
+#   mutate(staff_to_bed_ratio = staff_total/beds,
+#          carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
+#          bed_tot_category = cut(beds, 
+#            breaks = seq(20,100,25), right = F, 
+#            paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
+#          ) %>% 
+#   mutate(staff_to_bed_ratio_g5 = cut(
+#     staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
+#     quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
+#     max(staff_to_bed_ratio)))) %>% 
+#   left_join(home_mean_occupancy) %>% 
+#   mutate(pct_occupancy = mean_occupancy/beds,
+#          occ_bedroom_ratio = mean_occupancy/no_bedrooms) %>% 
+#   mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
+#          occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)))
+# 
+# ggplot(adhoc1, aes(staff_to_bed_ratio, confirmed_cases/denominator*100000)) +
+#   geom_point() +
+#   geom_smooth() +
+#   ylab("Cumulative incidence per 100,000") +
+#   xlab("Ratio staff : beds")
+# 
+# mod_staffing <- glm(confirmed_cases ~  
+#   staff_to_bed_ratio + offset(log(denominator)), 
+#   family = poisson(link = "log"), data = adhoc1)
+# 
+# 
+# ggplot(adhoc1, aes(occ_bedroom_ratio, confirmed_cases/denominator*100000)) +
+#   geom_point() +
+#   geom_smooth() +
+#   ylab("Cumulative incidence per 100,000") +
+#   xlab("Ratio occupants : bedrooms (Jan-Mar 2020)")
+# 
+# adhoc1 %>% 
+#   group_by(occ_bedroom_ratio_g5) %>% 
+#   summarise(confirmed_cases = sum(confirmed_cases),
+#             denominator = sum(denominator))  %>% 
+#   mutate(rate = confirmed_cases/denominator*100000)
+
+```
+
 # References

### 2020-10-14 R/database.R

@@ -371,6 +371,58 @@ extract_occupancy <- function() {
   usethis::ui_done("occupancy.rda")
 }
 
+
+#' @rdname extract_data
+extract_deprivation <- function() {
+
+  IMD_2019_source <- url("https://data.bris.ac.uk/datasets/1ef3q32gybk001v77c1ifmty7x/uk_imd_scores_data.txt")
+  IMD_2019_data <- read.delim(IMD_2019_source, stringsAsFactors = F)
+  IMD_2019_data <- dplyr::select(IMD_2019_data,
+                                 area_code,
+                                 uk_imd_england_score,
+                                 uk_imd_england_quintile)
+  IMD_2019_data
+}
+
+#' @rdname extract_data
+extract_lsoa <- function() {
+  x <- read.csv(url("http://infuse.ukdataservice.ac.uk/help/definitions/2011geographies/lsoa-area-list2011.csv"))
+  names(x) <- tolower(names(x))
+  x$lsoa_code <- trimws(x$area_id)
+  x$lsoa_label <- gsub("_", " ", x$area_label)
+  x[, c("lsoa_code", "lsoa_label")]
+}
+
+
+#' @rdname extract_data
+#' @import SPARQL
+extract_datazones_2001 <- function(postcodes) {
+  query <- paste0(
+    "PREFIX rdf: <http://www.w3.org/1999/02/22-rdf-syntax-ns#>
+PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
+SELECT ?postcode ?dtz01
+WHERE {
+  ?s rdf:type <http://data.ordnancesurvey.co.uk/ontology/postcode/PostcodeUnit> .
+  ?s rdfs:label ?postcode
+  FILTER(?postcode IN ('",
+    paste(postcodes, collapse = "', '"),
+    "')) .
+    ?s <http://publishmydata.com/def/ontology/foi/within> ?dtz .
+  ?dtz <http://publishmydata.com/def/ontology/foi/memberOf> <http://statistics.gov.scot/def/foi/collection/data-zones-2001> .
+  ?dtz <http://publishmydata.com/def/ontology/foi/code> ?dtz01
+  }"
+  )
+
+  query <- gsub("\n", " ", query)
+
+  results <- SPARQL::SPARQL(
+    url = "http://statistics.gov.scot/sparql",
+    query = query)
+
+  results$results
+}
+
+
 #' @rdname extract_data
 #' @import PostcodesioR
 extract_reference_homes <- function() {
[…]

### 2020-10-14 tests/testthat/test-data.R

@@ -24,9 +24,6 @@ test_that("age_groups", {
 
   expect_equal(as.character(cut_age_pentadecimal(102)), "100–114 years")
   expect_equal(as.character(cut_age_pentadecimal(150)), NA_character_)
-
-
-
 })
 
 
@@ -62,5 +59,11 @@ test_that("total_cases_in_home", {
   truncated_result <- rev(truncated_result)
   expect_equal(trunc(total_cases_in_home(data$occupancy, data$new_cases, data$date)),
                truncated_result)
+
+})
+
+test_that("lsoa data", {
+  example <- extract_datazones_2001(c('G21 3TX', 'AB10 1BD'))
+  expect_equivalent(example$dtz01, c("S01000109", "S01003644"))
 
 })

### 2020-11-27 manuscript/manuscript.Rmd
Revisions version 1.4

@@
 ---
-title: "Covid-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
+title: "COVID-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
 bibliography: references.bib
@@
 output:
+  word_document:
+    toc: no
+    reference_docx: template.docx
   html_document:
@@
 
-Peter Dutey Magni,$^{12}$ Haydn Williams,$^{3}$ Arnoupe Jhass,$^{4}$ Greta Rait,$^4$ Harry Hemingway,$^{125}$ Andrew Hayward,$^{6}$ Laura Shallcross$^{12*}$
+Peter Dutey Magni, Researcher$^{12}$ Haydn Williams, Data manager$^{3}$ Arnoupe Jhass, General Practitioner$^{4}$ Greta Rait, Professor$^4$ Fabiana Lorencatto, Senior researcher$^5$ Harry Hemingway,$^{126}$ Andrew Hayward,$^{7}$ Laura Shallcross$^{12*}$
 
@@
 
-$^6$ Institute of Epidemiology & Health Care, University College London, London, UK
+$^6$ Centre for Behaviour Change, University College London, London, UK
+
+$^7$ Institute of Epidemiology & Health Care, University College London, London, UK
 
 $*$ Corresponding author
+
+Peter Dutey-Magni p.dutey-magni@ucl.ac.uk
+
+Haydn Williams haydn.williams@fhscgroup.com
+
+Arnoupe Jhass a.jhass@ucl.ac.uk
+
+Greta Rait g.rait@ucl.ac.uk
+
+Lorencatto, Fabiana f.lorencatto@ucl.ac.uk
+
+Harry Hemingway h.hemingway@ucl.ac.uk
+
+Andrew Hayward a.hayward@ucl.ac.uk
+
+Laura Shallcross l.shallcross@ucl.ac.uk
 
@@
 }
-
 source("../.Rprofile")
@@
   )) %>% 
-  mutate(covid_status = factor(covid_status, 
-                               levels = c("Uninfected", "Symptomatic (not confirmed)",
-                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
-  left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
-                   beds,  pct_occupancy, no_bedrooms, uk_imd_england_quintile,
-                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
-  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))
+  mutate(
+    covid_status = factor(
+      covid_status, 
+      levels = c("Uninfected", 
+                 "Symptomatic (not confirmed)",
+                 "Asymptomatic confirmed", 
+                 "Symptomatic confirmed"))) %>% 
+  left_join(
+    select(reference_homes, home_code, country, 
+           region_UK, staff_to_bed_ratio,
+           staff_to_bed_ratio_g5, carer_to_bed_ratio,
+           bed_tot_category, beds,  pct_occupancy,
+           mean_occupancy, no_bedrooms, staff_total,
+           uk_imd_england_quintile,
+           occ_bedroom_ratio, occ_bedroom_ratio_g5,
+           occ_bedroom_ratio_g6)) %>% 
+  left_join(distinct(select(incidents, 
+                            resident_id,
+                            starts_with("covid_first"))))
 
@@
     dplyr::group_by(home_code) %>% 
-    dplyr::summarise(first_case_date = min(covid_first_confirmed, na.rm = T), source="DAT")
+    dplyr::summarise(
+      first_case_date = min(covid_first_confirmed, 
+                            na.rm = T), 
+      source="DAT")
   ) %>% 
@@
   dplyr::summarise(first_case_date = min(first_case_date))
-  # tidyr::crossing(tibble(outbreak = c(0, 1)))
 
@@
   )) %>% 
-  mutate(covid_status = factor(covid_status, 
+  mutate(
+    covid_status = factor(
+      covid_status, 
       levels = c("Uninfected", "Symptomatic (not confirmed)",
                  "Asymptomatic confirmed", "Symptomatic confirmed")),
-         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
-  left_join(select(reference_homes, home_code, country, region_UK,
-                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
+    ageg = cut_age_denary(compute_age(dob, date_min), 
+                          age.min = 75, age.max = 95))   %>% 
+  left_join(
+    select(reference_homes, home_code, country, region_UK,
+           staff_to_bed_ratio, staff_to_bed_ratio_g5, 
+           carer_to_bed_ratio, bed_tot_category, 
            beds, pct_occupancy, no_bedrooms, uk_imd_england_quintile,
@@
     0L
-  )) %>% 
-  ungroup()
+  )) %>% ungroup()
+
+residents_episodes <- mutate(
+  residents_episodes, 
+  gender = as.factor(gender),
+  uk_imd_england_quintile = factor(
+    uk_imd_england_quintile,
+    levels = c('3', '1 - least deprived', '2',
+               '4', '5 - most deprived')),
+  bed_to_staff_ratio = 1/staff_to_bed_ratio
+)
 
@@
     St_min = S_t^exp(-1.96*S_t_SElog),
-    event = "Covid-19-related death"
+    event = "COVID-19-related death"
   )
@@
   name = "incidence_CT_residents",
-  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections in residents according to FSHCG aggregate data (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 supp_epi_CT_oubreak_only <- table_nums(
   name = "incidence_CT_staff",
-  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections among staff according to FSHCG aggregate data (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_resident_descriptives <- table_nums(
   name = "tab_resident_descriptives",
-  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption = paste0("Characteristics of FSHCG private bed residents by type of LTCF, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_CFR_age_sex <- table_nums(
@@
   name = "tab_CoxPH_death",
-  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
+  caption =  paste0("Risk factors for all-cause mortality in private residents of LTCFs with and without COVID-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
   formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
@@
   name = "tab_attributable_deaths",
-  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
+  caption =  paste0("Model-based estimates of attributable death in private residents of LTCFs with and without COVID-19 outbreaks (n=", 
   formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
@@
   name = "fig_study_diagram",
-  caption = paste0("Study overview: location of FSHCG care homes and diagram of data sources")
+  caption = paste0("Study overview: location of FSHCG LTCFs and diagram of data sources")
 )
@@
   name = "fig_KM_CT_plots",
-  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-related deaths in (A) residents (n=",   formatbm(desc_residents$CT_total_residents),
-  ") and (B) care home staff (n=",   formatbm(desc_staff$total_staff),
-  ") according to FSHCG outbreak surveillance counts (", format(date_min_tallies, "%e %b %Y"), "-", format(date_max_tallies, "%e %b %Y"), ")"))
+  caption = paste0("Kaplan-Meier estimates of the cumulative incidence of symptomatic cases, confirmed infections and COVID-19 related deaths in (A) residents (n=",   formatbm(desc_residents$CT_total_residents),
+  ") and (B) staff (n=",   formatbm(desc_staff$total_staff),
+  ") according to FSHCG aggregate data (", format(date_min_tallies, "%e %b %Y"), "-", format(date_max_tallies, "%e %b %Y"), ")"))
 
@@
 
-**Background** Rates of Covid-19 infection have declined in many countries, but outbreaks persist in care home residents who are at high risk of severe outcomes.
-Epidemiological data from care homes are scarce.
-We used population-level active surveillance to estimate incidence of, and risk factors for Covid-19, and attributable mortality in care home residents.
-
-**Methods:** Cohort study using individual-level electronic health records from 8,713 residents and daily counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` UK care homes.
-We modelled risk factors for infection and mortality using Cox proportional hazards and estimated attributable fractions.
-
-**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections.
-Confirmed infection incidence in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days.
-`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one Covid-19 infection or death.
+**Background**: Information on the transmission and outcomes of COVID-19 infection in staff and residents in Long term care facilities (LTCFs) in the UK is scarce.
+This study aimed to estimate the incidence of and risk factors for suspected and confirmed COVID-19 infection and mortality during the first wave of the pandemic.
+
+**Methods and Findings**: 179 UK LTCFs were included in a cohort study using population-level active surveillance and linked electronic health records.
+Participants included 9,339 residents and 11,604 staff.
+We estimated the incidence of suspected and confirmed infection and mortality in staff and residents using daily tallies of infection (aggregate data) and used individual-level electronic health records from 8,713 residents to model risk factors for confirmed infection and mortality using Cox proportional hazards.
+We estimated mortality attributable to COVID-19 by comparing deaths in care homes with and without outbreaks.
+
+`r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections.
+The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs had at least one COVID-19 infection or COVID-19-related death.
 Lower staffing ratios and higher occupancy rates were independent risk factors for infection.
 
-`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]) residents.
-`r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
-Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1.9 [1.7; 2.2]).
-33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.
-
-**Interpretation:** 1 in 5 residents had symptoms of infection during the pandemic, but many cases were not tested.
+Daily tallies from LTCF managers recorded `r desc_residents$DAT_CFR$y` deaths in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
+Models estimated that 573/1,694 (33.8%) of all-cause deaths were attributable to COVID-19.
+Mortality in residents with no direct evidence of infection was 98% higher in care homes with at least one COVID-19 infection or death compared to care homes without (adjusted HR 2.0 [1.6; 2.4]).
+
+**Conclusions:** 1 in 5 residents had symptoms of infection during the pandemic, and excess mortality in LTCFs with outbreaks suggest many deaths occurred in people who were infected with COVID-19, but not tested.
 Higher occupancy and lower staffing levels increase infection risk.
-Disease control measures should integrate active surveillance and testing with fundamental changes in staffing and care home occupancy to protect staff and residents from infection.
-
-**Funding:** Economic and Social Research Council [ES/V003887/1].
+Disease control measures should integrate active surveillance and testing with fun-damental changes in staffing and care home occupancy to protect staff and residents from infection.
 
@@
 
-Globally the number of cases of Covid-19 continues to increase, but cases in Europe have declined since April 2020,[@ECDC2020] following the introduction of lockdown measures.[@Hale2020] Although the incidence of infection in the general population in England is low (0.04%),[@cisbulletin2020] new infections persist, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] This raises the possibility that these settings represent a reservoir for transmission of infection back to the community.
-
-In the UK, there are an estimated 400,000 residents living in approximately 11,000 homes for the elderly.[@LaingBuisson2019] Care home residents are particularly vulnerable to Covid-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care home setting.
-At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Care home staff also have higher aged-standardised rates of Covid-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved Covid-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
-Understanding the proportion of excess deaths that can be directly and indirectly attributed to Covid-19 infection is important, to fully assess the impact of the pandemic on care homes.
-
-The development of public health strategies to protect the public, residents and staff from Covid-19 requires knowledge of the burden of and risk factors for infection in care home residents and staff, linked to outcomes.
-However, there is no syndromic surveillance for infection in care homes in England, and widespread one-off testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for care home staff and residents until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England’s (PHE) outbreak investigations which permitted a maximum of five tests per care home.
-Consequently national estimates of incidence and prevalence will substantially underestimate the burden of infection in care home residents and staff.
-
-In the absence of cohort studies or active surveillance, outbreak investigations provide the most reliable estimates of the burden of infection and case-fatality.[@McMichael2020; @Dora2020] An estimated 44% of English care homes have had at least one outbreak, with a living systematic review[@Salcher-Konrad2020] reporting substantial variation in cumulative incidence of infection (0%-72%) and case fatality (0-34%) in care home residents.
-A major limitation of outbreak investigations is that follow-up is usually less than 30 days,[@Salcher-Konrad2020] and investigations are only conducted in settings in which outbreaks have been detected.
-Understanding the proportion of care homes with undetected cases is crucial for policy decisions around the frequency of and justification for regular testing in care homes.
-
-To our knowledge, there are no studies which have employed population-level active surveillance in care homes throughout this pandemic to measure outcomes of both suspected and confirmed infection in residents and staff.
-We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest independent provider of residential and nursing care, with the aim of identifying strategies to protect care home staff and residents from future waves of infection.
-Our objectives were to estimate incidence of and risk factors for infection and mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
-We also estimated mortality attributable to Covid-19.
+Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care facility.
+At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Staff in LTCFs also have higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
+Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on LTCFs.
+
+Public health strategies to protect the public, residents and staff from COVID-19 include rapid testing, restriction of visitors, and vaccination.
+Such strategies require knowledge of the burden of and risk factors for infection in residents and staff in LTCFs, linked to outcomes, which draw on evidence from the first wave of the pandemic.
+Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity.
+In England, there is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF.
+Consequently national estimates of incidence and prevalence based on the first wave of infection (February-July 2020) will substantially underestimate the burden of infection in residents and staff in LTCFs.
+
+To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in LTCFs to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections.
+We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest independent provider of residential and nursing care, with the aim of identifying strategies to protect staff and residents in LTCFs from future waves of infection.
+Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
+We also estimated mortality attributable to COVID-19.
 
@@
 
-Staff or residents living and working in care homes for the elderly between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`, which were run by the Four Seasons Healthcare Group (FSHCG) were eligible for study inclusion.
+Staff and residents living and working in LTCFs for the elderly between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`, which were run by the Four Seasons Healthcare Group (FSHCG) were eligible for study inclusion.
 The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is primarily state funded.
@@
 In 2020, there were 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 1).
-90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for Covid-19 at least once between 11 May and 22 June 2020.
+90% of FSHCG LTCFs participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.
 
@@
 
-Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in research.[@Smith2020]
+Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in previous research studies.[@Smith2020]
 
@@
 
-The care home chain collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
-The dataset includes: dates of care home entry and exit, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).
-Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, care home identifier, incident date/time, date of birth, sex, Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death.
-Information in Datix was used categorise residents’ infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1).
+FSHCG collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
+The dataset includes: dates of entry to and exit from the LTCF, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).
+Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, tests and test results, resident current location (LTCF/hospital), and death.
+Individual-level data on residents were linked to Datix reports (supplementary material 1), and used to categorise residents' infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1).
 The term 'confirmed' denoted a positive PCR test.
-Datix was also used to differentiate in-hospital and in care home deaths, and to identify Covid-19 related deaths.
-Individual level data on residents was linked to Datix reports by creating an index file based on a combination of forename, surname and the care home identifier (supplementary material 2).
-1492/1880 (79%) of Datix reports were successfully linked.
-Individual-level records were available between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`.
+Datix was also used to differentiate deaths in hospital from those in the LTCF, and to identify COVID-19 related deaths.
+1,492/1,880 (79%) of Datix reports were successfully linked.
 
@@
 
-On 24 March the care home chain introduced a new system to report Covid-19 cases and deaths.
-This required care home managers to report daily tallies in residents (new symptomatic cases, new confirmed infection in home, new confirmed infection in hospital, deaths related to Covid-19) and staff (symptomatic cases, new confirmed cases).
+On 24 March a new reporting system was introduced by FSHCG, which required managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (symptomatic cases, new confirmed cases).
+The number of occupied beds in each LTCF was reported weekly via the same mechanism.
+COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner.
 Data on staff deaths were not extracted due to the small number of cases.
-Covid-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to Covid-19 by the coroner.
-The number of occupied beds in each care home was reported weekly.
-
-Care home characteristics (number of beds, region, nursing versus residential care) were collected from organisational data.
-We therefore extracted organisational data, individual-level data for 8713 residents and aggregate data for all care home staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).
+
+Characteristics of each LTCF (number of beds, region, nursing versus residential care) were collected from organisational data.
+We therefore extracted organisational data, individual-level data for 8,713 residents and aggregate data for all staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).
 
@@
 
-Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, care home occupancy and staff to bed ratio).
-Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first Covid-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms.
-We also estimated the ratio of staff to beds.
-A care home outbreak was defined as at least one confirmed infection or Covid-19 related death.
+Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
+Baseline LTCF occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable.
+We also estimated the ratio of beds to staff as a continuous variable.
+An outbreak in a LTCF was defined as at least one confirmed infection or COVID-19 related death.
 
@@
 
-*Infection in care home staff and residents*
+*Infection in staff and residents in LTCFs*
 
 Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies.
-These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff.[@CQC2020stats] Infection incidence was also estimated from Datix, but was subject to under-reporting (supplementary material 2).
-In order to calculate infection and death incidence, we estimated the total number of residents in each care home by extrapolating estimates of care home occupancy from the individual-level dataset (because dates of care home entry and exit were not available for 855 beds, supplementary material 1).
+These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] In order to calculate infection and death incidence, we estimated the total number of residents in each care LTCF by extrapolating estimates of LTCF occupancy from the individual-level dataset to the aggregate dataset (because dates of entry and exit to/from the LTCF were not available for 855 beds, supplementary material 1).
 Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation.
 The total number of residents at risk of infection was unknown, so it was approximated in a multiple decrements life table (supplementary material 1).
-The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths by day.
-The rate ratio for care home versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
+The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset.
+The incidence rate ratio for LTCF (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
+
+Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (supplementary material 2).
+The individual-level dataset was used to estimate rates of infection by age and gender, and Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and the outcome of confirmed infection.
 
@@
 
-Individual-level data were used to estimate rates of infection, all-cause mortality and case-fatality by age and gender in residents.
-Aggregate data were also used to estimate the crude rate of Covid-19 related mortality.
-Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and confirmed infection.
-In order to investigate the relationship between Covid-19 infection and excess mortality, we assumed that residents in non-outbreak care homes had not been exposed to infection, and would therefore not experience excess Covid-19 related mortality.
-We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
-
-A Cox proportional hazards regression model tested the effect of individual and care home level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.
-We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes.
+The aggregate dataset was used to estimate the crude rate of COVID-19 related mortality in residents and individual-level data were used to estimate rates of all-cause mortality and case-fatality by age and gender.
+
+In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in non-outbreak LTCFs had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in LTCFs with and without outbreaks.
+
+A Cox proportional hazards regression model tested the effect of individual- and LTCF-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and LTCF outbreak status.
+We estimated the attributable fraction of deaths for each infection category in LTCFs with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak LTCFs.
 This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
+Huber sandwich estimators of variance accounted for the design effect of LTCF clustering in regression models.
 
@@
 
-# Findings
-
-## Cases of infection in residents and staff
+# Results
+
+## Study population
 
 The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff.
-`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one Covid-19 outbreak (including unconfirmed outbreaks) in either the individual-level or aggregate datasets.
-The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset.
-Mean and median duration of resident follow-up was `r round(mean(residents_overview$resident_days))` and `r median(residents_overview$resident_days)` days respectively in the individual-level dataset.
-
-Symptoms of infection were recorded in  `r formatbm(desc_residents$CT_total_symptomatic)` residents based on the aggregate dataset, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`).
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one COVID-19 outbreak in either the individual-level or aggregate datasets.
+The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset, and `r round(mean(residents_overview$resident_days))` days in the individual-level dataset.
+
+## Infection and COVID-19 related mortality (aggregate data)
+
+Symptoms of infection were recorded in  `r formatbm(desc_residents$CT_total_symptomatic)` residents based on counts reported by managers of LTCFs, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
 An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
-The confirmed infection incidence proportion was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-
-Kaplan-Meier estimates of the temporal trend in symptomatic and confirmed cases are shown in `r fig_nums("fig_KM_CT_plots", display = "cite")`.
-
-In England, `r england_results$FSHC_confirmed` infections were confirmed of a total `r formatbm(england_results$FSHC_resident_days)` residents-days between `r format((england_results$date_min), "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.
-In comparison, the survey of English community households[@cisbulletin2020] found a total of `r england_results$cis_confirmed` confirmed cases out of `r formatbm(england_results$cis_person_days)` person-days during the same period.
-This implies a confirmed infection rate ratio comparing care homes to the community of `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`].
+The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
+The rate of confirmed infections in LTCFs in England was 13-fold higher in LTCFs compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]
+
+`r desc_residents$CT_total_cdeath` COVID-19 related resident deaths were reported in the aggregate dataset, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
+`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).
 
@@
 
-| Residents     |  Symptomatic  | Confirmed  |    All homes <br>  Covid-19-related deaths  | Symptomatic  |  Confirmed  | Outbreak homes <br>  Covid-19-related deaths  |
+| Residents     |  Symptomatic  | Confirmed  |    All homes <br>  COVID-19-related deaths  | Symptomatic  |  Confirmed  | Outbreak homes <br>  COVID-19-related deaths  |
 | ------------- | -------------:| -------------:|-------------:|-------------:|-------------:|-------------:|
@@
     "Confirmed infection",
-    "Covid-19-related death"
+    "COVID-19-related death"
   )
@@
 
-*Note*: underlying data available from supplementary material 3.
+*Note*: underlying data available on request from authors, subject to permissions from FHSCG.
 
@@
 
-## Mortality in residents
-
-`r desc_residents$CT_total_cdeath` Covid-related resident deaths were reported in the aggregate dataset, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
-`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).
+## All-cause mortality (individual-level data)
 
 Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.
-`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents were in a nursing home bed, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
-`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in these residents, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
-The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
+`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents received nursing care, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
+`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in residents of LTCFs, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
+The proportion of resident deaths was two-fold higher in LTCFs with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
 
@@
 
-## Factors associated with confirmed infections in residents
+## Factors associated with confirmed infections (individual-level data)
 
 Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
-Male sex, age $\geq$ 85 years, and residence in a nursing home (adjusted HR = 1.5 [1.2; 1.8]) were all independently associated with increased risk of confirmed Covid-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-Large care homes had greater rates of infection (adjusted HR = 1.8 [1.4; 2.4] for care homes with $\geq$ 70 beds versus <35 beds).
-Care home baseline occupancy and staffing ratios had the greatest effect on residents risk of infection.
-For example, the adjusted hazard ratio for confirmed infection was 2.5 times [1.9; 3.3] greater in homes with 0.85-1 resident per room versus homes with 0.7-0.85 resident per room.
-In homes with more residents than bedrooms, hazards were 10 times [6.9; 15] greater.
-
-Higher staff to resident ratios were associated with lower risk of infection: an increase by 10 percentage points of the bed to staff ratio was associated with a 23% increase in infection (adjusted HR = 1.23 [1.17; 1.31]).
+Male sex, age $\geq$ 85 years, and residence in a nursing LTCF (adjusted HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+After accounting for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds).
+LTCF baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection.
+For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR for a 10% point increase = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR for a 10% point increase = 1.3 [1.1; 1.5]).
 
@@
 
-```{r hazard_ratios_confirmed,   results="asis"}
-# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+
-#                     home_type+user_type2+  bed_tot_category + bed_to_staff_ratio +occ_bedroom_ratio_g6, data = residents_overview)
-#  summary(cph_conf)
+```{r hazard_ratios_confirmed,  warning=F, results="asis"}
+residents_overview$uk_imd_england_quintile <- factor(
+  residents_overview$uk_imd_england_quintile,
+  levels = c('3', '1 - least deprived', '2', '4', '5 - most deprived'))
+residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
+
+residents_overview$occ_to_staff <- residents_overview$mean_occupancy/residents_overview$staff_total
+
+# residents_overview <- mutate(residents_overview, gender = as.factor(gender))
+# foreign::write.dta(residents_overview, file = "../residents_overview.dta", 
+#                    convert.factors = c("labels", "string", "numeric", "codes"))
+cph_conf <- coxph(
+  Surv(start_int, stop_int, confirmed, type="counting") ~ gender + ageg + home_type + 
+    user_type2 + uk_imd_england_quintile + bed_tot_category +
+    I(occ_bedroom_ratio-.9) +  I(occ_to_staff-.73)  , #
+  cluster = residents_overview$home_code,
+  data = residents_overview)
 # cox.zph(cph_conf)
 # survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")
-residents_overview$uk_imd_england_quintile <- factor(residents_overview$uk_imd_england_quintile,
-                                                     levels = c(
-                                                       '3', 
-                                                       '1 - least deprived', '2', '4', '5 - most deprived'
-                                                     ))
-residents_overview$bed_to_staff_ratio <- 1/residents_overview$staff_to_bed_ratio
+# I(occ_to_staff-.75) + I(pct_occupancy-0.877) 
 explanatory = c("gender", "ageg", "home_type", "user_type2",  "uk_imd_england_quintile",
-                "bed_tot_category", "bed_to_staff_ratio",  "occ_bedroom_ratio_g6")
-dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
+                "bed_tot_category", "I(occ_bedroom_ratio-.9)", "I(bed_to_staff_ratio-.85)",
+                "cluster(home_code)")
+dependent = "Surv(time = start_int, time2 = stop_int+0.01, event = confirmed)"
 CPH_confirmed <-  finalfit(residents_overview,
@@
 
-## Factors associated with all-cause mortality
+## Factors associated with all-cause mortality (individual-level data)
 
 The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality.
-After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.4 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.2; 1.5]).
-
-We estimated excess mortality in outbreak and non-outbreak homes, taking individuals with no evidence of infection (group A) in non-outbreak homes as the reference group.
-Risk of all-cause mortality was almost two-fold higher in residents in Group A (no direct evidence of infection) in outbreak versus non-outbreak homes (adjusted HR = 2.0 [1.7; 2.2]).
-Risk of death was also higher in group B (residents with symptoms but unconfirmed infection) in outbreak versus non-outbreak homes relative to the baseline group (adjusted HR = 4.3 [3.0; 6.2] versus 9.4 [7.6; 12]).
-All-cause mortality was strongly associated with group C - asymptomatic confirmed infection (adjusted HR = 3.3 [2.0; 5.7]) and group D - symptomatic confirmed infection (adjusted HR = 13 [11; 16]), compared to baseline.
-
-It is important to note these hazard ratio estimates do not give a comprehensive measure of effect: hazards were not proportional across these categories (`r fig_nums("fig_KM_mortality", display="cite")`).
+After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
+
+We estimated excess mortality in outbreak and non-outbreak LTCFs, taking individuals with no evidence of infection (group A) in non-outbreak LTCFs as the reference group.
+Hazards of all-cause mortality were almost two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak LTCFs (adjusted HR = 1.9 [1.6; 2.3]).
+All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.1 [1.9; 5.1]) or symptomatic (Group D: adjusted HR = 12 [9.0; 15]).
 
@@
   ),death_Apr = case_when(
-    status == "Deceased" & date == max(date) & date >= as.Date("2020-04-01") ~ 1L,
+    status == "Deceased" & date == max(date) & 
+      date >= as.Date("2020-04-01") ~ 1L,
     TRUE ~ 0L
@@
 
-```{r hazard_ratios_death, results="asis"}
-# cph_death2 <- coxph(Surv(date1, date2, death_status) ~ gender+ ageg + home_type + user_type2 +
-#                       bed_tot_category + bed_to_staff_ratio + occ_bedroom_ratio_g6 +
-#                       covid_status_outbreak  + cluster(resident_id), data = residents_episodes)
-# summary(cph_death2)
-# cox.zph(cph_death2)
-residents_episodes$uk_imd_england_quintile <- factor(residents_episodes$uk_imd_england_quintile,
-                                                     levels = c(
-                                                       '3', 
-                                                       '1 - least deprived', '2', '4', '5 - most deprived'
-                                                     ))
-
-residents_episodes$bed_to_staff_ratio <- 1/residents_episodes$staff_to_bed_ratio
-explanatory = c("gender", "ageg", "home_type", "user_type2", "uk_imd_england_quintile",
-                "bed_tot_category", "bed_to_staff_ratio", "occ_bedroom_ratio_g6",
-                "covid_status_outbreak",  "cluster(resident_id)")
-dependent = "Surv(time = date1, time2 = date2, event = death_status)"
-CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
-                       column = T, add_dependent_label = F)  %>%   ff_remove_p() 
-knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
-```
-
-*Note:* Baseline group = uninfected residents in non-outbreak care home
+```{r hazard_ratios_death,  warning=F, results="asis"}
+# foreign::write.dta(residents_episodes, 
+#                    file = "../residents_episodes.dta")
+# cph_death <- coxph(
+#   Surv(date1, date2, death_status) ~ gender + ageg +
+#     home_type + user_type2 + bed_tot_category +
+#     uk_imd_england_quintile +
+#     I(bed_to_staff_ratio-.85) + I(occ_bedroom_ratio-.9) +
+#     covid_status_outbreak_digit1 +
+#     covid_status_outbreak_digit2 +
+#     covid_status_outbreak_digit3 +
+#     covid_status_outbreak_digit4 +
+#     covid_status_outbreak_digit5 +
+#     + cluster(resident_id),
+#   data = residents_episodes)
+# cox.zph(cph_death)
+# for(i in 8:12){
+#   plot(cox.zph(cph_death)[i], col = 2)
+# }
+explanatory = c(
+  "gender", "ageg", "home_type", "user_type2", 
+  "bed_tot_category", "uk_imd_england_quintile",
+  "I(occ_bedroom_ratio-.9)", "I(bed_to_staff_ratio-.85)",
+  "covid_status_outbreak",  "cluster(home_code)")
+dependent = "Surv(time = date1, time2 = date2+0.1, event = death_status)"
+CPH_death <-  finalfit(residents_episodes, dependent, 
+                       explanatory, column = T, 
+                       add_dependent_label = F)  %>% 
+  ff_remove_p() 
+knitr::kable(CPH_death, row.names=FALSE, 
+             align=c("l", "l", "r", "r", "r", "r"))
+```
+
+*Note:* Baseline group = uninfected residents in non-outbreak LTCFs
 
@@
 
-## Attributable mortality
+## Attributable mortality (individual-level data)
 
 Model-based estimates of attributable mortality were derived from the individual-level data.
-Overall, 567/1,694 (33%) deaths were attributed to Covid-19.
-In homes with outbreaks only 28% (159 residents) of the mortality attributable to COVID-19 occurred in people with confirmed infection (Groups C and D), (`r table_nums("tab_attributable_deaths", display = "cite")`).
-Exclusion of the early pandemic period in sensitivity analysis increased attributable mortality to 560/1,343 (41.7%).
-Model-based estimates of deaths based on individual-level were slightly higher (8%) than counts from the aggregate data.
+Overall, 573/1,694 (34%) deaths were attributed to COVID-19.
+In LTCFs with outbreaks only, 160/547 (29%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, `r table_nums("tab_attributable_deaths", display = "cite")`).
+Model-based estimates of deaths attributable to COVID-19 based on individual-level data were similar (9% higher) to the number of deaths reported by managers of LTCFs ($n$ = 526) in the aggregate dataset.
 
@@
                     gender+ ageg + home_type + covid_status_outbreak_digit +
-                    staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
-                    cluster(resident_id), data =  residents_episodes)
+                    bed_tot_category + uk_imd_england_quintile +
+                    I(bed_to_staff_ratio-.85) + I(occ_bedroom_ratio-.9) +
+                    cluster(home_code), data =  residents_episodes)
 
@@
   # `Total deaths (April-June)` = formatbm(deaths_apr),
-  `Deaths attributable to Covid-19 (% all-cause deaths)` = paste0(
+  `Deaths attributable to COVID-19 (% all-cause deaths)` = paste0(
     round(AF * sum(deaths/2)), " (",
     tbrounding(AF*100, 2), ")"),
-  # `Deaths attributable to Covid-19 (% all-cause deaths April-June)` = paste0(
+  # `Deaths attributable to COVID-19 (% all-cause deaths April-June)` = paste0(
   #   round(AF_apr * sum(deaths_apr/2)), " (",
@@
 
-# Interpretation
+# Discussion
 
@@
 
-This population-level study demonstrates the major impact of Covid-19 on care homes, with 22% of residents and 16% of staff experiencing symptoms and overall case-fatality of 35.7%.
-Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks, implying substantial case under-ascertainment.
-Less than one-third of deaths attributable to COVID-19 in outbreak homes were confirmed which is likely due to poor availability of testing until late in the pandemic.
-In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy are critically important to reducing the spread of infection.
-
-Our estimates of the prevalence of confirmed Covid-19 infections and deaths in residents are comparable to a large survey of care home managers in England.[@onsVivaldi2020] However, both studies are likely to have underestimated the proportion of residents who became infected due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is higher than previous literature,[@Salcher-Konrad2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
-For example, an outbreak investigation[@Easter6] in 4 care homes in London, UK measured a case-fatality rate of 17% among 126 residents over a period of 62 days, while Stall et al.[@Stall2020] measured a rate of 28% in over a period of 53 days in a Canadian study.
-
-Whereas two-thirds of care homes in our study reported at least one case of infection or death, just 44% of care homes have notified an outbreak to PHE.[@PHE12020]  This suggests that nationally, local health protection teams may be unaware of Covid-19 infections in up to 1 in 5 care homes.
-Integration of data systems, so that test results can be accessed and acted upon by local public health teams is fundamental to the pandemic response.
-
-In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home occupancy.
+This population-level study demonstrates the major impact of COVID-19 on LTCFs, with 22% of residents and 16% of staff experiencing symptoms and a case-fatality rate of 35.7%.
+Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks, implying substantial case under-ascertainment.
+Less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed which is likely due to poor availability of testing until late in the pandemic.
+In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced LTCF occupancy are critically important to reducing the spread of infection.
+
+Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of LTCFs in England.[@onsVivaldi2020] However, both studies are likely to have underestimated the proportion of residents who became infected due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
+
+Two thirds of LTCFs in our study reported at least one case of infection or death, which is comparable to estimates from a recent study from the Lothian region of Scotland which reported that 61% of LTCFs for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system.
+
+In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and LTCF occupancy.
 We also identified lower staff to resident ratios as a risk factor for infection.
-These organisational factors, linked to chronic underfunding of the care home sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
-When staff care for fewer residents they also have reduced likelihood of spreading infection between patients.
-Higher staff to resident ratios may also decrease reliance on agency staff who may spread infection between homes, and indicate better resourced care homes with greater access to protective personal equipment.
+These organisational factors, linked to chronic underfunding of the care sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
+When staff care for fewer residents they also have reduced likelihood of spreading infection between residents.
+Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced LTCFs.
 
@@
 
-The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of care homes, and identify symptomatic as well as confirmed and asymptomatic cases.
-To our knowledge, this is the most compete reporting system for Covid-19 infections in care homes published to date.
-It is possible that care homes that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.
-
-A limitation is lack of access to information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in Covid-19.[@Zhou2020] However, we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
-We also lacked information on the overall rate of care home testing.
+The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of LTCFs, and identify symptomatic as well as confirmed and asymptomatic cases.
+To our knowledge, this is the most complete reporting system for COVID-19 infections in LTCFs published to date.
+It is possible that LTCFs that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.
+
+Limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses.
+We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
+The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each LTCF.
+In common with most LTCFs, 9% of beds were 'block contracted'.
+This meant that individual-level data were available for these individuals, although they were accounted for in the aggregate dataset.
+We made the assumption that individuals in these beds were comparable to all individuals in the facility, which may have introduced selection bias into our analysis of individual-level data.
+Finally, we estimated mortality attributable to COVID-19 by making the assumption that residents in non-outbreak LTCFs had not been exposed to infection.
+This is supported by the finding that two-thirds of LTCFS reported at least one COVID-19 case or related death, which suggests that FSHCG active surveillance system was effective in identifying LTCFs with outbreaks.
 
@@
 
-In the UK the number of infected residents and staff has been underestimated, due to limited availability of testing until late in the pandemic.
+In the UK the number of infected residents and staff in the first wave of the pandemic has been underestimated, due to limited availability of testing before May.
 High levels of asymptomatic infection will also lead to under-ascertainment.
-It is important to note however, that mortality rates were also increased in those recorded as asymptomatic infections suggesting that in an elderly cohort patients may have atypical presentations that do not conform to standard case definitions.
-Although our findings support increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because  the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
+Although our findings support increased use of testing to improve case ascertainment, frequent PCR testing in residents of LTCFs may not always be desirable if the risk of infection is low, because  the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
 Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short.
-Rapid early diagnosis of symptomatic cases in residents and staff and expansion of more widespread testing after a case is identified may also be effective strategies to prevent transmission.
-Such approaches depend on strengthened surveillance in care homes and would be greatly facilitated by the availability of near patient testing platforms, which may be achievable in larger homes.
-
-Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of Covid-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
-Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which Covid-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected .
-Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes.
-Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.
-
-Covid-19 cases are declining in Europe, but there is an opportunity to mitigate the impact of future waves of infection on care home staff and residents.
-Our findings suggest that countries can achieve this by adopting a holistic approach, which integrates surveillance and focused testing for Covid-19 with increased investment to reduce care home occupancy and increase staffing.
-
-<!-- Evidence from the USA suggests that frequent testing in care homes can reduce the spread of infection,[@Dora2020] since various studies have shown effective transmission of SARS-CoV-2 from people who are infected but not yet symptomatic.[@He2020; @Kimball2020] The UK government has recently announced plans[@dhscRegularTesting2020] to increase the frequency of testing in care home staff and residents.
-Although our findings strongly support increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, since the testing procedure (nasopharyngeal swabs) is invasive, and it may cause distress to vulnerable residents, particularly those with dementia.
-Focussing testing on staff and/or developing strategies to identify care homes that are at greatest risk of outbreaks infection could support efforts to target surveillance and infection prevention activities.
-Alternatively the development of saliva antigen tests could provide a less invasive solution to testing.
--->
+Strengthened surveillance in LTCFs would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays (LFIAs),[@Flower2020] provided the predictive value of these tests is adequate.
+
+## Unanswered questions and future research
+
+Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
+Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected.
+Detailed analysis of cause of death and reasons for hospital admission in residents of LTCFs will be important to understand how the pandemic has affected the quality of care in LTCFs.
+Our analysis provides a method that could be widely applied to estimate excess mortality, provided LTCFs with outbreaks can be reliably identified.
+
+# Conclusions
+
+Delays in establishing widespread testing for COVID-19 mean that studies based on laboratory confirmed cases are likely to substantially underestimate disease burden and COVID-19 related mortality in the first wave of the pandemic.
+
+This study found that overall case fatality was high, but the majority of attributable mortality was in residents without laboratory-confirmed infection and one fifth of residents and 16% of staff had symptoms.
+Increased care home occupancy and lower staff to resident ratios increase the risk of infection.
+Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in LTCFs to reduce occupancy and increase staffing.
 
@@
 This work was supported by the Economic and Social Research Council [ES/V003887/1].
-This work was also supported by a National Institute for Health Research (NIHR) Clinician Scientist award (CS-2016-007 to L.S.); In-Practice fellowship (NIHR300293 to A.J.); a NIHR School of Primary Care Research fellowship (SPCR-118 to A.J.); and a NIHR In-Practice Fellowship (to M.R.).
-AH and HH are NIHR Senior Investigators .
+This work was also supported by a National Institute for Health Research (NIHR) Clinician Scientist award (CS-2016-007 to L.S.); In-Practice fellowship (NIHR300293 to A.J.); the NIHR School of Primary Care Research fellowship (G.R.).
+AH and HH are NIHR Senior Investigators.
 AH and HH are supported by Health Data Research UK (grant No.
@@
 
-Figure 2 and 3 data are available on request to authors.
 This work uses data provided by residents and collected by the Four Seasons Health Care Group as part of their care and support.
@@
 ```{r effect_organisational_variables}
-# adhoc1 <- residents_ct %>% 
-#   filter(home_code %in% home_inclusion) %>% 
-#   group_by(home_code) %>% 
-#   summarise(confirmed_cases = sum(new_confirmed),
-#             symptomatic_cases = sum(new_symptomatic),
-#             denominator = sum(occupancy - I_confirmed)) %>% 
-#   left_join(reference_homes) %>% 
-#   left_join(select(filter(beds, year=="2020"), -year)) %>% 
-#   mutate(staff_to_bed_ratio = staff_total/beds,
-#          carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
-#          bed_tot_category = cut(beds, 
-#            breaks = seq(20,100,25), right = F, 
-#            paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
-#          ) %>% 
-#   mutate(staff_to_bed_ratio_g5 = cut(
-#     staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
-#     quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
-#     max(staff_to_bed_ratio)))) %>% 
-#   left_join(home_mean_occupancy) %>% 
-#   mutate(pct_occupancy = mean_occupancy/beds,
-#          occ_bedroom_ratio = mean_occupancy/no_bedrooms) %>% 
-#   mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
-#          occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)))
-# 
+adhoc1 <- residents_ct %>%
+  filter(home_code %in% home_inclusion) %>%
+  group_by(home_code) %>%
+  summarise(confirmed_cases = sum(new_confirmed),
+            symptomatic_cases = sum(new_symptomatic),
+            denominator = sum(occupancy - I_confirmed)) %>%
+  left_join(reference_homes) %>%
+  left_join(select(filter(beds, year=="2020"), -year)) %>%
+  mutate(staff_to_bed_ratio = staff_total/beds,
+         carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
+         bed_tot_category = cut(beds,
+           breaks = seq(20,100,25), right = F,
+           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds")),
+         uk_imd_england_quintile = factor(
+           uk_imd_england_quintile, 
+           levels = c('3', 
+                      '1 - least deprived', '2', '4', '5 - most deprived'
+                                                     ))
+         ) %>%
+  mutate(staff_to_bed_ratio_g5 = cut(
+    staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
+    quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
+    max(staff_to_bed_ratio)))) %>%
+  left_join(home_mean_occupancy) %>%
+  mutate(pct_occupancy = mean_occupancy/beds,
+         occ_bedroom_ratio = mean_occupancy/no_bedrooms,
+         bed_to_staff_ratio = 1/staff_to_bed_ratio) %>%
+  mutate(occ_bedroom_ratio_g5 = cut(occ_bedroom_ratio, seq(0.2, 1.2, .2)),
+         occ_bedroom_ratio_g6 = cut(occ_bedroom_ratio, seq(0.25, 1.2, .15)))
+
+mod_staffing <- glm(confirmed_cases ~   bed_tot_category + 
+    I(bed_to_staff_ratio-.85) * I(occ_bedroom_ratio-.9) + 
+    uk_imd_england_quintile + offset(log(denominator)), 
+   family = poisson(link = "log"), data = adhoc1)
+
 # ggplot(adhoc1, aes(staff_to_bed_ratio, confirmed_cases/denominator*100000)) +
@@
 
+```{r parametric_model}
+# Spiegeldoodah
+library(flexsurv)
+
+exp_conf <- flexsurv::flexsurvreg(
+  Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+
+    home_type+ user_type2+ bed_tot_category + 
+    I(bed_to_staff_ratio-.85) * I(occ_bedroom_ratio-.9) + 
+    uk_imd_england_quintile, data = residents_overview, dist = "exp" )
+
+pred_data1 <- data.frame(list(
+  ageg = structure(3L, .Label = c("<75 years", "75–84 years", 
+                                  "85–94 years", "95+ years"), class = "factor"), 
+  gender = "Male", 
+  home_type = structure(2L, .Label = c("Residential", "Nursing"), class = "factor"),
+  user_type2 = structure(1L, .Label = c("General/elderly", 
+                                        "Dementia"), class = "factor"),
+  start_int = 0, 
+  stop_int = 104, 
+  bed_tot_category = structure(
+    2L, .Label = c("20–34 beds", "45–59 beds", "70–84 beds"), 
+    class = "factor"), 
+  uk_imd_england_quintile = structure(
+    1L, .Label = c("3", "1 - least deprived", "2", 
+                   "4", "5 - most deprived"), class = "factor"), 
+  occ_bedroom_ratio = 0.9, 
+  bed_to_staff_ratio = 0.85))
+
+pred_data2 <- data.frame(list(
+  ageg = structure(3L, .Label = c("<75 years", "75–84 years", 
+                                  "85–94 years", "95+ years"), class = "factor"), 
+  gender = "Male", 
+  home_type = structure(2L, .Label = c("Residential", "Nursing"), class = "factor"),
+  user_type2 = structure(1L, .Label = c("General/elderly", 
+                                        "Dementia"), class = "factor"),
+  start_int = 0, 
+  stop_int = 104, 
+  bed_tot_category = structure(
+    2L, .Label = c("20–34 beds", "45–59 beds", "70–84 beds"), 
+    class = "factor"), 
+  uk_imd_england_quintile = structure(
+    1L, .Label = c("3", "1 - least deprived", "2", 
+                   "4", "5 - most deprived"), class = "factor"), 
+  occ_bedroom_ratio = .9,  
+  bed_to_staff_ratio = 0.95 ############ increased by .1
+  ))
+
+pred_data3 <- data.frame(list(
+  ageg = structure(3L, .Label = c("<75 years", "75–84 years", 
+                                  "85–94 years", "95+ years"), class = "factor"), 
+  gender = "Male", 
+  home_type = structure(2L, .Label = c("Residential", "Nursing"), class = "factor"),
+  user_type2 = structure(1L, .Label = c("General/elderly", 
+                                        "Dementia"), class = "factor"),
+  start_int = 0, 
+  stop_int = 104, 
+  bed_tot_category = structure(
+    2L, .Label = c("20–34 beds", "45–59 beds", "70–84 beds"), 
+    class = "factor"), 
+  uk_imd_england_quintile = structure(
+    1L, .Label = c("3", "1 - least deprived", "2", 
+                   "4", "5 - most deprived"), class = "factor"), 
+  occ_bedroom_ratio = 1.0,  ############ increased by .1
+  bed_to_staff_ratio = 0.85 
+  ))
+
+summary(exp_conf, newdata = pred_data1, t = 100, type = "cumhaz")
+summary(exp_conf, newdata = pred_data2, t = 100, type = "cumhaz")
+summary(exp_conf, newdata = pred_data3, t = 100, type = "cumhaz")
+
+```
+
 # References

### 2020-11-27 manuscript_appendix/manuscript_appendix.Rmd
Revisions version 1.4

@@
 Manager-reported outbreak surveillance counts were used together with occupancy estimated from weekly occupied bed censuses in order to estimate the incidence of cases across time using a Kaplan-Meier life table product limit estimator.
-As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents, as was suggested by the individual-level records.
-Admitted cases were accounted for in the daily new case count.
-Based on this $\hat{I_t}$, the estimated number or previously symptomatic/confirmed residents present in the home on day $t$ was computed as
 
-$$\hat{I_t} = min(n_{t}/n_{t-1}, 1)( \hat{E}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad I_{t_0} = 0$$
-where $d_{t}$ denotes the number of new cases for each day $t = \{0, \ldots, T\}$, while $min(n_{t}/n_{t-1}, 1)$ acts as an (underestimated) approximation of the discharge rate.
-The total exposure time was calculated as $\sum_{t=0}^{T} (N_t-\hat{I}_t)$ where $N_t$ denotes daily occupancy figures.
+The life table records for every day $t = \{0, \ldots, T\}$:
+
+* $n_t$, the number of residents present in the home on day $t$
+* $I_t$, the number of residents infected up to and including day $t-1$ who are still present in the home
+* $d_t$, the number of residents with a new infection on day $t$ (includes residents admitted on day $t$ who got infected outside the home, for instance in hospital)
+* $S_t$, the number of exposed-to-risk residents, that is, the number of residents who have not yet had an infection and are susceptible to an infection.
+
+Note that on any day, $n_t = S_t + I_t$.
+
+Neither $S_{t}$ or $I_t$ were recorded as such by care homes. Instead, these numbers were approximated in a multiple decrement life table. For this, we assumed that:
+
+* no residents were infected before 24 March 2020 ($I_{t=0} = 0$) when tallies began
+* residents previously infected (counted in $I_t$) are discharged at the same rate as exposed-to-risk residents (counted in $S_t$). This assumption was verified in the individual-level records.
+
+Admitted cases were accounted for in the daily new case count. Based on this
+
+Based on these assumptions, we estimated $I_t$ and $S_t$ as
+
+$$\hat{I_t} = min(n_{t}/n_{t-1}, 1)( \hat{I}_{t-1} + d_{t-1}); \qquad  \qquad  \qquad I_{t=0} = 0$$
+
+$$\hat{S_t} = n_t - \hat{I_t}$$
+
+where $min(n_{t}/n_{t-1}, 1)$ serve as an (underestimated) approximation of the discharge rate.
+
+The cumulative exposure time (used for computing cumulative incidence) was calculated as $\sum_{t=0}^{T} \hat{S}_t$.
 

### 2020-12-11 manuscript/manuscript.Rmd
resubmission

@@
 output:
-  word_document:
-    toc: no
-    reference_docx: template.docx
   html_document:
@@
     code_folding: hide
+  word_document:
+    toc: no
+    reference_docx: template.docx
 link-citations: yes
@@
 
-Peter Dutey Magni, Researcher$^{12}$ Haydn Williams, Data manager$^{3}$ Arnoupe Jhass, General Practitioner$^{4}$ Greta Rait, Professor$^4$ Fabiana Lorencatto, Senior researcher$^5$ Harry Hemingway,$^{126}$ Andrew Hayward,$^{7}$ Laura Shallcross$^{12*}$
+Peter Dutey Magni, Researcher$^{1,2}$ Haydn Williams, Data manager$^{3}$ Arnoupe Jhass, General Practitioner$^{4}$ Greta Rait, Professor$^4$ Fabiana Lorencatto, Senior researcher$^5$ Harry Hemingway,$^{1,2,6}$ Andrew Hayward,$^{7}$ Laura Shallcross$^{1,2*}$
 
@@
 
-**Background**: Information on the transmission and outcomes of COVID-19 infection in staff and residents in Long term care facilities (LTCFs) in the UK is scarce.
-This study aimed to estimate the incidence of and risk factors for suspected and confirmed COVID-19 infection and mortality during the first wave of the pandemic.
-
-**Methods and Findings**: 179 UK LTCFs were included in a cohort study using population-level active surveillance and linked electronic health records.
-Participants included 9,339 residents and 11,604 staff.
-We estimated the incidence of suspected and confirmed infection and mortality in staff and residents using daily tallies of infection (aggregate data) and used individual-level electronic health records from 8,713 residents to model risk factors for confirmed infection and mortality using Cox proportional hazards.
-We estimated mortality attributable to COVID-19 by comparing deaths in care homes with and without outbreaks.
-
-`r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections.
+**Background**: Epidemiological data on COVID-19 infection in long-term care facilities (LTCFs) are scarce.
+We analysed data from a large provider of long-term care to investigate infection and mortality during the first wave of the pandemic.
+
+**Methods**: Cohort study of 179 UK LTCFs with  9,339 residents and 11,604 staff.
+We used manager-reported daily tallies to estimate the incidence of suspected and confirmed infection and mortality in staff and residents.
+Individual-level electronic health records from 8,713 residents were used to model risk factors for confirmed infection, mortality, and estimate attributable mortality.
+
+**Results**: `r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory-confirmed infections.
 The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively.
@@
 
-Daily tallies from LTCF managers recorded `r desc_residents$DAT_CFR$y` deaths in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
-Models estimated that 573/1,694 (33.8%) of all-cause deaths were attributable to COVID-19.
-Mortality in residents with no direct evidence of infection was 98% higher in care homes with at least one COVID-19 infection or death compared to care homes without (adjusted HR 2.0 [1.6; 2.4]).
-
-**Conclusions:** 1 in 5 residents had symptoms of infection during the pandemic, and excess mortality in LTCFs with outbreaks suggest many deaths occurred in people who were infected with COVID-19, but not tested.
-Higher occupancy and lower staffing levels increase infection risk.
-Disease control measures should integrate active surveillance and testing with fun-damental changes in staffing and care home occupancy to protect staff and residents from infection.
+`r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
+Mortality in residents with no direct evidence of infection was 98% higher in care homes with outbreaks versus those without (adjusted HR 2.0 [1.6; 2.4]).
+
+**Conclusions:** Findings suggest many deaths occurred in people who were infected with COVID-19, but not tested.
+Higher occupancy and lower staffing levels were independently associated with risks of infection.
+Protecting staff and residents from infection requires regular testing for COVID-19 and fundamental changes to staffing and LTCF occupancy.
 
@@
 
-Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care facility.
-At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Staff in LTCFs also have higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
+Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
+At the peak of the pandemic, deaths recorded in LTCF residents were three times higher than during the preceding year.[@onsDR2020] Staff in LTCFs also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
 Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on LTCFs.
@@
 Public health strategies to protect the public, residents and staff from COVID-19 include rapid testing, restriction of visitors, and vaccination.
-Such strategies require knowledge of the burden of and risk factors for infection in residents and staff in LTCFs, linked to outcomes, which draw on evidence from the first wave of the pandemic.
+Such strategies require knowledge of the burden of and risk factors in residents and staff in LTCFs, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave.
 Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity.
-In England, there is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF.
-Consequently national estimates of incidence and prevalence based on the first wave of infection (February-July 2020) will substantially underestimate the burden of infection in residents and staff in LTCFs.
+There is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF.
+Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in LTCFs.
 
@@
 
-Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in previous research studies.[@Smith2020]
+We extracted organisational data, individual-level data for 8,713 residents and aggregate data for all staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).
+Electronic records collected by the FSHCG are primarily used for billing and monitoring, but have also been used in previous research.[@Smith2020]
 
@@
 
-FSHCG collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
+FSHCG collects electronic records on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
 The dataset includes: dates of entry to and exit from the LTCF, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).
-Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, tests and test results, resident current location (LTCF/hospital), and death.
+Individual-level data on incidents including infections are reported via 'Datix': resident names, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, test results, resident current location (LTCF/hospital), and death.
 Individual-level data on residents were linked to Datix reports (supplementary material 1), and used to categorise residents' infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1).
@@
 
-On 24 March a new reporting system was introduced by FSHCG, which required managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (symptomatic cases, new confirmed cases).
+On 24 March FSHCG introduced a new reporting system requiring managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases).
 The number of occupied beds in each LTCF was reported weekly via the same mechanism.
@@
 
-Characteristics of each LTCF (number of beds, region, nursing versus residential care) were collected from organisational data.
-We therefore extracted organisational data, individual-level data for 8,713 residents and aggregate data for all staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).
+Characteristics of each LTCF (number of beds, region, nursing versus residential care) were obtained from FSHCG.
 
@@
 
-## Risk factors for infection
-
-Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
+## Risk factors
+
+Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
 Baseline LTCF occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable.
@@
 Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies.
-These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] In order to calculate infection and death incidence, we estimated the total number of residents in each care LTCF by extrapolating estimates of LTCF occupancy from the individual-level dataset to the aggregate dataset (because dates of entry and exit to/from the LTCF were not available for 855 beds, supplementary material 1).
+These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary material 1.
 Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation.
-The total number of residents at risk of infection was unknown, so it was approximated in a multiple decrements life table (supplementary material 1).
+The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary material 1).
 The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset.
@@
 Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (supplementary material 2).
-The individual-level dataset was used to estimate rates of infection by age and gender, and Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and the outcome of confirmed infection.
+Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.
 
@@
 
-Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]
+Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. Computer scripts are available online.[@Dutey2020]
 
@@
 
-Symptoms of infection were recorded in  `r formatbm(desc_residents$CT_total_symptomatic)` residents based on counts reported by managers of LTCFs, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+LTCF managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
 An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
@@
 
-`r desc_residents$CT_total_cdeath` COVID-19 related resident deaths were reported in the aggregate dataset, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
+LTCF managers recorded `r desc_residents$CT_total_cdeath` COVID-19 related resident deaths, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
 `r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).
@@
 
-`r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) reported symptoms of infection during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+LTCF managers recorded `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) experiencing symptoms of infection during the study period, while `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
 
@@
 
-Estimates of incidence for private residents derived from Datix are reported in supplementary material 2.
+Incidence estimates derived from individual-level Datix records are reported in supplementary material 2.
 
@@
 
-Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents.
-`r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents received nursing care, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`).
-`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in residents of LTCFs, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
+Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents (`r table_nums("tab_resident_descriptives", display = "cite")`), who accounted for `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
 The proportion of resident deaths was two-fold higher in LTCFs with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
@@
 
-Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
+Factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
 Male sex, age $\geq$ 85 years, and residence in a nursing LTCF (adjusted HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-After accounting for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds).
+After controlling for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds).
 LTCF baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection.
-For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR for a 10% point increase = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR for a 10% point increase = 1.3 [1.1; 1.5]).
+For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).
 
@@
 
-The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality.
-After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
+Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality.
+After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
 
@@
 All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.1 [1.9; 5.1]) or symptomatic (Group D: adjusted HR = 12 [9.0; 15]).
+In confirmed infections, mortality was significantly higher in individuals a record of symptoms.
 
@@
 
-Model-based estimates of attributable mortality were derived from the individual-level data.
-Overall, 573/1,694 (34%) deaths were attributed to COVID-19.
+Model-based estimates indicate that 573/1,694 (34%) all-cause deaths were attributable to COVID-19 (`r table_nums("tab_attributable_deaths", display = "cite")`).
 In LTCFs with outbreaks only, 160/547 (29%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, `r table_nums("tab_attributable_deaths", display = "cite")`).
-Model-based estimates of deaths attributable to COVID-19 based on individual-level data were similar (9% higher) to the number of deaths reported by managers of LTCFs ($n$ = 526) in the aggregate dataset.
 
@@
 
-This population-level study demonstrates the major impact of COVID-19 on LTCFs, with 22% of residents and 16% of staff experiencing symptoms and a case-fatality rate of 35.7%.
-Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks, implying substantial case under-ascertainment.
-Less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed which is likely due to poor availability of testing until late in the pandemic.
+Across 179 LTCFs, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic.
+Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed due to insufficient testing capacity until late in the pandemic.
 In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced LTCF occupancy are critically important to reducing the spread of infection.
 
-Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of LTCFs in England.[@onsVivaldi2020] However, both studies are likely to have underestimated the proportion of residents who became infected due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
-
-Two thirds of LTCFs in our study reported at least one case of infection or death, which is comparable to estimates from a recent study from the Lothian region of Scotland which reported that 61% of LTCFs for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system.
+Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of LTCFs in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
+
+Two thirds of LTCFs in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of LTCFs for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak LTCFs had not been exposed to infection.
+This assumption made it possible to estimate mortality attributable to COVID-19.
+
+Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
+Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected.
+Detailed analysis of cause of death and reasons for hospital admission in residents of LTCFs will be important to understand how the pandemic has affected the quality of care in LTCFs.
+Our analysis provides a method that could be widely applied to estimate excess mortality, provided LTCFs with outbreaks can be reliably identified.
 
@@
 
-The unique surveillance system we established in partnership with FSHCG allowed us to track infections throughout the entire pandemic period across a large number of LTCFs, and identify symptomatic as well as confirmed and asymptomatic cases.
+The unique surveillance system we established in partnership with FSHCG tracked infections throughout the pandemic across a large number of LTCFs.
 To our knowledge, this is the most complete reporting system for COVID-19 infections in LTCFs published to date.
@@
 The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each LTCF.
-In common with most LTCFs, 9% of beds were 'block contracted'.
-This meant that individual-level data were available for these individuals, although they were accounted for in the aggregate dataset.
-We made the assumption that individuals in these beds were comparable to all individuals in the facility, which may have introduced selection bias into our analysis of individual-level data.
-Finally, we estimated mortality attributable to COVID-19 by making the assumption that residents in non-outbreak LTCFs had not been exposed to infection.
-This is supported by the finding that two-thirds of LTCFS reported at least one COVID-19 case or related death, which suggests that FSHCG active surveillance system was effective in identifying LTCFs with outbreaks.
-
-## Clinical, research and policy implications
-
-In the UK the number of infected residents and staff in the first wave of the pandemic has been underestimated, due to limited availability of testing before May.
-High levels of asymptomatic infection will also lead to under-ascertainment.
-Although our findings support increased use of testing to improve case ascertainment, frequent PCR testing in residents of LTCFs may not always be desirable if the risk of infection is low, because  the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
+
+## Conclusions
+
+UK numbers of infected residents and staff were underestimated during the first wave of the pandemic.
+Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in LTCFs to reduce occupancy and increase staffing.
+Although this presupposes increased use of testing to improve case ascertainment, frequent testing in residents of LTCFs may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
 Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short.
-Strengthened surveillance in LTCFs would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays (LFIAs),[@Flower2020] provided the predictive value of these tests is adequate.
-
-## Unanswered questions and future research
-
-Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
-Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected.
-Detailed analysis of cause of death and reasons for hospital admission in residents of LTCFs will be important to understand how the pandemic has affected the quality of care in LTCFs.
-Our analysis provides a method that could be widely applied to estimate excess mortality, provided LTCFs with outbreaks can be reliably identified.
-
-# Conclusions
-
-Delays in establishing widespread testing for COVID-19 mean that studies based on laboratory confirmed cases are likely to substantially underestimate disease burden and COVID-19 related mortality in the first wave of the pandemic.
-
-This study found that overall case fatality was high, but the majority of attributable mortality was in residents without laboratory-confirmed infection and one fifth of residents and 16% of staff had symptoms.
-Increased care home occupancy and lower staff to resident ratios increase the risk of infection.
-Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in LTCFs to reduce occupancy and increase staffing.
+Strengthened surveillance in LTCFs would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
 

### 2021-02-02 manuscript/manuscript.Rmd
first batch corrections from A&A

@@
 ---
-title: "COVID-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
+title: "COVID-19 infection and attributable mortality in UK care homes: Cohort study using active surveillance and electronic records (March-June 2020)"
 bibliography: references.bib
@@
     code_folding: hide
-  word_document:
-    toc: no
-    reference_docx: template.docx
 link-citations: yes
@@
 source("../.Rprofile")
-devtools::load_all("..")
+devtools::load_all()
 load_data()
@@
     dplyr::filter(tidyr::replace_na(new_deaths_hospital, 0) > 0 |
-                    tidyr::replace_na(new_deaths_home, 0)>0) %>% 
+                    tidyr::replace_na(new_deaths_home, 0)>0 |
+                    tidyr::replace_na(new_confirmed_home, 0)>0 |
+                    tidyr::replace_na(new_confirmed_hospital, 0)>0
+                  ) %>% 
     dplyr::group_by(home_code) %>% 
@@
 
+home_outbreak_sensitivity <-  new_cases %>% ungroup() %>% 
+  arrange(home_code, date) %>% 
+  group_by(home_code) %>% 
+  mutate(caserunningtot = cumsum(tidyr::replace_na(new_deaths_hospital, 0)+
+                    tidyr::replace_na(new_deaths_home, 0))) %>%
+  summarise(one_case = min_non_missing(if_else(caserunningtot>0, date, NA_Date_)),
+            two_case = min_non_missing(if_else(caserunningtot>1, date, NA_Date_)),
+            five_case = min_non_missing(if_else(caserunningtot>4, date, NA_Date_))) %>% 
+  left_join(home_first_case)
+
 residents_episodes <- tline %>% 
   filter(rday==1)  %>% 
-  left_join(home_first_case) %>%
-  mutate(outbreak = dplyr::case_when(
+  left_join(home_outbreak_sensitivity) %>%
+  mutate(
+  outbreak = dplyr::case_when(
     date >= first_case_date ~ 1L,
@@
     TRUE ~ 0L
-  ))  %>% 
+  )
+  # outbreak = dplyr::case_when(
+  #   date >= two_case ~ 1L,
+  #   date < two_case ~ 0L,
+  #   TRUE ~ 0L
+  # )
+  # outbreak = dplyr::case_when(
+  #   date >= five_case ~ 1L,
+  #   date < five_case ~ 0L,
+  #   TRUE ~ 0L)
+  ) %>% 
   group_by(resident_id, home_code, dob, gender, 
@@
 desc_residents_outbreak_only$outbreak_homes <- unique(
-  c(home_total_confirmed$home_code[home_total_confirmed$CT_confirmed != 0],
-  distinct(filter(new_cases, tidyr::replace_na(new_deaths_hospital, 0) > 0 |
-                    tidyr::replace_na(new_deaths_home, 0)>0), home_code)$home_code,
-  distinct(filter(residents_overview, confirmed == 1), home_code)$home_code)
+  filter(home_outbreak_sensitivity, !is.na(first_case_date) == 1)$home_code
 )
@@
   name = "tab_resident_descriptives",
-  caption = paste0("Characteristics of FSHCG private bed residents by type of LTCF, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
+  caption = paste0("Characteristics of FSHCG private bed residents by type of care home, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
 tab_CFR_age_sex <- table_nums(
@@
   name = "tab_CoxPH_death",
-  caption =  paste0("Risk factors for all-cause mortality in private residents of LTCFs with and without COVID-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
+  caption =  paste0("Risk factors for all-cause mortality in private care home residents with and without COVID-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
   formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
@@
   name = "tab_attributable_deaths",
-  caption =  paste0("Model-based estimates of attributable death in private residents of LTCFs with and without COVID-19 outbreaks (n=", 
+  caption =  paste0("Model-based estimates of attributable death in private care home residents with and without COVID-19 outbreaks (n=", 
   formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
@@
   name = "fig_study_diagram",
-  caption = paste0("Study overview: location of FSHCG LTCFs and diagram of data sources")
+  caption = paste0("Study overview: location of FSHCG care homes and diagram of data sources")
 )
@@
 
-**Background**: Epidemiological data on COVID-19 infection in long-term care facilities (LTCFs) are scarce.
-We analysed data from a large provider of long-term care to investigate infection and mortality during the first wave of the pandemic.
-
-**Methods**: Cohort study of 179 UK LTCFs with  9,339 residents and 11,604 staff.
+**Background**: Epidemiological data on COVID-19 infection in care homes are scarce.
+We analysed data from a large provider of long-term care for older people to investigate infection and mortality during the first wave of the pandemic.
+
+**Methods**: Cohort study of 179 UK care homes with  9,339 residents and 11,604 staff.
 We used manager-reported daily tallies to estimate the incidence of suspected and confirmed infection and mortality in staff and residents.
@@
 The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively.
-`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs had at least one COVID-19 infection or COVID-19-related death.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one COVID-19 infection or COVID-19-related death.
 Lower staffing ratios and higher occupancy rates were independent risk factors for infection.
@@
 `r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
-Mortality in residents with no direct evidence of infection was 98% higher in care homes with outbreaks versus those without (adjusted HR 2.0 [1.6; 2.4]).
+Mortality in residents with no direct evidence of infection was two-fold higher in care homes with outbreaks versus those without (adjusted HR 2.1 [1.7; 2.5]).
 
@@
 Higher occupancy and lower staffing levels were independently associated with risks of infection.
-Protecting staff and residents from infection requires regular testing for COVID-19 and fundamental changes to staffing and LTCF occupancy.
+Protecting staff and residents from infection requires regular testing for COVID-19 and fundamental changes to staffing and care home occupancy.
 
@@
 
-Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
-At the peak of the pandemic, deaths recorded in LTCF residents were three times higher than during the preceding year.[@onsDR2020] Staff in LTCFs also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
-Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on LTCFs.
-
-Public health strategies to protect the public, residents and staff from COVID-19 include rapid testing, restriction of visitors, and vaccination.
-Such strategies require knowledge of the burden of and risk factors in residents and staff in LTCFs, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave.
+Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people with or within on-site nursing care.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
+At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
+Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.
+
+Public health strategies to protect the public, residents and staff from SARS-CoV-2 include rapid testing, restriction of visitors, and vaccination.
+Such strategies require knowledge of the burden of and risk factors in residents and staff in care homes, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave.
 Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity.
-There is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF.
-Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in LTCFs.
-
-To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in LTCFs to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections.
-We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest independent provider of residential and nursing care, with the aim of identifying strategies to protect staff and residents in LTCFs from future waves of infection.
+There is no syndromic surveillance for infection in care homes in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in care homes until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per care home.
+Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in care homes.
+
+To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections.
+We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit provider of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection.
 Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
@@
 
-Staff and residents living and working in LTCFs for the elderly between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`, which were run by the Four Seasons Healthcare Group (FSHCG) were eligible for study inclusion.
-The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is primarily state funded.
+Staff and residents living/working in care homes for older people run by the Four Seasons Healthcare Group (FSHCG) between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")` were eligible for study inclusion.
+The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is predominantly state-funded.
 Most residents are permanent, but a small proportion receive temporary (respite) care.
 
-In 2020, there were 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 1).
-90% of FSHCG LTCFs participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.
+In 2020, there were 9,568 beds, representing 9% of all registered care home beds in England, Scotland and Northern Ireland (supplementary methods).
+90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.
 
@@
 
-FSHCG collects electronic records on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds).
-The dataset includes: dates of entry to and exit from the LTCF, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).
-Individual-level data on incidents including infections are reported via 'Datix': resident names, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, test results, resident current location (LTCF/hospital), and death.
-Individual-level data on residents were linked to Datix reports (supplementary material 1), and used to categorise residents' infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1).
+FSHCG collects electronic records on residents occupying 'private' beds, excluding those occupying beds that are 'block contracted' to the local authority (855 beds, see Figure 1).
+Records includes: dates of entry to and exit from the care home, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, older residents).
+Individual-level data on incidents including infections are reported via 'Datix': resident names, care home identifier, incident date/time, date of birth, sex, COVID-19 symptoms (9 multiple choices), test results, resident current location (care home/hospital), and death.
+Individual-level data on residents were linked to Datix reports (supplementary methods), and used to categorise residents' infection status into four groups: (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary methods).
 The term 'confirmed' denoted a positive PCR test.
-Datix was also used to differentiate deaths in hospital from those in the LTCF, and to identify COVID-19 related deaths.
+Datix was also used to differentiate deaths in hospital from those in the care home, and to identify COVID-19 related deaths.
 1,492/1,880 (79%) of Datix reports were successfully linked.
@@
 
-On 24 March FSHCG introduced a new reporting system requiring managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases).
-The number of occupied beds in each LTCF was reported weekly via the same mechanism.
+On 24 March FSHCG introduced a new reporting system requiring managers of each care home to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases).
+The number of occupied beds in each care home was reported weekly via the same mechanism.
 COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner.
@@
 
-Characteristics of each LTCF (number of beds, region, nursing versus residential care) were obtained from FSHCG.
+Characteristics of each care home (number of beds, region, nursing versus residential care) were obtained from FSHCG.
 
@@
 
-Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
-Baseline LTCF occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable.
+Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, occupancy, bed-to-staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
+Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable.
 We also estimated the ratio of beds to staff as a continuous variable.
-An outbreak in a LTCF was defined as at least one confirmed infection or COVID-19 related death.
+
+To adjust for case under-ascertainment, we created a dummy variable indicating 'affected care homes', which we define throughout the manuscript as care homes that recorded $\geq$1 confirmed infection or COVID-19 related death during the period.
 
@@
 
-*Infection in staff and residents in LTCFs*
+*Infection in staff and residents in care homes*
 
 Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies.
-These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary material 1.
+These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary methods.
 Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation.
-The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary material 1).
+The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary methods).
 The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset.
-The incidence rate ratio for LTCF (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
-
-Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (supplementary material 2).
+The incidence rate ratio for care home (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]
+
+Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (Table S1b).
 Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.
@@
 
-In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in non-outbreak LTCFs had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in LTCFs with and without outbreaks.
-
-A Cox proportional hazards regression model tested the effect of individual- and LTCF-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and LTCF outbreak status.
-We estimated the attributable fraction of deaths for each infection category in LTCFs with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak LTCFs.
+In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-affected' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
+
+A Cox proportional hazards regression model tested the effect of individual- and care home-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.
+We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes.
 This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.
-Huber sandwich estimators of variance accounted for the design effect of LTCF clustering in regression models.
+Huber sandwich estimators of variance accounted for the design effect of care home clustering in regression models.
 
@@
 The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff.
-`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one COVID-19 outbreak in either the individual-level or aggregate datasets.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one COVID-19 outbreak in either the individual-level or aggregate datasets.
 The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset, and `r round(mean(residents_overview$resident_days))` days in the individual-level dataset.
@@
 
-LTCF managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+Care home managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (Table S1a, Figure 2).
 An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
 The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-The rate of confirmed infections in LTCFs in England was 13-fold higher in LTCFs compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]
-
-LTCF managers recorded `r desc_residents$CT_total_cdeath` COVID-19 related resident deaths, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
-`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).
+The rate of confirmed infections in care homes in England was 13-fold higher in care homes compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]
+
+Care home managers recorded `r desc_residents$CT_total_cdeath` COVID-19 related resident deaths, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days.
+`r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (Table S1a).
 
@@
   scale_y_continuous("Cumulative incidence",
-                     breaks = seq(0,1,.02)) +
+                     breaks = seq(0,1,.04)) +
   ggsci::scale_color_lancet() +
@@
   theme_bw()  +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
+  theme(legend.position = c(0.2, .85),
+        legend.background = element_blank(),
+        axis.title = element_text(size = 11),
+        axis.text = element_text(size = 11),
+        axis.text.x = element_text(angle = 30, vjust = 1, hjust=1)) +
   coord_cartesian(ylim = c(0,0.27)) + ggtitle("(A) Residents")
@@
   scale_y_continuous("Cumulative incidence",
-                     breaks = seq(0,1,.02)) +
+                     breaks = seq(0,1,.04)) +
   scale_x_date(name = "Date", 
@@
   theme(legend.position = "none", 
-        axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
+        axis.title = element_text(size = 11),
+        axis.text = element_text(size = 11),
+        axis.text.x = element_text(angle = 30, vjust = 1, hjust=1)) +
   coord_cartesian(ylim = c(0,0.27)) +  ggtitle("(B) Staff")
@@
   km_residents_graph,
-  km_staff_graph,
-  widths = c(1.4, 1)
-)
-
-pdf("figure2.pdf", 12,4)
+  km_staff_graph 
+)
+
+pdf("figure2.pdf", 11,4)
 km_ct_plots
@@
 
-LTCF managers recorded `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) experiencing symptoms of infection during the study period, while `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
+Care home managers recorded `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) experiencing symptoms of infection during the study period, while `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (Table S2, `r fig_nums("fig_KM_CT_plots", display = "cite")`).
 
@@
 
-Incidence estimates derived from individual-level Datix records are reported in supplementary material 2.
+Incidence estimates derived from individual-level Datix records are reported in supplementary Table S1b.
 
@@
 Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents (`r table_nums("tab_resident_descriptives", display = "cite")`), who accounted for `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%].
-The proportion of resident deaths was two-fold higher in LTCFs with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
-
-`r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (`r table_nums("tab_CFR_age_sex", display = "cite")`).
+The proportion of resident deaths was two-fold higher in care homes with outbreaks compared to those without outbreaks (22.6% versus 11.2%).
+
+`r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (Table S4).
 The case-fatality rate increased with age and was higher in men compared to women.
@@
 Factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
-Male sex, age $\geq$ 85 years, and residence in a nursing LTCF (adjusted HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-After controlling for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds).
-LTCF baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection.
+Male sex, age $\geq$ 85 years, and residence in a nursing care home (adjusted hazard ratio HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+After controlling for organisational differences, care home size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for care homes with $\geq$ 70 beds versus <35 beds).
+Care home baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection.
 For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).
@@
 
-Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality.
-After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
-
-We estimated excess mortality in outbreak and non-outbreak LTCFs, taking individuals with no evidence of infection (group A) in non-outbreak LTCFs as the reference group.
-Hazards of all-cause mortality were almost two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak LTCFs (adjusted HR = 1.9 [1.6; 2.3]).
-All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.1 [1.9; 5.1]) or symptomatic (Group D: adjusted HR = 12 [9.0; 15]).
-In confirmed infections, mortality was significantly higher in individuals a record of symptoms.
+Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality mortality (Figure S3).
+After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.1; 1.6]).
+
+We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group.
+Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.1 [1.7; 2.5]).
+All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.4 [2.1; 5.7]) or symptomatic (Group D: adjusted HR = 13 [10; 16]).
+In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.
 
@@
 
-*Note:* Baseline group = uninfected residents in non-outbreak LTCFs
+*Note:* Baseline group = uninfected residents in non-outbreak care homes
 
@@
 
-Model-based estimates indicate that 573/1,694 (34%) all-cause deaths were attributable to COVID-19 (`r table_nums("tab_attributable_deaths", display = "cite")`).
-In LTCFs with outbreaks only, 160/547 (29%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, `r table_nums("tab_attributable_deaths", display = "cite")`).
+Model-based estimates indicate that 633/1,694 (37%) all-cause deaths were attributable to COVID-19 (Table S7).
+In care homes with outbreaks only, just 161/609 (26%) deaths attributa-ble to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).
 
@@
 ```{r attributable_deaths, warning=F, message=F, results = "asis"}
-cph_death3 <- coxph(Surv(date1, date2, death_status) ~
-                    gender+ ageg + home_type + covid_status_outbreak_digit +
+cph_death3 <- coxph(Surv(date1, date2+0.1, death_status) ~
+                    gender+ ageg + home_type + user_type2 + covid_status_outbreak_digit +
                     bed_tot_category + uk_imd_england_quintile +
-                    I(bed_to_staff_ratio-.85) + I(occ_bedroom_ratio-.9) +
+                    I(occ_bedroom_ratio-.9) + I(bed_to_staff_ratio-.85) +
                     cluster(home_code), data =  residents_episodes)
@@
 
-Across 179 LTCFs, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic.
-Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed due to insufficient testing capacity until late in the pandemic.
-In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced LTCF occupancy are critically important to reducing the spread of infection.
-
-Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of LTCFs in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
-
-Two thirds of LTCFs in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of LTCFs for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak LTCFs had not been exposed to infection.
+Across 179 care homes, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic.
+Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
+In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reducing the spread of infection.
+
+Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of care homes in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
+
+Two thirds of care homes in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of care homes for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak care homes had not been exposed to infection.
 This assumption made it possible to estimate mortality attributable to COVID-19.
@@
 Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected.
-Detailed analysis of cause of death and reasons for hospital admission in residents of LTCFs will be important to understand how the pandemic has affected the quality of care in LTCFs.
-Our analysis provides a method that could be widely applied to estimate excess mortality, provided LTCFs with outbreaks can be reliably identified.
-
-In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and LTCF occupancy.
-We also identified lower staff to resident ratios as a risk factor for infection.
+Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes.
+Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.
+
+In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home baseline occupancy.
+We also found staffing levels to be negatively associated with infection rates.
 These organisational factors, linked to chronic underfunding of the care sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
 When staff care for fewer residents they also have reduced likelihood of spreading infection between residents.
-Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced LTCFs.
+Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes.
+This association may also be confounded by other characteristics of homes with higher staffing levels, if those happen to be better resourced, for instance in relation to staff training, availability of personal protective equipment, building characteristics.
 
@@
 
-The unique surveillance system we established in partnership with FSHCG tracked infections throughout the pandemic across a large number of LTCFs.
-To our knowledge, this is the most complete reporting system for COVID-19 infections in LTCFs published to date.
-It is possible that LTCFs that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.
-
-Limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses.
+The unique surveillance system we established in partnership with FSHCG tracked infections throughout the pandemic across a large number of care homes.
+To our knowledge, this is the most complete reporting system for COVID-19 infections in care homes published to date.
+It is possible that care homes that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.
+
+Internal validity of our findings on mortality attributable to COVID-19 is sensitive to the assumption made that residents of care homes categorised as 'not affected' were not exposed to infections.
+Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses.
 We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
-The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each LTCF.
+The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home.
 
@@
 UK numbers of infected residents and staff were underestimated during the first wave of the pandemic.
-Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in LTCFs to reduce occupancy and increase staffing.
-Although this presupposes increased use of testing to improve case ascertainment, frequent testing in residents of LTCFs may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
+Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in care homes to reduce occupancy and increase staffing.
+Although this presupposes increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
 Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short.
-Strengthened surveillance in LTCFs would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
+Strengthened surveillance in care homes would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
 

### 2021-02-09 manuscript/manuscript.Rmd
final revisions

@@
   summarise(entry_date = min(date),
-            exit_date = max(date+1),
+            exit_date = max(date)+.99,
             resident_days = sum(rday),
@@
             susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
-  mutate(date1 = as.numeric( entry_date - date_min),
+  mutate(date1 = as.numeric(entry_date - date_min),
          date2 = as.numeric(exit_date - date_min),
@@
          symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
-         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
+         date2 = if_else(date1 == date2, date2 + .2, date2)) %>% 
   mutate(covid_status = case_when(
@@
 desc_residents_outbreak_only$outbreak_homes <- unique(
-  filter(home_outbreak_sensitivity, !is.na(first_case_date) == 1)$home_code
+  filter(home_outbreak_sensitivity, !is.na(
+    # two_case
+    first_case_date
+    ) == 1)$home_code
 )
@@
 **Results**: `r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory-confirmed infections.
-The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively.
+The incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively.
 `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one COVID-19 infection or COVID-19-related death.
@@
 `r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]).
-Mortality in residents with no direct evidence of infection was two-fold higher in care homes with outbreaks versus those without (adjusted HR 2.1 [1.7; 2.5]).
+Mortality in residents with no direct evidence of infection was two-fold higher in care homes with outbreaks versus those without (adjusted HR 2.2 [1.8; 2.6]).
 
@@
 
-Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people with or within on-site nursing care.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
+Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
 At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
@@
 To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections.
-We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit provider of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection.
+We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection.
 Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
@@
 
-To adjust for case under-ascertainment, we created a dummy variable indicating 'affected care homes', which we define throughout the manuscript as care homes that recorded $\geq$1 confirmed infection or COVID-19 related death during the period.
+To adjust for case under-ascertainment, we created a dummy variable indicating care homes that had experienced an outbreak, defined throughout the manuscript as care homes that recorded ≥1 confirmed infection or COVID-19 related death during the study period.
+This definition was preferred over a standard outbreak definition (≥2 cases linked in time and place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing.
+Sensitivity analysis was used to assess the impact on excess mortality of using a more specific outbreak definition (supplementary methods).
 
@@
 
-In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-affected' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
+In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
 
@@
 
-Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality mortality (Figure S3).
-After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.1; 1.6]).
+Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality (Figure S3).
+After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
 
 We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group.
-Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.1 [1.7; 2.5]).
-All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.4 [2.1; 5.7]) or symptomatic (Group D: adjusted HR = 13 [10; 16]).
+Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.2 [1.8; 2.6]).
+All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.8 [2.3; 6.4]) or symptomatic (Group D: adjusted HR = 14 [11; 18]).
 In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.
@@
   "covid_status_outbreak",  "cluster(home_code)")
-dependent = "Surv(time = date1, time2 = date2+0.1, event = death_status)"
+dependent = "Surv(time = date1, time2 = date2, event = death_status)"
+
 CPH_death <-  finalfit(residents_episodes, dependent, 
@@
 
-Model-based estimates indicate that 633/1,694 (37%) all-cause deaths were attributable to COVID-19 (Table S7).
-In care homes with outbreaks only, just 161/609 (26%) deaths attributa-ble to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).
+Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7).
+In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).
 
@@
 ```{r attributable_deaths, warning=F, message=F, results = "asis"}
-cph_death3 <- coxph(Surv(date1, date2+0.1, death_status) ~
+cph_death3 <- coxph(Surv(date1, date2, death_status) ~
                     gender+ ageg + home_type + user_type2 + covid_status_outbreak_digit +
@@
   summarise(
-    rdays = sum(as.integer(date2-date1)),
+    rdays = sum(date2-date1),
     deaths = sum(death_status == 1),
@@
   `Resident-days` = formatbm(rdays),
+  `Deaths attributable to COVID-19 (% all-cause deaths)` = round(AF * sum(deaths/2)),
   `Total deaths` = formatbm(deaths),
+  `%` = tbrounding(AF*100, 2)
+
   # `Total deaths (April-June)` = formatbm(deaths_apr),
-  `Deaths attributable to COVID-19 (% all-cause deaths)` = paste0(
-    round(AF * sum(deaths/2)), " (",
-    tbrounding(AF*100, 2), ")"),
   # `Deaths attributable to COVID-19 (% all-cause deaths April-June)` = paste0(
@@
 Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
-In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reducing the spread of infection.
+In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reduce the spread of infection.
 
@@
 We also found staffing levels to be negatively associated with infection rates.
-These organisational factors, linked to chronic underfunding of the care sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
+These organisational factors, linked to chronic underfunding of the care sector, are likely to hinder the implementation of robust infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning.
 When staff care for fewer residents they also have reduced likelihood of spreading infection between residents.
 Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes.
-This association may also be confounded by other characteristics of homes with higher staffing levels, if those happen to be better resourced, for instance in relation to staff training, availability of personal protective equipment, building characteristics.
+These associations may also be confounded by other care home characteristics which could not be measured in this study, such as access to personal protective equipment, or the structure and layout of the building.
 
@@
 
-Internal validity of our findings on mortality attributable to COVID-19 is sensitive to the assumption made that residents of care homes categorised as 'not affected' were not exposed to infections.
+Our estimates of mortality attributable to COVID-19 are dependent on our definition of 'outbreak' versus 'non-outbreak' care homes.
+We used a sensitive definition of outbreaks (≥1 case per care home) because case under-ascertainment in care homes was common due to lack of testing.
+However, it is possible that we incorrectly classified some care homes with one or a handful of cases throughout the pandemic as having experienced outbreaks.
 Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses.
@@
 The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home.
+In addition, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.
 

### 2021-02-10 manuscript/manuscript.Rmd
last minute word cutting

@@
 
-Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility.
-At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
+Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in care homes.[@PHE2020] In the UK, an estimated 400,000 residents live in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces.
+At the peak of the pandemic, deaths recorded in UK care homes were three times higher than during the preceding year.[@onsDR2020] Staff also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK statistics suggests two-thirds of excess deaths recorded in residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested.
 Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.
 
-Public health strategies to protect the public, residents and staff from SARS-CoV-2 include rapid testing, restriction of visitors, and vaccination.
-Such strategies require knowledge of the burden of and risk factors in residents and staff in care homes, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave.
-Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity.
-There is no syndromic surveillance for infection in care homes in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in care homes until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per care home.
-Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in care homes.
+Strategies to protect residents and staff from SARS-CoV-2 include rapid testing, restriction of visitors, and vaccination.
+These require knowledge of the burden of and risk factors for infection in residents and staff in care homes, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave.
+Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic due to limited testing capacity.
+There is no syndromic surveillance for infection in care homes in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's outbreak investigations which permitted a maximum of five tests per care home.
+Consequently, national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in care home residents and staff.
 
 To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections.
-We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection.
+We analysed electronic health records from the Four Seasons Health Care Group (FSHCG), one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection.
 Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection.
@@
 
-Staff and residents living/working in care homes for older people run by the Four Seasons Healthcare Group (FSHCG) between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")` were eligible for study inclusion.
-The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is predominantly state-funded.
+Staff and residents living/working in care homes for older people run by the FSHCG between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")` were eligible for study inclusion.
+FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is predominantly state-funded.
 Most residents are permanent, but a small proportion receive temporary (respite) care.
@@
 In 2020, there were 9,568 beds, representing 9% of all registered care home beds in England, Scotland and Northern Ireland (supplementary methods).
-90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.
+90% of FSHCG homes participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.
 
@@
 FSHCG collects electronic records on residents occupying 'private' beds, excluding those occupying beds that are 'block contracted' to the local authority (855 beds, see Figure 1).
-Records includes: dates of entry to and exit from the care home, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, older residents).
+Records include: dates of entry to and exit from the care home, sex, date of birth, type of stay (residential/nursing) and care (general/dementia/older residents).
 Individual-level data on incidents including infections are reported via 'Datix': resident names, care home identifier, incident date/time, date of birth, sex, COVID-19 symptoms (9 multiple choices), test results, resident current location (care home/hospital), and death.
@@
 COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner.
-Data on staff deaths were not extracted due to the small number of cases.
-
-Characteristics of each care home (number of beds, region, nursing versus residential care) were obtained from FSHCG.
 
@@
 
-Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, occupancy, bed-to-staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]).
+Risk factors included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (nursing/residential, number of beds, occupancy, bed-to-staff ratio, Index of Multiple Deprivation [@Abel2016]) obtained from FSHCG.
 Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable.
@@
 
-To adjust for case under-ascertainment, we created a dummy variable indicating care homes that had experienced an outbreak, defined throughout the manuscript as care homes that recorded ≥1 confirmed infection or COVID-19 related death during the study period.
-This definition was preferred over a standard outbreak definition (≥2 cases linked in time and place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing.
-Sensitivity analysis was used to assess the impact on excess mortality of using a more specific outbreak definition (supplementary methods).
+A dummy variable marked the time from which an outbreak occurred, defined throughout the manuscript as a care home recording ≥1 confirmed infection or COVID-19 related death.
+This definition was preferred over a standard outbreak definition (≥2 cases linked in time/place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing.
+Sensitivity analysis using a more specific outbreak definition can be found in supplementary methods.
 
@@
 
-Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies.
-These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary methods.
-Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation.
-The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary methods).
+Incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies, the trusted source used for national reporting of cases in all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach (supplementary methods).
 The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset.
@@
 
-Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (Table S1b).
+Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting.
 Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.
@@
 
-The aggregate dataset was used to estimate the crude rate of COVID-19 related mortality in residents and individual-level data were used to estimate rates of all-cause mortality and case-fatality by age and gender.
-
-In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.
-
-A Cox proportional hazards regression model tested the effect of individual- and care home-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.
+The aggregate dataset was used to estimate the crude rate of COVID-19 related mortality in residents.
+Individual-level data were used to estimate rates of all-cause mortality and case-fatality by age and gender.
+
+To investigate COVID-19 excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 mortality.[@Burton2020]
+
+A Cox proportional hazards regression model tested the effect of individual- and home-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.
 We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes.
@@
 
-Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. Computer scripts are available online.[@Dutey2020]
+Data were analysed in `R`3.5.0, `epitool`[@Aragon2020] and `survival`.[@therneau2015]. Computer scripts are available online.[@Dutey2020]
 
@@
 The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff.
-`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one COVID-19 outbreak in either the individual-level or aggregate datasets.
+`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded an outbreak.
 The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset, and `r round(mean(residents_overview$resident_days))` days in the individual-level dataset.
@@
 Care home managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (Table S1a, Figure 2).
-An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
+An additional `r desc_residents$CT_total_confirmed` residents had confirmed infections, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital.
 The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`].
-The rate of confirmed infections in care homes in England was 13-fold higher in care homes compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]
+The rate of confirmed infections in care homes in England was 13-fold higher in care homes compared to the community prevalence of infection derived from the ONS household infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]
 
@@
 
-Incidence estimates derived from individual-level Datix records are reported in supplementary Table S1b.
-
 ## All-cause mortality (individual-level data)
@@
 `r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (Table S4).
-The case-fatality rate increased with age and was higher in men compared to women.
+The case-fatality rate increased with age and was higher in men.
 
@@
 
-Factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model.
-Male sex, age $\geq$ 85 years, and residence in a nursing care home (adjusted hazard ratio HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
-After controlling for organisational differences, care home size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for care homes with $\geq$ 70 beds versus <35 beds).
-Care home baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection.
-For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).
+Factors affecting rates of confirmed infections were investigated in Cox Proportional Hazard models (`r table_nums("tab_CoxPH_confirmed", display = "cite")`).
+Male sex, age ≥85 years, and nursing care (adjusted hazard ratio HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed infection.
+After controlling for organisational differences, care home size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for care homes with ≥70 beds versus <35 beds).
+Care home baseline occupancy and staffing ratios had the greatest effect on residents’ risk of infection.
+For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).
 
@@
 
-Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality (Figure S3).
-After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).
-
-We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group.
+Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examined the relationship between infection status (groups A–D) and mortality (Figure S3).
+After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and nursing care (adjusted HR = 1.3 [1.1; 1.6]).
+
+We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (Group A) in non-outbreak care homes as the reference group.
 Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.2 [1.8; 2.6]).
@@
 Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7).
-In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).
+In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D).
 
@@
 
-Across 179 care homes, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic.
-Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
-In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reduce the spread of infection.
-
-Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of care homes in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
-
-Two thirds of care homes in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of care homes for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak care homes had not been exposed to infection.
+Across 179 care homes, 22% of residents experienced symptoms while 10% had laboratory-confirmed infections, with a case-fatality rate of 35.7% across the first wave of the pandemic.
+Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks.
+Only one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
+In addition to the need for active surveillance and increased testing capacity, higher staff-to-resident ratios and reduced occupancy may be important to reduce the spread of infection.
+
+Our estimates are comparable to a large survey of managers of care homes in England.[@onsVivaldi2020] Both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.
+
+Two-thirds of care homes in our study reported ≥1 infection or death, in agreement with a study from one region of Scotland which reported that 61% of care homes had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCG's active surveillance system, and supports our assumption that residents in non-outbreak care homes had not been exposed to infection.
 This assumption made it possible to estimate mortality attributable to COVID-19.
 
-Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.
-Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected.
-Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes.
+Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.[@Lai2020; @Weinberger2020] Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes.
 Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.
@@
 When staff care for fewer residents they also have reduced likelihood of spreading infection between residents.
-Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes.
-These associations may also be confounded by other care home characteristics which could not be measured in this study, such as access to personal protective equipment, or the structure and layout of the building.
+Higher staff-to-resident ratios may also decrease reliance on agency staff working across multiple settings, and indicate better resourced care homes.
+These associations may also be confounded by other characteristics which could not be measured in this study, such as access to personal protective equipment, or the building structure/layout.
 
@@
 
-The unique surveillance system we established in partnership with FSHCG tracked infections throughout the pandemic across a large number of care homes.
+The unique surveillance system we established in partnership with FSHCG tracked infections across a large number of care homes.
 To our knowledge, this is the most complete reporting system for COVID-19 infections in care homes published to date.
@@
 Our estimates of mortality attributable to COVID-19 are dependent on our definition of 'outbreak' versus 'non-outbreak' care homes.
-We used a sensitive definition of outbreaks (≥1 case per care home) because case under-ascertainment in care homes was common due to lack of testing.
-However, it is possible that we incorrectly classified some care homes with one or a handful of cases throughout the pandemic as having experienced outbreaks.
-Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses.
-We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
+We used a sensitive definition (≥1 case/home) due to under-ascertainment caused by the lack of testing.
+However, it is possible that we incorrectly classified some care homes with only a few cases throughout the pandemic as having experienced outbreaks.
+Other key limitations relate to the completeness of individual-level data.
+We lacked information on comorbidity and ethnicity, shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity.
 The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home.
-In addition, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.
+Finally, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.
 
@@
 Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in care homes to reduce occupancy and increase staffing.
-Although this presupposes increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
+Although testing will improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents.
 Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short.
-Strengthened surveillance in care homes would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
+Strengthened surveillance in care homes could be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
 

### 2021-02-10 manuscript/manuscript.Rmd
update one figure

@@
 Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7).
-In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D).
+In care homes with outbreaks only, just 161/1014 (16%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D).
 
@@
 Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks.
-Only one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
+Only one in six deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic.
 In addition to the need for active surveillance and increased testing capacity, higher staff-to-resident ratios and reduced occupancy may be important to reduce the spread of infection.
